### 2015-11-26 progReport/report.tex
Added the second chapter of the Progress Report

\documentclass[a4paper, onecolumn, 10pt]{article}

    %%--Input Português/(...)--%%
    \usepackage[english]{babel}
    \usepackage[utf8]{inputenc}

    %%--Packages Opcionais--%%
    \usepackage{graphicx}
    \usepackage{epstopdf}
    \usepackage{textcomp}
    \usepackage{amsmath}
    \usepackage{amstext}
    \usepackage{SIunits}
    \usepackage{pgfplots}
    \usepackage{float}
    \usepackage[scale=0.80]{geometry}
    \usepackage{enumerate}
    \usepackage{subfig}
    \usepackage{tabularx}
    \usepackage{xcolor}
    \usepackage[colorlinks=true,urlcolor=blue,linkcolor=magenta]{hyperref}
    \usepackage{listings}
    \DeclareMathOperator*{\argmax}{\mathbf{arg\,max\,\,}}

    %%--Packages de Tipos de Letra--$$
    \usepackage[mathscr]{euscript} 
    %\renewcommand*\ttdefault{pcr}
    \usepackage{libertine}
    %\usepackage{mathptmx}
    \usepackage{inconsolata}
    \usepackage{carlito}
    %%-- Customizing the Code Presentation -- %%
    \definecolor{codegreen}{rgb}{0,0.6,0}
    \definecolor{codegray}{rgb}{0.5,0.5,0.5}
    \definecolor{codepurple}{rgb}{0.58,0,0.82}
    \definecolor{backcolour}{rgb}{0.95,0.95,0.92}
    \definecolor{sectioncol}{HTML}{9C0906}
    \definecolor{subsectioncol}{HTML}{CC100C}

	\lstdefinestyle{mystyle}{
	backgroundcolor=\color{backcolour},   
	commentstyle=\color{codegreen},
	keywordstyle=\color{magenta},
	numberstyle=\ttfamily\tiny\color{codegray},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\scriptsize,
	breakatwhitespace=false,         
	breaklines=true,                 
	captionpos=b,                    
	keepspaces=true,                 
	numbers=left,                    
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2
	}
    \lstset{style=mystyle}

    %%--Cabeçalho de Rodapé--%%
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \cfoot{\large\thepage}
    \rfoot{}
    \lhead{}
    \chead{}
    \rhead{}
    \renewcommand{\headrulewidth}{0.5pt}
    \renewcommand{\footrulewidth}{0.5pt}
    \newcommand{\mc}[1]{\mathcal{#1}}	
    \newcommand{\mb}[1]{\mathbf{#1}}	
    \newcommand{\prob}[1]{\mathrm{P}\left( #1 \right)}
    \newcommand{\compon}{X_1 = x_1^{(k)}, X_2 = x_2^{(k)}, \dots, X_D = x_D^{(k)}}
    \newcommand{\prodBay}{\prod_{i=1}^{D} \prob{X_i = x_i^{(k)} | Y = c_j }}
    	%%--Estilo das Secções--%%
    \usepackage[sf]{titlesec}

    \titlespacing{\section}{0em}{2.5em}{2em}
    \titleformat{\section}
    {\color{sectioncol}\sffamily\scshape\Large\bfseries}
    {\color{sectioncol}\LARGE\thesection}{1em}{}

    \titlespacing{\subsection}{1em}{1.5em}{1em}
    \titleformat{\subsection}
    {\color{subsectioncol}\sffamily\slshape\normalsize\bfseries}
    {\color{subsectioncol}\thesubsection \hspace{5pt} -- }{5pt}{}

    \titlespacing{\subsubsection}{2em}{1.5em}{1em}
    \titleformat{\subsubsection}
    {\color{subsectioncol}\sffamily\slshape\normalsize}
    {\color{subsectioncol}\thesubsubsection \hspace{5pt} -- }{5pt}{}

    \title{\Huge PDI -- Relatório de Progresso}
    \author{\itshape \href{mailto:ee11126@fe.up.pt}{José Pedro Castro Fonseca}   }

\begin{document}

\pagestyle{plain}

\begin{figure}
	\centering
	\includegraphics[scale=0.3]{logo_feup.eps}
\end{figure}
\sffamily
\maketitle

%\vspace{80pt}

%\begin{titlepage}
%\sffamily
%	{\huge\bfseries Machine Learning -- Homework 2} \\
%	\vspace{20pt}
%	{\large\itshape  José Pedro Castro Fonseca} \\
%	\vspace{15pt}
%	{\bfseries November 2, 2015 -- Porto, Portugal} \\
%\end{titlepage}

%\maketitle
\rmfamily\pagestyle{fancy}
\section{Análise Introdutória do Problema a Tratar}

\section{Trabalho Realizado}
Nesta secção é feita uma pequena demonstração das interações com o Orientador, assim como os eventos anteriores à atribuição do tema de dissertação que motivaram a proposta do mesmo.

	\subsection{Pessoas Envolvidas na Dissertação}
	Para além de mim, o Candidato ao grau de Mestre, incluem-se duas pessoas, nomeadamente

	\begin{itemize}
		\item
			\textbf{Orientador --} O \href{https://sigarra.up.pt/feup/pt/func_geral.formview?p_codigo=210963}{Professor João Canas Ferreira}, professor Auxiliar do Departamento de Engenharia Electrotécnica e de Computadores da Faculdade de Engenharia da Universidade do Porto, e meu antigo docente da UC de \href{https://sigarra.up.pt/feup/pt/UCURR_GERAL.FICHA_UC_VIEW?pv_ocorrencia_id=352359}{Projecto de Circuitos VLSI}.

			\textbf{Co-Orientador --} O \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, candidato a PhD em Ciência de Computadores, na área de Inteligência Artificial, na Universidade de Cambridge, Reino Unido.

	\end{itemize}

	\subsection{Instâncias de Interação com o Orientador}
	As principais instâncias de interação com o orientador resumem-se nos seguintes pontos.
	\begin{itemize}
		\item
			\textbf{Sondagem dos Temas Disponíveis (4 de Setembro de 2015) --} Após me ter inscrito à UC de \href{https://sigarra.up.pt/feup/pt/ucurr_geral.ficha_uc_view?pv_ocorrencia_id=374787}{Aprendizagem Computacional}, comecei a navegar um pouco pelo mundo das aplicações do Machine Learning e da forma como este pode ser acelerado com estruturas de Hardware dedicadas, ao invés de utilizar Processadores convencionais. Adicionalmente, colegas de anos anteriores sugeriram-me que o Prof. João Canas Ferreira poderia ter algum trabalho disponível nessa área. Assim, enviei um email ao Prof. João Canas Ferreira para sondar os temas que teria disponíveis nesta área, e obtive resposta positiva. Combinamos, então, uma reunião Skype em que discutiriamos com mais detalhe esta hipótese, no dia 15 de Setembro.

		\item
			\textbf{Reunião por Skype (15 de Setembro de 2015) --} Nesta reunião, o Prof. Canas deu-me a conhecer alguns projectos em concreto que tinha em mente, e a abordagem ao problema que deveria ser seguida. Acordamos, então, que ele iria submeter uma \href{https://sigarra.up.pt/feup/pt/estagios_empresas.ver_dados_proposta?p_id=191431&pv_perfil=ALU&p_aluno_id=116261}{Proposta de Tema de Dissertação}, mas não seria imediatamente alocada para mim, dado que eu gostaria de dar uma vista de olhos pelas restantes Temas, e só depois tomar uma decisão final e porque a minha média me permitiria escolher o tema em concurso normal, sem o problema de eu ser afastado pelo processo de seriação.

		\item
			\textbf{Envio da Proposta de Dissertação (17 de Setembro de 2015) --} Envio, por email, da proposta de dissertação redigida pelo Prof. Canas, para minha análise. Esta proposta é uma indicação geral do trabalho a ser desenvolvido, sendo que o algoritmo específico e o equipamento utilizado poderiam sofrer modificações com o desenrolar da pesquisa.

		\item
			\textbf{Submissão da minha lista de preferências no Concurso Normal (27 de Setembro de 2015) --} Neste dia, submeto a minha lista de preferências de escolha de tema de dissertação, em que me decido definitivamente por este tema. Informei, por email,  o Prof. Canas da mesma escolha. Sugiro a possibilidade de se incluir como co-orientador o Ivo Timóteo.

		\item
			\textbf{Reunião Presencial (5 de Novembro de 2015) --} Discussão do plano de trabalhos, dos objectivos a serem atingidos e das metodologias a serem desenvolvidas. Nesta reunião, apresentei também alguns resultados preliminares da minha primeira pesquisa sobre a matéria, e o Orientador sugere, adicionalmente, a incorporação dos eventuais resultados da dissertação num projecto em desenvolvimento no INESC, como uma possibilidade.

		\item
			\textbf{Reunião Presencial (26 de Novembro de 2015) --} Neste dia, apresento uma versão preliminar deste relatório ao Orientador para discussão, assim como uma primeira ideia de trabalho, explicada na Secção~\ref{sec:analiseProb}.
	\end{itemize}

\end{document}

### 2015-11-27 progReport/report.tex
Added more subsections to the Report Structure. Hopefully, this will be the final shape
Started writing the First Chapter. Also added bibliography file that still does not work!

@@
 
+    %%--Gestão de Referências Bibliográficas--%%
+    \usepackage{biblatex}
+    \bibliography{references}
     %%--Packages de Tipos de Letra--$$
@@
 \section{Análise Introdutória do Problema a Tratar}
+Esta Tese de Dissertação consistirá no projecto e implementação de um Acelerador de Hardware reconfigurável -- baseado em FPGA\footnote{\textit{Field-Programmable Gate Array}, um tipo de Circuitos Integrados reconfiguráveis em Software, cuja estrutura é \textbf{descrita} usando linguagens como Verilog ou VHDL} -- para estruturas de Aprendizagem Computacional, nomeadamente \textbf{Redes Neuronais Recursivas}.
+	\subsection{Contexto}
+	As \textbf{Redes Neuronais Artificiais} são, desde o seu aparecimento nos anos 80, um dos algoritmos de classificação mais populares do mundo da Inteligência Artificial.
+Tal como o nome sugere, são estruturas cuja operação é inspirada na forma como os blocos constituintes do nosso cérebro, os neurónios, operam.
+Apesar da sua elevada performance, um dos seus \textit{shortcomings} é o facto de não conseguir reter informação de eventos anteriores na sua decisão actual, i.e.
+perante um vector de entrada $x_i$, a decisão sobre o vector de saída $y_i$ não possui influência directa de entradas anteriores $x_{i-1}, x_{i-2}, ...$, sendo assim desajustada para o processamento de séries temporais de dados, em que entradas actuais têm uma elevada dependência informacional e sintatica de entradas anteriores, como a análise de texto, sinais deterministicos e a predição de novos valores dessas séries.
+	Uma das formas de melhorar o aspecto anterior, foi a introdução nos anos 80 de \textbf{elementos de memória}, sendo isto feito atrvés do \textit{feedback} das camadas exteriores de decisão para as camadas interiores, ou escondidas.
+A estrutura que utilizarei neste trabalho serão as \textit{Long Short-Term Memory Networks} \cite{hochreiter-lstm}, apresentada em que permite a
+
+	\subsection{Motivação}
+
+	\subsection{Objectivos}
 
@@
 			\textbf{Orientador --} O \href{https://sigarra.up.pt/feup/pt/func_geral.formview?p_codigo=210963}{Professor João Canas Ferreira}, professor Auxiliar do Departamento de Engenharia Electrotécnica e de Computadores da Faculdade de Engenharia da Universidade do Porto, e meu antigo docente da UC de \href{https://sigarra.up.pt/feup/pt/UCURR_GERAL.FICHA_UC_VIEW?pv_ocorrencia_id=352359}{Projecto de Circuitos VLSI}.
-
+		\item
 			\textbf{Co-Orientador --} O \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, candidato a PhD em Ciência de Computadores, na área de Inteligência Artificial, na Universidade de Cambridge, Reino Unido.
@@
 
+	\section{Pontos a Concretizar}
+
+	\printbibliography
 \end{document}

### 2015-11-28 progReport/report.tex
Complete version of the Report, awaiting corrections. Corrected bibliography problem. Added .bib file with references

@@
     %%--Gestão de Referências Bibliográficas--%%
-    \usepackage{biblatex}
-    \bibliography{references}
+    \usepackage[backend=biber, style=ieee]{biblatex}
+    \addbibresource{references.bib}
     %%--Packages de Tipos de Letra--$$
@@
 \rmfamily\pagestyle{fancy}
-\section{Análise Introdutória do Problema a Tratar}
+\section{Análise Introdutória do Problema a Tratar}\label{sec:analiseProb}
 Esta Tese de Dissertação consistirá no projecto e implementação de um Acelerador de Hardware reconfigurável -- baseado em FPGA\footnote{\textit{Field-Programmable Gate Array}, um tipo de Circuitos Integrados reconfiguráveis em Software, cuja estrutura é \textbf{descrita} usando linguagens como Verilog ou VHDL} -- para estruturas de Aprendizagem Computacional, nomeadamente \textbf{Redes Neuronais Recursivas}.
 	\subsection{Contexto}
-	As \textbf{Redes Neuronais Artificiais} são, desde o seu aparecimento nos anos 80, um dos algoritmos de classificação mais populares do mundo da Inteligência Artificial.
+	As \textbf{Redes Neuronais Artificiais} são  um dos algoritmos de classificação mais populares do mundo da Inteligência Artificial.
 Tal como o nome sugere, são estruturas cuja operação é inspirada na forma como os blocos constituintes do nosso cérebro, os neurónios, operam.
@@
 perante um vector de entrada $x_i$, a decisão sobre o vector de saída $y_i$ não possui influência directa de entradas anteriores $x_{i-1}, x_{i-2}, ...$, sendo assim desajustada para o processamento de séries temporais de dados, em que entradas actuais têm uma elevada dependência informacional e sintatica de entradas anteriores, como a análise de texto, sinais deterministicos e a predição de novos valores dessas séries.
+
 	Uma das formas de melhorar o aspecto anterior, foi a introdução nos anos 80 de \textbf{elementos de memória}, sendo isto feito atrvés do \textit{feedback} das camadas exteriores de decisão para as camadas interiores, ou escondidas.
-A estrutura que utilizarei neste trabalho serão as \textit{Long Short-Term Memory Networks} \cite{hochreiter-lstm}, apresentada em que permite a
+A estrutura que utilizarei neste trabalho serão as \textit{Long Short-Term Memory Networks} \autocite{hochreiter-lstm}, propostas por Hochreiter et al.
+em 1997, e que desde então é uma das técnicas \textit{state-of-the-art} na área da Aprendizagem Profunda (\textit{Deep Learning}, tendo obtido a melhor classificação de sempre no reconhecimento de caligrafia \autocite{hochreiter-hrr} em 2009, reconhecimento de fala \autocite{speechRec-lstm}, entre outros.
 
 	\subsection{Motivação}
+	Todas estas técnicas são geralmente implementadas em Processadores Convencionais, fazendo uso de linguagens de alto ou baixo nível.
+Desta forma, todo o paralelismo real inerente às estruturas de cálculo fica limitado ao número de \textit{cores} físicos presentes na máquina, que hoje em dia não excedem os 4 ou 8 nos dispositivos convencionais usados pela maioria das pessoas (Computadores Pessoais ou dispositivos móveis).
+Assim, de forma a paralelizar completamente a estrutura e, por conseguinte, aumentar consideravelmente a performance, uma solução é a utilização de uma GPU's ou FPGA, perimitindo a realização de aprendizagens cada vez mais complexas, e cada vez mais depressa.
+Além disso, quando a taxa de amostragem de dados de entrada é muito elevada, a solução convencional de CPU não é escalavel e apresenta sérias limitações.
+
+	Apesar de já haver trabalho de adaptação de outros algoritmos de aprendizagem para Hardware, a área do \textit{Deep Learning} ainda é um campo pouco explorado no Hardware, sendo que no caso específico de LSTM, ainda existe apenas um artigo \autocite{fpgalstm}, submetido ainda este mês, que explora uma primeira implementação desta técnica em FPGA para a geração automática de texto.
+
+	Ainda que os CPU's convencionais sejam insubstituíveis para as funções normais a que geralmente estão afectos, é possível que possam haver co-processadores baseados nestas estruturas que acelerem certas tarefas de reconhecimento automático, sem comprometer a performance do sistema global.
+Além disso, foi provado  que as Redes Neuronais Recorrentes são \textit{Turing Complete} \autocite{turingcompletenessRNN}\autocite{turingcompletenessRNN2}, significando, por isso, que podem simular o funcionamento de qualquer programa que possa ser executado num CPU convencional.
 
 	\subsection{Objectivos}
+	Assim, tendo como base o trabalho apresentado em \autocite{fpgalstm}, a ideia é construir um sistema que, além daquele, permita uma reconfiguração completa, em tempo real, da topologia da rede, e a subsituição das células básicas por outras com variações na sua estrutura.
+
+	Será feito, em primeiro lugar, um modelo conceptual do sistema numa linguagem de alto-nível que permita validar o funcionamento dos IP Cores que forem posteriormente desenvolvidos em Verilog, e que depois serão sintetizados para a plataforma-alvo FPGA.
+Além disso, será utilizado um ou dois \textit{Benchmarks} conhecidos para, por um lado, comparar o desempenho do sistema FPGA versus a implementação em CPU e, por outro, comparar o desempenho da minha implementação com outros desempenhos de referência.
+Seria igualmente interessante escolher uma aplicação-tipo deste tipo de Redes, e testá-la na minha implementaçõa, para poder observar o seru funcionamento prático.
 
@@
 
-	\section{Pontos a Concretizar}
+	\subsection{Sequência Temporal de Trabalhos}
+	Toda a frequência da UC de Aprendizagem de Computacional foi uma oportunidade de me inteirar das diferentes técnicas da área.
+Após este conhecimento básico, passei uma boa parte do mês de Outubro a fazer uma pesquisa sobre implementações destas técnicas em Hardware, e descobri que as técnicas básicas inicias de SVM e Médias k-NN -- presentes na proposta inicial -- são já bastante trabalhadas, sendo pertinente escolher uma nova área de foco.
+	Assim, depois da primeira reunião presencial com o Orientador, foquei-me durante o mês de Novembro em encontrar técnicas que, apesar de serem estado-da-arte, não tivessem sido exploradas na vertente de implementação em hardware, ou em caso positivo, isso estivesse ainda no começo.
+Assim, depois de uma pesquisa bibliográfica nos sites \href{http://www.engineeringvillage.com/}{\textit{Engineering Village}}, fui conduzido a estas técnicas de Redes Neuronais Recorrentes.
+Procurei, também, informação de alto nível no \href{http://people.idsia.ch/~juergen/rnn.html}{site} do Prof.
+Jürgen Schmidhuber (\textit{Università della Svizzera italiana}, Suiça), um dos autores deste algoritmo.
+Após esta fase preliminar de recolha de informação, e de ter trocado impressões com o Orientador e o Co-Orientador, sintetizei toda essa informação neste Relatório de Progresso.
+
+	\section{Pontos a Concretizar no Relatório Seguinte}
+	Para o Relatório Final de PDI, é necessário fazer um levantamento exaustivo de todo o estado da arte neste tipo de redes, bem como uma pesquisa bibliográfica que permita compreender mais a fundo os seus detalhes teóricos, o que será de extrema importância para conseguir reproduzi-las eficazmente em hardware.
+
+	Além disso, é necessário identificar a melhor \textit{framework} para desenvolver o modelo conceptual de alto-nível, permitindo assim uma implementação eficiente e comparável com todo o trabalho na área, assim como dos \textit{benchmarks} mais adequados para testar a eficiência da implementação hardware.
+Por último, convém seleccionar uma aplicação (numa área de reconhecimento de fala, produção músical, entre outras) que permita justificar a pertinência da aplicação prática desta implementação.
 

### 2015-11-29 progReport/report.tex
Revised version of the report

--- 2015-11-26 progReport/report.tex
@@
     %%--Input Português/(...)--%%
-    \usepackage[english]{babel}
+    \usepackage[portuguese]{babel}
     \usepackage[utf8]{inputenc}
@@
 
+    %%--Gestão de Referências Bibliográficas--%%
+    \usepackage[backend=biber, style=ieee]{biblatex}
+    \addbibresource{references.bib}
     %%--Packages de Tipos de Letra--$$
@@
 \rmfamily\pagestyle{fancy}
-\section{Análise Introdutória do Problema a Tratar}
+\begin{center}
+	\Large \textit{Dynamically Reconfigurable Multi-Classifier Architecture on FPGA}
+\end{center}
+\section{Análise Introdutória do Problema a Tratar}\label{sec:analiseProb}
+Esta Tese de Dissertação consistirá no projeto e implementação de um Acelerador de Hardware reconfigurável -- baseado em FPGA\footnote{\textit{Field-Programmable Gate Array}, um tipo de Circuitos Integrados reconfiguráveis em Software, cuja estrutura é \textbf{descrita} usando linguagens como Verilog ou VHDL} -- para estruturas de Aprendizagem Computacional, nomeadamente \textbf{Redes Neuronais Recursivas}.
+	\subsection{Contexto}
+	As \textbf{Redes Neuronais Artificiais} são uma das estruturas de aprendizagem mais populares do mundo da Inteligência Artificial.
+Tal como o nome sugere, são estruturas cuja operação é inspirada na forma como os blocos constituintes do nosso cérebro, os neurónios, operam.
+Apesar da sua elevada performance, um dos seus \textit{shortcomings} é o facto de não conseguirem reter informação de eventos anteriores na sua decisão atual, i.e.
+perante um vetor de entrada $x_i$, a decisão sobre o vetor de saída $y_i$ não possui influência direta de entradas anteriores $x_{i-1}, x_{i-2}, ...$, sendo assim desajustada para o processamento de séries temporais de dados, em que entradas atuais têm uma elevada dependência informacional e sintatica de entradas anteriores, como na análise de texto, e a predição de valores futuros de sinais deterministicos.
+
+	Uma das formas de melhorar o aspeto anterior, foi a introdução nos anos 80 de \textbf{elementos de memória}, sendo isto feito atrvés do \textit{feedback} das camadas exteriores de decisão para as camadas interiores, ou escondidas.
+A estrutura que utilizarei neste trabalho serão as \textit{Long Short-Term Memory Networks} \autocite{hochreiter-lstm}, propostas por Hochreiter et al.
+em 1997, e que desde então é uma das técnicas \textit{state-of-the-art} na área da Aprendizagem Profunda (\textit{Deep Learning}), tendo obtido a melhor classificação de sempre no reconhecimento de caligrafia \autocite{hochreiter-hrr} em 2009, reconhecimento de fala \autocite{speechRec-lstm}, entre outros.
+
+	\subsection{Motivação}
+	Todas estas técnicas são geralmente implementadas em Processadores Convencionais, fazendo uso de linguagens de alto ou baixo nível.
+Desta forma, todo o paralelismo real inerente às estruturas de cálculo fica limitado ao número de \textit{cores} físicos presentes na máquina, que hoje em dia não excedem os 4 ou 8 nos dispositivos convencionais usados pela maioria das pessoas (Computadores Pessoais ou dispositivos móveis).
+Assim, de forma a paralelizar completamente a estrutura e, por conseguinte, aumentar consideravelmente a performance, uma solução é a utilização de uma GPU ou FPGA, perimitindo a realização de aprendizagens cada vez mais complexas, e cada vez mais depressa.
+Além disso, quando a taxa de amostragem de dados de entrada é muito elevada, a solução convencional de CPU não é escalavel e apresenta sérias limitações.
+
+	Apesar de já haver trabalho de adaptação de outros algoritmos de aprendizagem para Hardware, a área do \textit{Deep Learning} ainda é um campo pouco explorado no Hardware, sendo que no caso específico de LSTM, ainda existe apenas um artigo \autocite{fpgalstm}, submetido ainda este mês, que explora uma primeira implementação desta técnica em FPGA para a geração automática de texto.
+
+	Ainda que os CPU's convencionais sejam insubstituíveis para as funções normais a que geralmente estão afetos, é possível que possam haver co-processadores baseados nestas estruturas que acelerem certas tarefas de reconhecimento automático, sem comprometer a performance do sistema global.
+Além disso, foi provado  que as Redes Neuronais Recorrentes são \textit{Turing Complete} \autocite{turingcompletenessRNN}\autocite{turingcompletenessRNN2}, significando, por isso, que podem simular o funcionamento de qualquer programa que possa ser executado num CPU convencional.
+
+	\subsection{Objetivos}
+	Assim, tendo como base o trabalho apresentado em \autocite{fpgalstm}, a ideia é construir um sistema que, além daquele, permita uma reconfiguração completa, em tempo real, da topologia da rede, e a subsituição das células básicas por outras com variações na sua estrutura.
+
+	Será feito, em primeiro lugar, um modelo conceptual do sistema numa linguagem de alto-nível que permita validar o funcionamento dos IP Cores que forem posteriormente desenvolvidos em Verilog, e que depois serão sintetizados para a plataforma-alvo FPGA.
+Além disso, será utilizado um ou dois \textit{Benchmarks} conhecidos para, por um lado, comparar o desempenho do sistema FPGA versus a implementação em CPU e, por outro, comparar o desempenho da minha implementação com outros desempenhos de referência.
+Seria igualmente interessante escolher uma aplicação-tipo destas Redes, e testá-la na minha implementação, para poder observar o seu funcionamento prático.
 
@@
 		\item
-			\textbf{Orientador --} O \href{https://sigarra.up.pt/feup/pt/func_geral.formview?p_codigo=210963}{Professor João Canas Ferreira}, professor Auxiliar do Departamento de Engenharia Electrotécnica e de Computadores da Faculdade de Engenharia da Universidade do Porto, e meu antigo docente da UC de \href{https://sigarra.up.pt/feup/pt/UCURR_GERAL.FICHA_UC_VIEW?pv_ocorrencia_id=352359}{Projecto de Circuitos VLSI}.
-
+			\textbf{Orientador --} O \href{https://sigarra.up.pt/feup/pt/func_geral.formview?p_codigo=210963}{Professor João Canas Ferreira}, professor Auxiliar do Departamento de Engenharia Eletrotécnica e de Computadores da Faculdade de Engenharia da Universidade do Porto, e meu antigo docente da UC de \href{https://sigarra.up.pt/feup/pt/UCURR_GERAL.FICHA_UC_VIEW?pv_ocorrencia_id=352359}{Projeto de Circuitos VLSI}.
+		\item
 			\textbf{Co-Orientador --} O \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, candidato a PhD em Ciência de Computadores, na área de Inteligência Artificial, na Universidade de Cambridge, Reino Unido.
@@
 			\textbf{Reunião por Skype (15 de Setembro de 2015) --} Nesta reunião, o Prof.
-Canas deu-me a conhecer alguns projectos em concreto que tinha em mente, e a abordagem ao problema que deveria ser seguida.
-Acordamos, então, que ele iria submeter uma \href{https://sigarra.up.pt/feup/pt/estagios_empresas.ver_dados_proposta?p_id=191431&pv_perfil=ALU&p_aluno_id=116261}{Proposta de Tema de Dissertação}, mas não seria imediatamente alocada para mim, dado que eu gostaria de dar uma vista de olhos pelas restantes Temas, e só depois tomar uma decisão final e porque a minha média me permitiria escolher o tema em concurso normal, sem o problema de eu ser afastado pelo processo de seriação.
+Canas deu-me a conhecer alguns projetos em concreto que tinha em mente, e a abordagem ao problema que deveria ser seguida.
+Acordamos, então, que ele iria submeter uma \href{https://sigarra.up.pt/feup/pt/estagios_empresas.ver_dados_proposta?p_id=191431&pv_perfil=ALU&p_aluno_id=116261}{Proposta de Tema de Dissertação}, mas não seria imediatamente alocada para mim, dado que eu gostaria de dar uma vista de olhos pelos restantes Temas, e só depois tomar uma decisão final, dado que a minha média me permitiria escolher o tema em concurso normal, sem o problema de eu ser afastado pelo processo de seriação.
 
@@
 			\textbf{Submissão da minha lista de preferências no Concurso Normal (27 de Setembro de 2015) --} Neste dia, submeto a minha lista de preferências de escolha de tema de dissertação, em que me decido definitivamente por este tema.
-Informei, por email,  o Prof.
-Canas da mesma escolha.
+Informei o Prof.
+Canas, por email, da minha escolha.
 Sugiro a possibilidade de se incluir como co-orientador o Ivo Timóteo.
@@
 		\item
-			\textbf{Reunião Presencial (5 de Novembro de 2015) --} Discussão do plano de trabalhos, dos objectivos a serem atingidos e das metodologias a serem desenvolvidas.
-Nesta reunião, apresentei também alguns resultados preliminares da minha primeira pesquisa sobre a matéria, e o Orientador sugere, adicionalmente, a incorporação dos eventuais resultados da dissertação num projecto em desenvolvimento no INESC, como uma possibilidade.
+			\textbf{Reunião Presencial (5 de Novembro de 2015) --} Discussão do plano de trabalhos, dos objetivos a serem atingidos e das metodologias a serem desenvolvidas.
+Nesta reunião, apresentei também alguns resultados preliminares da minha primeira pesquisa sobre a matéria, e o Orientador sugere, adicionalmente, a incorporação dos eventuais resultados da dissertação num projeto em desenvolvimento no INESC, como uma possibilidade.
 
@@
 
+	\subsection{Sequência Temporal de Trabalhos}
+	Toda a frequência da UC de Aprendizagem de Computacional foi uma oportunidade de me inteirar das diferentes técnicas da área.
+Após este conhecimento básico, passei uma boa parte do mês de Outubro a fazer uma pesquisa sobre implementações destas técnicas em Hardware, e descobri que as técnicas básicas iniciais de SVM e Médias k-NN -- presentes na proposta inicial -- são já bastante trabalhadas, sendo científicamente mais interessante escolher uma nova área de foco.
+
+	Assim, depois da primeira reunião presencial com o Orientador, foquei-me durante o mês de Novembro em encontrar técnicas que, apesar de serem estado-da-arte, não tivessem sido exploradas na vertente de implementação em hardware, ou em caso positivo, isso estivesse ainda no começo: depois de uma pesquisa no site \href{http://www.engineeringvillage.com/}{\textit{Engineering Village}}, fui conduzido a estas técnicas de Redes Neuronais Recorrentes.
+Procurei, também, informação de alto nível no \href{http://people.idsia.ch/~juergen/rnn.html}{site} do Prof.
+Jürgen Schmidhuber (\textit{Università della Svizzera italiana}, Suiça), um dos autores das \textit{Long Short-Term Memory Networks}.
+Após esta fase preliminar de recolha de informação, e de ter trocado impressões com o Orientador e o Co-Orientador, sintetizei toda essa informação neste Relatório de Progresso.
+
+	\section{Pontos a Concretizar no Relatório Seguinte}
+	Para o Relatório Final de PDI, é necessário fazer um levantamento exaustivo de todo o estado da arte neste tipo de redes, bem como uma pesquisa bibliográfica que permita compreender mais a fundo os seus detalhes teóricos, o que será de extrema importância para conseguir reproduzi-las eficazmente em hardware.
+
+	Além disso, é necessário identificar a melhor \textit{framework} para desenvolver o modelo conceptual de alto-nível, permitindo assim uma implementação eficiente e comparável com o restante trabalho já existente na área, assim como dos \textit{benchmarks} mais adequados para testar a eficiência da implementação hardware.
+Por último, convém seleccionar uma aplicação (numa área de reconhecimento de fala, produção músical, entre outras) que permita justificar a pertinência da aplicação prática desta implementação.
+
+	\printbibliography
 \end{document}

### 2016-01-11 prepReport/chapter1.tex
Advanced Preparation Report. Python module ready, implements BPTT. Testfile sums two integers

\chapter{Introdução} \label{chap:intro}

Este documento ilustra o formato a usar em dissertações na \Feup.
São dados exemplos de margens, cabeçalhos, títulos, paginação, estilos
de índices, etc. 
São ainda dados exemplos de formatação de citações, figuras e tabelas,
equações, referências cruzadas, lista de referências e índices.
Este documento não pretende exemplificar conteúdos a usar.

Uma recolha sobre as normas existentes pode ser encontrada em~\citet{kn:Mat93}.

Neste primeiro capítulo ilustra-se a utilização de citações e de
referências bibliográficas.

\section{Secção Exemplo} \label{sec:context}

Para além de dar um exemplo de utilização de uma citação, o parágrafo
seguinte introduz uma referência que pode ser consultada, entre muitas
outras referências bibliográficas interessantes~\citep{kn:Tha01,kn:PP05}.

\begin{quote}
  ``Like the Abstract, the Introduction should be written to engage the
  interest of the reader. It should also give the reader an idea of
  how the dissertation is structured, and in doing so, define the
  thread of the contents.''~\citep[chap.\ Introduction]{kn:Tha01} 
\end{quote}

Lorem ipsum~\citep{kn:Lip08} dolor sit amet, consectetuer adipiscing
elit. 
Sed eget nunc. Phasellus interdum, risus viverra mollis laoreet, felis
justo iaculis ante, eget ornare purus augue non urna. Nam in magna. In a
est. Phasellus a tellus vitae enim vehicula imperdiet. Etiam sit amet
elit. In hac habitasse platea dictumst. Quisque eget turpis vel felis
elementum tempus. Curabitur sit amet tortor id libero dapibus
pretium. Integer mattis eros eu lorem. Duis erat tellus, porttitor
sed, blandit eget, fringilla et, lacus. Phasellus tristique nibh nec
orci. Mauris sed leo. Suspendisse fringilla tempor dolor. Donec sapien
enim, congue in, porta et, sollicitudin in, quam. Curabitur semper,
mauris ut vestibulum eleifend, diam ipsum tincidunt quam, et
vestibulum velit mauris ut risus.

Sed eget libero. Nulla facilisi. Proin eget tortor. Morbi
gravida. Donec arcu risus, blandit a, rutrum at, ornare ut,
nisl. Etiam consectetuer tortor eu odio. Etiam blandit molestie
ligula. Nulla facilisi. Nam a augue non justo laoreet hendrerit. Nam
aliquam, purus eu ultricies dictum, urna purus posuere neque, vel
tempus tellus enim a arcu.

Pellentesque congue sapien in ligula. Nulla nec mi sed augue congue
tristique. Cras pretium. Pellentesque lobortis, libero id adipiscing
auctor, orci massa vehicula nulla, vel ullamcorper tortor ipsum ut
elit. Morbi rhoncus, dui sed tristique volutpat, lorem felis euismod
lorem, vel tristique nibh arcu vel eros. Nunc tempor. Sed et erat a
tortor fermentum consequat. Cras varius nisl accumsan libero. Aliquam
faucibus, justo ut pharetra blandit, est nibh ultrices est, vel
facilisis quam lectus ut enim. Aliquam convallis, nibh non bibendum
pharetra, nisl velit lobortis orci, ac ultrices neque sem viverra
massa. Donec malesuada. Aliquam ligula. Fusce in nisl. Etiam lacinia
est quis velit. Maecenas massa. Maecenas sed pede. Nulla
sodales. Etiam vitae erat. Duis tristique sem sit amet libero. Ut a
libero nec ligula tempus mattis.

\section{Outra Secção Exemplo} \label{sec:goals}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Morbi sit
amet nibh. Fusce faucibus, enim vel ultrices ornare, est mauris
ultricies velit, vitae consequat sem erat vel nunc. Nam libero eros,
mattis eget, sagittis nec, imperdiet at, sapien. Aliquam lacus. Aenean
adipiscing nibh in orci. Aliquam vestibulum, elit at fringilla
dignissim, metus diam lobortis urna, a laoreet nunc odio ac ipsum. Sed
at urna. Integer vehicula fringilla augue. Nulla lacus eros, rhoncus
sit amet, posuere ut, vehicula ac, nibh. Ut eleifend, eros eu placerat
vehicula, justo turpis blandit dolor, eu tincidunt felis risus at
ante. Aenean suscipit nisl eget eros. Ut laoreet libero eget
enim. Cras tempus pellentesque felis. Vestibulum vitae erat ac nibh
posuere eleifend.

Integer nec quam. Sed fermentum. Nunc vitae leo. Etiam sit amet
quam. Nunc vestibulum massa in mauris. Duis eget nulla. Fusce
ultricies arcu eu nibh volutpat feugiat. Maecenas urna pede, commodo
quis, porta eu, bibendum elementum, pede. Sed eros massa, molestie
eget, mattis non, rutrum ac, magna. Duis dui. Maecenas eget tortor ut
dolor semper mattis. Maecenas auctor, tellus et ultricies tempor, elit
est placerat lacus, in posuere mauris lorem et arcu.

Nulla nec eros et pede vehicula aliquam. Aenean sodales pede vel
ante. Fusce sollicitudin sodales lacus. Maecenas justo mauris,
adipiscing vitae, ornare quis, convallis nec, eros. Etiam laoreet
venenatis ipsum. In tellus odio, eleifend ac, ultrices vel, lobortis
sed, nibh. Fusce nunc augue, dictum non, pulvinar sed, consectetuer
eu, ipsum. Vivamus nec pede. Pellentesque pulvinar fringilla dolor. In
sit amet pede. Proin orci justo, semper vel, vulputate quis, convallis
ac, nulla. Nulla at justo. Mauris feugiat dolor. Etiam posuere
fermentum eros. Morbi nisl ipsum, tempus id, ornare quis, mattis id,
dolor. Aenean molestie metus suscipit dolor. Aliquam id lectus sed
nisl lobortis rhoncus. Curabitur vitae diam sed sem aliquet
tempus. Sed scelerisque nisi nec sem.

\section{Estrutura da Dissertação} \label{sec:struct}

Para além da introdução, esta dissertação contém mais x capítulos.
No capítulo~\ref{chap:sota}, é descrito o estado da arte e são
apresentados trabalhos relacionados. 
No capítulo~\ref{chap:chap3}, ipsum dolor sit amet, consectetuer
adipiscing elit.
No capítulo~\ref{chap:chap4} praesent sit amet sem. 
No capítulo~\ref{chap:concl}  posuere, ante non tristique
consectetuer, dui elit scelerisque augue, eu vehicula nibh nisi ac
est.

### 2016-01-11 prepReport/chapter2.tex
Advanced Preparation Report. Python module ready, implements BPTT. Testfile sums two integers

\chapter{Revisão Bibliográfica} \label{chap:sota}

Neste capítulo é ilustrada a introdução de entradas no índice
remissivo e são feitas diversas referências bibliográficas.

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Donec a
eros. Phasellus non nulla non massa venenatis convallis. In
porta. Mauris quis magna.

\section{Introdução}

Neste capítulo usa-se texto de um artigo apresentado na Conferência
XATA2006~\citep{kn:MVL06-xata}.

Nos últimos tempos têm surgido diversas soluções, apresentadas por
empresas do sector Auto\-ma\-ção de Sistemas para a disponibilização de
sistemas \scadadms{} na \textit{Web}.

Fusce risus mi, tristique eu, consectetuer id, auctor sed, elit. Donec
laoreet. Duis consectetuer interdum libero. Etiam eu orci. In eu
arcu. Fusce luctus diam eget lectus. Duis interdum lacus sed
ligula. Proin vestibulum felis eget lacus. Vivamus vestibulum, tellus
ut congue viverra, mauris lacus tempor turpis, eu congue nisi magna at
dolor. Ut molestie vehicula libero. Praesent in neque sed risus tempus
ornare. Donec hendrerit, erat eu semper aliquam, pede nulla dapibus
risus, ut pretium orci pede et neque.
Etiam eget tortor a metus convallis viverra. Quisque eget nisi sed
orci facilisis interdum. Aliquam non felis.

\section{Secção Exemplo}\label{sec:dialecto}

\emph{Scalable Vector Graphics}\index{SVG}\index{XML!SVG} é uma
linguagem em formato XML que descreve gráficos de duas dimensões. 
Este formato padronizado pela W3C (\emph{World Wide Web Consortium})
é livre de patentes ou direitos de autor e está totalmente
documentado, à semelhança de outros W3C
standards~\citep{kn:svgdoc}.

Sendo uma linguagem XML, o \svg{} herda uma série de vantagens: a
possibilidade de transformar \svg{} usando técnicas como
XSLT\index{XML!XSLT}, de embeber \svg{} em qualquer documento
XML\index{XML} usando \textit{namespaces} ou até de  
estilizar \svg{} recorrendo a CSS\index{CSS} (\emph{Cascade Style Sheets}). 
De uma forma geral, pode dizer-se que \svg{}s interagem bem com as
atuais tecnologias ligadas ao XML e à Web, tal como referido
em~\citep{kn:svgibm,kn:svgw3c}.

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Donec a
eros. Phasellus non nulla non massa venenatis convallis. In
porta. Mauris quis magna. Proin mauris eros, aliquet id, eleifend
vitae, semper quis, erat. Aliquam id lectus non odio dignissim
blandit. Vestibulum porttitor arcu ut ligula. Nunc quis
erat. Curabitur ipsum tortor, ornare vitae, dapibus pretium, hendrerit
sed, urna. Vestibulum ante ipsum primis in faucibus orci luctus et
ultrices posuere cubilia Curae; Phasellus bibendum, nulla eget varius
aliquam, tortor nulla sollicitudin quam, vel vestibulum nisl magna at
sem. Aliquam velit sapien, ultrices viverra, tempus quis, ultrices at,
dui. Aliquam sit amet justo. Quisque tristique, metus eu iaculis
sagittis, urna leo bibendum diam, a ultricies sem diam a augue. Mauris
consectetuer, libero vel euismod tincidunt, nisi metus viverra ante,
quis pretium sapien odio nec risus. Nunc semper auctor
nulla\footnote{Exemplo de nota de rodapé.}.

\subsection{Subsecção Exemplo} \label{batik}

Batik é um conjunto de bibliotecas baseadas em \textit{Java} que
permitem o uso de imagens \svg{} (visualização, geração ou
manipulação) em aplicações ou \textit{applets}~\citep{kn:batik}.  
O projeto Batik\index{Batik} destina-se a fornecer ao programador
alguns módulos que permitem desenvolver soluções especificas usando
\svg~\citep{kn:svgdoc}.

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Nunc eu
nulla. Pellentesque vitae nibh ultrices quam iaculis
convallis. Aliquam purus eros, varius eget, volutpat sodales,
imperdiet nec, lacus. Curabitur in elit sed sem rutrum posuere. Class
aptent taciti sociosqu ad litora torquent per conubia nostra, per
inceptos himenaeos. Duis sem. Praesent ultricies odio vel
sapien. Integer faucibus malesuada libero. Cras semper, dolor id
ullamcorper varius, magna risus volutpat felis, id pellentesque nulla
ante at erat. Integer sodales.

Quisque sit amet odio. In at risus sit amet turpis interdum
posuere. Maecenas iaculis vehicula sem. Ut leo arcu, malesuada vel,
imperdiet id, dignissim a, purus. Duis eleifend, lectus non venenatis
dignissim, risus libero imperdiet mi, nec gravida massa libero sed
mauris. Nullam lobortis libero non sapien. Integer convallis iaculis
erat. Morbi dictum. Ut ultrices pellentesque velit. Cras ac
ante. Etiam in neque tincidunt lacus gravida vehicula. Proin et nisi.

Vivamus non nunc nec risus tempor varius. Quisque bibendum mi at
dolor. Aliquam consectetuer condimentum risus. Aliquam luctus pulvinar
sem. Duis aliquam, urna et vulputate tristique, dui elit aliquet nibh,
vel dignissim magna turpis id sapien. Duis commodo sem id
quam. Phasellus dolor. Class aptent taciti sociosqu ad litora torquent
per conubia nostra, per inceptos himenaeos.

\subsection{Subsecção Exemplo}

Loren ipsum dolor sit amet, consectetuer adipiscing elit. 
Praesent sit amet sem. Maecenas eleifend facilisis leo. Vestibulum et
mi. Aliquam posuere, ante non tristique consectetuer, dui elit
scelerisque augue, eu vehicula nibh nisi ac est. Suspendisse elementum
sodales felis. Nullam laoreet fermentum urna.

Duis eget diam. In est justo, tristique in, lacinia vel, feugiat eget,
quam. Pellentesque habitant morbi tristique senectus et netus et
malesuada fames ac turpis egestas. Fusce feugiat, elit ac placerat
fermentum, augue nisl ultricies eros, id fringilla enim sapien eu
felis. Vestibulum ante ipsum primis in faucibus orci luctus et
ultrices posuere cubilia Curae; Sed dolor mi, porttitor quis,
condimentum sed, luctus in.

\section{Resumo ou Conclusões}

Aliquam erat volutpat. Nunc pede ipsum, porttitor eu, bibendum non,
bibendum nec, nisl. Maecenas eget mauris. Nullam pulvinar. Curabitur
rutrum commodo est. Nam sapien pede, interdum eu, accumsan ultrices,
venenatis sit amet, tellus. Praesent ac ante bibendum enim varius
suscipit. Donec enim. Proin nisi. Quisque libero turpis, varius ut,
elementum vel, pulvinar sed, nunc.

### 2016-01-11 prepReport/chapter3.tex
Advanced Preparation Report. Python module ready, implements BPTT. Testfile sums two integers

\chapter{Capítulo Exemplo}\label{chap:chap3}

Neste capítulo apresentam-se exemplos de formatação de figuras e
tabelas, equações e referências cruzadas.

Maecenas eleifend facilisis leo. Vestibulum et
mi. Aliquam posuere, ante non tristique consectetuer, dui elit
scelerisque augue, eu vehicula nibh nisi ac est. 
Suspendisse elementum sodales felis. Nullam laoreet fermentum urna.

\section{Introdução}

Apresenta-se de seguida um exemplo de equação, completamente fora do contexto:
\begin{eqnarray}
CIF_1: \hspace*{5mm}F_0^j(a) &=& \frac{1}{2\pi \iota} \oint_{\gamma} \frac{F_0^j(z)}{z - a} dz\\
CIF_2: \hspace*{5mm}F_1^j(a) &=& \frac{1}{2\pi \iota} \oint_{\gamma} \frac{F_0^j(x)}{x - a} dx \label{eq:cif}
\end{eqnarray}

Na Equação~\ref{eq:cif} lorem ipsum dolor sit amet, consectetuer
adipiscing elit. Suspendisse tincidunt viverra elit. Donec tempus
vulputate mauris. Donec arcu. Vestibulum condimentum porta
justo. Curabitur ornare tincidunt lacus. Curabitur ac massa vel ante
tincidunt placerat. Cras vehicula semper elit. Curabitur gravida, est
a elementum suscipit, est eros ullamcorper quam, sed cursus velit
velit tempor neque. Duis tempor condimentum ante. Nam
sollicitudin. Vestibulum adipiscing, orci eu tempor dapibus, risus
sapien porta metus, et cursus leo metus eget nibh.

Pellentesque rutrum, sapien at viverra facilisis, metus eros blandit
sem, quis dictum erat metus eget erat. Vivamus malesuada dapibus
nulla. Maecenas nec purus. Suspendisse auctor mattis augue. Phasellus
enim nisi, iaculis sit amet, pellentesque a, iaculis in, dui. Integer
risus.

\section{Secção Exemplo}

A arquitectura do visualizador assenta sobre os seguintes conceitos
base~\citep{kn:ZPMD97}:

\begin{itemize}
\item \textbf{Componentes} --- Suspendisse auctor mattis augue \emph{push};
\item \textbf{Praesent} --- Sit amet sem maecenas eleifend facilisis leo;
\item \textbf{Pellentesque} --- Habitant morbi tristique senectus et netus.
\end{itemize}

\subsection{Exemplo de Figura}

É apresentado na Figura~\ref{fig:arch} da página~\pageref{fig:arch} um
exemplo de figura flutuante que deverá ficar no topo da página.

\begin{figure}[t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.86\textwidth]{puzzle}
    \caption{Arquitectura da Solução Proposta}
    \label{fig:arch}
  \end{center}
\end{figure}

Loren ipsum dolor sit amet, consectetuer adipiscing elit. 
Praesent sit amet sem. Maecenas eleifend facilisis leo. Vestibulum et
mi. Aliquam posuere, ante non tristique consectetuer, dui elit
scelerisque augue, eu vehicula nibh nisi ac est. Suspendisse elementum
sodales felis. Nullam laoreet fermentum urna.

Duis eget diam. In est justo, tristique in, lacinia vel, feugiat eget,
quam. Pellentesque habitant morbi tristique senectus et netus et
malesuada fames ac turpis egestas. Fusce feugiat, elit ac placerat
fermentum, augue nisl ultricies eros, id fringilla enim sapien eu
felis. Vestibulum ante ipsum primis in faucibus orci luctus et
ultrices posuere cubilia Curae; Sed dolor mi, porttitor quis,
condimentum sed, luctus in.

\subsection{Exemplo de Tabela}

É apresentado na Tabela~\ref{tab:exemplo} um exemplo de tabela
flutuante que deverá ficar no topo da página.

\begin{table}[t]
  \centering
  \caption{Tabela Exemplo}
\begin{tabular}{|c|r@{.}lr@{.}lr@{.}l||r|}
	\hline
\multicolumn{8}{|c|}
	{\rule[-3mm]{0mm}{8mm}Iteração $k$ de $f(x_n)$} \\
\textbf{\em k}
	& \multicolumn{2}{c}{$x_1^k$}
	& \multicolumn{2}{c}{$x_2^k$}
	& \multicolumn{2}{c||}{$x_3^k$}
	& comentários \\ \hline \hline
0   & -0&3                 & 0&6                 &  0&7   & - \\
1   &  0&47102965 & 0&04883157 & -0&53345964  & $\delta<\epsilon$ \\
2   &  0&49988691 & 0&00228830 & -0&52246185  & $\delta < \varepsilon$ \\
3   &  0&49999976 & 0&00005380 & -0&523656   &   $N$ \\
4   &  0&5                 & 0&00000307 & -0&52359743  & \\
\vdots	& \multicolumn{2}{c}{\vdots}
	& \multicolumn{2}{c}{$\ddots$}
	& \multicolumn{2}{c||}{\vdots}  & \\
7   &  0&5   & 0&0    & \textbf{-0}&\textbf{52359878}
		 & $\delta<10^{-8}$ \\ \hline
\end{tabular}
  \label{tab:exemplo}
\end{table}

Loren ipsum dolor sit amet, consectetuer adipiscing elit. 
Praesent sit amet sem. Maecenas eleifend facilisis leo. Vestibulum et
mi. Aliquam posuere, ante non tristique consectetuer, dui elit
scelerisque augue, eu vehicula nibh nisi ac est. Suspendisse elementum
sodales felis. Nullam laoreet fermentum urna.

Duis eget diam. In est justo, tristique in, lacinia vel, feugiat eget,
quam. Pellentesque habitant morbi tristique senectus et netus et
malesuada fames ac turpis egestas. Fusce feugiat, elit ac placerat
fermentum, augue nisl ultricies eros, id fringilla enim sapien eu
felis. Vestibulum ante ipsum primis in faucibus orci luctus et
ultrices posuere cubilia Curae; Sed dolor mi, porttitor quis,
condimentum sed, luctus in.

\section{Secção Exemplo}

Loren ipsum dolor sit amet, consectetuer adipiscing elit. 
Praesent sit amet sem. Maecenas eleifend facilisis leo. Vestibulum et
mi. Aliquam posuere, ante non tristique consectetuer, dui elit
scelerisque augue, eu vehicula nibh nisi ac est. Suspendisse elementum
sodales felis. Nullam laoreet fermentum urna.

Duis eget diam. In est justo, tristique in, lacinia vel, feugiat eget,
quam. Pellentesque habitant morbi tristique senectus et netus et
malesuada fames ac turpis egestas. Fusce feugiat, elit ac placerat
fermentum, augue nisl ultricies eros, id fringilla enim sapien eu
felis. Vestibulum ante ipsum primis in faucibus orci luctus et
ultrices posuere cubilia Curae; Sed dolor mi, porttitor quis,
condimentum sed, luctus in.

\section{Resumo}

Pellentesque habitant morbi tristique senectus et netus et
malesuada fames ac turpis egestas. Fusce feugiat, elit ac placerat
fermentum, augue nisl ultricies eros, id fringilla enim sapien eu
felis. Vestibulum ante ipsum primis in faucibus orci luctus et
ultrices posuere cubilia Curae; Sed dolor mi, porttitor quis,
condimentum sed, luctus in.

### 2016-01-11 prepReport/chapter4.tex
Advanced Preparation Report. Python module ready, implements BPTT. Testfile sums two integers

\chapter{Mais um Capítulo}\label{chap:chap4}

Neste capítulo mostra-se apenas o formato da dissertação.

Ipsum dolor sit amet, consectetuer
adipiscing elit.  Praesent sit amet sem. 
Maecenas eleifend facilisis leo. Vestibulum et
mi. Aliquam posuere, ante non tristique consectetuer, dui elit
scelerisque augue, eu vehicula nibh nisi ac est. 
Suspendisse elementum sodales felis. 
Nullam laoreet fermentum urna.

\section{Secção Exemplo}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Integer
hendrerit commodo ante. Pellentesque nibh libero, aliquam at, faucibus
id, commodo a, velit. Duis eleifend sem eget leo. Morbi in
est. Suspendisse magna sem, varius nec, hendrerit non, tincidunt quis,
quam. Aenean congue. Vivamus vel est sit amet sem iaculis
posuere. Cras mollis, enim vel gravida aliquam, libero nunc
ullamcorper dui, ullamcorper sodales lectus nulla sed urna. Morbi
aliquet porta risus. Proin vestibulum ligula a purus. Maecenas a
nulla. Maecenas mattis est vitae neque auctor tempus. Etiam nulla dui,
mattis vitae, porttitor sed, aliquet ut, enim. Cras nisl magna,
aliquet et, laoreet at, gravida ac, neque. Sed id est. Nulla dapibus
dolor quis ipsum rhoncus cursus.

Etiam nisi est, dignissim sodales, fermentum id, pulvinar ac,
eros. Duis id orci. Nam pretium nisl ac augue. Ut adipiscing magna
eget est. Curabitur varius. Nulla facilisi. Pellentesque sit amet
neque ac dui accumsan blandit. Donec mauris felis, egestas sit amet,
convallis ac, dignissim quis, dolor. Maecenas cursus tortor vel
leo. Quisque tristique. Nunc augue odio, tincidunt in, dapibus sed,
ultricies sit amet, lorem. In hac habitasse platea dictumst. Praesent
iaculis, lacus hendrerit tempor sodales, libero tellus aliquet orci,
ut rhoncus massa lectus quis erat. Pellentesque quis dolor nec tortor
rhoncus convallis. Aliquam erat volutpat. Fusce placerat, magna eu
imperdiet lobortis, augue massa blandit turpis, a consectetuer quam
arcu sit amet risus. Suspendisse potenti. Praesent sapien metus,
interdum vitae, fermentum id, faucibus ut, lorem. Nunc iaculis purus
id tortor. Aenean risus pede, laoreet ac, tristique sed, lobortis in,
turpis.

Vestibulum et lorem in ligula viverra pharetra. Curabitur quis purus
in urna facilisis bibendum. Pellentesque at arcu accumsan velit
bibendum ornare. Praesent massa. Quisque dolor. In libero. Vestibulum
ac diam id leo feugiat blandit. Donec porta, tellus ac pellentesque
molestie, felis mauris viverra lacus, sed dignissim purus justo eu
justo. Proin iaculis, nunc eu volutpat volutpat, libero purus rutrum
enim, id euismod lacus lorem nec augue. Donec hendrerit lacinia
ante. Integer mollis vulputate orci. In pellentesque, metus pharetra
elementum pharetra, est purus bibendum turpis, eu pretium sapien
libero convallis odio. Cras sodales bibendum risus. Sed mattis nulla
non leo. Nulla nunc. Phasellus egestas sodales massa. Class aptent
taciti sociosqu ad litora torquent per conubia nostra, per inceptos
himenaeos. Pellentesque habitant morbi tristique senectus et netus et
malesuada fames ac turpis egestas. Etiam mi.

\section{Mais uma Secção}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Quisque
purus sapien, interdum ut, vestibulum a, accumsan ullamcorper,
erat. Mauris a magna ut leo porta imperdiet. Donec dui odio, porta in,
pretium non, semper quis, orci. Quisque erat diam, pharetra vel,
laoreet ac, hendrerit vel, enim. Donec tristique luctus risus. Fusce
dolor est, eleifend id, elementum sit amet, varius vitae, neque. Morbi
at augue. Ut sem ligula, auctor vitae, facilisis id, pharetra non,
lectus. Nulla lacus augue, aliquam eget, sollicitudin sed, hendrerit
eu, leo. Suspendisse ac tortor. Mauris at odio. Etiam vehicula. Nam
lacinia purus at nibh. Aliquam fringilla lorem ac justo. Ut nec
enim. Nunc ornare, eros eu facilisis tristique, nisl lorem lacinia
risus, non ullamcorper tellus urna et eros. Quisque eleifend tempus
metus. Nunc ipsum.

Phasellus ullamcorper justo id risus. Nunc in leo. Mauris auctor
lectus vitae est lacinia egestas. Nulla faucibus erat sit amet lectus
varius semper. Praesent ultrices vehicula orci. Nam at metus. Aenean
eget lorem nec purus feugiat molestie. Phasellus fringilla nulla ac
risus. Aliquam elementum aliquam velit. Aenean nunc odio, lobortis id,
dictum et, rutrum ac, ipsum. Aenean tellus magna, lacinia eget,
bibendum ut, interdum sit amet, ipsum. Class aptent taciti sociosqu ad
litora torquent per conubia nostra, per inceptos himenaeos. Mauris
felis lacus, dapibus sit amet, pretium feugiat, aliquet non,
purus. Aliquam elementum, diam quis porttitor gravida, sem sapien
iaculis nulla, ut pharetra odio felis a metus. Nulla lacus ipsum,
tristique ut, dapibus sed, mollis et, justo. Vivamus non ipsum sed
ligula placerat ultrices. Maecenas dictum leo adipiscing
mauris. Vestibulum tristique, lacus a consequat suscipit, nunc dui
sollicitudin arcu, non interdum libero est eget tortor. Ut eget neque
quis leo tempor dictum.

Quisque ullamcorper. Aliquam vel magna. Sed pulvinar dictum
ligula. Sed ultrices dolor ut turpis. Vivamus sagittis orci malesuada
arcu venenatis auctor. Proin vehicula pharetra urna. Aliquam egestas
nunc quis nisl. Donec ullamcorper. Nulla purus. Ut suscipit lacus
vitae dui. Mauris semper. Ut eget sem. Integer orci. Nam vitae dui
eget nisi placerat convallis.

Sed id lorem. Proin gravida bibendum lacus. Sed molestie, urna quis
euismod laoreet, diam dolor dictum diam, vitae consectetuer leo ipsum
id ante. Integer eu lectus non mauris pharetra viverra. In feugiat
libero ut massa. Morbi cursus, lorem sollicitudin blandit semper,
felis magna pellentesque lacus, ut rhoncus leo neque at tellus. Sed
mattis, diam eget eleifend tincidunt, ligula eros tincidunt diam,
vitae auctor turpis est vel nunc. In eu magna. Donec dolor metus,
egestas sit amet, ultrices in, faucibus sed, lectus. Etiam est enim,
vehicula pharetra, porta non, viverra vel, nunc. Ut non sem. Etiam nec
neque. Sed rhoncus, justo id imperdiet pharetra, mi tellus accumsan
neque, vitae volutpat tortor enim in odio. Nunc porta justo a
lorem. Nulla hendrerit odio vitae dolor. Suspendisse eu nisl.

\section{Resumo ou Conclusões}

Proin vehicula pharetra urna. Aliquam egestas
nunc quis nisl. Donec ullamcorper. Nulla purus. Ut suscipit lacus
vitae dui. Mauris semper. Ut eget sem. Integer orci. Nam vitae dui
eget nisi placerat convallis.

### 2016-01-11 prepReport/theorBack.tex
Advanced Preparation Report. Python module ready, implements BPTT. Testfile sums two integers

\chapter{Theoretical Background}\label{chap:theorBack}

\section{Basic Concepts of Machine Learning}
Machine Learning is a subject of Computer Science that studies the development of mathematical formulations that allows software to learn autonomously, without an explicit description of each rule of operation. Its goal is to extract latent characteristic features in datasets that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters automatically from the input data it receives.

Let us consider a practical example. For instance, suppose we want to build a device that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary. We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from it being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language. The approach taken by Machine Learning is diametrically different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 \; \mb{x}_2 \;\mb{x}_2 \; \dots \mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model. Each of the $i$-th data point was previously labelled, and in fact we feed each training data point $\mb{x}_i$ \emph{along} with its corresponding label $t_i$, so that the model can adapt its parameters accordingly to the \emph{target value} it is supposed to classify. This set of labels $\mb{t} = \left[ t_1 \; t_2 \; t_3 \; \dots \; t_N \right]$ is called the \textbf{target vector}.

We are, then, left with the following question: how can the model quantitatively evaluate the quality of its current set of parameters? That could be achieved in a number of ways, but the most usual is using a \textbf{Cost Function} that, as the name suggests, measures the cost of each wrong classification of the model. The model then evolves in a way that minimizes the cost function. A usual choice for the cost function is the \textbf{sum of squares error}. Mathematically, if $y_{\theta}^i = y_{\theta}(x_i)$ is the prediction for the input data point $x_i$ with label $t_i$, given the current set of parameters $\theta$, the cost function using this metric is given by

\begin{equation}\label{eq:costfunctionFund}
	J(\theta) = \frac{1}{2} \sum_{i=1}^N \left( y_{\theta}^i - t_i \right)^2.
\end{equation}

Furthermore, sometimes instead of applying the raw data to the model, we can apply some sort of preprocessing to the data to extract the relevant features from it. For instance, instead of just feeding a raw image, we can perform several operations like edge detection, and apply them in parallel.

The problems described above are, in fact, a subset of the problems that Machine Learning tries to address. These problems are called \textbf{classification problems}, because for each input data point, our model tries to fit it into the most appropriate class. But we can also address \textbf{regression problems} where the output is not limited to a discrete set of values but rather a continuous interval.

In summary, the most typical setting for a Machine Learning problem is having a large \textit{input dataset} which we use to \textit{train} our model (i.e. allowing him to dynamically adapt its set of parameters $\theta$), in order to produce an output label $y_i$ for each of them that minimizes a quality metric, typically defined as a \textit{cost function}. Artificial Neural Networks are one of the tools used in Machine Learning to perform this task, and will be discussed in Section~\ref{sec:theorBack_ann} as somewhat of a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}). These two last networks branch even further from these set of problems, and are usually employed in \textit{Deep Learning} tasks, where we try to extract even higher level information from data at the expense of increased model complexity.

\section{Artificial Neural Networks}\label{sec:theorBack_ann}

Artificial Neural Networks (ANN) are mathematical structures that, as the name suggests, try to mimic the Human Brain. ANN's building blocks, like their biological counterpart, model the high-level behaviour of biological neurons, in the sense that they neglect unnecessary biological aspects (such as modeling all the voltages across the neuron and all its electromagnetic interactions), and only retain its fundamental underlying mathematical function, which is a weighted linear combination of its inputs subject to a \emph{activation function}, i.e. a function that outputs a decision value depending on its inputs.  Mathematically, we have

\begin{equation}
	y = f(\mb{w}^T \mb{x} + b_0)
\end{equation}
where $\mb{w} = \left[ w_1 \; w_2 \;  w_3 \;  \dots \; w_n \right]$ is the input weight vector, $\mb{x} = \left[ x_1 \; x_2 \; x_3 \; \dots \; x_n \right]$ the input vector, $b_0$ is the bias factor and $f(\cdot)$ is the chosen activation function. Furthermore, we call the scalar quantity $a = \mb{w}^T \mb{x} + b_0$ the \textbf{activation}, since its value determines how the activation function will behave. Figure~\ref{fig:modelNeuron} exemplifies the roles of these variables within our neuron model, and compares each part of it with the biological counterpart.

\begin{figure}[!H]
	\begin{subfigure}{0.5\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{figures/neuron.eps}
		\caption{Neural Network Node}
		\label{fig:modelNeuron_a}
	\end{subfigure}
	\begin{subfigure}{0.5\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{figures/multipolarNeuron.png}
		\caption{Biological Neuron Diagram}
		\label{fig:modelNeuron_b}
	\end{subfigure}

	\caption{In Figure~\ref{fig:modelNeuron_a}. each input feature $x_i$ is weighted by its corresponding weight $w_i$. During the training procedure, these weights are adjusted so that the output $y$ approaches the target value. In Figure~\ref{fig:modelNeuron_b}, we see the diagram of an actual multi polar neuron. The dendrites, where the stimuli are received, plays a role similar to that of the input nodes. The axon transmits the signal to the synaptic terminals, that are similar to the $y$ output}
	\label{fig:modelNeuron}
\end{figure}

As far as the activation function is concerned, we can have several types. An immediate choice would be the \textbf{Binary Step Function} that  outputs -1 if the activation is \textbf{below} a given threshold and 1 otherwise. There can also be \textbf{real valued activation functions}, whose output is not binary, but rather that of a continuously differentiable function, such as the logistic sigmoid $\sigma(a) = \frac{1}{1 + \mathrm{e}^{-a}}$ or the hyperbolic tangent $\tanh(a)$. This aspect will prove useful for the usual training methods, that involve the computation of derivatives. In Figure~\ref{fig:activFunc} these activation functions are plotted.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/activFunc.eps}
	\caption{Three different activation functions. As you can see, the hyperbolic tangent has the same extreme value as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision. On the other hand, the sigmoid function goes from zero to one, and is also smooth like the hyperbolic tangent}.
	\label{fig:activFunc}
\end{figure}

A neuron by itself can be thought of as a simple linear regression, where we optimize the weight of each feature according to a target value, or function. While important in some applications, the main interest in ANN is to evaluate increasingly more complex models, and not a simple linear regression. This is achieved by \emph{chaining} nodes to one another, connecting the output of a given node to one of the inputs of another. We call \emph{layers} to a group of these nodes that occupy the same hierarchical position. There can be any number of layers, with any number of nodes, but most implementations generally have 3 layers: the \emph{initial} layer, the \emph{hidden} layer (in the middle) and the \emph{output} layer. Figure~\ref{fig:neuralnet} suggests a possible structure for a 3 layer ANN.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/ann.eps}
	\caption{A three layer ANN. We have omitted some of the connections in the hidden layer, for simplification purposes. $\mb{w}_1$ represents the weight matrix of the input layer, $\mb{w}_2$ the weight matrix of the connections between the input layer and the hidden layer, and $\mb{w}_3$ the weight connections between the hidden and the output layer. $f_{ij}(\dots)$ is the activation function of the $j$-th neuron of the $i$-th layer. Since they can be different, I chose different indexes to each.}
	\label{fig:neuralnet}
\end{figure}

Regarding the training of ANNs, is is performed through a two-step process: first, a \textbf{feed-forward} step where the input is applied, and the activations are evaluated in succession up to the output neurons; then, the perform the \textbf{backpropagation} step, where we calculate the errors in each of the nodes, but now from the output to the input: the weights are updated and optimized using an iterative  method called \textit{Gradient Descent}, where if $\tau$ is the current time step, the next update on the weight matrix $\mb{w}^{(\tau+1)}$ is given by

\begin{equation}\label{eq:gradDesc}
	\mb{w}^{(\tau+1)} = \mb{w}^{(\tau)} - \eta \nabla E(\mb{w}^{(\tau)})
\end{equation}
where $E(\cdot)$ is the error function. As we can see, the weight matrix is moved in the direction that minimizes the error function the most, and $\eta$ controls how fast this is achieved, being the reason why it is called the \textbf{learning rate}.

The computation of gradient of the error function comprises the evaluation of its derivatives with respect to each weight of all network connections, $w_{ij}$. They are

\begin{equation}\label{eq:partialE}
	\frac{\partial E}{\partial w_{ji}} = \delta_j f(a_i)
\end{equation}
where $f(\cdot)$ is the activation function of the neuron and

\begin{equation}\label{eq:deltaj}
	\delta_j = f'(a_j) \sum_k w_{kj} \delta_k .
\end{equation}
The interpretation of these equations is simple. If $w_{ji}$ is the weight of the connection between the neuron $j$ we are considering and a neuron $i$ in a previous layer, then the sum over $k$ relates to all the neurons in the \emph{next} layer to which $j$ connects: this way, since the update on $w_{ji}$, according to~\ref{eq:gradDesc}, is given by

\begin{equation}\label{eq:update}
	w_{ji}^{(\tau+1)} = w_{ji}^{(\tau+1)} - \eta \frac{\partial E}{w_{ji}}
\end{equation}
we see that, from~\ref{eq:partialE}  it simply is the product of the error of the current neuron, $\delta_j$, with the output of the previous neuron $f(a_i)$. In turn, from~\ref{eq:deltaj}, we see the recurrence relationship between it and the weighted sum of all posterior neurons that connect to it. Hence, the name backpropagation is now clear: we are, in fact, propagating the errors backwards into the neuron of interest, weighted by the corresponding weight, but now \emph{backwards} instead of forward, as before. For the output units, the $\delta_j$ is simply the difference between the produced output and the corresponding label for that sample. This two-step process is performed for every data point in out dataset. %%Place where I can include an algorithmic overview?
For a complete proof of the above formulas, see~\citep[chap. 5.3.1]{Bishop2006}.

\section{Recurrent Neural Networks}\label{sec:theorBack_rnn}
A Recurrent Neural Network (RNN) is, essentially, a regular ANN where some neurons (especially in the hidden layer) have \emph{feedback connections} to themselves, i.e. their outputs are fed as inputs. The relevance of this different structure is the possibility to retain \emph{sequence} information about the data. Before, each incoming data point only contributed to the training of the network, but no information about the correlation between themselves and the data points that preceded them did not influence the training step. They were regarded as if no temporal relationship existed, and therefore each data point is conditionally independent of any other. This is obviously not necessarily truth, and in fact there are many cases, where the correlation between data points is high for those closely spaced in time, in which it is actually completely false, as in video signals, audio signals, or other kinds of \emph{temporal sequences} of data. Therefore, the feedback connection of the neuron to himself acts as a kind of \emph{memory element} that takes into account in the present decision, the history of decisions previously taken, and hence the previous data. 
Figure~\ref{fig:recneuron} suggests a possible structure for a neuron of a hidden layer in an RNN, and also an alternate representation, where the structure is unfold through time.

%% METE AQUI A FIGURA

FALAR EM BPTT

Even though RNNs outperform static ANNs in sequence recognition problems~\citep{Bengio1991}, they fail to retain long-term dependencies. Of course that the weight training process is itself a form of memory, but the problem is that the weight update is much slower than the activations~\citep{Yoshua01}, and therefore this memory only retains short-term dependencies. This is because of the so-called \textbf{Vanishing Gradients Problem}~\citep{Yoshua94,Yoshua01}, where the error decays exponentially through time, and the impact of previous incoming data points on the training of the weights quickly decrease.

To overcome this issue, Hochreiter and Schmidhüber proposed, in 1997, a novel approach, called the Long-Short Term Network~\citep{Hoch97}. The next section explains the main structure of this approach, and also how it is trained, serving as a support basis for the work of this thesis.

\section{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}

### 2016-01-11 prepReport/compile.sh

@@ -1,0 +1,4 @@
+pdflatex thesis.tex
+bibtex thesis
+pdflatex thesis.tex
+pdflatex thesis.tex

### 2016-01-11 python/testLSTM.py

@@ -1,0 +1,163 @@
+# -*- coding: utf-8 -*-
+"""
+LSTM Layer Class
+@author: José Pedro Castro Fonseca, University of Porto, PORTUGAL
+"""
+import numpy as np
+import time
+
+# Defines the seed for random() as the current time in seconds
+np.random.seed(round(time.time()))
+
+# Evaluates the Sigmoid Function
+def sigmoid(x):
+    output = 1/(1+np.exp(-x))
+    return output
+
+# Evaluates the derivative of the Sigmoid Func, based on a previous sigmoid() call
+def sigmoidPrime(output):
+    return output*(1-output)
+
+# Evaluates the derivative of the TanH Func, based on a previous np.tanh() call
+def tanhPrime(output):
+    return (1-output**2)
+
+class LSTMlayer :
+
+    def __init__(self, inputUnits, hiddenUnits, learnRate, T):
+        # The Network Parameters, passed by the user
+        self.inputUnits  = inputUnits
+        self.hiddenUnits = hiddenUnits
+        self.learnRate   = learnRate
+        self.T           = T
+        self.t           = 0
+
+        # Initializing the matrix weights
+        #LSTM Block
+        self.Wz = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wi = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wf = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wo = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        
+        self.Rz = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        self.Ri = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        self.Rf = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        self.Ro = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        
+        self.pi = np.random.random((hiddenUnits)) - 0.5
+        self.pf = np.random.random((hiddenUnits)) - 0.5
+        self.po = np.random.random((hiddenUnits)) - 0.5
+        
+        self.bz = np.random.random((hiddenUnits)) - 0.5
+        self.bi = np.random.random((hiddenUnits)) - 0.5
[…]

### 2016-01-11 python/testLSTM.py
Corrected Test File

@@ -1,163 +1,130 @@
-# -*- coding: utf-8 -*-
-"""
-LSTM Layer Class
-@author: José Pedro Castro Fonseca, University of Porto, PORTUGAL
-"""
 import numpy as np
+import matplotlib.pyplot as plt
 import time
+import LSTMlayer
 
-# Defines the seed for random() as the current time in seconds
-np.random.seed(round(time.time()))
+np.random.seed(5)
 
-# Evaluates the Sigmoid Function
+# compute sigmoid nonlinearity
 def sigmoid(x):
     output = 1/(1+np.exp(-x))
     return output
 
-# Evaluates the derivative of the Sigmoid Func, based on a previous sigmoid() call
+# convert output of sigmoid function to its derivative
+def sigmoid_output_to_derivative(output):
+    return output*(1-output)
+
 def sigmoidPrime(output):
     return output*(1-output)
 
-# Evaluates the derivative of the TanH Func, based on a previous np.tanh() call
 def tanhPrime(output):
     return (1-output**2)
 
-class LSTMlayer :
+# training dataset generation
+int2binary = {}
+binary_dim = 32 
 
-    def __init__(self, inputUnits, hiddenUnits, learnRate, T):
-        # The Network Parameters, passed by the user
-        self.inputUnits  = inputUnits
-        self.hiddenUnits = hiddenUnits
-        self.learnRate   = learnRate
-        self.T           = T
-        self.t           = 0
+largest_number = pow(2,binary_dim)
 
-        # Initializing the matrix weights
-        #LSTM Block
-        self.Wz = np.random.random((hiddenUnits, inputUnits)) - 0.5
-        self.Wi = np.random.random((hiddenUnits, inputUnits)) - 0.5
-        self.Wf = np.random.random((hiddenUnits, inputUnits)) - 0.5
-        self.Wo = np.random.random((hiddenUnits, inputUnits)) - 0.5
-        
-        self.Rz = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
-        self.Ri = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
-        self.Rf = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
-        self.Ro = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
-        
[…]

### 2016-01-13 python/LSTMlayer.py
Improved the LSTM Layer. Now returns the vector of deltas to train previous layers. Also, new script that fully uses LSTM cells (testLSTM2.py), doesn't quite work yet

@@ -91,6 +91,9 @@ class LSTMlayer :
         
         self.future_y = np.zeros_like(self.y_prev[:,1])
         self.future_c = np.zeros_like(self.c_prev[:,1])
+
+        # Delta vectors for a previous layer
+        self.delta_x = np.zeros((inputUnits,self.T))
     
     def forwardPropagate(self, X):    
 
@@ -127,12 +130,12 @@ class LSTMlayer :
         # EVALUATING THE DELTAS
         for t in reversed(range(self.T)):
             if (t == self.T-1):
-                self.delta_y_list[:,t] = (upperLayerDeltas[t])
-                self.delta_o_list[:,t] = (np.multiply( np.multiply(upperLayerDeltas[t],np.tanh(self.c_prev[:,t])), sigmoidPrime(self.o_prev[:,t])))
+                self.delta_y_list[:,t] = (upperLayerDeltas[:,t])
+                self.delta_o_list[:,t] = (np.multiply( np.multiply(upperLayerDeltas[:,t],np.tanh(self.c_prev[:,t])), sigmoidPrime(self.o_prev[:,t])))
                 self.delta_c_list[:,t] = (np.multiply(np.multiply(self.delta_y_list[:,t],tanhPrime(np.tanh(self.c_prev[:,t]))), self.o_prev[:,t]))
             else:
-                self.delta_y_list[:,t] = (upperLayerDeltas[t] + np.dot(self.Rz,self.delta_z_list[:,t+1]) + np.dot(self.Ri,self.delta_i_list[:,t+1]) + np.dot(self.Rf,self.delta_f_list[:,t+1]) + np.dot(self.Ro,self.delta_o_list[:,t+1]))
-                self.delta_o_list[:,t] = (np.multiply( np.multiply(upperLayerDeltas[t],np.tanh(self.c_prev[:,t])), sigmoidPrime(self.o_prev[:,t])))
+                self.delta_y_list[:,t] = (upperLayerDeltas[:,t] + np.dot(self.Rz,self.delta_z_list[:,t+1]) + np.dot(self.Ri,self.delta_i_list[:,t+1]) + np.dot(self.Rf,self.delta_f_list[:,t+1]) + np.dot(self.Ro,self.delta_o_list[:,t+1]))
+                self.delta_o_list[:,t] = (np.multiply( np.multiply(upperLayerDeltas[:,t],np.tanh(self.c_prev[:,t])), sigmoidPrime(self.o_prev[:,t])))
[…]

### 2016-01-13 python/testLSTM2.py

@@ -22,14 +22,14 @@ def tanhPrime(output):
 
 # training dataset generation
 int2binary = {}
-binary_dim = 32 
+binary_dim = 8 
 
 largest_number = pow(2,binary_dim)
 
 # input variables
-alpha = 0.22
+alpha = 0.1
 input_dim = 2
-hidden_dim = 18 
+hidden_dim = 4 
 output_dim = 1
 
 # initialize neural network weights
@@ -37,7 +37,8 @@ synapse_1        = 2*np.random.random(hidden_dim) - 1
 synapse_1_update = 2*np.random.random(hidden_dim) - 1
 y_prev = np.zeros((hidden_dim,binary_dim))
 lstmLayer1 = LSTMlayer.LSTMlayer(input_dim, hidden_dim, alpha, binary_dim)
-plt.axis([0, 30000, 0, 1000000])
+lstmLayerOut = LSTMlayer.LSTMlayer(hidden_dim, output_dim, alpha+0.05, binary_dim)
+plt.axis([0, 30000, 0, 200000])
 plt.ion()
 plt.show()
 
@@ -45,6 +46,7 @@ total_error = 0
 prev_total_error = 0
 synapse_1_update = 0
 layer_2 = np.ndarray((1,1))
+layer_2_errors = np.ndarray((1, binary_dim))
 
 # training logic
 for j in range(100000):
@@ -65,11 +67,6 @@ for j in range(100000):
 
     overallError = 0
     
-    layer_2_deltas = np.zeros((hidden_dim,binary_dim))
-    layer_1_values = list()
-    layer_1_values.append(np.zeros(hidden_dim))
-
-
     # -------------- THE FORWARD PROPAGATION STEP -------------- #
     for position in range(binary_dim):
         
@@ -78,39 +75,24 @@ for j in range(100000):
         y = np.array([int(c[binary_dim - position - 1])]).T
 
         # Perform a forward propagation through the network
-        y_prev[:,position] = lstmLayer1.forwardPropagate(X)
+        tempIntermediateLayer = lstmLayer1.forwardPropagate(X)
+        y_network   = lstmLayerOut.forwardPropagate(tempIntermediateLayer)
         
-        # Output Layer (uses output from LSTM, i.e y_curr).
-        layer_2 = sigmoid(np.dot(y_prev[:,position], synapse_1))
-
         # did we miss?... if so, by how much?
-        layer_2_error = y[0] - layer_2
[…]

### 2016-02-07 prepReport/theorBack.tex
Corrected some mistakes in the report

@@
 \section{Basic Concepts of Machine Learning}
-Machine Learning is a subject of Computer Science that studies the development of mathematical formulations that allows software to learn autonomously, without an explicit description of each rule of operation.
-Its goal is to extract latent characteristic features in datasets that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters automatically from the input data it receives.
+Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation.
+Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters accordingly with the input data it receives, improving the estimates it yields as it receives new input samples.
 
 Let us consider a practical example.
-For instance, suppose we want to build a device that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary.
+For instance, suppose we want to build a program that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary.
 We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from it being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language.
@@
 
-Furthermore, sometimes instead of applying the raw data to the model, we can apply some sort of preprocessing to the data to extract the relevant features from it.
-For instance, instead of just feeding a raw image, we can perform several operations like edge detection, and apply them in parallel.
+Sometimes, instead of applying the raw data to the model, we can apply some sort of \emph{preprocessing} to the data to extract the relevant features from it.
+For instance, instead of just feeding a raw image, we can perform several operations like edge detection or low-pass filtering, and apply them in parallel.
+In cases of highly dimensional data (i.e.
+each data vector has a very high number of features), we can apply techniques like \textbf{Principal Component Analysis} to reduce the feature space to a smaller dimension one, where the previous features were combined into two or three new features that pose themselves as the most relevant.
 
@@
 But we can also address \textbf{regression problems} where the output is not limited to a discrete set of values but rather a continuous interval.
+On the other hand, the Neural Network that this work will implement addresses a special kind of classification problem, where the classification decision is influenced not only by the current input sample, but also by a given \emph{window of samples} that trail the current sample.
 
 In summary, the most typical setting for a Machine Learning problem is having a large \textit{input dataset} which we use to \textit{train} our model (i.e.
-allowing him to dynamically adapt its set of parameters $\theta$), in order to produce an output label $y_i$ for each of them that minimizes a quality metric, typically defined as a \textit{cost function}.
-Artificial Neural Networks are one of the tools used in Machine Learning to perform this task, and will be discussed in Section~\ref{sec:theorBack_ann} as somewhat of a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}).
+allowing him to dynamically adapt its set of parameters $\theta$), in order to produce an output label $y_i$ for each of them that minimizes a quality metric, the \textit{Cost Funcion}, which can be chosen to the sum of squared differences, the log-loss, or any other appropriate  mathematical relationship  between the estimate $y_i$ and the correct label $t_i$.
+
+Now that the basic Machine Learning concepts have been presented, I will discuss, henceforth, one of the most important algorithms that address the supervised classification problems, the \emph{Artificial Neural Networks} that will be discussed in Section~\ref{sec:theorBack_ann} as somewhat of a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}), both of which are improvements over the initial formulation of the ANNs.
 These two last networks branch even further from these set of problems, and are usually employed in \textit{Deep Learning} tasks, where we try to extract even higher level information from data at the expense of increased model complexity.
@@
 \begin{equation}\label{eq:update}
-	w_{ji}^{(\tau+1)} = w_{ji}^{(\tau+1)} - \eta \frac{\partial E}{w_{ji}}
+	w_{ji}^{(\tau+1)} = w_{ji}^{(\tau)} - \eta \frac{\partial E}{\partial w_{ji}}
 \end{equation}
@@
 %%Place where I can include an algorithmic overview?
-For a complete proof of the above formulas, see~\citep[chap. 5.3.1]{Bishop2006}.
+For a complete proof of the above formulas, see~\cite[chap. 5.3.1]{Bishop2006}.
 
@@
 
-Even though RNNs outperform static ANNs in sequence recognition problems~\citep{Bengio1991}, they fail to retain long-term dependencies.
-Of course that the weight training process is itself a form of memory, but the problem is that the weight update is much slower than the activations~\citep{Yoshua01}, and therefore this memory only retains short-term dependencies.
-This is because of the so-called \textbf{Vanishing Gradients Problem}~\citep{Yoshua94,Yoshua01}, where the error decays exponentially through time, and the impact of previous incoming data points on the training of the weights quickly decrease.
+Even though RNNs outperform static ANNs in sequence recognition problems~\cite{Bengio1991}, they fail to retain long-term dependencies.
+Of course that the weight training process is itself a form of memory, but the problem is that the weight update is much slower than the activations~\cite{Yoshua01}, and therefore this memory only retains short-term dependencies.
+This is because of the so-called \textbf{Vanishing Gradients Problem}~\cite{Yoshua94,Yoshua01}, where the error decays exponentially through time, and the impact of previous incoming data points on the training of the weights quickly decrease.
 
-To overcome this issue, Hochreiter and Schmidhüber proposed, in 1997, a novel approach, called the Long-Short Term Network~\citep{Hoch97}.
+To overcome this issue, Hochreiter and Schmidhüber proposed, in 1997, a novel approach, called the Long-Short Term Network~\cite{Hoch97}.
 The next section explains the main structure of this approach, and also how it is trained, serving as a support basis for the work of this thesis.

### 2016-02-07 prepReport/theorBack.tex
Started LSTM Theoretical Chapter

@@
 
-FALAR EM BPTT
+The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron}.
+Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simulatneous Perturbation Stochastic Approximation}.
 
@@
 Of course that the weight training process is itself a form of memory, but the problem is that the weight update is much slower than the activations~\cite{Yoshua01}, and therefore this memory only retains short-term dependencies.
-This is because of the so-called \textbf{Vanishing Gradients Problem}~\cite{Yoshua94,Yoshua01}, where the error decays exponentially through time, and the impact of previous incoming data points on the training of the weights quickly decrease.
+This is because of the so-called \textbf{Vanishing Gradients Problem}~\cite{Yoshua94,Yoshua01}, where the error decays exponentially through time, and the impact of previous incoming data points on the training of the weights, and thus the current decision, quickly decreases.
 
 To overcome this issue, Hochreiter and Schmidhüber proposed, in 1997, a novel approach, called the Long-Short Term Network~\cite{Hoch97}.
-The next section explains the main structure of this approach, and also how it is trained, serving as a support basis for the work of this thesis.
+The next section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support basis for the work of this thesis.
 
 \section{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
+
+\subsection{Structure, Operation and Equations}\label{sec:struct_lstm}
+
+\subsection{Training -- SPSA}\label{sec:training_lstm}

### 2016-02-08 prepReport/theorBack.tex
Wrote all the Theoretical Part

@@
 
-To overcome this issue, Hochreiter and Schmidhüber proposed, in 1997, a novel approach, called the Long-Short Term Network~\cite{Hoch97}.
-The next section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support basis for the work of this thesis.
-
 \section{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
+To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}.
+This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support basis for the work of this thesis.
+Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}.
+One of the inital proposers of LSTM, Prof.
+Jürgen Schmidhüber, did a survey on the most common variations of the structure~\cite{Greff15} last year, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.
 
 \subsection{Structure, Operation and Equations}\label{sec:struct_lstm}
+A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}.
+As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network.
+Although ommited from the  picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}.
+The main components, their role and relevance, are explained as follows
+
+\begin{itemize}
+
+    \item \textbf{Input Gate} -- this is the input gate, where the importance of each feature of the input vector at time $t$, $\mb{x}^t$, and the output vector at the previous time step $\mb{y}^{t-1}$ is weighed in, producing an output $\mb{i}^{t}$.
+
+    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information from the input gate to the memory cell.
+It also receives the input vector and the previous output vector as inputs, but it does not have peephole connections and its dynamics are controlled by a different set of weights.
+The \textbf{activation function of this gate can be any}, but the most common choice is the \textbf{Hyperbolic Tangent}.
+
+    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to set them or reset them, using the \textit{Hadamard Elementwise} matrix multiplication of its output at time $t$, $\mb{c}^{(t)}$, with the contents of the memory unit at the previous time step, $\mb{c}^{(t-1)}$.
+The activation function of this gate is \textbf{always sigmoid}.
+
+    \item \textbf{Output Block Gate} -- this gate has a role very similar to that of the Block Input Gate, but now it controls the information flow \textit{out} of the LSTM neuron, namely the activated Memory Cell output.
+
+    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron.
+This is the memory element of the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates that connect to it.
+Also, this is where the peephole connections come from:
+
+    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation function that, as the gate activation function, can be any, but the \textit{hyperbolic tangent} is the most common choice.
+
+    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates to `peep' the states of the memory cell.
+They were added after the initial 1997 formulation, and their absence was proven to have a minimal performance impact~\cite{Greff15}.
+\end{itemize}
+
+After this small conceptual definitions, that allow us to grasp some intuition on the operation of a \textit{single} LSTM cell, I can present the overview of a \textit{layer} of LSTM neurons and their formal mathematical formulation, that will be needed both for the high-level model and the HDL description.
+The operation of each gate of the layer is given by the following set of equations
+
+\begin{align}
+    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
+    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
+    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
+    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
+    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
+    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
+\end{align}
+The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer.
+Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port.
+In that case, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single output, and we only use the appropriate value in the weight matrices $\mb{W}_{\cdot}$ and $\mb{R}_{\cdot}$
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
+    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}}
+	\label{fig:lstmneuron}
+\end{figure}
 
 \subsection{Training -- SPSA}\label{sec:training_lstm}
+Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme.
+Since I am aiming at an hardware implementation, and although the memory resources of nowadays FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system.
+That being said, we see that the usual training algorithms for LSTM cells -- i.e.
+Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance.
+A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.
+The weight update for the $i$-th weight at the time step $t$ is given by
+
+\begin{equation}\label{eq:spsa_weightUpdate}
+    \Delta w_i^{(t)} = \frac{J(\mb{w}^{(t)} + \beta \mb{s}^{(t)}) - J(\mb{w}^{(t)})}{\beta s_i^{(t)}}
+\end{equation}
+where $\beta$ is the magnitude of the perturbation to be introduced, $\mb{s}^{(t)}$ is a \textit{sign vector} whose $i$-th element, $s_i^{(t)}$, is either $-1$ or $1$.
+This way, we see that every weight is \textbf{randomly} incremented either by $-\beta$ or $\beta$, and we only need to keep a duplicate of the weight matrix with the perturbations, and we only need to evaluate the cost function twice per incoming sample.
+As for the \textit{update rule} is concerned, we have
+
+\begin{equation}\label{eq:spsa_updateRule}
+    w_i^{(t+1)} = w_i^{(t)} - \eta \Delta w_i^{(t)}
+\end{equation}
+where $\eta$ is the learning rate, and $\Delta w_i^{(t)}$ is the update for the $i$-th weight evaluated in~\ref{eq:spsa_weightUpdate}.
+According to the analysis performed in~\cite{Maeda05}, performing a second-order Taylor expansion at $\mb{w} = \mb{w}^{(t)}$, and taking the expected value of the equation, we get that
+
+\begin{equation}\label{eq:spsa_gradProof}
+    E(\Delta w_i^{(t)}) = \frac{\partial J(\mb{w}^{(t)})}{\partial w_i^{(t)}}
+\end{equation}
+that is, the weight update approximates the gradient of the cost function relative to that weight, and so the learning rule is a special form of \textit{Stochastic Gradient Descent}.
+
+One last comment to be made concerning the update rule, is the hipothetical need for \textbf{limits} to the weight values, when the update rule exceeds $|w_{\text{max}}|$ (in that case, we set $w_i^{(t+1)} = \pm w_{\text{max}}$), which sometimes might be needed if the behaviour of the weight update is not appropriate~\cite{Maeda05}.

### 2016-02-08 prepReport/intro.tex
Structured final document

\chapter{Introduction}\label{chap:intro}

\section{Background}

\section{Objectives}

\section{Motivation}

### 2016-02-08 prepReport/theorBack.tex
Structured final document

@@
-\chapter{Theoretical Background}\label{chap:theorBack}
+\chapter{Problem Characterization}\label{chap:theorBack}
 
-\section{Basic Concepts of Machine Learning}
+\section{Theoretical Background}
+
+\subsection{Basic Concepts of Machine Learning}
 Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation.
@@
 
-\section{Artificial Neural Networks}\label{sec:theorBack_ann}
+\subsection{Artificial Neural Networks}\label{sec:theorBack_ann}
 
@@
 
-\section{Recurrent Neural Networks}\label{sec:theorBack_rnn}
+\subsection{Recurrent Neural Networks}\label{sec:theorBack_rnn}
 A Recurrent Neural Network (RNN) is, essentially, a regular ANN where some neurons (especially in the hidden layer) have \emph{feedback connections} to themselves, i.e.
@@
 
-\section{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
+\subsection{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
 To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}.
@@
 
-\subsection{Structure, Operation and Equations}\label{sec:struct_lstm}
+\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
 A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}.
@@
 
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
+    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}}
+	\label{fig:lstmneuron}
+\end{figure}
+
 \begin{itemize}
-
     \item \textbf{Input Gate} -- this is the input gate, where the importance of each feature of the input vector at time $t$, $\mb{x}^t$, and the output vector at the previous time step $\mb{y}^{t-1}$ is weighed in, producing an output $\mb{i}^{t}$.
@@
 
-After this small conceptual definitions, that allow us to grasp some intuition on the operation of a \textit{single} LSTM cell, I can present the overview of a \textit{layer} of LSTM neurons and their formal mathematical formulation, that will be needed both for the high-level model and the HDL description.
-The operation of each gate of the layer is given by the following set of equations
+After these small conceptual definitions, that allow us to grasp some intuition on the operation of a \textit{single} LSTM cell, I can present the overview of a \textit{layer} of LSTM neurons and their formal mathematical formulation, that will be needed both for the high-level model and the HDL description.
+The operation of each set of gates of the layer is given by the following set of equations
 
-\begin{align}
+\begin{align}\label{eq:equationsLSTM}
     \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
@@
 The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer.
+Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e.
+the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$.
 Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port.
-In that case, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single output, and we only use the appropriate value in the weight matrices $\mb{W}_{\cdot}$ and $\mb{R}_{\cdot}$
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
-    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}}
-	\label{fig:lstmneuron}
-\end{figure}
+In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{\cdot}$, $\mb{R}_{\cdot}$ and the remaining.
 
-\subsection{Training -- SPSA}\label{sec:training_lstm}
+\subsubsection{Training -- SPSA}\label{sec:training_lstm}
 Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme.

### 2016-02-08 prepReport/workPlan.tex
Structured final document

\chapter{Work Plan}\label{chap:workPlan}

\section{Main Tasks}

\section{Approach}

\section{Tentative Scheduling}

\section{Software and Hardware Resources to be used}

### 2016-02-08 prepReport/stateArt.tex
Finished Teoretical Part. Started State-of-the-art. Cleaned up uncessary inital chaps

\chapter{State of the Art}\label{chap:stateArt}

\section{LSTM Applications (non-FPGA related)}

\section{Hardware Implementations of LSTM}

\section{Training Algorithms and Implementation}

\section{Final Overview}

### 2016-02-08 prepReport/theorBack.tex
Corrected reference error
Finished Teoretical Part. Started State-of-the-art. Cleaned up uncessary inital chaps

@@
 
-\begin{figure}[!H]
+\begin{figure}[H]
 	\begin{subfigure}{0.5\linewidth}
@@
 		\centering
-		\includegraphics[width=0.9\linewidth]{figures/multipolarNeuron.png}
+		\includegraphics[width=0.9\linewidth]{figures/MultipolarNeuron.png}
 		\caption{Biological Neuron Diagram}
@@
 The dendrites, where the stimuli are received, plays a role similar to that of the input nodes.
-The axon transmits the signal to the synaptic terminals, that are similar to the $y$ output}
+The axon transmits the signal to the synaptic terminals, that are similar to the $y$ output.
+Source: \href{https://en.wikipedia.org/wiki/Neuron\#/media/File:Neuron\_Hand-tuned.sv://en.wikipedia.org/wiki/Neuron\#/media/File:Blausen\_0657\_MultipolarNeuron.png}{Wikipedia}}
 	\label{fig:modelNeuron}
@@
 	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
-    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}}
+    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
 	\label{fig:lstmneuron}
@@
 One last comment to be made concerning the update rule, is the hipothetical need for \textbf{limits} to the weight values, when the update rule exceeds $|w_{\text{max}}|$ (in that case, we set $w_i^{(t+1)} = \pm w_{\text{max}}$), which sometimes might be needed if the behaviour of the weight update is not appropriate~\cite{Maeda05}.
+
+\section{Proposed Solution}\label{sec:propSol}
+Given the previous discusson of how LSTM neurons work, their benefits when compared to other RNN architectures, this work aims to conceive an efficient dedicated FPGA implementation with on-chip learning that, hopefully, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or from a framework.
+
+As discussed in Chapter~\ref{chap:stateArt}, hitherto there is only one FPGA implementation of LSTM cells~\cite{Chang15}, but there is no on-chip learning on the architecture (the weights are learned offline, and loaded in the FPGA before operation, and they are not changed while the FPGA system is operating), and my proposed work will implement an efficient LSTM implementation \emph{with} the ability to learn on-chip, which is an innovative and useful feature for the LSTM architecture on a dedicated hardware implementation.
+I will achieve that using the SPSA method described in~\ref{sec:training_lstm}, and inspired in the previous work of~\cite{Maeda05}.
+
+This solution will combine the benefits of a dedicated hardware implementation, where the inherent parallelism so obviously visible in the network architecture can be fully exploited something that, predominantly, yields a much higher performance in terms of computation time than a naïve CPU implementation, as in~\cite{Maeda05} where a $21\times$ speed-up against an ARM CPU was acheived.

### 2016-02-09 prepReport/stateArt.tex
Completed Hardware LSTM state of the art chapter.

@@
 
-\section{LSTM Applications (non-FPGA related)}
+Over the course of this chapter, I am going to present an overview of the most recent developments related to the work of this thesis, both in terms of existing dedicated hardware implementations~\ref{sec:sa_hardware}, the most relevant work in adapting suitable training algorithms to hardware~\ref{sec:sa_training} and also some of the most relevant applications of LSTM~\ref{sec:sa_apps}, which are not FPGA-based, but demonstrate how LSTM is useful by itself, and how well it competes with other Machine Learning algorithms in terms of long time-series dependences in data.
 
-\section{Hardware Implementations of LSTM}
+\section{LSTM Applications (non-FPGA related)}\label{sec:sa_apps}
 
-\section{Training Algorithms and Implementation}
+\section{Hardware Implementations of LSTM}\label{sec:sa_hardware}
+Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently (November 2015) by Chang et al.~\cite{Chang15} in the Computing Research Repository (CoRR).
+Is consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered Zedboard implementation.
+It uses a character-level language model from Andrej Karpathy, written in Lua using the Torch7 framework (the Lua calls are implemented in C, so no performance is lost).
 
-\section{Final Overview}
+Although the article is not clear on whether there is active learning by the ARM CPU -- the authors refer that the CPU loads the weights before operation, and that it changes them during operation, although how and why that change is done is not even clearly explained, neither mathematically nor conceptually --, \textbf{there is no on-chip learning module} in the FPGA according to the description provided.
+
+The implementation itself makes an extensive use of the Multiply-and-Accumulate units (MAC) of the FPGA, which since they are limited in number in the target platform, limits the number of neurons that we can deploy in parallel.
+The authors report a nearly 75\% usage for only 8 LSTM neurons.
+Although an apparently excessive number, I am left to scrutinize whether the usage of MACs is compulsory to obtain a better performance than a CPU or if it can be discarded, allowing for smaller cells.
+This way, we can have a layer of more neurons occupying the same area and, hopefully, the same, or even less, resources within the FPGA.
+
+\section{Training Algorithms and Implementation}\label{sec:sa_training}
+
+\section{Final Overview}\label{sec:overview}

### 2016-02-09 prepReport/stateArt.tex
Halfway Through State of the Art. Applications almost complete

@@
 \section{LSTM Applications (non-FPGA related)}\label{sec:sa_apps} 
+LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address the same set of problems where LSTM are employed, namely predicting and producing classification decisions from time-series data \textbf{citation needed!}.
+A very comprehensive description of applications can be found in one of the initial authors webpage dedicated to the subject~\footnote{\href{http://people.idsia.ch/~juergen/rnn.html}{http://people.idsia.ch/~juergen/rnn.html}}.
+I will now enumerate some of the bleeding edge applications of LSTM.
+
+\begin{itemize}
+    \item \textbf{Handwriting Recognition} -- and LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}.
+    \item \textbf{Speech Recognition} -- an architecture\cite{Graves13} proposed by Graves et al.
+in 2013 achieved an astonishing 17.7\% of accuracy on the TIMT Phoneme Recognition Benchmark, which up to the date is a new record.
+Furthermore, it has also been used for large scale acoustic modelling of speech~\cite{Sak14}.
+    \item \textbf{Handwriting Synthesis} --
+    \item \textbf{Speech Synthesis} --
+    \item \textbf{Biomedical Applications} --
+    \item \textbf{Music Analysis and Generation} --
+    \item \textbf{Video data Analysis} --
+
+\end{itemize}
 
@@
 Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently (November 2015) by Chang et al.~\cite{Chang15} in the Computing Research Repository (CoRR).
-Is consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered Zedboard implementation.
+It consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered~ Zedboard implementation.
 It uses a character-level language model from Andrej Karpathy, written in Lua using the Torch7 framework (the Lua calls are implemented in C, so no performance is lost).
@@
 \section{Training Algorithms and Implementation}\label{sec:sa_training}
+As stated in~\ref{sec:sa_hardware}, the work of~\cite{Chang15} does not feature on-chip learning at FPGA level, although there are a handful proposed solutions for it in recent literature.
+I will particularly look into the ones that use SPSA (see Section~\ref{sec:training_lstm}), since that is the training algorithm of choice for my proposed solution, and also to the ones that particularly apply SPSA to the training of Neural Networks.
+
+The SPSA algorithm was initially published by Spall in~\cite{Spall98}, and its theoretical details are outlined in Section~\ref{sec:training_lstm}.
+As of applications of it to the training of general Neural Networks, the earliest examples come from 1995 and 1996 on~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time where the memory resources of digital circuitry were limited, and so most of these structures were analog-based.
+Its adequacy was also established for \textbf{control problems}, such as those proposed in~\cite{Figueiredo97}, where a Time Delay Neural Network is used to control an unknown plant in a linear feedback system.
+
+In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network in an FPGA (and thus a digital system), achieving promising results in an Altera \textregistered~ FPGA.
+The article carefully delineates the approach taken, and also the hardware architecture designed, so it is a very good reference for the design that I will have to implement.
+
+Furthermore, a 2013 article by Tavear et al.~\cite{Tavear13} proposes, for the first time, using SPSA to train LSTM Neurons, although the article focuses on proving the suitability of SPSA to LSTM, and no factual hardware implementation is done or proposed.
+The authors simply demonstrate the suitability using conceptual arguments and by building a software model of an SPSA-trained LSTM network, and by comparing both the performance and computing speed of their model with the results achieved by Hochreiter et al.
+in~\cite{Hochreiter07}.
+Since the forward phase in both regular LSTM and SPSA-trained LSTM is the same, the computation time suffers no performance penalty whatsoever and the learning ability is preserved to a high degree, showing that SPSA is a valid alternative do BPTT and other similar and more common training schemes.
 

### 2016-02-09 prepReport/stateArt.tex
Completed some examples on State of the Art. Almost ready

@@
     \item \textbf{Speech Synthesis} --
-    \item \textbf{Biomedical Applications} --
-    \item \textbf{Music Analysis and Generation} --
-    \item \textbf{Video data Analysis} --
+    \item \textbf{Translation} -- LSTM was used by Sutskever et al.
+(Google) to perform sequence-to-sequence translation on the WMT'14 dataset, translating English to French with a close to state of the art score~\cite{Sustkever14}.
+
+    \item \textbf{Biomedical Applications} -- this network architecture was used in a protein homology detection scheme~\cite{Hochreiter07} using the SCOP 1.53 benchmark dataset, displaying a staggering performance when compared to other methods.
+Similarly, a recent article from 2015 by Sønderby et al.
+suggested the use of standalone LSTM and also Convolutional LSTM to perform subcellular localization of proteins, given solely their protein sequence, with an astounding accuracy of 0.902~\cite{Sonderby15}
+
+    \item \textbf{Music Analysis and Composition} -- Lehner et al.
+proposed a low-latecy solution based on LSTM, suitable for real-time detection of a singing voice in music recordings~\cite{Lehner15} whose performance surpassed other baseline methods, with lower latency.
+        In terms of music transcription from audio data, there is a study that proposes the use of LSTM cells to perform a transcription of piano recordings to musical notes~\cite{Bock12}, in order to automate music transcription.
+The model was trained with recordings of both acoustic pianos and synthesized pianos, and the labelling was performed using an associated MIDI file for each piece that was used in the training, showing promising results.
+\cite{Coca13} suggests the use of LSTM to perform autonomous computer music composition, and Eck and Schmidhüber proposed LSTM to perform Blues improvisation in~\cite{Eck02}.
+
+    \item \textbf{Video and Image Analysis} --
 

### 2016-02-09 prepReport/stateArt.tex
Only the overview is missing on the State of the Art chapter

@@
     \item \textbf{Handwriting Recognition} -- and LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}.
+LSTMs have also proven to surpass HMM based models in terms of optical character recognition in printed text, as~\cite{Breuel13} suggests.
+
     \item \textbf{Speech Recognition} -- an architecture\cite{Graves13} proposed by Graves et al.
@@
 Furthermore, it has also been used for large scale acoustic modelling of speech~\cite{Sak14}.
-    \item \textbf{Handwriting Synthesis} -- 
-    \item \textbf{Speech Synthesis} --
+
+    \item \textbf{Handwriting Synthesis} -- a comprehensive study by Graves shows, among other sequence generation tasks such as Text Prediction, the use of LSTM to produce synthetic handwriting, that looks incredibly similar to human-produced handwriting~\cite{Graves13_2}.
+
     \item \textbf{Translation} -- LSTM was used by Sutskever et al.
@@
 
-    \item \textbf{Video and Image Analysis} --
+    \item \textbf{Video and Image Analysis} -- Vinyals et al., at Google, proposed and LSTM network for \textbf{image captioning}, preceded by a Convolutional Neural Network to apply preprocessing to the images~\cite{Vinyals14}.
+This way, the LSTM works as \emph{sentence generator}, that captions the images, with state of the art performance.
+        Besides image captioning, video captioning is also an interesting topic.
+Venugopalan et al.
+recently proposed a CNN + LSTM architecture to translate video sequences to natural language~\cite{Donahue14}, using the Microsoft Research Video Description Corpus as a dataset.Tthere are other similar studies that combine image and video captioning~\cite{Donahue14_2}.
 

### 2016-02-09 prepReport/stateArt.tex
State of the art chapter completed

@@
 \section{Final Overview}\label{sec:overview}
+As we could attest from this small literature survey, although there is already an hardware implementation os LSTM, there is still a good deal of room for improvement by adding on chip-learning to the system, and also to restric it to a smaller use of the FPGA resources, allowing it to accomodate a more complex network.
+Furthermore,~\cite{Tavear13} shows that, at least for that particular case, the LSTM network doesn't suffer a great performance impact from using SPSA training, as opposed to the more common BPTT, and~\cite{Maeda05} showed that an SPSA hardware implementation is feasible.
+These three conclusions combined indicate that the idea of a hardware LSTM network with on-chip learning using SPSA is possible.
+
+Besides the proposed hardware implementation, it would be advantageous to perform \emph{benchmarks} to see how well it compares with software solutions, and Section~\ref{sec:sa_apps} suggests a handful of examples to test my final solution.

### 2016-02-10 prepReport/workPlan.tex
Work Plan Chapter concluded

@@
 
-\section{Main Tasks}
+\section{Approach and  Main Tasks}
+Dwight Eisenhower, United States' President from 1953 to 1961 and Army Officer, said that `no battle was ever won according to plan, but no battle was ever won without one'.
+Hence, although this initial planning draft might not be religiously followed in all its extent, it is important to break down the work in \textbf{tasks} so that I and my supervisors can keep track of what has to be done, and establishing deadlines for the small achievements that, together, will build up to the succesful completion of the proposed work of this thesis.
+The detailed explanation of each task is provided in Section~\ref{sec:wp_tasks_td} and the \textbf{schedule} for them, where their approximate begin and end weeks are marked, is represented by a \emph{Gantt Chart} in Section~\ref{sec:wp_tasks_gantt}.
 
-\section{Approach}
+\subsection{Task Description}\label{sec:wp_tasks_td}
+The work will be partitioned in \emph{five} large groups of tasks, and each of these groups will be further partitioned in smaller subtaskswhich are the actual assignements that I will be completing.
+We can think of the group of task as a sort of \emph{logical} grouping of the subtasks, according both to their relatedness and also to their precedence relationships.
+The task groups, and their related subtasks, are as follows
 
-\section{Tentative Scheduling}
+\begin{itemize}
+    \item \textbf{T1 -- Preparation} -- This stage is the initial phase of the work, where I will essentialy setup my working environment and finish the Python modules for the LSTM network that will be used in the validation of my Verilog modules
 
-\section{Software and Hardware Resources to be used}
+        \begin{itemize}
+            \item \textbf{T1.1 -- Complete Python Class with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Throgh Time version.
+
+            \item \textbf{T1.2 -- Reproduce the Benchmark Results with the Python Class} -- after developing and testing the Python Class with SPSA, I will try to reproduce with it the results of~\cite{Tavear13}, in order for me to have a high-level model whose performance is comparable to~\cite{Tavear13, Hochreiter07}, and thus provide a good performance baseline for this and other future benchmark tests that I would like to run in order to compare my FPGA solution with a software solution.
+
+            \item \textbf{T1.3 -- Setup FPGA Development Board} -- this is a farily technical step, where I setup my working environment, and make sure that everything is working correctly and is properly configured.
+As will be pointed out in Section~\ref{sec:wp_sh}, the FPGA development board of choice will be Xilinx~\textregistered Zedboard.
+
+            \item \textbf{T1.4 -- Familiarize with the Zedboard} -- before any actual use, I should synthesize and run some sample designs, and perhaps write a small Verilog design, and go through the full design flow, from behavioural simulation, to synthesis.
+This will allow me to know exactly what steps to take when I will need to develop and test my Dissertation related modules, and focus more on the solving the problem at hand rather than solving technical details.
+        \end{itemize}
+
+    \item \textbf{T2 -- HDL Module Development} -- this is where I will port the ideas from Section~\ref{chap:theorBack} to fully functional HDL blocks, using Verilog and targeting a Zedboard technology.
+
+        \begin{itemize}
+            \item \textbf{T2.1 -- Design Planning} -- before performing and HDL coding, it is favourable to first assemble the system design strategy in paper.
+This is where, together with my supervisor, I will outline the internal organization of my system, and make the fundamental design decisions.
+Of course, this task will prolong during all the \textbf{T2} extent, since some design changes to the original plan will have to be performed as I find \emph{bugs} in my early Verilog implementations that force a design correction.
+
+            \item \textbf{T2.2 -- Design SPSA Module} -- this task comprises the Verilog description of the module that will update the network weights using SPSA, and also the development of a testbench to validate my description.
+Furthermore, I will compare the output of this model with the golden output produced by a standalone SPSA Python Class previously developed in \textbf{T1}.
+            \item \textbf{T2.3 -- Design LSTM Neuron/Layer Module} -- similar to the previous task, but now I am aiming to implement the actual LSTM Neuron or a layer of LSTM Neurons.
+This is a design decision that will need to be set in \textbf{T2.1}, to asses whether it is justifiable to have dedicated multiplication units assigned to each neuron, or we can perform matrix-like batch multiplications and treat the LSTM as a batch of Neurons, i.e.
+as a layer.
+Like in the previous task, a testbench will also be developed, and the module validate both with it and the Python golden output.
+
+            \item \textbf{T2.4 -- Integrate and Validate both LSTM and SPSA modules} -- here, after the modules designed in \textbf{T2.2} and \textbf{T2.3} are validated individually, I will integrate them together in a higher level module, and perform a behavioural simulations of the new integrated module, as well as a verification procedure where the results from~\cite{Tavear13} will hopefully be reproduced and compared against the Python golden output.
+
+        \end{itemize}
+
+    \item \textbf{T3 -- FPGA Synthesis of the Verilog Code} -- after having Verilog code that passed all the validation tests and whose correct functionality is, therefore, asserted, I will synthesize the Verilog modules to FPGA RTL blocks, and perform a new simulation on these RTL syntesized code, that unlike the previous behavioural simulation, now has timing information resulting from the FPGA block mapping, and the place and routing of those mapped blocks.
+
+        \begin{itemize}
+            \item \textbf{T3.1 -- Synthesis of the SPSA Module} -- the module developed in \textbf{T2.2} is now syet
+                {bar height=.6}thesized and validated
+            \item \textbf{T3.2 -- Synthesis of the LSTM Neuron/Layer Module} -- same as the previous task, but now on the module of \textbf{T2.3}
+            \item \textbf{T3.3 -- Integration of the Modules and Validation} -- the modules synthesized in the previous tasks are integrated and validated using the same testbench as in \textbf{T2.4}.
+        \end{itemize}
+
+    \item \textbf{T4 -- Benchmarking the Design} -- at this stage, we have a fully functional FPGA design of an LSTM network with SPSA on-chip training.
+Henceforth, we will need to benchmark its performance by comparing it to~\cite{Tavear13, Hochreiter07}, and prove its usefulness by using it in one or two usecases from~\ref{chap:stateArt}, which are interesting uses of Deep Learning techniques such as these.
+
+        \begin{itemize}
+            \item \textbf{T4.1 -- Benchmark the Performance/Computing Time} -- perform a side-by-side comparison of our model with the software counterpart of\cite{Tavear13, Hochreiter07} and with the only hardware implementation of LSTM of~\cite{Chang15}.
+            \item \textbf{T4.2 -- Reproduce some of the state of the art applications of LSTM} -- in order to be able to show real-life, user-oriented, applications, I understand it is a good idea to try to reproduce some of the state of the art usages of LSTM, outlined in Section~\ref{chap:sa_apps}.
+It is too early to commit myself to which of these applications I will use, and how many, because that will depend on how well the work on tasks \textbf{T2} and \textbf{T3} is flowing.
+Nevertheless, I intend to at least impement one or two of them.
+        \end{itemize}
+
+    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are developing, this task will obviously be active, by documenting all the design decisions, milestones acheieved and logging all the results that come from the implementation.
+A couple of weeks before the deadline, this will be the only active task, since producing a clear and well-written document requires concentration and undivided attention.
+
+\end{itemize}
+
+\subsection{Tentative Task Scheduling}\label{sec:wp_tasks_gantt}
+Here is the Gantt Chart that depicts the temporal extension of the tasks described in Section~\ref{sec:wp_tasks_td}. 
+\begin{landscape}
+
+    \ganttset{bar height=.2}
+    \ganttset{calendar week text= \small {\startday/\startmonth}}
+    \begin{ganttchart}[vgrid, x unit=0.16cm, y unit chart=0.65cm, time slot format=isodate]{2016-02-15}{2016-06-26}
+
+        \gantttitlecalendar{year, month, week} \\
+        \ganttgroup{\textbf{T1}}{2016-02-15}{2016-02-29} \\
+        \ganttbar{\textbf{T1.1}}{2016-02-15}{2016-02-22} \\
+        \ganttbar{\textbf{T1.2}}{2016-02-22}{2016-02-29} \\
+        \ganttbar{\textbf{T1.3}}{2016-02-15}{2016-02-22} \\
+        \ganttbar{\textbf{T1.4}}{2016-02-22}{2016-02-29} \\
+
+        \ganttgroup{\textbf{T2}}{2016-02-29}{2016-04-11} \\
+        \ganttbar{\textbf{T2.1}}{2016-02-29}{2016-03-28} \\
+        \ganttbar{\textbf{T2.2}}{2016-02-29}{2016-03-07} \\
+        \ganttbar{\textbf{T2.3}}{2016-03-07}{2016-03-21} \\
+        \ganttbar{\textbf{T2.4}}{2016-03-21}{2016-04-11} \\
+
+        \ganttgroup{\textbf{T3}}{2016-04-11}{2016-05-02} \\
+        \ganttbar{\textbf{T3.1}}{2016-04-11}{2016-04-18} \\
+        \ganttbar{\textbf{T3.2}}{2016-04-18}{2016-04-25} \\
+        \ganttbar{\textbf{T3.3}}{2016-04-25}{2016-05-02} \\
+
+        \ganttgroup{\textbf{T4}}{2016-05-02}{2016-06-06} \\
+        \ganttbar{\textbf{T4.1}}{2016-05-02}{2016-05-16} \\
+        \ganttbar{\textbf{T4.2}}{2016-05-09}{2016-06-06} \\
+
+        \ganttgroup{\textbf{T5}}{2016-02-29}{2016-06-26} \\
+
+    \end{ganttchart}
+
+\end{landscape}
+\section{Software and Hardware resources to be used}\label{sec:wp_sh}
+
+\subsection{Hardware Resources}
+The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard}, whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and an ARM \textregistered~ Cortex\texttrademark~-A9 processor.
+It is an FPGA development board that features several I/O interfaces such as Etherner, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilites, making it easy to prototype additional features to the core design (i.e.
+for building the note transcription system of~\cite{Bock12}, and to make it operate in real-time, I would need an audio input, which I can easily get from the Audio IO, without investing time in connecting the codec and programming the FPGA to interface it).
+
+\subsection{Software Resources}
+In terms of software resources related to the FPGA development, I will need an \textbf{HDL simulation tool} to simulate and verify the Verilog code that I will develop, and a \textbf{Synthesis tool} to synthesize the Verilog files into RTL code, and to map that description into the resources available in the FPGA.
+The simulation tool used will be Questasim\texttrademark, the successor of the well-known Modelsim\texttrademark, which is an industry-standard tool in the area.
+For the synthesis tool, I will use Xilinx\textregistered Vivado Suite, also an industry-standard tool, and that is the only alternative, since the Zedboard is also produced by Xilinx\textregistered.
+These two tools are proprietary, but there are licenses available at FEUP.
+Furthermore, I am planning to use FloPoCo~\footnote{\href{http://flopoco.gforge.inria.fr/}{http://flopoco.gforge.inria.fr/}} to generate some of the modules that perform arithmetic operations, to hasten the developing effort/time, and to ensure that they are the most efficient possible, and don't pose a performance bottleneck to my design; this software is freeware, and a standard in Academia.
+
+For the Python Classes, I will make use of the 3.3 version of the Python distribution, the high-performance numerical computation library Numpy\footnote{\href{http://www.numpy.org/}{http://www.numpy.org/}}, which is a standard in scientific computing, and the Matplotlib\footnote{\href{http://matplotlib.org/}{http://matplotlib.org/}} plotting library.
+All of this software is also free to use, and open-source.
+I may also need to use the PyBrain\footnote{\href{http://pybrain.org/}{http://pybrain.org/}} framework, since it has an LSTM implementation.
+
+To write the final Disseration report, I will use \LaTeX (this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings.
+Once again, all of this is free to use.

### 2016-02-11 prepReport/intro.tex
Report Completed. Needs further polishing

@@
 \chapter{Introduction}\label{chap:intro}
+Before start reading this report, let me provide you with a quick overview of its mains parts.
 
-\section{Background}
+In Chapter~\ref{chap:intro}, I will present a conceptual background to the problem at hand, exposing the background that surrounds it~\ref{sec:intro_back}, the motivation for the desired solution~\ref{sec:intro_mot}, the objectives of it~\ref{sec:intro_obj} and, lastly, the people that will help me achieve it~\ref{sec:intro_people}.
 
-\section{Objectives}
+In Chapter~\ref{chap:theorBack}, the theoreical foundations will be layed out.
+Section~\ref{sec:tb_basicCon} presents the basic concepts of Machine Learning for the unsuspicious reader that is not versed in that area of knowledge, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs are presented, in consecutive order.
+Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that I will use in the final solution, that will be outlined in Section~\ref{sec:propSol}.
 
-\section{Motivation}
+In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications~\ref{sec:sa_apps}, their hardware implementations~\ref{sec:sa_hardware} and the current work regarding the chosen training algorithm~\ref{sec:sa_training} are presented.
+
+In Chapter~\ref{chap:workPlan}, the work plan for the Dissertation work is exhaustively detailed in \emph{tasks} in Section~\ref{sec:wp_tasks_td}, and their temporal arrangement is pictured in a Gantt Chart in Section~\ref{sec:wp_tasks_gantt}.
+Furthermore, the description of the Hardware and Software resources that I will be utilizing during the course of my work is done in Section~\ref{sec:wp_sh}
+
+Finally, in Chapter~\ref{chap:earlyWork} I will present some of the work that I have already done to prepare this thesis.
+
+\section{Background}\label{sec:intro_back}
+The Artificial Neural Networks are one of the most popular learning structures in the field of Machine Learning.
+As the name, itself, suggests, their operation is inspired by the operation of the building blocks of our brain, the \textbf{neurons}.
+In spite of its high performance, one of their shortcomings is the fact that they cannot retain temporal dependences among incoming data samples, thus not beign suitable to process time-serie data, such as audio, video or other kinds of time-varying data streams, where current inputs have a high temporal dependence with previous and future inputs.
+
+To address this issue, several algorithms have been used, such as Hidden Markov Models (HMM) or Recurrent Neural Networks (RNN), but both of these methods fail to recall temporal dependences that extend over an extended period of time, for the reasons that we will understand in Section \textbf{faltacitação}.
+Thus, in 1997 Hochreiter et al.
+proposed a novel RNN structure~\cite{Hoch97}, the Long Short-Term Network (LSTM), where a memory cell was introduced, and the input/output/read/write access are controlled by individual gates that are activated both by the incoming data samples, but also by the output from the previous time-step (it is an RNN after all).
+They are one of the state of the art methods in Deep Learning nowadays, as we can attest in Section~\ref{sec:sa_apps}.
+
+\section{Motivation}\label{sec:intro_mot}
+Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization, that can dramatically increase its performance, something that, hitherto, has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process still needs to be performed offline, in a normal CPU, when it also could be speeded up in a dedicated hardware structure.
+
+All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to the date generally have between 2-8 cores (mobile devices and general use personal computers),
+
+In order to parallelize the computations to the fullest extent possible, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the portin process is not entierely automatic, and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
+This way, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a supposedly lower computation time, and unburdening the CPU from running it, and saving performance room for running other tasks related, for instance, with the Operating System.
+Furthermore, when processing an incoming stream of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.
+
+\section{Objectives}\label{sec:intro_obj}
+Taking into account the considerations done in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
+Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate the some of the applications portrayed in~\ref{sec:sa_apps}.
+Lastly, to make justice to the name of the thesis, I will work to make this structure reconfigurable on the go, minimizing the reconfiguration time.
+
+\section{People Involved}\label{sec:intro_people}
+Besides me, the candidate to the Master's Degree, there are two more people involved, namely
+
+\begin{itemize}
+    \item \textbf{Supervisor} -- The \href{https://sigarra.up.pt/feup/pt/func\_geral.formview?p\_codigo=210963}{Professor João Canas Ferreira}, auxiliary professor at the Faculty of Engineering of the University of Porto.
+    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
+\end{itemize}

### 2016-02-11 prepReport/theorBack.tex
Report Completed. Needs further polishing

@@
 
-\section{Theoretical Background}
+\section{Theoretical Background}\label{sec:tb}
 
-\subsection{Basic Concepts of Machine Learning}
+\subsection{Basic Concepts of Machine Learning}\label{sec:tb_basicCon}
 Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation.
@@
 Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port.
-In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{\cdot}$, $\mb{R}_{\cdot}$ and the remaining.
+In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{*}$, $\mb{R}_{*}$ and the remaining.
 

### 2016-02-11 prepReport/earlyWork.tex
Added Chapter 5. Still needs polishing

\chapter{Early Work}\label{chap:earlyWork}
In order to better undesrtand LSTM,  gain some useful insight on their operation, and to have a software implementation, that is not part of any framework, with which I can validate in the future my FPGA implementation, I have developed an fully functional LSTM Layer in Python and Numpy, that still uses BPTT as a training algorithm (the governing equations are in appendix A of~\cite{Greff15}). The idea is to adapt it to use SPSA in the beginning of my Master's Thesis. The code is present in

\lstinputlisting[language=Python, caption=Implementation of a LSTM Layer]{LSTMlayer.py}

### 2016-02-11 prepReport/theorBack.tex
Added Chapter 5. Still needs polishing

@@
 A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.
+The main idea of this technique is, instead of explicitly evalute the gradients for the cost function at each time step, perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
 The weight update for the $i$-th weight at the time step $t$ is given by
@@
 where $\eta$ is the learning rate, and $\Delta w_i^{(t)}$ is the update for the $i$-th weight evaluated in~\ref{eq:spsa_weightUpdate}.
-According to the analysis performed in~\cite{Maeda05}, performing a second-order Taylor expansion at $\mb{w} = \mb{w}^{(t)}$, and taking the expected value of the equation, we get that
+According to the analysis performed in~\cite{Maeda05}, performing a two-term Taylor expansion at $\mb{w} = \mb{w}^{(t)}$, and taking the expected value of the equation, we get that there is a $\mb{w}_{S1}$ so that
+
+\begin{equation}\label{eq:taylor}
+    \Delta w_i^{(t)} = s_i^{(t)} (\mb{s}^{(t)})^T \frac{\partial J(\mb{w}^{(t)})}{\partial \mb{w}} +
+                        \frac{c s_i^{(t)}}{2} (\mb{s}^{(t)})^T \frac{\partial^2 J(\mb{w}^{(t)})}{\partial \mb{w}^2}  (\mb{s}^{(t)}) 
+\end{equation}
+If the random sign vector $\mb{s}^{(t)}$ is chosen carefully so that the expected value of the vector at the $i$-th position is zero, $E(s_i^{(t)})$, and the signs of two different components $i$ and $j$, at different time-steps, are independent (i.e.
+$E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, equation~\ref{eq:taylor} reduces to
 
@@
 \end{equation}
-that is, the weight update approximates the gradient of the cost function relative to that weight, and so the learning rule is a special form of \textit{Stochastic Gradient Descent}.
+that is, the \emph{expected value} of the weight update approximates the gradient of the cost function relative to that weight, and so the learning rule is a special form of \textit{Stochastic Gradient Descent}.
 

### 2016-02-11 prepReport/workPlan.tex
Added Chapter 5. Still needs polishing

@@
 \subsection{Hardware Resources}
-The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard}, whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and an ARM \textregistered~ Cortex\texttrademark~-A9 processor.
+The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard} (Figure~\ref{fig:zedboard}, whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and an ARM \textregistered~ Cortex\texttrademark~-A9 processor.
 It is an FPGA development board that features several I/O interfaces such as Etherner, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilites, making it easy to prototype additional features to the core design (i.e.
 for building the note transcription system of~\cite{Bock12}, and to make it operate in real-time, I would need an audio input, which I can easily get from the Audio IO, without investing time in connecting the codec and programming the FPGA to interface it).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/zedboard.jpg}
+    \caption{The Zedboard FPGA Development Board. Source: \href{http://www.xilinx.com}{Xilinx}}
+    \label{fig:zedboard}
+\end{figure}
 

### 2016-02-11 prepReport/LSTMlayer.py

@@ -1,0 +1,169 @@
+# -*- coding: utf-8 -*-
+"""
+LSTM Layer Class
+@author: Jose Pedro Castro Fonseca, University of Porto, PORTUGAL
+"""
+import numpy as np
+import time
+
+# Defines the seed for random() as the current time in seconds
+np.random.seed(round(time.time()))
+
+# Evaluates the Sigmoid Function
+def sigmoid(x):
+    output = 1/(1+np.exp(-x))
+    return output
+
+# Evaluates the derivative of the Sigmoid Func, based on a previous sigmoid() call
+def sigmoidPrime(output):
+    return output*(1-output)
+
+# Evaluates the derivative of the TanH Func, based on a previous np.tanh() call
+def tanhPrime(output):
+    return (1-output**2)
+
+class LSTMlayer :
+
+    def __init__(self, inputUnits, hiddenUnits, learnRate, T):
+        # The Network Parameters, passed by the user
+        self.inputUnits  = inputUnits
+        self.hiddenUnits = hiddenUnits
+        self.learnRate   = learnRate
+        self.T           = T
+        self.t           = 0
+
+        # Initializing the matrix weights
+        #LSTM Block
+        self.Wz = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wi = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wf = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wo = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        
+        self.Rz = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        self.Ri = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        self.Rf = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        self.Ro = np.random.random((hiddenUnits, hiddenUnits)) - 0.5
+        
+        self.pi = np.random.random((hiddenUnits)) - 0.5
+        self.pf = np.random.random((hiddenUnits)) - 0.5
+        self.po = np.random.random((hiddenUnits)) - 0.5
+        
+        self.bz = np.random.random((hiddenUnits)) - 0.5
+        self.bi = np.random.random((hiddenUnits)) - 0.5
[…]

### 2016-02-11 prepReport_IT_11Feb/suggestions.tex
Suggestions for the “Dissertation preparation” document. IT

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{scrartcl} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{titling}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% MATH PACKAGES
\usepackage{amsmath}

%%% END Article customizations

%%% The "real" document content comes below...

\title{Suggestions for the ``Dissertation preparation'' document.}
\author{Ivo Timóteo \\11$^{\text{th}}$ February, 2016}
\date{}
\pretitle{\begin{flushleft}\LARGE}
\posttitle{\end{flushleft}}
\preauthor{\begin{flushleft}}
\postauthor{\end{flushleft}\rule{\linewidth}{0.5mm}}
\predate{}
\postdate{}

\begin{document}
\maketitle

Next time make the .tex compile without errors so that I can suggest inline.

Also, careful with commas and sentence structure. It's better to have multiple shorter, clear sentences than nested monsters requiring the reader to keep a stack of depth 4. And don't separate conjunctions with commas, please!

Furthermore, I am not correcting everything because I am too lazy to type. Make sure to give another very good passage over the text afterwards.

\subsection*{List of figures}
\begin{itemize}
	\item Use short captions on the figures otherwise LoF looks awful.
	\begin{verbatim}\caption[Short version for LoF]{Full caption}\end{verbatim}
\end{itemize}

\subsection*{Introduction}
\begin{itemize}
	\item {\bf REPLACE} [first sentence]\\{\bf BY} Brief overview of the report:
	\item {\bf REPLACE} the objectives of it\\{\bf BY} its objectives
	\item {\bf REPLACE} help me achieve it \\{\bf BY} help me achieve them
	\item {\bf SPELLING} theoreical
	\item {\bf REMOVE} for the unsuspicious reader that is not versed in that area of knowledge
	\item {\bf REMOVE} are presented, in consecutive order.
	\item {\bf REPLACE} I will use in the final solution, that will be outlined \\{\bf BY} will be used in the final solution, which is outlined
	\item {\bf REMOVE} exhaustively
	\item {\bf REPLACE} in tasks \\{\bf BY} as tasks
	\item {\bf REPLACE} Chart \\{\bf BY} chart
	\item {\bf REPLACE} utilizing during the course of my work \\{\bf BY} using in my work
	\item {\bf REPLACE} that I have already done to prepare this thesis \\{\bf BY} already concluded
\end{itemize}

\subsection*{Background}
\begin{itemize}
	\item {\bf REMOVE} The
	\item {\bf REPLACE} learning structures \\{\bf BY} models
	\item {\bf REMOVE} , itself,
	\item {\bf REPLACE} its \\{\bf BY} their
	\item {\bf SPELLING} beign
	\item {\bf REMOVE} that extend
	\item {\bf WTF} faltacitação
	\item {\bf COMMENT} Not really clear at all from the short explanation why LSTM improves over HMM and RNN which is what your sentence kind of makes us expect... 
\end{itemize}

\subsection*{Motivation}
\begin{itemize}
	\item {\bf REMOVE} [comma after realization]
	\item {\bf REMOVE} , hitherto,
	\item {\bf REPLACE} still needs to be \\{\bf BY} is
	\item {\bf REMOVE} possible
	\item {\bf REMOVE} [comma after automatic]
	\item {\bf REPLACE} , and saving \\{\bf BY} thus saving
	\item {\bf WTF} performance room
\end{itemize}

\subsection*{Objectives}
\begin{itemize}
	\item {\bf REPLACE} done \\{\bf BY} presented
	\item {\bf REMOVE} , to make justice to the name thesis, 
\end{itemize}

\subsection*{Theoretical background}
\begin{itemize}
	\item {\bf COMMENT} Not all learning is for classification; not all learning wants to extract features (e.g., we might just want a simple blackbox predictor); not all learning is supervised. Explicit vector representations have commas separating different components.
	\item {\bf REMOVE} diametrically
	\item {\bf COMMENT} refer that the sum of squares error is common because it's what appears under the assumption of Gaussian noise.
	\item {\bf COMMENT} ``try to extract higher level information from data''. Can you explain? Weren't you just looking for $y_i$? What is a higher level prediction?
\end{itemize}

\subsection*{Artificial Neural Networks}
\begin{itemize}
	\item {\bf COMMENT} ``try to mimic the Human Brain'' -- I don't think this is actually accepted at all; no one I know believes this to be true. Also, why the capitalization? Also, right after that you claim to know the underlying mathematical function of a human neuron. Finally, you present a linear model as being equivalent to a function which is incredibly reductive (can't you have a function that is a non-linear model of its inputs?).
	\item {\bf COMMENT} $b_0$ is usually called $w_0$ and you define $x = [1, x_1, \dots, x_n]$ giving you $y=f(\bf{w}^T\bf{x})$. Also, you have $N$ as number of data points and $n$ as number of features of the input which can be confusing.
	\item {\bf COMMENT} Step functions are usually between 0 and 1. On the plot it would be nicer if every function had the same range.
	\item {\bf COMMENT} Why are layers important?
	\item {\bf COMMENT} When discussing the equations (2.3) to (2.6) it would be nice to have the schematics of at least three chained neurons.
\end{itemize}

\subsection*{Recurrent Neural Networks}
\begin{itemize}
	\item {\bf REMOVE} , essentially,
	\item {\bf REPLACE} especially \\{\bf BY} commonly
	\item {\bf COMMENT} ``but no information about the correlation between themselves and the data points that preceded them did not influence the training step'' -- this is a double negative which I don't think you wanted.
	\item {\bf REPLACE} They were regarded as if no temporal relationship existed, and therefore each data point is conditionally independent of any other. \\{\bf BY} Temporal relationship is disregarded and  each data point considered conditionally independent of any other.
	\item {\bf REPLACE} truth \\{\bf BY} true
	\item {\bf REPLACE} , in which it is actually completely false, \\{\bf BY} such
	\item {\bf SPELLING} Simulatneous
	\item {\bf COMMENT} Stop the weird unnecessary capitalization of stuff :)
\end{itemize}

\subsection*{LSTMN}
\begin{itemize}
	\item {\bf REMOVE} basis
	\item {\bf REPLACE} its formulation \\{\bf BY} it
	\item {\bf REPLACE} structure [2] last year \\{\bf BY} model last year [2]
	\item {\bf SPELLING} presenation
	\item {\bf REPLACE} as well as the work \\{\bf BY} as well as of the work
	\item {\bf COMMENT} ``Hadamard elementwise matrix multiplication'' is {\bf very} redundant. Why is the activation function of the forget gate always sigmoid? Btw, even though sigmoid is usually seen as the logistic function it actually refers to the entire family of S-shaped functions.
	\item {\bf REMOVE} [commas around ``can be any'']
	\item {\bf REPLACE} As for the update rule is concerned, we have \\{\bf BY} The update rule is
	\item {\bf SPELLING} hipothetical
	\item {\bf COMMENT} how do you assess whether the behaviour of the weight is appropriate or not?
\end{itemize}

\subsection*{Proposed Solution}
\begin{itemize}
	\item {\bf REPLACE} work, their \\{\bf BY} work and their
	\item {\bf SPELLING} acheived 
\end{itemize}

\subsection*{State of the art}
\begin{itemize}
	\item {\bf COMMENT} ``bleeding edge'' is a bit too informal. Also, too many adjectives in 3.1; staggering, incredibly, astonishing, astounding, etc  
	\item {\bf SPELLING} Tthere
	\item {\bf SPELLING} restric
\end{itemize}

\subsection*{Work plan}
\begin{itemize}
	\item {\bf COMMENT} I don't know what to say about the first paragraph...
	\item {\bf SPELLING} subtaskswhich
	\item {\bf SPELLING} throgh
	\item {\bf SPELLING} asses
	\item {\bf SPELLING} [T3.1 is badly typed]
\end{itemize}

\end{document}

### 2016-02-12 prepReport/intro.tex
Implemented the Ivo suggested corrections

--- 2016-02-08 prepReport/intro.tex
@@
 \chapter{Introduction}\label{chap:intro}
+Brief overview of the report:
 
-\section{Background}
+In Chapter~\ref{chap:intro}, I will present a conceptual background to the problem at hand, exposing the background that surrounds it~\ref{sec:intro_back}, the motivation for the desired solution~\ref{sec:intro_mot}, its objectives~\ref{sec:intro_obj} and, lastly, the people that will help me achieve them~\ref{sec:intro_people}.
 
-\section{Objectives}
+In Chapter~\ref{chap:theorBack}, the theoretical foundations will be layed out.
+Section~\ref{sec:tb_basicCon} presents the basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs.
+Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final solution, which will be outlined in Section~\ref{sec:propSol}.
 
-\section{Motivation}
+In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications~\ref{sec:sa_apps}, their hardware implementations~\ref{sec:sa_hardware} and the current work regarding the chosen training algorithm~\ref{sec:sa_training} are presented.
+
+In Chapter~\ref{chap:workPlan}, the work plan for the Dissertation work is detailed as \emph{tasks} in Section~\ref{sec:wp_tasks_td}, and their temporal arrangement is pictured in a Gantt chart in Section~\ref{sec:wp_tasks_gantt}.
+Furthermore, the description of the Hardware and Software resources that I will be utilizing during the course of my work is done in Section~\ref{sec:wp_sh}
+
+Finally, in Chapter~\ref{chap:earlyWork} I will present some of the work that I have already concluded.
+
+\section{Background}\label{sec:intro_back}
+Artificial Neural Networks are one of the most popular models in the field of Machine Learning.
+As the name suggests, their operation is inspired by the operation of the building blocks of our brain, the \textbf{neurons}.
+In spite of its high performance, one of their shortcomings is the fact that they cannot retain temporal dependences among incoming data samples, thus not being suitable to process time-series data, such as audio, video or other kinds of time-varying data streams, where current inputs have a high temporal dependence with previous and future inputs.
+
+To address this issue, several algorithms have been used, such as Hidden Markov Models (HMM) or Recurrent Neural Networks (RNN), but both of these methods fail to recall temporal dependences that extend over an large period of time, for the reasons that we will understand in Sections~\ref{sec:theorBack_rnn} and~\ref{sec:theorBack_lstm}.Thus, in 1997 Hochreiter et al.
+proposed a novel RNN structure~\cite{Hoch97}, the Long Short-Term Network (LSTM), where a memory cell was introduced, and the input/output/read/write access is controlled by individual gates that are activated both by the incoming data samples, but also by the outputs from the previous time-step (it is an RNN after all).
+They are one of the state of the art methods in Deep Learning nowadays, as we can attest in Section~\ref{sec:sa_apps}.
+
+\section{Motivation}\label{sec:intro_mot}
+Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be speeded up in a dedicated hardware structure.
+
+All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to the date generally have between 2-8 cores (mobile devices and general use personal computers),
+
+In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the portin process is not entierely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
+This way, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a supposedly lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance.
+Furthermore, when processing an incoming stream of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.
+
+\section{Objectives}\label{sec:intro_obj}
+Taking into account the considerations presented in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
+Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate the some of the applications portrayed in~\ref{sec:sa_apps}.
+Lastly, I will work to make this structure reconfigurable on the go, and trying to minimize the reconfiguration time.
+
+\section{People Involved}\label{sec:intro_people}
+Besides me, the candidate to the Master's Degree, there are two more people involved, namely
+
+\begin{itemize}
+    \item \textbf{Supervisor} -- The \href{https://sigarra.up.pt/feup/pt/func\_geral.formview?p\_codigo=210963}{Professor João Canas Ferreira}, auxiliary professor at the Faculty of Engineering of the University of Porto.
+    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
+\end{itemize}

### 2016-02-12 prepReport/stateArt.tex
Implemented the Ivo suggested corrections

@@
 Similarly, a recent article from 2015 by Sønderby et al.
-suggested the use of standalone LSTM and also Convolutional LSTM to perform subcellular localization of proteins, given solely their protein sequence, with an astounding accuracy of 0.902~\cite{Sonderby15}
+suggested the use of standalone LSTM and also Convolutional LSTM to perform sub cellular localization of proteins, given solely their protein sequence, with an astounding accuracy of 0.902~\cite{Sonderby15}
 
@@
 Venugopalan et al.
-recently proposed a CNN + LSTM architecture to translate video sequences to natural language~\cite{Donahue14}, using the Microsoft Research Video Description Corpus as a dataset.Tthere are other similar studies that combine image and video captioning~\cite{Donahue14_2}.
+recently proposed a CNN + LSTM architecture to translate video sequences to natural language~\cite{Donahue14}, using the Microsoft Research Video Description Corpus as a dataset.There are other similar studies that combine image and video captioning~\cite{Donahue14_2}.
 
@@
 \section{Final Overview}\label{sec:overview}
-As we could attest from this small literature survey, although there is already an hardware implementation os LSTM, there is still a good deal of room for improvement by adding on chip-learning to the system, and also to restric it to a smaller use of the FPGA resources, allowing it to accomodate a more complex network.
+As we could attest from this small literature survey, although there is already an hardware implementation os LSTM, there is still a good deal of room for improvement by adding on chip-learning to the system, and also to restrict it to a smaller use of the FPGA resources, allowing it to accomodate a more complex network.
 Furthermore,~\cite{Tavear13} shows that, at least for that particular case, the LSTM network doesn't suffer a great performance impact from using SPSA training, as opposed to the more common BPTT, and~\cite{Maeda05} showed that an SPSA hardware implementation is feasible.

### 2016-02-12 prepReport/theorBack.tex
Added a recurrent neuron figure
Implemented the Ivo suggested corrections

@@
 We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from it being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language.
-The approach taken by Machine Learning is diametrically different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 \; \mb{x}_2 \;\mb{x}_2 \; \dots \mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model.
+The approach taken by Machine Learning is different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 \; \mb{x}_2 \;\mb{x}_2 \; \dots \mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model.
 Each of the $i$-th data point was previously labelled, and in fact we feed each training data point $\mb{x}_i$ \emph{along} with its corresponding label $t_i$, so that the model can adapt its parameters accordingly to the \emph{target value} it is supposed to classify.
@@
 The model then evolves in a way that minimizes the cost function.
-A usual choice for the cost function is the \textbf{sum of squares error}.
+A usual choice for the cost function is the \textbf{sum of squares error}, given the Gaussian Noise assumption.
 Mathematically, if $y_{\theta}^i = y_{\theta}(x_i)$ is the prediction for the input data point $x_i$ with label $t_i$, given the current set of parameters $\theta$, the cost function using this metric is given by
@@
 
-Artificial Neural Networks (ANN) are mathematical structures that, as the name suggests, try to mimic the Human Brain.
+Artificial Neural Networks (ANN) are mathematical structures that, as the name suggests, try to mimic the basic way of how a human brain works.
 ANN's building blocks, like their biological counterpart, model the high-level behaviour of biological neurons, in the sense that they neglect unnecessary biological aspects (such as modeling all the voltages across the neuron and all its electromagnetic interactions), and only retain its fundamental underlying mathematical function, which is a weighted linear combination of its inputs subject to a \emph{activation function}, i.e.
@@
 \begin{equation}
-	y = f(\mb{w}^T \mb{x} + b_0)
+	y = f(\mb{w}^T \mb{x})
 \end{equation}
-where $\mb{w} = \left[ w_1 \; w_2 \;  w_3 \;  \dots \; w_n \right]$ is the input weight vector, $\mb{x} = \left[ x_1 \; x_2 \; x_3 \; \dots \; x_n \right]$ the input vector, $b_0$ is the bias factor and $f(\cdot)$ is the chosen activation function.
-Furthermore, we call the scalar quantity $a = \mb{w}^T \mb{x} + b_0$ the \textbf{activation}, since its value determines how the activation function will behave.
+where $\mb{w} = \left[ w_0 \; w_1 \; w_2 \;  w_3 \;  \dots \; w_n \right]$ is the input weight vector, $\mb{x} = \left[ x_0 \; x_1 \; x_2 \; x_3 \; \dots \; x_n \right]$ the input vector, $b_0$ is the bias factor and $f(\cdot)$ is the chosen activation function.
+Furthermore, we call the scalar quantity $a = \mb{w}^T \mb{x}$ the \textbf{activation}, since its value determines how the activation function will behave.
 Figure~\ref{fig:modelNeuron} exemplifies the roles of these variables within our neuron model, and compares each part of it with the biological counterpart.
@@
 
-    \caption{In Figure~\ref{fig:modelNeuron_a}.
+    \caption[Representations of a Neuron in an ANN~\ref{fig:modelNeuron_a} and a biological multi polar neuron~\ref{fig:modelNeuron_b}]{In Figure~\ref{fig:modelNeuron_a}.
 each input feature $x_i$ is weighted by its corresponding weight $w_i$.
@@
 	\includegraphics[width=0.9\linewidth]{figures/activFunc.eps}
-	\caption{Three different activation functions.
+    \caption[Possible activation functions of choice for a neuron]{Three different activation functions.
 As you can see, the hyperbolic tangent has the same extreme value as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision.
@@
 	\includegraphics[width=0.9\linewidth]{figures/ann.eps}
-	\caption{A three layer ANN.
+    \caption[Three layer ANN]{A three layer ANN.
 We have omitted some of the connections in the hidden layer, for simplification purposes.
@@
 
-Regarding the training of ANNs, is is performed through a two-step process: first, a \textbf{feed-forward} step where the input is applied, and the activations are evaluated in succession up to the output neurons; then, the perform the \textbf{backpropagation} step, where we calculate the errors in each of the nodes, but now from the output to the input: the weights are updated and optimized using an iterative  method called \textit{Gradient Descent}, where if $\tau$ is the current time step, the next update on the weight matrix $\mb{w}^{(\tau+1)}$ is given by
+Regarding the training of ANNs, it is performed through a two-step process: first, a \textbf{feed-forward} step where the input is applied, and the activations are evaluated in succession up to the output neurons; then, we perform the \textbf{backpropagation} step, where we calculate the errors in each of the nodes (the so-called \textbf{deltas} of equation~\ref{eq:deltaj}), but now from the output to the input: the weights are updated and optimized using an iterative  method called \textit{Gradient Descent}, where if $\tau$ is the current time step, the next update on the weight matrix $\mb{w}^{(\tau+1)}$ is given by
 
@@
 The relevance of this different structure is the possibility to retain \emph{sequence} information about the data.
-Before, each incoming data point only contributed to the training of the network, but no information about the correlation between themselves and the data points that preceded them did not influence the training step.
-They were regarded as if no temporal relationship existed, and therefore each data point is conditionally independent of any other.
-This is obviously not necessarily truth, and in fact there are many cases, where the correlation between data points is high for those closely spaced in time, in which it is actually completely false, as in video signals, audio signals, or other kinds of \emph{temporal sequences} of data.
+Before, each incoming data point only contributed to the training of the network, but the information about the correlation between themselves and the data points that preceded them did not influence the training step.
+Temporal relationship is disregarded and each data point considered conditionally independent of any other.
+This is obviously not necessarily true, and in fact there are many cases where the correlation between data points is high for those closely spaced in time, such as in video signals, audio signals, or other kinds of \emph{temporal sequences} of data.
 Therefore, the feedback connection of the neuron to himself acts as a kind of \emph{memory element} that takes into account in the present decision, the history of decisions previously taken, and hence the previous data.
-Figure~\ref{fig:recneuron} suggests a possible structure for a neuron of a hidden layer in an RNN, and also an alternate representation, where the structure is unfold through time.
 
-%% METE AQUI A FIGURA
+Figure~\ref{fig:recurrneuron} suggests a possible structure for a neuron of a hidden layer in an RNN, and also an alternate representation, where the structure is unfold through time.
 
-The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron}.
-Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simulatneous Perturbation Stochastic Approximation}.
+\begin{figure}[H]
+	\begin{subfigure}{0.5\linewidth}
+		\centering
+		\includegraphics[width=0.34\linewidth]{figures/recneuron.eps}
+		\caption{Recurrent Neuron}
+		\label{fig:recneuron}
+	\end{subfigure}
+	\begin{subfigure}{0.5\linewidth}
+		\centering
+		\includegraphics[width=0.9\linewidth]{figures/recneuron_unf.eps}
+		\caption{Recurrent Neuron, unfolded through time}
+		\label{fig:recneuron_unf}
+	\end{subfigure}
+
+    \caption[Representation of a Recurrent Neuron (Figure~\ref{fig:recneuron}) and a Recurrent Neuron unfolded through time (Figure~\ref{recneuron_unf})]{In Figure~\ref{fig:recneuron}, a recurrent neuron is depicted, where the output is fed back to the input, weighted by $\theta_h$.
+Figure~\ref{fig:recneuron_unf}, an unfolding through time of that same neuron is performed.
+The basic idea is that feeding the output back to the input is similar to feeding the output to the input of the neuron \emph{at a previous time step}: this way, we linearize the structure.
+Considering a training epoch of $T$ samples, corresponds to having $T$ unfolded neurons/layers}
+	\label{fig:recurrneuron}
+\end{figure}
+
+The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron_unf}.
+Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simultaneous Perturbation Stochastic Approximation}.
 
@@
 To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}.
-This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support basis for the work of this thesis.
+This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support for the work of this thesis.
 Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}.
 One of the inital proposers of LSTM, Prof.
-Jürgen Schmidhüber, did a survey on the most common variations of the structure~\cite{Greff15} last year, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.
+Jürgen Schmidhüber, did a survey on the most common variations of the model last year~\cite{Greff15}, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.
 
@@
 A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.
-The main idea of this technique is, instead of explicitly evalute the gradients for the cost function at each time step, perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
+The main idea of this technique is, instead of explicitly evaluate the gradients for the cost function at each time step, perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
 The weight update for the $i$-th weight at the time step $t$ is given by
@@
 
-One last comment to be made concerning the update rule, is the hipothetical need for \textbf{limits} to the weight values, when the update rule exceeds $|w_{\text{max}}|$ (in that case, we set $w_i^{(t+1)} = \pm w_{\text{max}}$), which sometimes might be needed if the behaviour of the weight update is not appropriate~\cite{Maeda05}.
+One last comment to be made concerning the update rule, is the hypothetical need for \textbf{limits} to the weight values, when the update rule exceeds $|w_{\text{max}}|$ (in that case, we set $w_i^{(t+1)} = \pm w_{\text{max}}$), which sometimes might be needed if the behaviour of the weight update is not appropriate~\cite{Maeda05}.
 
 \section{Proposed Solution}\label{sec:propSol}
-Given the previous discusson of how LSTM neurons work, their benefits when compared to other RNN architectures, this work aims to conceive an efficient dedicated FPGA implementation with on-chip learning that, hopefully, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or from a framework.
+Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive an efficient dedicated FPGA implementation with on-chip learning that, hopefully, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework.
 
@@
 
-This solution will combine the benefits of a dedicated hardware implementation, where the inherent parallelism so obviously visible in the network architecture can be fully exploited something that, predominantly, yields a much higher performance in terms of computation time than a naïve CPU implementation, as in~\cite{Maeda05} where a $21\times$ speed-up against an ARM CPU was acheived.
+This solution will combine the benefits of a dedicated hardware implementation, where the inherent parallelism so obviously visible in the network architecture can be fully exploited something that, predominantly, yields a much higher performance in terms of computation time than a naïve CPU implementation (as in~\cite{Maeda05} where a $21\times$ speed-up against an ARM CPU was acheived) with the flexibility that a reconfigurable platform such as an FPGA provides.

### 2016-02-12 prepReport/workPlan.tex
Implemented the Ivo suggested corrections

@@
 Dwight Eisenhower, United States' President from 1953 to 1961 and Army Officer, said that `no battle was ever won according to plan, but no battle was ever won without one'.
-Hence, although this initial planning draft might not be religiously followed in all its extent, it is important to break down the work in \textbf{tasks} so that I and my supervisors can keep track of what has to be done, and establishing deadlines for the small achievements that, together, will build up to the succesful completion of the proposed work of this thesis.
+Hence, although this initial planning draft might not be religiously followed in all its extent, it is important to break down the work in \textbf{tasks} so that I and my supervisors can keep track of what has to be done, and establishing deadlines for the small achievements that, together, will build up to the successful completion of the proposed work of this thesis.
 The detailed explanation of each task is provided in Section~\ref{sec:wp_tasks_td} and the \textbf{schedule} for them, where their approximate begin and end weeks are marked, is represented by a \emph{Gantt Chart} in Section~\ref{sec:wp_tasks_gantt}.
@@
 \subsection{Task Description}\label{sec:wp_tasks_td}
-The work will be partitioned in \emph{five} large groups of tasks, and each of these groups will be further partitioned in smaller subtaskswhich are the actual assignements that I will be completing.
+The work will be partitioned in \emph{five} large groups of tasks, and each of these groups will be further partitioned in smaller subtasks which are the actual assignments that I will be completing.
 We can think of the group of task as a sort of \emph{logical} grouping of the subtasks, according both to their relatedness and also to their precedence relationships.
@@
 \begin{itemize}
-    \item \textbf{T1 -- Preparation} -- This stage is the initial phase of the work, where I will essentialy setup my working environment and finish the Python modules for the LSTM network that will be used in the validation of my Verilog modules
+    \item \textbf{T1 -- Preparation} -- This stage is the initial phase of the work, where I will essentially setup my working environment and finish the Python modules for the LSTM network that will be used in the validation of my Verilog modules
 
         \begin{itemize}
-            \item \textbf{T1.1 -- Complete Python Class with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Throgh Time version.
+            \item \textbf{T1.1 -- Complete Python Class with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Through Time version.
 
@@
 
-            \item \textbf{T1.3 -- Setup FPGA Development Board} -- this is a farily technical step, where I setup my working environment, and make sure that everything is working correctly and is properly configured.
+            \item \textbf{T1.3 -- Setup FPGA Development Board} -- this is a fairly technical step, where I setup my working environment, and make sure that everything is working correctly and is properly configured.
 As will be pointed out in Section~\ref{sec:wp_sh}, the FPGA development board of choice will be Xilinx~\textregistered Zedboard.
@@
             \item \textbf{T2.3 -- Design LSTM Neuron/Layer Module} -- similar to the previous task, but now I am aiming to implement the actual LSTM Neuron or a layer of LSTM Neurons.
-This is a design decision that will need to be set in \textbf{T2.1}, to asses whether it is justifiable to have dedicated multiplication units assigned to each neuron, or we can perform matrix-like batch multiplications and treat the LSTM as a batch of Neurons, i.e.
+This is a design decision that will need to be set in \textbf{T2.1}, to assess whether it is justifiable to have dedicated multiplication units assigned to each neuron, or we can perform matrix-like batch multiplications and treat the LSTM as a batch of Neurons, i.e.
 as a layer.
@@
         \begin{itemize}
-            \item \textbf{T3.1 -- Synthesis of the SPSA Module} -- the module developed in \textbf{T2.2} is now syet
-                {bar height=.6}thesized and validated
+            \item \textbf{T3.1 -- Synthesis of the SPSA Module} -- the module developed in \textbf{T2.2} is now synthesized and validated.
             \item \textbf{T3.2 -- Synthesis of the LSTM Neuron/Layer Module} -- same as the previous task, but now on the module of \textbf{T2.3}
@@
     \item \textbf{T4 -- Benchmarking the Design} -- at this stage, we have a fully functional FPGA design of an LSTM network with SPSA on-chip training.
-Henceforth, we will need to benchmark its performance by comparing it to~\cite{Tavear13, Hochreiter07}, and prove its usefulness by using it in one or two usecases from~\ref{chap:stateArt}, which are interesting uses of Deep Learning techniques such as these.
+Henceforth, we will need to benchmark its performance by comparing it to~\cite{Tavear13, Hochreiter07}, and prove its usefulness by using it in one or two use-cases from~\ref{chap:stateArt}, which are interesting uses of Deep Learning techniques such as these.
 
         \begin{itemize}
-            \item \textbf{T4.1 -- Benchmark the Performance/Computing Time} -- perform a side-by-side comparison of our model with the software counterpart of\cite{Tavear13, Hochreiter07} and with the only hardware implementation of LSTM of~\cite{Chang15}.
-            \item \textbf{T4.2 -- Reproduce some of the state of the art applications of LSTM} -- in order to be able to show real-life, user-oriented, applications, I understand it is a good idea to try to reproduce some of the state of the art usages of LSTM, outlined in Section~\ref{chap:sa_apps}.
+            \item \textbf{T4.1 -- Benchmark the Performance/Computing Time} -- perform a side-by-side comparison of our model with the software counterpart of~\cite{Tavear13, Hochreiter07} and with the only hardware implementation of LSTM of~\cite{Chang15}.
+            \item \textbf{T4.2 -- Reproduce some of the state of the art applications of LSTM} -- in order to be able to show real-life, user-oriented, applications, I understand it is a good idea to try to reproduce some of the state of the art usages of LSTM, outlined in Section~\ref{sec:sa_apps}.
 It is too early to commit myself to which of these applications I will use, and how many, because that will depend on how well the work on tasks \textbf{T2} and \textbf{T3} is flowing.
-Nevertheless, I intend to at least impement one or two of them.
+Nevertheless, I intend to at least implement one or two of them.
         \end{itemize}
 
-    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are developing, this task will obviously be active, by documenting all the design decisions, milestones acheieved and logging all the results that come from the implementation.
+    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are developing, this task will obviously be active, by documenting all the design decisions, milestones achieved and logging all the results that come from the implementation.
 A couple of weeks before the deadline, this will be the only active task, since producing a clear and well-written document requires concentration and undivided attention.
@@
 The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard} (Figure~\ref{fig:zedboard}, whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and an ARM \textregistered~ Cortex\texttrademark~-A9 processor.
-It is an FPGA development board that features several I/O interfaces such as Etherner, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilites, making it easy to prototype additional features to the core design (i.e.
+It is an FPGA development board that features several I/O interfaces such as Ethernet, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilities, making it easy to prototype additional features to the core design (i.e.
 for building the note transcription system of~\cite{Bock12}, and to make it operate in real-time, I would need an audio input, which I can easily get from the Audio IO, without investing time in connecting the codec and programming the FPGA to interface it).
@@
 
-To write the final Disseration report, I will use \LaTeX (this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings.
+To write the final Dissertation report, I will use \LaTeX (this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings.
 Once again, all of this is free to use.

### 2016-02-12 prepReport/intro.tex
Implemented half of the suggestions of JCF

@@
 \section{Motivation}\label{sec:intro_mot}
-Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be speeded up in a dedicated hardware structure.
+Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be sped up by a dedicated hardware structure.
 
-All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to the date generally have between 2-8 cores (mobile devices and general use personal computers),
+All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to now generally have between 2-8 cores (mobile devices and general use personal computers),
 
-In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the portin process is not entierely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
-This way, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a supposedly lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance.
+In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the porting process is not entierely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
+Therefore, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance.
 Furthermore, when processing an incoming stream of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.
@@
 \section{Objectives}\label{sec:intro_obj}
-Taking into account the considerations presented in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
-Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate the some of the applications portrayed in~\ref{sec:sa_apps}.
+Taking into account the considerations done in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
+Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate some of the applications portrayed in~\ref{sec:sa_apps}.
 Lastly, I will work to make this structure reconfigurable on the go, and trying to minimize the reconfiguration time.
@@
     \item \textbf{Supervisor} -- The \href{https://sigarra.up.pt/feup/pt/func\_geral.formview?p\_codigo=210963}{Professor João Canas Ferreira}, auxiliary professor at the Faculty of Engineering of the University of Porto.
-    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
+    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, MSc, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
 \end{itemize}

### 2016-02-12 prepReport/stateArt.tex
Implemented half of the suggestions of JCF

@@
 
-Over the course of this chapter, I am going to present an overview of the most recent developments related to the work of this thesis, both in terms of existing dedicated hardware implementations~\ref{sec:sa_hardware}, the most relevant work in adapting suitable training algorithms to hardware~\ref{sec:sa_training} and also some of the most relevant applications of LSTM~\ref{sec:sa_apps}, which are not FPGA-based, but demonstrate how LSTM is useful by itself, and how well it competes with other Machine Learning algorithms in terms of long time-series dependences in data.
+Over the course of this chapter, I am going to present an overview of the most recent developments related to the work of this thesis, both in terms of existing dedicated hardware implementations~\ref{sec:sa_hardware}, the most relevant work in adapting suitable training algorithms to hardware~\ref{sec:sa_training} and also some of the most relevant applications of LSTM~\ref{sec:sa_apps}, which are not FPGA-based, but demonstrate how LSTM are useful by themselves, and how well it competes with other Machine Learning algorithms in terms of long time-series dependences in data.
 
 \section{LSTM Applications (non-FPGA related)}\label{sec:sa_apps} 
-LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address the same set of problems where LSTM are employed, namely predicting and producing classification decisions from time-series data \textbf{citation needed!}.
+LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address the same set of problems where LSTM are employed, namely predicting and producing classification decisions from time-series data.
 A very comprehensive description of applications can be found in one of the initial authors webpage dedicated to the subject~\footnote{\href{http://people.idsia.ch/~juergen/rnn.html}{http://people.idsia.ch/~juergen/rnn.html}}.
-I will now enumerate some of the bleeding edge applications of LSTM.
+I will now enumerate some of the leading edge applications of LSTM.
 
 \begin{itemize}
-    \item \textbf{Handwriting Recognition} -- and LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}.
-LSTMs have also proven to surpass HMM based models in terms of optical character recognition in printed text, as~\cite{Breuel13} suggests.
+    \item \textbf{Handwriting Recognition} -- an LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}.
+LSTMs have also proven to surpass HMM-based models in terms of optical character recognition of printed text, as~\cite{Breuel13} suggests.
 
@@
 
-    \item \textbf{Handwriting Synthesis} -- a comprehensive study by Graves shows, among other sequence generation tasks such as Text Prediction, the use of LSTM to produce synthetic handwriting, that looks incredibly similar to human-produced handwriting~\cite{Graves13_2}.
+    \item \textbf{Handwriting Synthesis} -- a comprehensive study by Graves shows, among other sequence generation tasks such as Text Prediction, that use of LSTM to produce synthetic handwriting, that looks incredibly similar to human-produced handwriting~\cite{Graves13_2}.
 
@@
 
-    \item \textbf{Biomedical Applications} -- this network architecture was used in a protein homology detection scheme~\cite{Hochreiter07} using the SCOP 1.53 benchmark dataset, displaying a staggering performance when compared to other methods.
+    \item \textbf{Biomedical Applications} -- this network architecture was used in a protein homology detection scheme~\cite{Hochreiter07} using the SCOP 1.53 benchmark dataset, displaying a good performance when compared to other methods.
 Similarly, a recent article from 2015 by Sønderby et al.
@@
 
-    \item \textbf{Video and Image Analysis} -- Vinyals et al., at Google, proposed and LSTM network for \textbf{image captioning}, preceded by a Convolutional Neural Network to apply preprocessing to the images~\cite{Vinyals14}.
-This way, the LSTM works as \emph{sentence generator}, that captions the images, with state of the art performance.
+    \item \textbf{Video and Image Analysis} -- Vinyals et al., at Google, proposed an LSTM network for \textbf{image captioning}, preceded by a Convolutional Neural Network (CNN) to apply preprocessing to the images~\cite{Vinyals14}.
+The LSTM works as \emph{sentence generator}, that captions the images with state of the art performance.
         Besides image captioning, video captioning is also an interesting topic.
@@
 recently proposed a CNN + LSTM architecture to translate video sequences to natural language~\cite{Donahue14}, using the Microsoft Research Video Description Corpus as a dataset.There are other similar studies that combine image and video captioning~\cite{Donahue14_2}.
-
 \end{itemize}
@@
 It consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered~ Zedboard implementation.
-It uses a character-level language model from Andrej Karpathy, written in Lua using the Torch7 framework (the Lua calls are implemented in C, so no performance is lost).
+It uses a character-level language model from Andrej Karpathy~\footnote{\href{https://github.com/karpathy/char-rnn}{https://github.com/karpathy/char-rnn}}, written in Lua using the Torch7 framework~\footnote{\href{http://torch.ch/}{http://torch.ch/}} (the Lua calls are implemented in C, so no performance is lost).
 
@@
 Although an apparently excessive number, I am left to scrutinize whether the usage of MACs is compulsory to obtain a better performance than a CPU or if it can be discarded, allowing for smaller cells.
-This way, we can have a layer of more neurons occupying the same area and, hopefully, the same, or even less, resources within the FPGA.
+This way, we can have a layer of more neurons occupying the same area and, hopefully, the same, or even fewer, resources within the FPGA.
 
@@
 The SPSA algorithm was initially published by Spall in~\cite{Spall98}, and its theoretical details are outlined in Section~\ref{sec:training_lstm}.
-As of applications of it to the training of general Neural Networks, the earliest examples come from 1995 and 1996 on~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time where the memory resources of digital circuitry were limited, and so most of these structures were analog-based.
+As to its applications to the training of general Neural Networks, the earliest examples come from 1995 and 1996 in~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time where the memory resources of digital circuitry were limited, and so most of these structures were analog-based.
 Its adequacy was also established for \textbf{control problems}, such as those proposed in~\cite{Figueiredo97}, where a Time Delay Neural Network is used to control an unknown plant in a linear feedback system.
 
-In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network in an FPGA (and thus a digital system), achieving promising results in an Altera \textregistered~ FPGA.
+In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network in an FPGA (and thus a digital system), achieving promising results in an Altera FPGA.
 The article carefully delineates the approach taken, and also the hardware architecture designed, so it is a very good reference for the design that I will have to implement.

### 2016-02-12 prepReport/theorBack.tex
Implemented half of the suggestions of JCF

@@
 Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation.
-Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters accordingly with the input data it receives, improving the estimates it yields as it receives new input samples.
+Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters according to the input data it receives, improving the estimates it yields as it receives new input samples.
 
@@
 For instance, suppose we want to build a program that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary.
-We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from it being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language.
-The approach taken by Machine Learning is different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 \; \mb{x}_2 \;\mb{x}_2 \; \dots \mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model.
+We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language.
+The approach taken by Machine Learning is different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 ,\; \mb{x}_2 ,\;\mb{x}_2 ,\; \dots ,\;\mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model.
 Each of the $i$-th data point was previously labelled, and in fact we feed each training data point $\mb{x}_i$ \emph{along} with its corresponding label $t_i$, so that the model can adapt its parameters accordingly to the \emph{target value} it is supposed to classify.
-This set of labels $\mb{t} = \left[ t_1 \; t_2 \; t_3 \; \dots \; t_N \right]$ is called the \textbf{target vector}.
+This set of labels $\mb{t} = \left[ t_1 ,\; t_2 ,\; t_3 ,\; \dots ,\; t_N \right]$ is called the \textbf{target vector}.
 
@@
 
-Now that the basic Machine Learning concepts have been presented, I will discuss, henceforth, one of the most important algorithms that address the supervised classification problems, the \emph{Artificial Neural Networks} that will be discussed in Section~\ref{sec:theorBack_ann} as somewhat of a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}), both of which are improvements over the initial formulation of the ANNs.
+Now that the basic Machine Learning concepts have been presented, I will discuss, henceforth, one of the most important algorithms that address the supervised classification problems, the \emph{Artificial Neural Networks} that will be discussed in Section~\ref{sec:theorBack_ann} as a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}), both of which are improvements over the initial formulation of the ANNs.
 These two last networks branch even further from these set of problems, and are usually employed in \textit{Deep Learning} tasks, where we try to extract even higher level information from data at the expense of increased model complexity.
@@
     \caption[Possible activation functions of choice for a neuron]{Three different activation functions.
-As you can see, the hyperbolic tangent has the same extreme value as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision.
-On the other hand, the sigmoid function goes from zero to one, and is also smooth like the hyperbolic tangent}.
+As you can see, the hyperbolic tangent has the same extreme values as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision.
+On the other hand, the sigmoid function goes from zero to one, and is also smooth like the hyperbolic tangent.}
 	\label{fig:activFunc}
@@
 We call \emph{layers} to a group of these nodes that occupy the same hierarchical position.
-There can be any number of layers, with any number of nodes, but most implementations generally have 3 layers: the \emph{initial} layer, the \emph{hidden} layer (in the middle) and the \emph{output} layer.
+There can be any number of layers, with any number of nodes, but most implementations generally have 3 layers: the \emph{input} layer, the \emph{hidden} layer (in the middle) and the \emph{output} layer.
 Figure~\ref{fig:neuralnet} suggests a possible structure for a 3 layer ANN.
@@
 The interpretation of these equations is simple.
-If $w_{ji}$ is the weight of the connection between the neuron $j$ we are considering and a neuron $i$ in a previous layer, then the sum over $k$ relates to all the neurons in the \emph{next} layer to which $j$ connects: this way, since the update on $w_{ji}$, according to~\ref{eq:gradDesc}, is given by
+If $w_{ji}$ is the weight of the connection between the neuron $j$ we are considering and a neuron $i$ in a previous layer, then the sum over $k$ relates to all the neurons in the \emph{next} layer to which $j$ connects: this way, since the update of $w_{ji}$, according to~\ref{eq:gradDesc}, is given by
 
@@
 The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron_unf}.
-Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simultaneous Perturbation Stochastic Approximation}.
+Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simultaneous Perturbation Stochastic Approximation} will be evaluated.
 
@@
 To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}.
-This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support for the work of this thesis.
+This section explains the main idea of this approach (Section ~\ref{sec:struct_lstm}), and also how it is trained (Section~\ref{sec:training_lstm}), serving as a support for the work of this thesis.
 Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}.
 One of the inital proposers of LSTM, Prof.
-Jürgen Schmidhüber, did a survey on the most common variations of the model last year~\cite{Greff15}, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.
+Jürgen Schmidhüber, did a survey on the most common variations of the model last year~\cite{Greff15}, and this will the basis of this short theoretical presentation, as well as the work that will be developed in this thesis.
 
@@
 As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network.
-Although ommited from the  picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}.
+Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}.
 The main components, their role and relevance, are explained as follows
@@
 It also receives the input vector and the previous output vector as inputs, but it does not have peephole connections and its dynamics are controlled by a different set of weights.
-The \textbf{activation function of this gate can be any}, but the most common choice is the \textbf{Hyperbolic Tangent}.
+The \textbf{activation function of this gate can be either}, but the most common choice is the \textbf{Hyperbolic Tangent}.
 
@@
 This is the memory element of the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates that connect to it.
-Also, this is where the peephole connections come from:
+Also, this is where the peephole connections come from.
 
@@
 
-    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates to `peep' the states of the memory cell.
+    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates to `peep' at the states of the memory cell.
 They were added after the initial 1997 formulation, and their absence was proven to have a minimal performance impact~\cite{Greff15}.
@@
 \end{align}
+where $\odot$ is the Hadamard multiplication.
 The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer.
@@
 Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme.
-Since I am aiming at an hardware implementation, and although the memory resources of nowadays FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system.
+Since I am aiming at an hardware implementation, and although the memory resources of current FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system.
 That being said, we see that the usual training algorithms for LSTM cells -- i.e.
@@
 A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.
-The main idea of this technique is, instead of explicitly evaluate the gradients for the cost function at each time step, perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
+The main idea of this technique is, instead of explicitly evaluating the gradients for the cost function at each time step, to perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
 The weight update for the $i$-th weight at the time step $t$ is given by
@@
 This way, we see that every weight is \textbf{randomly} incremented either by $-\beta$ or $\beta$, and we only need to keep a duplicate of the weight matrix with the perturbations, and we only need to evaluate the cost function twice per incoming sample.
-As for the \textit{update rule} is concerned, we have
+As for the \textit{update rule}, we have
 
@@
 \section{Proposed Solution}\label{sec:propSol}
-Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive an efficient dedicated FPGA implementation with on-chip learning that, hopefully, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework.
+Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\re (see Section~\ref{sec:wp_sh}).
 
-As discussed in Chapter~\ref{chap:stateArt}, hitherto there is only one FPGA implementation of LSTM cells~\cite{Chang15}, but there is no on-chip learning on the architecture (the weights are learned offline, and loaded in the FPGA before operation, and they are not changed while the FPGA system is operating), and my proposed work will implement an efficient LSTM implementation \emph{with} the ability to learn on-chip, which is an innovative and useful feature for the LSTM architecture on a dedicated hardware implementation.
+As discussed in Chapter~\ref{chap:stateArt}, up to now there is only one FPGA implementation of LSTM cells~\cite{Chang15}, but there is no on-chip learning on the architecture (the weights are learned offline, and loaded in the FPGA before operation, and they are not changed while the FPGA system is operating), and my proposed work will implement an efficient LSTM implementation \emph{with} the ability to learn on-chip, which is an innovative and useful feature for the LSTM architecture on a dedicated hardware implementation.
 I will achieve that using the SPSA method described in~\ref{sec:training_lstm}, and inspired in the previous work of~\cite{Maeda05}.

### 2016-02-12 prepReport/earlyWork.tex
Applied all improvements suggested

@@
 \chapter{Early Work}\label{chap:earlyWork}
-In order to better undesrtand LSTM,  gain some useful insight on their operation, and to have a software implementation, that is not part of any framework, with which I can validate in the future my FPGA implementation, I have developed an fully functional LSTM Layer in Python and Numpy, that still uses BPTT as a training algorithm (the governing equations are in appendix A of~\cite{Greff15}).
-The idea is to adapt it to use SPSA in the beginning of my Master's Thesis.
+In order to better undesrtand LSTM,  gain some useful insight on their operation, and to have a software implementation, that is not part of any framework, with which I can validate in the future my FPGA implementation, I have developed a fully functional LSTM Layer in Python and Numpy, that still uses BPTT as a training algorithm (the governing equations are in appendix A of~\cite{Greff15}).
+The idea is to adapt it to use SPSA in the beginning of my Master's work.
 The code is present in

### 2016-02-12 prepReport/theorBack.tex
Applied all improvements suggested

@@
 
-    \caption[Representation of a Recurrent Neuron (Figure~\ref{fig:recneuron}) and a Recurrent Neuron unfolded through time (Figure~\ref{recneuron_unf})]{In Figure~\ref{fig:recneuron}, a recurrent neuron is depicted, where the output is fed back to the input, weighted by $\theta_h$.
+    \caption[Representation of a Recurrent Neuron (Figure~\ref{fig:recneuron}) and a Recurrent Neuron unfolded through time (Figure~\ref{fig:recneuron_unf})]{In Figure~\ref{fig:recneuron}, a recurrent neuron is depicted, where the output is fed back to the input, weighted by $\theta_h$.
 Figure~\ref{fig:recneuron_unf}, an unfolding through time of that same neuron is performed.
@@
 As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network.
-Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}.
+Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}.
 The main components, their role and relevance, are explained as follows
@@
 
-\begin{align}\label{eq:equationsLSTM}
-    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
-    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
-    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
-    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
-    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
-    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
+\begin{align}
+    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \notag\\
+    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \notag\\
+    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \notag\\
+    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \notag \\
+    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \notag\\
+    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
 \end{align}
@@
 If the random sign vector $\mb{s}^{(t)}$ is chosen carefully so that the expected value of the vector at the $i$-th position is zero, $E(s_i^{(t)})$, and the signs of two different components $i$ and $j$, at different time-steps, are independent (i.e.
-$E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, equation~\ref{eq:taylor} reduces to
+$E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, where $\delta$ is the Kronecker fuction), taking the \emph{Expected Value} of equation~\ref{eq:taylor} yields
 
@@
 \section{Proposed Solution}\label{sec:propSol}
-Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\re (see Section~\ref{sec:wp_sh}).
+Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\ref{sec:wp_sh}).
 

### 2016-02-12 prepReport/workPlan.tex
Applied all improvements suggested

@@
 Hence, although this initial planning draft might not be religiously followed in all its extent, it is important to break down the work in \textbf{tasks} so that I and my supervisors can keep track of what has to be done, and establishing deadlines for the small achievements that, together, will build up to the successful completion of the proposed work of this thesis.
-The detailed explanation of each task is provided in Section~\ref{sec:wp_tasks_td} and the \textbf{schedule} for them, where their approximate begin and end weeks are marked, is represented by a \emph{Gantt Chart} in Section~\ref{sec:wp_tasks_gantt}.
+The detailed explanation of each task is provided in Section~\ref{sec:wp_tasks_td} and the \textbf{schedule} for them, where their estimated begin and end weeks are marked, is represented by a \emph{Gantt Chart} in Section~\ref{sec:wp_tasks_gantt}.
 
@@
 We can think of the group of task as a sort of \emph{logical} grouping of the subtasks, according both to their relatedness and also to their precedence relationships.
-The task groups, and their related subtasks, are as follows
+The task groups, and their related subtasks, are as follows:
 
@@
         \begin{itemize}
-            \item \textbf{T1.1 -- Complete Python Class with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Through Time version.
+            \item \textbf{T1.1 -- Complete Python Implementation with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Through Time version.
 
-            \item \textbf{T1.2 -- Reproduce the Benchmark Results with the Python Class} -- after developing and testing the Python Class with SPSA, I will try to reproduce with it the results of~\cite{Tavear13}, in order for me to have a high-level model whose performance is comparable to~\cite{Tavear13, Hochreiter07}, and thus provide a good performance baseline for this and other future benchmark tests that I would like to run in order to compare my FPGA solution with a software solution.
+            \item \textbf{T1.2 -- Reproduce the Benchmark Results with the Python Implementation} -- after developing and testing the Python Class with SPSA, I will try to reproduce with it the results of~\cite{Tavear13}, in order for me to have a high-level model whose performance is comparable to~\cite{Tavear13, Hochreiter07}, and thus provide a good performance baseline for this and other future benchmark tests that I would like to run in order to compare my FPGA solution with a software solution.
 
@@
         \begin{itemize}
-            \item \textbf{T2.1 -- Design Planning} -- before performing and HDL coding, it is favourable to first assemble the system design strategy in paper.
+            \item \textbf{T2.1 -- Design Planning} -- before performing and HDL coding, it is desirable first assemble the system design strategy in paper.
 This is where, together with my supervisor, I will outline the internal organization of my system, and make the fundamental design decisions.
-Of course, this task will prolong during all the \textbf{T2} extent, since some design changes to the original plan will have to be performed as I find \emph{bugs} in my early Verilog implementations that force a design correction.
+Of course, this task will continue during the entire of \textbf{T2} extent, since some design changes to the original plan will have to be performed as I find \emph{bugs} in my early Verilog implementations that force a design correction.
 
@@
 Furthermore, I will compare the output of this model with the golden output produced by a standalone SPSA Python Class previously developed in \textbf{T1}.
-            \item \textbf{T2.3 -- Design LSTM Neuron/Layer Module} -- similar to the previous task, but now I am aiming to implement the actual LSTM Neuron or a layer of LSTM Neurons.
+
+            \item \textbf{T2.3 -- Design LSTM Neuron/Layer Module} -- similarly to the previous task, but now I am aiming to implement the actual LSTM Neuron or a layer of LSTM Neurons.
 This is a design decision that will need to be set in \textbf{T2.1}, to assess whether it is justifiable to have dedicated multiplication units assigned to each neuron, or we can perform matrix-like batch multiplications and treat the LSTM as a batch of Neurons, i.e.
@@
 
-    \item \textbf{T3 -- FPGA Synthesis of the Verilog Code} -- after having Verilog code that passed all the validation tests and whose correct functionality is, therefore, asserted, I will synthesize the Verilog modules to FPGA RTL blocks, and perform a new simulation on these RTL syntesized code, that unlike the previous behavioural simulation, now has timing information resulting from the FPGA block mapping, and the place and routing of those mapped blocks.
+    \item \textbf{T3 -- FPGA Synthesis of the Verilog Code} -- after having Verilog code that passed all the validation tests and whose correct functionality is, therefore, ensured, I will synthesize the Verilog modules to FPGA RTL blocks, and perform a new simulation on these RTL syntesized code, that unlike the previous behavioural simulation, now has timing information resulting from the FPGA block mapping, and the place and routing of those mapped blocks.
 
@@
 
-    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are developing, this task will obviously be active, by documenting all the design decisions, milestones achieved and logging all the results that come from the implementation.
+    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are progressing, this task will obviously be active, by documenting all the design decisions, milestones achieved and logging all the results that come from the implementation.
 A couple of weeks before the deadline, this will be the only active task, since producing a clear and well-written document requires concentration and undivided attention.
@@
 \subsection{Hardware Resources}
-The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard} (Figure~\ref{fig:zedboard}, whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and an ARM \textregistered~ Cortex\texttrademark~-A9 processor.
+The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard} (Figure~\ref{fig:zedboard}), whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and two ARM \textregistered~ Cortex\texttrademark~-A9 processors.
 It is an FPGA development board that features several I/O interfaces such as Ethernet, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilities, making it easy to prototype additional features to the core design (i.e.
@@
 
-To write the final Dissertation report, I will use \LaTeX (this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings.
-Once again, all of this is free to use.
+To write the final Dissertation report, I will use \LaTeX~(this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings.
+Once again, all of these are free to use.

### 2016-02-12 prepReport/workPlan.tex
Corrected References

@@
 
-For the Python Classes, I will make use of the 3.3 version of the Python distribution, the high-performance numerical computation library Numpy\footnote{\href{http://www.numpy.org/}{http://www.numpy.org/}}, which is a standard in scientific computing, and the Matplotlib\footnote{\href{http://matplotlib.org/}{http://matplotlib.org/}} plotting library.
+For the Python implementation, I will make use of the 3.3 version of the Python distribution, the high-performance numerical computation library Numpy\footnote{\href{http://www.numpy.org/}{http://www.numpy.org/}}, which is a standard in scientific computing, and the Matplotlib\footnote{\href{http://matplotlib.org/}{http://matplotlib.org/}} plotting library.
 All of this software is also free to use, and open-source.

### 2016-02-12 prepReport/intro.tex
Corrected References
FINAL Version

@@
 
-In Chapter~\ref{chap:workPlan}, the work plan for the Dissertation work is detailed as \emph{tasks} in Section~\ref{sec:wp_tasks_td}, and their temporal arrangement is pictured in a Gantt chart in Section~\ref{sec:wp_tasks_gantt}.
+In Chapter~\ref{chap:earlyWork} I will present some of the work that I have already concluded.
+
+Finally, in Chapter~\ref{chap:workPlan}, the work plan for the Dissertation work is detailed as \emph{tasks} in Section~\ref{sec:wp_tasks_td}, and their temporal arrangement is pictured in a Gantt chart in Section~\ref{sec:wp_tasks_gantt}.
 Furthermore, the description of the Hardware and Software resources that I will be utilizing during the course of my work is done in Section~\ref{sec:wp_sh}
-
-Finally, in Chapter~\ref{chap:earlyWork} I will present some of the work that I have already concluded.
 

### 2016-06-19 finalText/proposedSol.tex
Proposed Solution chapter almost done

\chapter{Proposed Architecture}\label{chap:propSol}

\section{Network Organization}

\section{Forward Propagation Circuit Overview}

\subsection{Finite-precision representation system}
Before discussing any details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding to how we numbers envolved in the calculations will be represented. For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point representation systems, as well as the bitwidth and precision chosen for then number representation system of the network, and Section~\ref{sec:convrulesfp} states how the conversion between real and fixed-point numbers is performed. Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic imposes. The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.

\subsubsection{Precision bitwidth}\label{sec:precbit} 
Since we are dealing with real numbers, and I plan to make use of the DSP48 slices within the FPGA, I chose to use an 18-bit signed fixed-point system, with the sign information coded as 2's complement. Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit. This way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$. In terms of range of representation, the maximum positive number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real number (the decimal point is at the $m$-th bit),

\begin{equation}\label{eq:maxpos}
    \text{Max. Positive Number} = \frac{2^{N-1}-1}{2^m} = 2^{N-1-m} - \frac{1}{2^m}
\end{equation}
and the smallest negative number is simply the MSB set to one (and also shifted appropriately),

\begin{equation}\label{eq:minneg}
    \text{Small. Negative Number} = -\frac{2^{N-1}}{2^m} = -2^{N-1-m}
\end{equation}

Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, our fractional part precision should be, at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$. On the other hand, given that in the Python experiments I have attested that the weight values generaly do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to not be very small in order to accomodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$. Since $18=n+m+1$, then $n = 17 - m = 6$ and the representation system to be used is $Q6.11$. According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented with this choice of $n$ and $m$ are in the range

\begin{equation}\label{eq:rangeQ611}
    -64 \leq x \leq 63.99951172
\end{equation}
with a minimum resolution of $2^{-11} = 0.00048828125$.

\subsubsection{Conversion between real and fixed-point}\label{sec:convrulesfp}
In order to find the real number equivalent of a $Qn.m$ fixed-point system, and \textit{vice-versa}, we need to take into account that, according to Section~\ref{sec:precbit}, the $i$-th position bit is worth $2^{i-m}$, and therefore the decimal point in this fixed-point system is at position $i=m$, since $2^{m-m} = 1$. This way, the rules are as follows

\begin{itemize}
    \item \textbf{Positive Real to $Qn.m$} -- since the $1$ is at bit $m$, we simply multiply the real number by $2^m$, and discard the fractional part of the result
    \item \textbf{Negative Real to $Qn.m$} -- we disregard the sign in the real number, and perform the same operation as before, but we then convert the resulting binary number to two's complement, i.e. by performing bitwise negation, followed by summing 1.
    \item \textbf{Positive $Qn.m$ to real} -- multiply the fixed-point number by $2^{-m}$, to shift the decimal point back by $m$ positions. 
    \item \textbf{Negative $Qn.m$ to real} -- convert from two's complement by performing bitwise logic negation, followed by summing one; then scale the decimal point back by $m$ positions by multiplying by $2^{-m}$. 
\end{itemize}

\subsubsection{Fixed-point arithmetic rules}\label{sec:arithrulesfp}
The three main operations needed in my network design are \textbf{signed sums}, \textbf{signed multiplications} and \textbf{arithmetic shifts} (i.e. the ones that preserve the sign of the MSB) to implement divisions/multiplications by powers of two. In terms of signed sums, the rule is simple: both numbers need to be scaled to the same base, with their $m$'s being same, so that the decimal point is in the same place in both numbers ($n$, however, can be different, since that only means that one number is longer than the other, and the missing bits in the smaller one can be interpreted as zeros).

For signed multiplication, since both operands are in fixed-point $Qn.m$, and are thus scaled by $2^m$, we need to scale the result by multiplying it by $2^{-m}$ (or perform an arithmetic right shift of $m$ bits). This is because, if $a$ and $b$ the real numbers to be multiplied, we have that in fixed-point arithmetic, the result $c$ is

\begin{equation}\label{eq:multfp}
    (a\cdot2^m) \cdot (b\cdot2^m) = c \cdot 2^{2m}.
\end{equation}
Since in $Qn.m$, all numbers are represented as $r 2^{m}$ with respect to their real counterpart $r$, we need to scale back the decimal point in the result of Equation~\ref{eq:multfp}. We can see that it can be easily achieved by dividing by $2^m$, as stated in the previous paragraph.

\subsection{Non-linearity Calculator}
In order to evaluate the non-linear activation functions $\sigma(x)$ and $\tanh(x)$, since there is no algorithm that can directly compute them, I had to find a suitable way to compute them accurately using a finite number of elementary operations, multiplications and additions, that can be performed efficiently by specially tailored blocks within the FPGA (DSP48 slices for multiplication, for instance). For that purpose, after using~\cite{Muller05} as a reference on elementary function approximation algorithms, I decided to use \textbf{Polynomial Approximations}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).

\subsubsection{Theoretical considerations on the approximation method}
The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of degree $n$, in order to meet the optimization criteria chosen \textit{a priori}. This optimization criteria can be either the well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation. Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error does not exceed the precision limit of this representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded. Weierstrass's Theorem~\cite{Weierstrass1885}, from 1885, guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$. Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynimal approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}. This leads to a linear system of $n+2$ equations whose $i$-th line is given by

\begin{equation}\label{eq:remezline}
    p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
\end{equation}

The optimal coefficients of this \textit{minimax} polynomial are found using \textit{Remez Algorithm}, which provides an iterative approach to solve the linear system given by Equation~\ref{eq:remezline} by finding, in each iteration, the $n+2$ set of points $x_i$ of Chebyshev's Theorem that minimize the error function to $\epsilon$. The algorithm operations are as follows

\begin{enumerate}
    \item Initializing the set $x_i$ of points to $x_i = \frac{a+b}{2} + \frac{(b-a)}{2}\cos\left(\frac{i\pi}{n+1}\right), \, 0 \leq i \leq n+1$ -- 
    \item Solve the system in~\ref{eq:remezline}
    \item Given the polynomial coefficients yielded by step 2, compute the $y_i$ points that minimize $p(x)-f(x)$, and replace the $x_i$s of the next iteration by these $y_i$s. Go to step 2 until $\epsilon$ does not decrease.
\end{enumerate}

On one hand, the system of~\ref{eq:remezline} of Step 2 can be written in matricial notation as

\begin{equation}
\begin{bmatrix} 1 & x_0 & x_0^2 & \cdots & x_0^n & - 1 \\ 1 & x_1 & x_1^2 & \cdots & x_1^n & + 1 \\  &  &  & \vdots & & \\ 1 & x_{n+1} & x_{n+1}^2 & \cdots & x_{n+1}^n & (-1)^{n+1} \end{bmatrix}
\begin{bmatrix} p_0 \\ p_1 \\ p_2 \\ \vdots \\ p_n \\ \epsilon \end{bmatrix} = \begin{bmatrix} f(x_0) \\ f(x_1) \\ \vdots \\ f(x_{n+1}) \end{bmatrix}
\end{equation}
and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the $x_i$s matrix and $b$ is the $f(x_i)$s vector. The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy}, as follows
\lstinputlisting[language=Python, caption=Python script that implements Remez's algorithm, label={lst:remezpy}]{remez.py}

Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, I chose to partition the domain of the activation functions in \textbf{6 intervals}, and approximate each of those intervals using polynomials of degree $n=2$. This proved to yield a lower overall approximation error, as expected (the interval on which to perform the approximation is smaller). Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal assymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding assymptote. The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are

\begin{equation}\label{eq:coefs_sigm}
\hat{\sigma(x)} = \left\{ 
\begin{array}{lc} 
0 & x  \leq -6 \\
0.20323428 + 0.0717631x + 0.00642858x^2 & -6 \leq x \leq -3 \\
0.50195831 + 0.27269294x + 0.04059181x^2 & -3 \leq  x \leq 0 \\
0.49805785 + 0.27266221x - 0.04058115x^2 &  0 \leq  x \leq 3 \\
0.7967568 + 0.07175359x - 0.00642671x^2 & 3 \leq  x \leq 6 \\
1 & x > 6
\end{array}
\right.
\end{equation}
for the sigmoid function and

\begin{equation}\label{eq:coefs_tanh}
\hat{\tanh(x)} = \left\{ 
\begin{array}{lc} 
-1 & x  \leq -3 \\
-0.39814608 + 0.46527859x + 0.09007576x^2 & -3 \leq x \leq -1 \\
0.0031444 + 1.08381219x + 0.31592922x^2 & -1 \leq  x \leq 0 \\
-0.00349517 + 1.08538355x -0.31676793x^2 &  0 \leq  x \leq 1 \\
0.39878032 + 0.46509003x - 0.09013554x^2 & 1 \leq  x \leq 3 \\
1 & x > 3
\end{array}
\right.
\end{equation}
for the hyperbolic tangent function. These constants were converted to $Qn.m$ using the rules in~\ref{sec:convrulesfp}, and embedded in the HDL model.

\subsubsection{Hardware Implementation}
The hardware module that implements these non-linearities is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$. This last functionality
is implemented using a simple multiplexer that loads the signals $p_0$, $p_1$ and $p_2$ with the coefficients of Equations~\ref{eq:coefs_sigm} and~\ref{eq:coefs_tanh} based on the value of the input operand $x$. As for the polynomial calculator is concerned, although we could use a full-pipelined evaluator, that would require two DSP slices -- to perform the two simultaneous multiplications -- but the DSP slices will be heavily used in the matrix-vector calculators, so it is advisable to save them for that purpose. A simpler approach is to note that by factoring $x$ we get

\begin{equation}\label{eq:factorPol}
p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
\end{equation}
where we can note that this operation can be divided in a two-step procedure of a simultaneous mutiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$. The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}. Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Numpy as reference) for both activation functions.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/non-lin.eps}
    \caption[Block Diagram of the Non-linearity Calculator]{Block Diagram of the Non-linearity Calculator using a single multiplication. The multiplexer state is controlled by the flip-flip and sum block, that change state every clock cycle. When reset is applied, the selector is set to zero.}
    \label{fig:nonlin}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.85\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
        \caption[Plot of the output of the Sigmoid Calculator HDL module]{Plot of the output of the Sigmoid Calculator HDL module}
        \label{fig:nonlin-out-sig}
    \end{subfigure}
    \begin{subfigure}{0.85\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
        \caption[Plot of the output of the Hyperbolic Tangent HDL module]{Plot of the output of the Hyperbolic Tangent HDL module}
        \label{fig:nonlin-out-tanh}
    \end{subfigure}
    \caption[The output of the HDL implementation of the activation functions]{The output of the HDL implementation of the activation functions}
    \label{fig:nonlin-out}
\end{figure}

\subsection{Matrix-vector Dot Product Unit}\label{sec:dotprod_sec}
From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accomodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has lenght $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$. Plus, if we need a layer with different parameters, either in terms of number of inputs or number of neurons, it advisable to only change the respective parameter at the synthesis stage instead of having to redesign the whole block for that particular size.

The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. It can be seen that each of the $i$-th component of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix. If we follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.

Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: before, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, if we define a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles. Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.

In Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} are depicted a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before. Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}. The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/mem-array-prod.eps}
    \caption[A column of the matrix that serves as input to the module]{A column of the matrix that serves as input to the module. The dark shaded part is for the first multiplier, and the light shaded is for the other, in parallel. The rowMux signal addresses the position within each shaded area}.
    \label{fig:mem-arrayprod}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/array-prod.eps}
    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure}
In summary, for an $N \times M$ weight matrix, this module outputs the result vector in $K_G \cdot M$ clock cycles.

\subsection{Dual-port Block RAM}
Although the bias weights, being simply a vector, I have chosen to keep them in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different. Since the
network size can be big, it is good to make sure that they are placed in the Block RAMs available in the FPGA. In Section~\ref{sec:dotprod_sec}, we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column by column access to the weight matrix, and therefore this block outputs a given column in the \verb+rowOut+ output port terminal, at the \emph{negative edge} of the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock. As far as \textbf{writing} to the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{negative edge} of the clock, and is placed at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.

Note that, for an $N \times M$ matrix, since the memory outputs and writes one column at a time, both the input and output port terminals will carry $N$ weights, and a total bitwidth of $\verb+BITWIDTH+ \cdot N$.

The Verilog coding followed the Verilog Coding Guidelines in Xilinx's \href{http://www.xilinx.com/support/documentation/sw_manuals/xilinx2015_2/ug901-vivado-synthesis.pdf}{UG901} that, in page 50, recommends that the \verb+RAM_STYLE+ parameter be set to \verb+block+. This can be done by adding the follwing compiler directive before the register definition, as follows

\begin{verbatim}
(* ram_style = "block" *) reg [PORT_BITWIDTH-1:0] RAM_matrix [NCOL-1:0];
\end{verbatim}
where we can see that each register contains the respective column of the matrix, with $\verb+PORT_BITWIDTH+ = \verb+BITWDITH+ \cdot M$ and $\verb+NCOL+ = M$.

\subsection{Gate Module}

\subsection{Timing Diagrams of Operation}

### 2016-06-19 finalText/stateArt.tex
Proposed Solution chapter almost done

--- 2016-02-12 prepReport/stateArt.tex
@@
 
-Furthermore, a 2013 article by Tavear et al.~\cite{Tavear13} proposes, for the first time, using SPSA to train LSTM Neurons, although the article focuses on proving the suitability of SPSA to LSTM, and no factual hardware implementation is done or proposed.
+Furthermore, a 2013 article by Tavear et al.~\cite{Tavear13} proposes, for the first time, using SPSA to train LSTM Neurons, although the article focuses on proving the suitability of SPSA to LSTM, and no actual hardware implementation is done or proposed.
 The authors simply demonstrate the suitability using conceptual arguments and by building a software model of an SPSA-trained LSTM network, and by comparing both the performance and computing speed of their model with the results achieved by Hochreiter et al.
@@
 \section{Final Overview}\label{sec:overview}
-As we could attest from this small literature survey, although there is already an hardware implementation os LSTM, there is still a good deal of room for improvement by adding on chip-learning to the system, and also to restrict it to a smaller use of the FPGA resources, allowing it to accomodate a more complex network.
+As we could attest from this small literature survey, although there is already an hardware implementation of LSTM, there is still a good deal of room for improvement by adding on chip-learning to the system, and also to restrict it to a smaller use of the FPGA resources, allowing it to accomodate a more complex network.
 Furthermore,~\cite{Tavear13} shows that, at least for that particular case, the LSTM network doesn't suffer a great performance impact from using SPSA training, as opposed to the more common BPTT, and~\cite{Maeda05} showed that an SPSA hardware implementation is feasible.
-These three conclusions combined indicate that the idea of a hardware LSTM network with on-chip learning using SPSA is possible.
+These three conclusions combined indicate that the idea of a hardware LSTM network with on-chip learning using SPSA is viable.
 
-Besides the proposed hardware implementation, it would be advantageous to perform \emph{benchmarks} to see how well it compares with software solutions, and Section~\ref{sec:sa_apps} suggests a handful of examples to test my final solution.
+Besides the proposed hardware implementation, it would be advantageous to perform \emph{benchmarking} to see how well it compares with software solutions, and Section~\ref{sec:sa_apps} suggests a handful of examples to test my final solution.

### 2016-06-19 finalText/code/remez.py

@@ -1,0 +1,84 @@
+"""
+Remez Algorithm for minimax polynomial approximation of transcendental functions
+
+José Pedro Castro Fonseca, 2016
+
+University of Porto, Portugal
+
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.signal import argrelmax
+
+# The target approximation function
+def f(x):
+    return np.tanh(x)
+
+# Nber of iterations
+numIter = 8
+ 
+# Approximation Interval
+b = 3
+a = 1
+
+# Fixed-point system precision
+prec = 2**(-10)
+t = np.arange(a, b, prec)
+
+# Polynomial Degree
+n = 2 
+
+# Initial set of points
+x = np.zeros((n+2,1))
+for i in reversed(range(n+2)): x[n+1-i,0] = (a+b)/2+(b-a)/2*np.cos(i*np.pi/(n+1)) 
+
+# The numerical matrices
+A = np.zeros((n+2, n+2))
+A[:,0] = 1
+for i in range(n+2): A[i,n+1] = (-1)**(i+1)
+
+c = np.zeros((n+2,1))
+
+# The algorithm iterations
+for it in range(numIter):
+    # Compute the A matrix    
+    for i in range(1,n+1):
+        A[:,i] = (x**i).T   
+    # Compute the b matrix
+    for i in range(n+2): c[i] = f(x[i])
+
+    # Solve the system
+    p = np.dot( np.linalg.inv(A), c )
+
+    # Recompute the x point vector
+    pol = np.poly1d(np.reshape(np.flipud(p[0:n+1]).T, (n+1)))
+    diff = abs(pol(t) - f(t))
+
+    extremaIndices = argrelmax(abs(diff), axis=0, mode='wrap')[0]
+    if len(extremaIndices) < (n+2) : 
+        extremaIndices = np.resize(extremaIndices, (1,n+2))
+        if abs(diff[t[len(t)-1]]) > abs(diff[t[len(t)-2]]):
+            extremaIndices[0,n+1] = len(t)-1
+        else:
+            extremaIndices = np.roll(extremaIndices, 1)
+            extremaIndices[0,0] = 0
+    
+    x = t[extremaIndices].T
+    #Prints Progress
+    print("Error: ", max(abs(diff)))
+    
+
+print("*************************") 
+print("Coefficients: ", p[0:n+1].T)
[…]

### 2016-06-19 finalText/remez.py

@@ -1,0 +1,84 @@
+"""
+Remez Algorithm for minimax polynomial approximation of transcendental functions
+
+Jose Pedro Castro Fonseca, 2016
+
+University of Porto, Portugal
+
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.signal import argrelmax
+
+# The target approximation function
+def f(x):
+    return np.tanh(x)
+
+# Nber of iterations
+numIter = 8
+ 
+# Approximation Interval
+b = 3
+a = 1
+
+# Fixed-point system precision
+prec = 2**(-10)
+t = np.arange(a, b, prec)
+
+# Polynomial Degree
+n = 2 
+
+# Initial set of points
+x = np.zeros((n+2,1))
+for i in reversed(range(n+2)): x[n+1-i,0] = (a+b)/2+(b-a)/2*np.cos(i*np.pi/(n+1)) 
+
+# The numerical matrices
+A = np.zeros((n+2, n+2))
+A[:,0] = 1
+for i in range(n+2): A[i,n+1] = (-1)**(i+1)
+
+c = np.zeros((n+2,1))
+
+# The algorithm iterations
+for it in range(numIter):
+    # Compute the A matrix    
+    for i in range(1,n+1):
+        A[:,i] = (x**i).T   
+    # Compute the b matrix
+    for i in range(n+2): c[i] = f(x[i])
+
+    # Solve the system
+    p = np.dot( np.linalg.inv(A), c )
+
+    # Recompute the x point vector
+    pol = np.poly1d(np.reshape(np.flipud(p[0:n+1]).T, (n+1)))
+    diff = abs(pol(t) - f(t))
+
+    extremaIndices = argrelmax(abs(diff), axis=0, mode='wrap')[0]
+    if len(extremaIndices) < (n+2) : 
+        extremaIndices = np.resize(extremaIndices, (1,n+2))
+        if abs(diff[t[len(t)-1]]) > abs(diff[t[len(t)-2]]):
+            extremaIndices[0,n+1] = len(t)-1
+        else:
+            extremaIndices = np.roll(extremaIndices, 1)
+            extremaIndices[0,0] = 0
+    
+    x = t[extremaIndices].T
+    #Prints Progress
+    print("Error: ", max(abs(diff)))
+    
+
+print("*************************") 
+print("Coefficients: ", p[0:n+1].T)
[…]

### 2016-06-19 prepReport/figures/activFunc.py

@@ -20,7 +20,7 @@ plt.grid(True);
 plt.xlabel(r"Activation $a$");
 plt.ylabel(r"$f(a)$");
 plt.legend(loc=2);
-plt.savefig(filename='activFunc.eps', ransparent=True);
+plt.savefig(filename='activFunc.eps', transparent=True);
 plt.show();
 
 

### 2016-06-19 verilog/sigmoid/plot.py

@@ -1,15 +1,28 @@
 import numpy as np
 import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
 
 f = open("myout.hex", "r")
 
 t = np.arange(-10, 10, 2**-11)
 y = np.zeros(len(t)+100)
 i = 0
+
 for line in f:
     if( "x" not in line):  
         y[i] = eval(line)/(2**11)
         i += 1
+absError = abs(y[1:len(t)+1] - 1/(1+np.exp(-t)))
+print("Max Error: ", max(absError))
 
-plt.plot(y[0:len(t)])
+plt.plot(t, y[0:len(t)], 'r', label="Verilog Model Output")
+plt.plot(t, 1/(1+np.exp(-t)), 'g', label=r"Numpy $\sigma(x)$")
+plt.grid(True);
+plt.xlabel(r"$x$");
+plt.ylabel(r"$\sigma(x)$");
+plt.legend(loc=2);
+plt.savefig(filename='../../finalText/figures/nonlin-out.eps', transparent=True);
 plt.show()

### 2016-06-19 verilog/tanh/plot.py

@@ -1,5 +1,9 @@
 import numpy as np
 import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
 
 f = open("myout.hex", "r")
 
@@ -8,8 +12,19 @@ y = np.zeros(len(t)+100)
 i = 0
 for line in f:
     if( "x" not in line):  
-        y[i] = (eval(line) - (2**18))/(2**11) if  eval(line) >= (2**17) else eval(line)/(2**11)
+        y[i] = (int(line,2) - (2**18))/(2**11) if  int(line,2) >= (2**17) else int(line,2)/(2**11)
         i += 1
 
-plt.plot(t, y[0:len(t)])
+plt.plot(t, y[0:len(t)],'r', label="Verilog Model Output")
+plt.plot(t, np.tanh(t), 'g', label=r"Numpy $\tanh(x)$")
+plt.grid(True);
+plt.xlabel(r"$x$");
+plt.ylabel(r"$\tanh(x)$");
+axes = plt.gca()
+axes.set_xlim([-10,10])
+axes.set_ylim([-1.02,1.02])
+plt.legend(loc=2);
+plt.savefig(filename='../../finalText/figures/nonlin-out-tanh.eps', transparent=True)
 plt.show()
+
+f.close()

### 2016-06-20 finalText/proposedSol.tex
Completed chaper 4

@@
 
-\section{Network Organization}
-
-\section{Forward Propagation Circuit Overview}
+\section{Constituent Modules}
 
@@
 
-\subsection{Non-linearity Calculator}
+\subsection{Non-linearity Calculator}\label{sec:nonlincalc}
 In order to evaluate the non-linear activation functions $\sigma(x)$ and $\tanh(x)$, since there is no algorithm that can directly compute them, I had to find a suitable way to compute them accurately using a finite number of elementary operations, multiplications and additions, that can be performed efficiently by specially tailored blocks within the FPGA (DSP48 slices for multiplication, for instance).
@@
 
-\begin{figure}[H]
+\begin{figure}
     \centering
@@
 
-\subsection{Gate Module}
+\subsection{Gate Module}\label{sec:gatemod} 
+Note that, according to~\cite{Greff15} the usage of peepholes does not compromise the performance of the network~\ref{sec:struct_lstm}, and therefore they have been ommited in order to simplify the Gate Module.
+
+\section{Hardware LSTM Network -- Forward Propagation}\label{sec:hLSTM-fp}
+After detailing the design of each of the constituent modules of the LSTM Network, I will now show how they are interconnected
+to implement a hardware version of Equations~\ref{eq:equationsLSTM}.
+This will be performed in Section~\ref{sec:hLSTM-struct}, where the design decisions and compromises that were taken will be detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, the naïve solution (~\ref{sec:struct-fullpar}) and an optimized version of the former (~\ref{sec:struct-shared}) that exploits some of the hardware redundancies by multiplexing in time the usage of those redundant structures.
+
+\subsection{Network Structure}\label{sec:hLSTM-struct}
+Looking at equations~\ref{eq:equationsLSTM}, we note that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- we see that they can be calculated in parallel, meaning that we need four Gate Modules (see Section~\ref{sec:gatemod} working in parallel, each one with its respective two Block RAMs for $\mb{W}_*$ and $\mb{R}_*$, each followed by the respective activation function calculator (detailed in Section~\ref{sec:nonlincalc}).
+There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
+
+\subsubsection{Fully Parallel}\label{sec:struct-fullpar}
+By implementing directly the ideas outlined previously, we get the network of Figure~\ref{fig:network}.
+The memory element of the network is the array of flip-flops that keep the value of the vector $\mb{c}^{(t)}$, which is activated everytime we have a new incoming signal in order to store the last value, which is now $\mb{c}^{(t)}$
+
+(Talk about RAM usage)
+
+In terms of DSP slice usage, this proposed design comprises 3 elementwise multiplications and 5 activation function Modules.
+Since the number of elementwise multiplications is equal to the network size $N$, we need $3N$ DSP slices for each elementwise block.
+As far as the activation function module is concerned, since we need to apply it to each one of the elements of the layer, $N$, and each module uses one multiplication, the ensemble of all 5 modules needs $5N$ DSP slices.
+Therefore, and noting that each Gate has $\frac{2N}{K_G}$ DSP slices, the total number of DSP slices used is
+
+\begin{equation}\label{eq:numdsp_network}
+    4\frac{2N}{K_G} + 5N + 3N = N \left( \frac{8}{K_G} + 8 \right).
+\end{equation}
+In terms of time performance, we can measure it by estimating the number of clock cycles needed to perform a complete forward propagation.
+Since the gate module outputs a result in $K_G \cdot N$ cycles, the activation function evaluators need always 2, and the elementwise multiplicators need only 1 cycles, and noting that the two elementwise multiplicators that sum to produce $\mb{c}^{(t)}$ can work in parallel, the estimated number of clock cycles needed is
+
+\begin{equation}\label{eq:numcc_network}
+    (N \cdot K_G + 2) + (2 + 1) + (2 + 1) = 8 + N\cdot K_G
+\end{equation}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/network.eps}
+    \caption[Hardware LSTM Network with full level of parallelism]{Hardware LSTM Network with full level of parallelism}
+    \label{fig:network}
+\end{figure}
+
+\subsubsection{Shared Elementwise Multiplication Block}\label{sec:struct-fullpar}
+In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplicator preceeding the flip-flops, all of them follow a similar structure: one of the operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$.
+A clever improvement over the last network architecture is to instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
+Besides, the right elementwise multiplier of Figure~\ref{fig:network} is not used as the same time as any other, so it is a perfect waste of resources.
+The issue about the elementwise multiplier that precedes the flip-flops can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$.
+These ideas resulted in the improved network design of Figure~\ref{fig:network-opt}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
+    \caption[Optimised version of~\ref{fig:network} that shares the elementwise multiplication and the activation function blocks]{Optimised version of~\ref{fig:network} that shares the elementwise multiplication and the activation function blocks}
+    \label{fig:network-opt}
+\end{figure}
+The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise  structure: this is where the time multiplexing of the structure takes place.
+Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplicator (the one that preceeded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, I added another multiplexer before the elementwise multiplication that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
+
+The flip-flops on the righthand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
+The first activated flip-flop is the middle one, which keeps the result from the operation of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the elementwise multiplier, the bottom flip-flop to save the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal.
+Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.
+
+Now, since there is only a single elementwise multiplier and only two activation function calculators, the total requirement for DSP slices is simply
+\begin{equation}\label{eq:numdsp_network-opt}
+    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right).
+\end{equation}
+where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact.
+In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion.
+The estimated number of clock cycles it takes to produce an output is
+
+\begin{equation}\label{eq:numcc_network-opt}
+    (N \cdot K_G + 2) + 3*(2 + 1)  = 11 + N\cdot K_G
+\end{equation}
+which is only 3 clock cycles more than the fully-parallel architecture.
+For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 3 more clock cycles.
 
 \subsection{Timing Diagrams of Operation}
+
+\section{Hardware LSTM Network -- SPSA Training}
+
+\subsection{Timing Diagrams of Operation}

### 2016-06-20 finalText/proposedSol.tex
Added Gate subchapter and figure

@@
 \subsection{Gate Module}\label{sec:gatemod} 
-Note that, according to~\cite{Greff15} the usage of peepholes does not compromise the performance of the network~\ref{sec:struct_lstm}, and therefore they have been ommited in order to simplify the Gate Module.
+The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$.
+If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices.
+This way, a Gate module needs to perform three tasks
+
+\begin{enumerate}
+    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
+    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
+    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
+\end{enumerate}
+Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:dotprod_sec}, the multiplication in task 1 takes $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
+This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/gate.eps}
+    \caption[Diagram of the hardware block that implements the Gate]{Diagram of the hardware block that implements the Gate}
+    \label{fig:gate}
+\end{figure}
 

### 2016-06-21 verilog/dot_prod/runsim.sh
Pipelined the design. Now syntesizes to almost 100 MHz

@@ -1,4 +1,4 @@
 rm *.bin
 python3 genData.py
 vlog *.v
-vsim -c -do "run -all" tb_dot_prod
+vsim -voptargs=+acc -c -do "run -all" tb_dot_prod

### 2016-06-21 verilog/gate/runsim.sh

@@ -1,5 +1,5 @@
 rm *.bin
 python3 genData.py
 vlog *.v
-vsim -c -do "run -all" tb_gate
+vsim -voptargs=+acc -c -do "run -all" tb_gate
 

### 2016-06-21 verilog/network/Synth/Synth.runs/impl_1/rundef.js

@@ -1,0 +1,44 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+// pre-commands:
+ISETouchFile( "init_design", "begin" );
+ISEStep( "vivado",
+         "-log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace" );
+
+
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-06-21 verilog/network/Synth/Synth.runs/impl_1/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-06-21 verilog/network/Synth/Synth.runs/impl_2/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-06-21 verilog/network/Synth/Synth.runs/impl_3/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_3'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-06-21 verilog/network/Synth/Synth.runs/synth_1/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/synth_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl

### 2016-06-21 verilog/network/Synth/Synth.runs/synth_2/rundef.js

@@ -1,0 +1,40 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+ISEStep( "vivado",
+         "-log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl" );
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-06-21 verilog/network/Synth/Synth.runs/synth_2/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/synth_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl

### 2016-06-21 verilog/network/Synth/Synth.runs/synth_5/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/synth_5'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl

### 2016-06-21 verilog/network/readres2.py

@@ -23,7 +23,7 @@ largest_number = pow(2,binary_dim)
 hiddenSz = 8
 QN = 6
 QM = 11
-numTrain = 1000
+numTrain = 10000
 prevX = list()
 outputGolden  = list()
 outputVerilog = list()
@@ -56,7 +56,7 @@ f_in.close()
 
 # Compiles and runs the Verilog simulation
 os.system("vlog *.v")
-os.system("vsim -c -novopt -do \"run -all\" tb_network")
+os.system("vsim -voptargs=+acc -c -do \"run -all\" tb_top_network")
 
 # Loads the pickled layer
 f_pkl = open("layer.pickle", "rb")

### 2016-06-23 verilog/network/Synth/Synth.runs/impl_5/runme.sh
Nonlin Module Pipelined. Bug introduced in Network

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_5'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-06-23 verilog/network/readres2.py

@@ -23,7 +23,7 @@ largest_number = pow(2,binary_dim)
 hiddenSz = 8
 QN = 6
 QM = 11
-numTrain = 10000
+numTrain = 1000
 prevX = list()
 outputGolden  = list()
 outputVerilog = list()
@@ -56,7 +56,7 @@ f_in.close()
 
 # Compiles and runs the Verilog simulation
 os.system("vlog *.v")
-os.system("vsim -voptargs=+acc -c -do \"run -all\" tb_top_network")
+os.system("vsim -voptargs=+acc -c -do \"run -all\" tb_network")
 
 # Loads the pickled layer
 f_pkl = open("layer.pickle", "rb")

### 2016-06-23 verilog/sigmoid/sigm_synth/sigm_synth.runs/impl_1/rundef.js

@@ -1,0 +1,44 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+// pre-commands:
+ISETouchFile( "init_design", "begin" );
+ISEStep( "vivado",
+         "-log sigmoid.vdi -applog -m64 -messageDb vivado.pb -mode batch -source sigmoid.tcl -notrace" );
+
+
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-06-23 verilog/sigmoid/sigm_synth/sigm_synth.runs/impl_1/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/sigmoid/sigm_synth/sigm_synth.runs/impl_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log sigmoid.vdi -applog -m64 -messageDb vivado.pb -mode batch -source sigmoid.tcl -notrace
+
+

### 2016-06-23 verilog/sigmoid/sigm_synth/sigm_synth.runs/synth_1/rundef.js

@@ -1,0 +1,40 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+ISEStep( "vivado",
+         "-log sigmoid.vds -m64 -mode batch -messageDb vivado.pb -notrace -source sigmoid.tcl" );
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-06-23 verilog/sigmoid/sigm_synth/sigm_synth.runs/synth_1/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/sigmoid/sigm_synth/sigm_synth.runs/synth_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log sigmoid.vds -m64 -mode batch -messageDb vivado.pb -notrace -source sigmoid.tcl

### 2016-06-23 verilog/tanh/tanh_synth/tanh_synth.runs/impl_1/runme.sh

@@ -20,7 +20,7 @@ else
 fi
 export LD_LIBRARY_PATH
 
-HD_PWD='/home/jfonseca/thesis/verilog/tanh/tanh_synth/tanh_synth.runs/impl_1'
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/tanh/tanh_synth/tanh_synth.runs/impl_1'
 cd "$HD_PWD"
 
 HD_LOG=runme.log

### 2016-06-23 verilog/tanh/tanh_synth/tanh_synth.runs/synth_1/runme.sh

@@ -20,7 +20,7 @@ else
 fi
 export LD_LIBRARY_PATH
 
-HD_PWD='/home/jfonseca/thesis/verilog/tanh/tanh_synth/tanh_synth.runs/synth_1'
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/tanh/tanh_synth/tanh_synth.runs/synth_1'
 cd "$HD_PWD"
 
 HD_LOG=runme.log

### 2016-07-01 finalText/abstract.tex
Preliminary Thesis Version Completed

\chapter*{Resumo}
\addcontentsline{toc}{chapter}{Resumo}

Esta dissertação tem como objetivo a implementação em hardware, numa plataforma
FPGA, de uma Rede Neuronal \textit{Long Short-Term Memory} (LSTM).

As redes neuronais são uma das técnicas mais usadas na área da Aprendizagem Computacional Profunda para tarefas em
que a programação explícita de \textit{software}, para além de ser impraticavelmente complexa, é necessário que haja uma qualidade
de tomada de decisão semelhante ou superior à inteligência Humana. Esta rede em particular, a LSTM, é uma rede recursiva, na medida em que a saída
da cada neurónio, num instante de tempo, \emph{também} serve de entrada no instante seguinte, e assim
os elementos de memória nela presentes conseguem encontrar padrões também em sequências de dados,
apresentando vantagens claras, neste campo, face a redes neuronais convencionais.

As aplicações desta rede são inúmeras, como se poderá atestar no capítulo
que faz o levantamento do estado-da-arte. Embora a implementação em \textit{software}
seja a solução comum para todas estas arquitecturas, as implementações em \textit{hardware}
são ainda poucas e os benefícios do paralelismo inerente a uma plataforma de \textit{hardware} dedicado (no caso deste trabalho, uma FPGA) não
são aproveitados. É nesse enquadramento que este trabalho se posiciona, apresentando uma implementação inédita de LSTM em FPGA, fazendo apenas uso de 
recursos internos à mesma, que tem uma melhoria de rapidez de processamento de cerca de 195 vezes face 
a uma implementação \textit{software}, assim como a proposta de um circuito que permita fazer o 
treino \textit{on-chip} da rede, usando um método de perturbações estocásticas simultâneas (SPSA).

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

The objective of this Dissertation work is to implement a Long Short-Term Memory (LSTM) Neural
Network in an FPGA platform.

Neural Networks are one of the most commonly used techniques in Deep Learning for tasks
where explicit programming of software, besides impractically complex, it is necessary that
the quality of the decision making is similar, or even better, than that of Human intelligence.
This particular type of network, LSTM, is a recursive network, given that the neuron outputs in a 
given time is \emph{also} fed as the input in the next time instant, and that way their memory elements can
make sense of patterns within data sequences, unlike conventional neural networks.

This type of network, as well as many other kinds of networks, have been profusely implemented
in software, and their practical applications are plentiful, as one can attest
in the State of the Art chapter. However, the benefits of the inherent parallelism offered by a dedicated hardware platform
(in this work, an FPGA) are not availed, and there are relatively few the implementation of Machine Learning algorithms in
these kind of platforms. It is within this grounding that this work positions itself, implementing an LSTM Network in FPGA, using only its internal resources, and achieving an $\times195$ increase in processing time when compared to a software implementation, and also proposes a circuit that will allow on-chip training for this network, using simultaneous stochastic perturbations (SPSA).

### 2016-07-01 finalText/acknows.tex
Preliminary Thesis Version Completed

\chapter*{Agradecimentos}
\addcontentsline{toc}{chapter}{Acknowledgements}

Agradeço a todos os professores que se cruzaram comigo na minha vida e que, através da sua ação
me transmitiram os conhecimentos necessários para me formar como Engenheiro e produzir este trabalho,
que espero que seja de uso para a comunidade científica e para a Humanidade. Uma palavra especial
ao meu orientador deste trabalho, o Prof. João Canas Ferreira, que desde o primeiro momento acolheu
esta proposta de trabalho e soube ver nela o seu potencial devido, e ao co-orientador, o Ivo Timóteo
que mais do que minha referência para a parte teórica deste trabalho, é um bom amigo.

Agradeço todo o esforço da comunidade \textit{open-source} que desenvolveu grande parte das ferramentas
com que trabalhei, desde o \LaTeX~ em que escrevi este documento, ao \href{https://www.python.org/}{Python} e \href{http://www.numpy.org/}{Numpy}
usado na automatização de tarefas e no teste duma primeira versão software da rede. Agradeço também ao autor das \href{https://ece.uwaterloo.ca/~aplevich/Circuit_macros/}{Circuit Macros},
J.D. Aplevich, que muito foram úteis para a elaboração de quase todos os diagramas deste trabalho.

Uma última palavra de apreço para as pessoas especiais da minha vida, os meus pais, que me formaram como Pessoa e me
incutiram força em todos os momentos da vida, do qual esta tese não é excepção, e à minha namorada, a Mafalda, que sempre 
se entusiasmou e interessou por todos os pequenos progressos que fui atingindo ao longo destes meses.

\vspace{10mm}
\flushleft{José Pedro Castro Fonseca}

### 2016-07-01 finalText/intro.tex
Preliminary Thesis Version Completed

--- 2016-02-12 prepReport/intro.tex
@@
 \chapter{Introduction}\label{chap:intro}
-Brief overview of the report:
 
-In Chapter~\ref{chap:intro}, I will present a conceptual background to the problem at hand, exposing the background that surrounds it~\ref{sec:intro_back}, the motivation for the desired solution~\ref{sec:intro_mot}, its objectives~\ref{sec:intro_obj} and, lastly, the people that will help me achieve them~\ref{sec:intro_people}.
-
-In Chapter~\ref{chap:theorBack}, the theoretical foundations will be layed out.
-Section~\ref{sec:tb_basicCon} presents the basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs.
-Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final solution, which will be outlined in Section~\ref{sec:propSol}.
-
-In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications~\ref{sec:sa_apps}, their hardware implementations~\ref{sec:sa_hardware} and the current work regarding the chosen training algorithm~\ref{sec:sa_training} are presented.
-
-In Chapter~\ref{chap:earlyWork} I will present some of the work that I have already concluded.
-
-Finally, in Chapter~\ref{chap:workPlan}, the work plan for the Dissertation work is detailed as \emph{tasks} in Section~\ref{sec:wp_tasks_td}, and their temporal arrangement is pictured in a Gantt chart in Section~\ref{sec:wp_tasks_gantt}.
-Furthermore, the description of the Hardware and Software resources that I will be utilizing during the course of my work is done in Section~\ref{sec:wp_sh}
+In this work, a hardware implementation of an LSTM Neural Network is presented, along with a proposal for a digital circuit that performs on-chip training.
+In the reference example, the proposed hardware network was 195 times
+faster than the reference software implementation. The training convergence, although simulated, is not conclusive.
 
@@
 
-In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the porting process is not entierely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
+In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the porting process is not entirely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
 Therefore, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance.
@@
     \item \textbf{Supervisor} -- The \href{https://sigarra.up.pt/feup/pt/func\_geral.formview?p\_codigo=210963}{Professor João Canas Ferreira}, auxiliary professor at the Faculty of Engineering of the University of Porto.
-    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, MSc, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
+    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, MSc, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Intelligence.
 \end{itemize}
+
+\section{Overview of the Document}\label{sec:intro_overview}
+Chapter~\ref{chap:theorBack}, the theoretical foundations will be layed out.
+Section~\ref{sec:tb_basicCon} presents the basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs.
+Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final solution, which will be outlined in Section~\ref{chap:propSol}.
+
+In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications~\ref{sec:sa_apps}, their hardware implementations~\ref{sec:sa_hardware} and the current work regarding the chosen training algorithm~\ref{sec:sa_training} are presented.
+
+In Chapter~\ref{chap:propSol} the proposed architecture of the hardware LSTM network is explained in detail, as well as details related to each of its constituent blocks.
+
+Finally, in Chapter~\ref{chap:results}, I present the results achieved with the hardware network of Chapter~\ref{chap:propSol}.
+
+\textbf{Disclaimer}: since this is a very innovative work, I will not present here any Verilog source code to protect my Intellectual Property.
+However, you can ask permission to have it by \href{mailto:ee11126@fe.up.pt}{emailing me}.

### 2016-07-01 finalText/proposedSol.tex
Preliminary Thesis Version Completed

@@
 \subsection{Finite-precision representation system}
-Before discussing any details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding to how we numbers envolved in the calculations will be represented.
+Before discussing any other details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding how the numbers involved in the calculations are represented.
 For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point representation systems, as well as the bitwidth and precision chosen for then number representation system of the network, and Section~\ref{sec:convrulesfp} states how the conversion between real and fixed-point numbers is performed.
@@
 Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit.
-This way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$.
+In this way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$.
 In terms of range of representation, the maximum positive number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real number (the decimal point is at the $m$-th bit),
@@
 Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, our fractional part precision should be, at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$.
-On the other hand, given that in the Python experiments I have attested that the weight values generaly do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to not be very small in order to accomodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$.
+On the other hand, given that in the Python experiments I have attested that the weight values generally do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to be large enough to accommodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$.
 Since $18=n+m+1$, then $n = 17 - m = 6$ and the representation system to be used is $Q6.11$.
@@
 Weierstrass's Theorem~\cite{Weierstrass1885}, from 1885, guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$.
-Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynimal approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}.
+Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynomial approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}.
 This leads to a linear system of $n+2$ equations whose $i$-th line is given by
@@
 
-The optimal coefficients of this \textit{minimax} polynomial are found using \textit{Remez Algorithm}, which provides an iterative approach to solve the linear system given by Equation~\ref{eq:remezline} by finding, in each iteration, the $n+2$ set of points $x_i$ of Chebyshev's Theorem that minimize the error function to $\epsilon$.
+The optimal coefficients of this \textit{minimax} polynomial are found using the \textit{Remez Algorithm}, which provides an iterative approach to solve the linear system given by Equation~\ref{eq:remezline} by finding, in each iteration, the $n+2$ set of points $x_i$ of Chebyshev's Theorem that minimize the error function to $\epsilon$.
 The algorithm operations are as follows
@@
 \begin{enumerate}
-    \item Initializing the set $x_i$ of points to $x_i = \frac{a+b}{2} + \frac{(b-a)}{2}\cos\left(\frac{i\pi}{n+1}\right), \, 0 \leq i \leq n+1$ -- 
+    \item Initializing the set $x_i$ of points to $x_i = \frac{a+b}{2} + \frac{(b-a)}{2}\cos\left(\frac{i\pi}{n+1}\right), \, 0 \leq i \leq n+1$ 
     \item Solve the system in~\ref{eq:remezline}
@@
 \end{equation}
-and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the $x_i$s matrix and $b$ is the $f(x_i)$s vector.
-The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy}, as follows
-\lstinputlisting[language=Python, caption=Python script that implements Remez's algorithm, label={lst:remezpy}]{remez.py}
+and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the matrix on the left-hand side of the equation and $b$ is the vector on the right-hand side.
+The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy} of Appendix~\ref{ap1:remez-pyth}.
 
@@
 This proved to yield a lower overall approximation error, as expected (the interval on which to perform the approximation is smaller).
-Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal assymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding assymptote.
+Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal asymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding asymptote.
 The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are
@@
 As for the polynomial calculator is concerned, although we could use a full-pipelined evaluator, that would require two DSP slices -- to perform the two simultaneous multiplications -- but the DSP slices will be heavily used in the matrix-vector calculators, so it is advisable to save them for that purpose.
-A simpler approach is to note that by factoring $x$ we get
+A simpler approach is to note that, according to Horner's Rule, we get
 
@@
 \end{equation}
-where we can note that this operation can be divided in a two-step procedure of a simultaneous mutiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$.
+where we can note that this operation can be divided in a two-step procedure of a simultaneous multiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$.
 The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}.
-Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Numpy as reference) for both activation functions.
+Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Python3's Numpy as reference) for both activation functions.
 
@@
 From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively.
-This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accomodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has lenght $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$.
+This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accommodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has length $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$.
 Plus, if we need a layer with different parameters, either in terms of number of inputs or number of neurons, it advisable to only change the respective parameter at the synthesis stage instead of having to redesign the whole block for that particular size.
@@
 $N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions.
-The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: before, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e.
+The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e.
 the number of columns); now, if we define a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
-Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
+Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; in this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
 
-In Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} are depicted a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before.
+Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before.
 Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
@@
 After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.
+Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency, we should add a 3 clock cycle overhead to the previous estimate.
 
@@
 \end{figure}
-In summary, for an $N \times M$ weight matrix, this module outputs the result vector in $K_G \cdot M$ clock cycles.
 
-\subsection{Dual-port Block RAM}
-Although the bias weights, being simply a vector, I have chosen to keep them in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different.
-Since the
-network size can be big, it is good to make sure that they are placed in the Block RAMs available in the FPGA.
+\subsection{Dual-port weight RAM}
+Even though I have chosen to keep the bias weight vectors in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different.
+Since the network size can be large, it is good to make sure that they are placed in the RAMs available in the FPGA.
 In Section~\ref{sec:dotprod_sec}, we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column by column access to the weight matrix, and therefore this block outputs a given column in the \verb+rowOut+ output port terminal, at the \emph{negative edge} of the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock.
@@
 
+After performing synthesis, I have noticed that using block RAM led to a slightly poorer performance, so I have settled with the use of \emph{Distributed RAM}, storing the weight matrices in LUTRAM.
+
 \subsection{Gate Module}\label{sec:gatemod} 
@@
 \end{enumerate}
-Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:dotprod_sec}, the multiplication in task 1 takes $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
+Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:dotprod_sec}, the multiplication in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
 This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.
+The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result.
+After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this module takes to output a result is $6+K_G\cdot N$.
 
@@
 to implement a hardware version of Equations~\ref{eq:equationsLSTM}.
-This will be performed in Section~\ref{sec:hLSTM-struct}, where the design decisions and compromises that were taken will be detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, the naïve solution (~\ref{sec:struct-fullpar}) and an optimized version of the former (~\ref{sec:struct-shared}) that exploits some of the hardware redundancies by multiplexing in time the usage of those redundant structures.
+This is performed in Section~\ref{sec:hLSTM-struct}, where the design decisions and compromises that were taken are detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, the naïve solution (~\ref{sec:struct-fullpar}) and an optimized version of the former (~\ref{sec:struct-shared}) that exploits some of the hardware redundancies by multiplexing in time the usage of those redundant structures.
 
@@
 In terms of time performance, we can measure it by estimating the number of clock cycles needed to perform a complete forward propagation.
-Since the gate module outputs a result in $K_G \cdot N$ cycles, the activation function evaluators need always 2, and the elementwise multiplicators need only 1 cycles, and noting that the two elementwise multiplicators that sum to produce $\mb{c}^{(t)}$ can work in parallel, the estimated number of clock cycles needed is
+Since the gate module outputs a result in $K_G \cdot N$ cycles, the activation function evaluators need always 5, and the elementwise multiplicators need only 2 cycles (one cycle for the pipeline, and another for the actual calculation), and noting that the two elementwise multiplicators that sum to produce $\mb{c}^{(t)}$ can work in parallel, the estimated number of clock cycles needed would be proportional to
 
 \begin{equation}\label{eq:numcc_network}
-    (N \cdot K_G + 2) + (2 + 1) + (2 + 1) = 8 + N\cdot K_G
+    (N \cdot K_G + 6) + (5 + 2) + (5 + 2) = 20 + N\cdot K_G
 \end{equation}
@@
 
-\subsubsection{Shared Elementwise Multiplication Block}\label{sec:struct-fullpar}
-In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplicator preceeding the flip-flops, all of them follow a similar structure: one of the operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$.
-A clever improvement over the last network architecture is to instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
+\subsubsection{Shared Elementwise Multiplication Block}\label{sec:struct-shared}
+In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplier preceeding the flip-flops, all of them follow a similar structure: one of the operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$.
+An improvement over the last network architecture is to instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
 Besides, the right elementwise multiplier of Figure~\ref{fig:network} is not used as the same time as any other, so it is a perfect waste of resources.
@@
 The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise  structure: this is where the time multiplexing of the structure takes place.
-Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplicator (the one that preceeded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, I added another multiplexer before the elementwise multiplication that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
+Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplicator (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, I added another multiplexer before the elementwise multiplication that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
 
-The flip-flops on the righthand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
+The flip-flops on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
 The first activated flip-flop is the middle one, which keeps the result from the operation of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the elementwise multiplier, the bottom flip-flop to save the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal.
@@
 In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion.
-The estimated number of clock cycles it takes to produce an output is
+After adjustments to the state machine, and accounting for pipelining and synchronization within the datapath, the clock cycles needed after the gate calculations are 27, so the total clock cycles needed are
 
 \begin{equation}\label{eq:numcc_network-opt}
-    (N \cdot K_G + 2) + 3*(2 + 1)  = 11 + N\cdot K_G
+    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
 \end{equation}
-which is only 3 clock cycles more than the fully-parallel architecture.
-For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 3 more clock cycles.
-
-\subsection{Timing Diagrams of Operation}
+which is only 13 clock cycles more than the fully-parallel architecture.
+For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
 
 \section{Hardware LSTM Network -- SPSA Training}
+In this section, I propose a digital circuit that aims to train the Network whose details were explained in the previous Section by implementing the training algorithm of Section~\ref{sec:training_lstm}.
+The training consists in changing the weight values in the RAMs accordingly with the error of the current prediction against the correct outcome.
+This envolves \textbf{two} forward propagations: one, where we run the network with the weights unchanged, and another where we \emph{perturbate} the weights by a parameter $\pm \beta$ (the actual sign for each weight is generated at the beginning of the first forward propagation).
+After these two forward propagations, there is a small time frame where the weights are updated, row by row, before the next incoming sample.
+This is explained in Section~\ref{sec:train-circ}.
+The way how the numbers are randomly generated in hardware is presented in Section~\ref{sec:randnum-circ}.
 
-\subsection{Timing Diagrams of Operation}
+\subsection{Pseudo-random Number Generator}\label{sec:randnum-circ}
+While a True Random Number Generator (TRNG) generates data from a physical entropy source, a Pseudo-random Number Generator (PRNG) generates a sequence of numbers by applying a given
+function to an initial seed. This means that, if one knows that seed and the function that generates the sequence, one can predict the previous and future numbers given the current state. While this
+is not acceptable for high security cryptographic systems, it serves the purpose of providing a sequence of random sign vectors (from whom both the signal of the perturbation and the update will be choosen from).
+The remaining issue is to find a PRNG with good statistical qualities, so instead of implementing a simple linear-feedback shift register (LFSR), I did some research on the topic, and eventually chose the solution
+of~\cite{Tkacik2003}, based on a 43-bit LFSR and a 37-bit Cellular Automata, by selecting 32 bits from each, permuting the bits and performing a bitwise XOR -- the output has, also, 32 bits.
+The quality of the
+statistical properties are guaranteed, since it passed the DIEHARD test battery, as well as others, according to~\cite{Tkacik2003}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/prng.eps}
+    \caption[The 32 bit Pseudo-random Number Generator. Source~\cite{Tkacik2003}]{The 32 bit Pseudo-random Number Generator. Source~\cite{Tkacik2003}}
+    \label{fig:prng}
+\end{figure}
+The generator polynomial of the LFSR is a maximal length polynomial, $x^{43}+x^{41}+x^{20}+x+1$, and thus has a period equal to $2^{43}-1$. The CA is also a register, but each bit is updated
+according to a given \textbf{rule} based on the values of the neighboring bits.
+This particular CA follows Rule 150 for the bit in position $i=27$, given by $c_{t+1}(i) = c_{t}(i-1) + c_{t}(i) + c_{t}(i+1)$, and
+the remaining bits are updated by Rule 90, given by $c_{t+1}(i) = c_{t}(i-1) + c_{t}(i+1)$. The bits positioned at the borders also follow rule 90, but treat as zeros the non-existing $i$ positions.
+
+The output of my hardware implementation of this PRNG is presented in Figure~\ref{fig:prng_out}.
+It is a state-time diagram, where the contents of the registers at each time are represented by black (ones) and white (zeros), and the left column represents the output of the LFSR, the center column is the output of the Cellular Automata and the right column is their combined output.
+As one can see, the time dependence of the LFSR is very high, which is not desirable.
+The Cellular Automata, although uncorrelated, still retains some of the patterns that emerge from CAs.
+On the other hand, the combined output is the one that most resembles white noise, which is a desirable characteristic.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/prng_out.eps}
+    \caption[The output of the implementation of the presented PRNG (right).
+The LFSR is the left column, and the CA is at the center]{The output of the implementation of the presented PRNG (right).
+The LFSR is the left column, and the CA is at the center.
+The zeros are represented by white, and the ones by black}
+    \label{fig:prng_out}
+\end{figure}
+
+\subsection{Training Algorithm Circuit}\label{sec:train-circ}
+Although the convergence results are not conclusive, here is presented the proposed digital circuit of the SPSA training for the network.
+The circuit of Figure~\ref{fig:train} is replicated for each of the eight weight RAMs, and the Sign RAM holds a single bit for each of the weight positions of the main weight RAM.
+This bit keeps the sign of the perturbation to be applied to that particular weight, both on the second forward propagation and the training step.
+During the first forward propagation, and since the weights are not perturbed in this run, we can use this time to propagate the sign RAM with new values for the next training stage.
+For this purpose, the PRNG is activated by the signal \verb+genRandNum+, generated by the state machine, and we write columns of bits to the sign RAM, re-using the addressing signal produced by the Gate, namely \verb+colAddressRead+ (\verb+weightUpdate+ is set to zero at this stage).
+At the second forward propagation, \verb+pertWeights+ is set to one, and the Perturbation Mux selects the input where the weights are summed with the perturbation $\beta$, whose sign is selected by the sign RAM.
+
+After this stage is completed, and before we start this new training cycle for the new incoming sample, we perform the actual training.
+For that, \verb+weightUpdate+ is set to one, and we enable writing to the Weight RAM.
+Also, the machine state is now responsible for addressing both RAMs, and we begin a simultaneous \emph{pipelined} write to the Weight RAM: this is done by delaying the the Write address in respect to the Read address.
+This way, at the first clock cyle while we read Column 0, in the next clock cycle we will read Column 1 \emph{and} write the sign-changed column 0 plus the update coming from the difference between the outputs in the two forward propagations.
+In order to simplify the update rule of Equation~\ref{eq:spsa_updateRule}, I have fixed $\alpha$ and $\beta$ as negative powers of two, and this replaces the multiplication and division by a simple arithmetic right shift -- this idea is also proposed by~\cite{Maeda05}.
+
+The learning, although working in the Python model developed, is still not converging in the simulations of the Verilog code for this circuit.
+This might have to deal with an overflow problem at the weight update.
+This is a good question to be addressed in future work.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/train.eps}
+    \caption[Proposed Hardware Implementation of SPSA Training]{Proposed Hardware Implementation of SPSA Training}
+    \label{fig:train}
+\end{figure}

### 2016-07-01 finalText/results.tex
Preliminary Thesis Version Completed

\chapter{Results}\label{chap:results}
The optimized design of Section~\ref{sec:struct-shared} was synthesized for different network parameters, using Xilinx's Vivado 2015.4 Design Suite, and the synthesis results are reported in Section~\ref{sec:res-synth}, along with their discussion. The performance of the synthesized networks is very promising, showing an approximate $195\times$ speed-up over the LSTM Python model, which was running on a high-end desktop computer with an over clocked CPU. Although the FPGA aims for an Embedded System environment usage, the fairness of this comparison still holds, since a Python/Numpy implementation is slower than a lower-level language implementation (such as in C) and this level of performance would be similar to a C implementation in a low-power, embedded CPU platform. These results are reported in Section~\ref{sec:res-valid}.

\section{Synthesis}\label{sec:res-synth}
In order to assess the impact of the network size on the performance and resource usage of the proposed design, I have synthesized the Verilog description of the proposed network for sizes $N$ of 4, 8, 16 and 32, and for each of these the resource sharing parameter $K_G$ was also varied (as will be explained further ahead, the number of inputs is set to $M=2$). Since performance is the key factor for using an FPGA implementation, the synthesis strategy used was the \verb+Flow_PerfOptimized_high+, and the implementation strategy was \verb+Performance_ExplorePostRoutePhysOpt+ to perform a post-route optimization stage that proved useful in many cases to meet the clock specifications when the first Place\&Route pass failed to meet them. For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were successfully synthesized.

\subsection{Maximum Frequency}\label{sec:res-synth-maxfreq}
Using the synthesis strategy outlined in the previous paragraph, the maximum clock frequency that does not cause timing violations of any sort is plotted in Figure~\ref{fig:maxfreq}. For $N=4$, since there are only 4 rows to be multiplied, the maximum value of $K_G$ is 4, and hence no synthesis was performed for $K_G = 8$; also, since $K_G = 2$ and $N=32$ would use $32(8/2+3) = 224$ DSP slices, that exceeds the 220 slices available in the Zynq 7020, so there is no synthesis data for that value.

First, for all network sizes, it is clear that with increasing $K_G$, the maximum clock speed decreases. This means that there is a critical path in the Matrix-Vector multiplication unit, whose multiplexer becomes increasingly complex for higher values of $K_G$. On the other hand, when $K_G$ is the same, smaller networks are faster than larger networks. The fastest design is an $N=4$ and $K_G = 2$ network, with a clock frequency of $140.154$ MHz, and the slowest one is an $N=32$ and $K_G=4$ network, clocked at $90.498$ MHz. The reference design used for validation in Section~\ref{sec:res-valid} is an $N=8$ and $K_G=2$ network is clocked at $132.89$ MHz, which yields a clock period of $7.525$ ns.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/maxfreq.eps}
    \caption[The maximum achievable clock frequency for several Network Sizes $N$ and resource sharing level $K_G$]{The maximum achievable clock frequency for several Network Sizes $N$ and resource sharing level $K_G$}
    \label{fig:maxfreq}
\end{figure}

\subsection{DSP Slices Usage}\label{sec:res-synth-dsp}
The estimates made in Equation~\ref{eq:numdsp_network-opt} were proven to be accurate, as Figure~\ref{fig:dspused} confirms. The reference network design uses 56 DSP slices, which corresponds to $25.45\%$ of the total number of DSP slices available.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/dspuse.eps}
    \caption[The number of DSP slices used for several Network Sizes $N$ and resource sharing level $K_G$]{The number of DSP slices used for several Network Sizes $N$ and resource sharing level $K_G$}
    \label{fig:dspused}
\end{figure}

\subsection{Power Usage}\label{sec:res-synth-power}
Another important metric of the performance of a design is its \textbf{power consumption}. All designs yielded a constant baseline value for the \emph{static} power consumption of around 120 mW, and the power consumption reported in Figure~\ref{fig:power} refers to the \emph{total} power consumptions, i.e. both the baseline static power and the dynamic power consumption. It is clear that the smaller the network is, the less power is consumed, as one would expect. Furthermore, an increasing level of resource sharing yields a substantially lower power consumption figure: this is because we use less DSP slices when we increase $K_G$. Of course, even though the power consumption is lower in that case, that comes at the expense of a lower clock frequency and more clock cycles elapsed per forward propagation.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/power.eps}
    \caption[Power consumption estimates for several Network Sizes $N$ and resource sharing level $K_G$]{Power Consumption estimates for several Network Sizes $N$ and resource sharing level $K_G$}
    \label{fig:power}
\end{figure}

\subsection{Other Resources Usage}\label{sec:res-synth-otheres}
Besides DSP slices, it is also worth mentioning the usage of LUT, LUTRAMs and Flip-Flops. In Table~\ref{tab:lut}, the usage of LUTs is reported, where we can see that although there is not a clear trend on how the LUT usage varies with increasing $K_G$, it is clear, and expectable, that the LUT usage increases with the size of the network by an approximate factor of 2, from $N=2$ to $N=16$. As for $N=32$, the usage does not follow this apparent trend, and rises sharply to 91\%. For $K_G = 8$ the usage even surpasses the maximum amount of LUTs available.
\begin{table}
    \centering
  \begin{tabular}{ | l | c | c | c |}
    \hline
    & $K_G=2$  & $K_G=4$ & $K_G=8$ \\ 
    \hline
    $N=4$ & 6.87\% & 6.04\% & N.A. \\    
    \hline
    $N=8$ & 14.64\% & 13.03\% & 14.11\% \\    
    \hline
    $N=16$ & 28.97\% & 27.72\% & 29.85\% \\    
    \hline
    $N=32$ & N.A. & 91.09\% & N.D. \\    
\hline
  \end{tabular}
    \caption{LUT usage for different $N$ and $K_G$}
    \label{tab:lut}
\end{table}

As for LUT, the FF usage also scales accordingly to a $2\times$ factor. Furthermore the LUTRAM usage scales well and does not pose a limitation on the network size. The usage results are reported in Table~\ref{tab:ff}.

\begin{table}
    \centering
  \begin{tabular}{ | l | c | c |}
    \hline
         & LUTRAM  & FF  \\ 
    \hline
    $N=4$ & 0.18\%  & 3.39\% \\    
    \hline
    $N=8$ & 4.41\%  & 6.7\% \\    
    \hline
    $N=16$ & 8.83\%  & 13.36\% \\    
    \hline
    $N=32$ & 17.66\% &  26.45\% \\    
\hline
  \end{tabular}
    \caption{Flip-flop usage for different $N$ and $K_G$}
    \label{tab:ff}
\end{table}

\section{Validation and Comparison}\label{sec:res-valid}
Over the course of this Section, the functionality of the network is verified against the Python reference module of~\ref{sec:res-pyth} that I have developed as a reference, both for the forward propagation of the network, as well as for the training algorithm. The methodology of this procedure is outlined in Section~\ref{sec:res-meth}. On Section~\ref{sec:res-valid-fpps}, I will assess how the performance results of Section~\ref{sec:res-synth-maxfreq} translate into how many classifications per second this design achieves, and how that figure compares with the capabilities of the Python module.

\subsection{Reference Python Module}\label{sec:res-pyth}
Before performing the Verilog description of the network, I have tested the ideas outlined in Section~\ref{chap:theorBack} first by building a software version of an LSTM Network. Since I wanted to quickly test the ideas without much effort, I decided to use Python and Numpy rather than MATLAB, since the former have higher performance at the same level of code complexity. The code for this Python class is listed in Appendix~\ref{ap1:lstm-pyth}. It supports both the forward propagation of an already trained network, but also provides functions to train a model using the SPSA method outlined in Section~\ref{sec:training_lstm} and also Backpropagation Through-Time as presented in~\cite{Greff15}. As the reader can attest by running the code himself, the SPSA proves to be a valid approach for training these networks, achieving convergence for the Machine Learning task that will be presented in the next paragraph, although it converges slower than Backpropagation Through Time.

The learning problem presented to both the software and hardware network is the sum of two binary numbers of 8 bits. The bits of each number at position $i$ are input to the network as a vector, and the network outputs its prediction of the correct value of the $i$-th bit of the \emph{result} number. After the whole number number is processed, the memory cells of the LSTM network are reset and a new addition task can be presented to the network. Even though this seems a rather simple problem, it accounts for all the essential issues at which this network excels: first, this is a \emph{classification} problem in which the Machine Learning algorithm needs to output a prediction based on the input feature vector, and that prediction has to take into account the \emph{history} of predictions so far, because the current bit is the sum of not only the bits of the two operands, but also the \emph{carry} generated at the last few positions -- this is where the memory cells of this special Neural Network come into effect.

\subsection{Methodology}\label{sec:res-meth}
The Python testbench that both trains and verifies the predictions of the software network of Appendix~\ref{ap1:lstm-pyth} is listed in Appendix~\ref{ap1:testbench-pyth}. This script, after achieving convergence, stores the trained network weights in \href{https://docs.python.org/3/library/pickle.html}{Pickle} format. These weights are, then, converted to $Q6.11$ fixed-point format and directly loaded into the Verilog model by the Verilog testbench. The Verilog testbench then applies in order the bits of randomly generated numbers, captures the output bit and checks whether it is correct or not, and counts the number of incorrect bits.

This network has, then, two inputs ($M=2$) and eight neurons ($N=8$). Furthermore, the output of the eight LSTM neurons is reduced to a single output by means of a simple neuron, as the one described in Figure~\ref{fig:modelNeuron_a}. This introduces a further 5 clock cycle overhead that is not accounted in Section~\ref{sec:res-valid-fpps}, since it is not actually part of the LSTM Network and, although it was placed in series in the datapath, after the LSTM network, it can run in parallel with the LSTM network, introducing no overhead whatsoever.

A Python script was written (see Appendix~\ref{ap1:runtb-lstm}) to automatically compile the Verilog code, generate the input and output data, and call Questasim to run the simulation of the HDL code. Figure~\ref{fig:res-meth-wave} shows the waveform viewer window of a run of the Verilog testbench, and in Appendix~\ref{ap1:transcript-cmd} shows the transcript of a successful run of this testbench, proving that the learning ability of the Hardware network still retains the predicting capability of the software network.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/wave.png}
    \caption[Screenshot of the simulation run of a complete forward propagation for $N=8$ and $K_G=2$]{Screenshot of the simulation run of a complete forward propagation for $N=8$ and $K_G=2$. The period between the two time cursors is the same as the one reported in Table~\ref{tab:process-time}}
    \label{fig:res-meth-wave}
\end{figure}
\subsection{Performance}\label{sec:res-valid-fpps}
To evaluate the performance of the network in terms of how fast it can produce data, I have defined a metric based on how many predictions it can produce per second (i.e. produce a new result bit in the output sequence), in \textbf{millions}. The prediction time is the time elapsed since a new input vector is applied \emph{to} the moment the LSTM Network produces an output vector (as discussed previously, the five clock cycles of the simple neuron at the output are \textbf{not} considered). To do this, we multiply the number of clock cycles yielded by Equation~\ref{eq:numcc_network-opt} by the equivalent clock period from the synthesis clock report of Figure~\ref{fig:maxfreq}. This result is epitomized in Table~\ref{tab:process-time}, where the calculation time of the Python module forward propagation function is also included. This time was measured using the \href{https://docs.python.org/3.5/library/timeit.html}{\textbf{timeit}} module, that allows the evaluation of the execution time of small pieces of code as well as complete functions with arguments. The Python code was run on a Linux System, powered by an i7-3770k Intel Processor, running at 4.2GHz.

\begin{table}
    \centering
    \begin{tabular}{ | l | c | c | c | c | c | }
    \hline
    & $K_G=8$  & $K_G=4$ & $K_G=2$ & Python & Speed-up \\ 
    \hline
    $N=4$ & N.A.  & 360.15 ns  & 292.535 ns & 65 $\mu$s & $\times180$ \\   
    \hline 
    $N=8$ & 960.3 ns  & 494 ns  &  368.725 ns & 72 $\mu$s & $\times194$ \\    
    \hline
    $N=16$ & 1.715 $\mu$s  & 921.5 ns  & 518.05 ns & 96 $\mu$s & $\times185$ \\
    \hline
    $N=32$ & N.D. & 1.78 $\mu$s  & N.A. & 185 $\mu$s  &  $\times104$ \\    
\hline
  \end{tabular}
    \caption{Total processing time for a single forward propagation, for hardware networks of various $K_G$ and the Python software network of equivalent size. The speed-up figure is calculated between the Python processing time versus the best hardware network time of the same size}
    \label{tab:process-time}
\end{table}
The performance increase is impressive, even for the slowest of the designs (the $N=32$ network). The hardware network is, at best, $\times200$ faster than the software counterpart, and at worst $\times100$ faster. Also, it is noticeable that increasing the level of resource sharing increases the computation time, since the level of parallelism is lower.

 Since the previous values are the time needed for a \emph{single} forward propagation, to know how many forward propagations we can perform \emph{per second}, we only need to invert the previous values. These values are presented in Figure~\ref{fig:Mclass-psec}. While the $N=8$ and $K_G=2$ network is able to perform around 3.2 million predictions per second, the Python model can only output around 14 thousand predictions, which is a very significant result that proves the relevance of this implementation.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/Mclass-psec.eps}
    \caption[Millions of classifications per second of each design according to the network size $N$]{Millions of classifications per second of each design according to the network size $N$. The comparison is between the software Python model and 3 networks of different levels of resource sharing $K_G$.}
    \label{fig:Mclass-psec}
\end{figure}

### 2016-07-01 finalText/theorBack.tex
Proposed Solution chapter almost done
Preliminary Thesis Version Completed

--- 2016-02-12 prepReport/theorBack.tex
@@
     \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \notag\\
+    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \notag\\
     \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \notag \\
-    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \notag\\
-    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
+    \mb{y}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
 \end{align}
@@
 One last comment to be made concerning the update rule, is the hypothetical need for \textbf{limits} to the weight values, when the update rule exceeds $|w_{\text{max}}|$ (in that case, we set $w_i^{(t+1)} = \pm w_{\text{max}}$), which sometimes might be needed if the behaviour of the weight update is not appropriate~\cite{Maeda05}.
-
-\section{Proposed Solution}\label{sec:propSol}
-Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\ref{sec:wp_sh}).
-
-As discussed in Chapter~\ref{chap:stateArt}, up to now there is only one FPGA implementation of LSTM cells~\cite{Chang15}, but there is no on-chip learning on the architecture (the weights are learned offline, and loaded in the FPGA before operation, and they are not changed while the FPGA system is operating), and my proposed work will implement an efficient LSTM implementation \emph{with} the ability to learn on-chip, which is an innovative and useful feature for the LSTM architecture on a dedicated hardware implementation.
-I will achieve that using the SPSA method described in~\ref{sec:training_lstm}, and inspired in the previous work of~\cite{Maeda05}.
-
-This solution will combine the benefits of a dedicated hardware implementation, where the inherent parallelism so obviously visible in the network architecture can be fully exploited something that, predominantly, yields a much higher performance in terms of computation time than a naïve CPU implementation (as in~\cite{Maeda05} where a $21\times$ speed-up against an ARM CPU was acheived) with the flexibility that a reconfigurable platform such as an FPGA provides.

### 2016-07-01 finalText/figures/powerplot.py

@@ -1,0 +1,30 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
+
+n4 = np.array([176, 149])
+x_n4 = np.array([1,2])
+n8 = np.array([189, 174, 161])
+x_n8 = np.array([1,2,3])
+n16 = np.array([386, 296, 200])
+x_n16 = np.array([1,2,3])
+n32 = np.array([482])
+x_n32 = np.array([2])
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+plt.plot(x_n4, n4, 'yo-', label="$N=4$")
+plt.plot(x_n8, n8, 'ro-', label="$N=8$")
+plt.plot(x_n16, n16, 'bo-', label="$N=16$")
+plt.plot(x_n32, n32, 'go-', label="$N=32$")
+ax.set_xticks([0,1,2,3,4])
+plt.axis([0.8,3.2,140,500])
+plt.grid(True);
+plt.xlabel(r"$\log_2(K_G$)");
+plt.ylabel(r"Power in mW");
+plt.legend(loc=1);
+plt.savefig(filename='power.eps', transparent=True);
+plt.show()

### 2016-07-01 finalText/readres2.py

@@ -1,0 +1,99 @@
+import pickle
+import numpy as np
+import os
+
+def real_to_Qnm(real, n, m):
+    if(real >= 0):
+        return int(np.round(real*(2**m)).astype(int)) & int(2**(n+m+1)-1)
+    else:
+        return int(2**(n+m+1) + np.round(real*(2**m)).astype(int)) & int(2**(n+m+1)-1)
+def Qnm_to_real(real, n, m):
+    real = int(real) & int(2**(n+m+1)-1)
+    if(real >= 2**(n+m)):
+        return (real-2**(n+m+1))/(2**m)
+    else:
+        return real/(2**m)
+def sigmoid(x):
+	output = 1/(1+np.exp(-x))
+	return output
+
+numBits  = 8
+binary_dim = numBits
+largest_number = pow(2,binary_dim)
+hiddenSz = 8
+QN = 6
+QM = 11
+numTrain = 1000
+prevX = list()
+outputGolden  = list()
+outputVerilog = list()
+outputPython  = list()
+wrongBits = 0
+f_in  = open("goldenIn_x.bin", "w")
+
+
+ 
+
+for i in range(numTrain):
+	# Generates the golden input and output
+	a_int = np.random.randint(largest_number/2) # int version
+	a = np.binary_repr(a_int, width=binary_dim) # binary encoding
+
+	b_int = np.random.randint(largest_number/2) # int version
+	b = np.binary_repr(b_int, width=binary_dim) # binary encoding
+
+	c_int = a_int + b_int
+	c = np.binary_repr(c_int, width=binary_dim)
+	
+	for position in range(binary_dim):
+		X  = np.array([[int(a[binary_dim - position - 1]), int(b[binary_dim - position - 1])]]).T
+		y  = np.array([int(c[binary_dim - position - 1])]).T
+		prevX.append(X)
+		f_in.write("{0:018b}\n".format(real_to_Qnm(X[0,0], QN, QM))) 
+		f_in.write("{0:018b}\n".format(real_to_Qnm(X[1,0], QN, QM))) 
+
+f_in.close()
+
+# Compiles and runs the Verilog simulation
+os.system("vlog *.v")
+os.system("vsim -voptargs=+acc -c -do \"run -all\" tb_network")
+
+# Loads the pickled layer
+f_pkl = open("layer.pickle", "rb")
+layer = pickle.load(f_pkl);
+
+# Loads the output values
+layerOut = np.zeros((8,1));
[…]

### 2016-07-01 finalText/testLSTM.py

@@ -1,0 +1,129 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import time
+import sys
+import pickle
+import LSTMlayer
+
+np.random.seed(round(time.time()))
+
+# compute sigmoid nonlinearity
+def sigmoid(x):
+    output = 1/(1+np.exp(-x))
+    return output
+
+# convert output of sigmoid function to its derivative
+def sigmoid_output_to_derivative(output):
+    return output*(1-output)
+
+def sigmoidPrime(output):
+    return output*(1-output)
+
+def tanhPrime(output):
+    return (1-output**2)
+
+# training dataset generation
+binary_dim = 8 
+largest_number = pow(2,binary_dim)
+
+# Simulation Parameters
+pert  = float(sys.argv[1])
+alpha = float(sys.argv[2])
+wmax  = float(sys.argv[3])
+samplesPerEpoch = 500
+input_dim  = 2
+hidden_dim = int(sys.argv[4])
+output_dim = 1
+maxEpoch   = 100 
+trainSamp  = 5000
+testSamp   = 100
+
+# initialize neural network weights
+lstmLayer1 = LSTMlayer.LSTMlayer(input_dim, hidden_dim, output_dim, alpha, 'SPSA', pert, wmax, binary_dim)
+plt.axis([0, maxEpoch, 0, binary_dim])
+plt.title("Alpha={0} -- Perturbation={1} -- Wmax={2}".format(alpha, pert, wmax))
+plt.ion()
+plt.show()
+
+epochError = 0
+correct    = 0
+# training logic
+for i in range(maxEpoch):
+   
+    print("Training Epoch: ", i)
+
+    for j in range(trainSamp):
+        # generate a simple addition problem (a + b = c)
+        a_int = np.random.randint(largest_number/2) # int version
+        a = np.binary_repr(a_int, width=binary_dim) # binary encoding
+
+        b_int = np.random.randint(largest_number/2) # int version
+        b = np.binary_repr(b_int, width=binary_dim) # binary encoding
+
+        # true answer
+        c_int = a_int + b_int
+        c = np.binary_repr(c_int, width=binary_dim) 
+
+        # -------------- THE FORWARD PROPAGATION STEP -------------- #
+        for position in range(binary_dim):
+            
+            # generate input and output
+            X = np.array([[int(a[binary_dim - position - 1]), int(b[binary_dim - position - 1])]]).T
+            y = np.array([int(c[binary_dim - position - 1])]).T
+
[…]

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_1/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log dot_prod.vdi -applog -m64 -messageDb vivado.pb -mode batch -source dot_prod.tcl -notrace
+
+

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_2/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log dot_prod.vdi -applog -m64 -messageDb vivado.pb -mode batch -source dot_prod.tcl -notrace
+
+

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_3/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_3'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log dot_prod.vdi -applog -m64 -messageDb vivado.pb -mode batch -source dot_prod.tcl -notrace
+
+

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_4/rundef.js

@@ -1,0 +1,44 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+// pre-commands:
+ISETouchFile( "init_design", "begin" );
+ISEStep( "vivado",
+         "-log dot_prod.vdi -applog -m64 -messageDb vivado.pb -mode batch -source dot_prod.tcl -notrace" );
+
+
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_4/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_4'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log dot_prod.vdi -applog -m64 -messageDb vivado.pb -mode batch -source dot_prod.tcl -notrace
+
+

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_6/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/impl_6'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log dot_prod.vdi -applog -m64 -messageDb vivado.pb -mode batch -source dot_prod.tcl -notrace
+
+

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_1/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log dot_prod.vds -m64 -mode batch -messageDb vivado.pb -notrace -source dot_prod.tcl

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_2/rundef.js

@@ -1,0 +1,40 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+ISEStep( "vivado",
+         "-log dot_prod.vds -m64 -mode batch -messageDb vivado.pb -notrace -source dot_prod.tcl" );
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_2/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log dot_prod.vds -m64 -mode batch -messageDb vivado.pb -notrace -source dot_prod.tcl

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_3/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_3'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log dot_prod.vds -m64 -mode batch -messageDb vivado.pb -notrace -source dot_prod.tcl

### 2016-07-01 verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_4/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/dot_prod/Synth_dot_prod/Synth_dot_prod.runs/synth_4'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log dot_prod.vds -m64 -mode batch -messageDb vivado.pb -notrace -source dot_prod.tcl

### 2016-07-01 verilog/gate/Synth_gate/Synth_gate.runs/impl_1/rundef.js

@@ -1,0 +1,44 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+// pre-commands:
+ISETouchFile( "init_design", "begin" );
+ISEStep( "vivado",
+         "-log gate.vdi -applog -m64 -messageDb vivado.pb -mode batch -source gate.tcl -notrace" );
+
+
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-07-01 verilog/gate/Synth_gate/Synth_gate.runs/impl_1/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/gate/Synth_gate/Synth_gate.runs/impl_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log gate.vdi -applog -m64 -messageDb vivado.pb -mode batch -source gate.tcl -notrace
+
+

### 2016-07-01 verilog/gate/Synth_gate/Synth_gate.runs/synth_1/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/gate/Synth_gate/Synth_gate.runs/synth_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log gate.vds -m64 -mode batch -messageDb vivado.pb -notrace -source gate.tcl

### 2016-07-01 verilog/network/Synth/Synth.runs/impl_7/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_7'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-07-01 verilog/network/Synth/Synth.runs/impl_8/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_8'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-07-01 verilog/network/Synth/Synth.runs/synth_6/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/synth_6'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl

### 2016-07-01 verilog/network/Synth/Synth.runs/synth_7/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/synth_7'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl

### 2016-07-01 verilog/rng/plotOut.py

@@ -41,11 +41,14 @@ print("Combined Output   : ", SR_ones/(NUM_ITER*43))
 print("Combined Output   : ", CA_ones/(NUM_ITER*37))
 
 plt.figure(1)	
-plt.subplot(1,3,1)	
-plt.imshow(prevSRout,cmap=plt.cm.gray, interpolation='none', aspect=0.2)		
-plt.subplot(1,3,2)
-plt.imshow(prevCAout,cmap=plt.cm.gray, interpolation='none', aspect=0.2)	
-plt.subplot(1,3,3)
-plt.imshow(prevPRNGout,cmap=plt.cm.gray,interpolation='none',  aspect=0.2)	
+ax1 = plt.subplot(1,3,1)	
+ax1.set_xticklabels([])
+plt.imshow(prevSRout,cmap=plt.cm.gray, interpolation='none', aspect=0.1)		
+ax2 = plt.subplot(1,3,2)
+ax2.set_xticklabels([])
+plt.imshow(prevCAout,cmap=plt.cm.gray, interpolation='none', aspect=0.1)	
+ax3 = plt.subplot(1,3,3)
+ax3.set_xticklabels([])
+plt.imshow(prevPRNGout,cmap=plt.cm.gray,interpolation='none',  aspect=0.1)	
 
 plt.show()

### 2016-07-01 verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/impl_1/rundef.js

@@ -1,0 +1,44 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+// pre-commands:
+ISETouchFile( "init_design", "begin" );
+ISEStep( "vivado",
+         "-log weightRAM.vdi -applog -m64 -messageDb vivado.pb -mode batch -source weightRAM.tcl -notrace" );
+
+
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-07-01 verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/impl_1/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/impl_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log weightRAM.vdi -applog -m64 -messageDb vivado.pb -mode batch -source weightRAM.tcl -notrace
+
+

### 2016-07-01 verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/impl_3/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/impl_3'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log weightRAM.vdi -applog -m64 -messageDb vivado.pb -mode batch -source weightRAM.tcl -notrace
+
+

### 2016-07-01 verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/synth_1/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/synth_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log weightRAM.vds -m64 -mode batch -messageDb vivado.pb -notrace -source weightRAM.tcl

### 2016-07-01 verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/synth_2/rundef.js

@@ -1,0 +1,40 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+ISEStep( "vivado",
+         "-log weightRAM.vds -m64 -mode batch -messageDb vivado.pb -notrace -source weightRAM.tcl" );
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-07-01 verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/synth_2/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/weight_ram/Synth_weight_ram/Synth_weight_ram.runs/synth_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log weightRAM.vds -m64 -mode batch -messageDb vivado.pb -notrace -source weightRAM.tcl

### 2016-07-03 verilog/network_train/readres2.py
Added Complete Regularization

@@ -23,7 +23,7 @@ largest_number = pow(2,binary_dim)
 hiddenSz = 8
 QN = 6
 QM = 11
-numTrain = 100000
+numTrain = 200000
 prevX = list()
 outputGolden  = list()
 outputVerilog = list()

### 2016-07-05 paper/paper.tex
Added the IEEE paper for Reconfig

--- 2016-07-01 finalText/theorBack.tex
@@
-\chapter{Problem Characterization}\label{chap:theorBack}
+\documentclass{IEEEtran}
 
-\section{Theoretical Background}\label{sec:tb}
+%% Additional Packages
+\usepackage{cite}
+\usepackage{graphicx}
+%% Macros
+\newcommand{\mb}[1]{\mathbf{#1}}
 
-\subsection{Basic Concepts of Machine Learning}\label{sec:tb_basicCon}
-Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation.
-Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters according to the input data it receives, improving the estimates it yields as it receives new input samples.
+\title{An FPGA Implementation of a Long Short-Term Memory Neural Network}
 
-Let us consider a practical example.
-For instance, suppose we want to build a program that given an input audio waveform representation of a spoken word, it matches it into a particular word of a dictionary.
-We could, of course, devise a set of rules and exceptions for each word analysing some of its features (perhaps the Fourier representation of each one, and, from it, manually finding the appropriate rules for each), but apart from being a very complex task, it wouldn't be a scalable solution, given the enormous number of words in each language.
-The approach taken by Machine Learning is different, and instead of manually processing each waveform, we build a large dataset, of size $N$, containing the waveforms of several words $\left[ \mb{x}_1 ,\; \mb{x}_2 ,\;\mb{x}_2 ,\; \dots ,\;\mb{x}_N \right]$ -- we call this dataset the \textbf{training dataset} -- and we feed it to our model.
-Each of the $i$-th data point was previously labelled, and in fact we feed each training data point $\mb{x}_i$ \emph{along} with its corresponding label $t_i$, so that the model can adapt its parameters accordingly to the \emph{target value} it is supposed to classify.
-This set of labels $\mb{t} = \left[ t_1 ,\; t_2 ,\; t_3 ,\; \dots ,\; t_N \right]$ is called the \textbf{target vector}.
+%% COMMENT THIS FOR SUBMISSION
+\author{\IEEEauthorblockN{Jose Pedro Castro Fonseca} \\ 
+\IEEEauthorblockA{Faculty of Engineering University of Porto \\
+Porto, Portugal}}
+\IEEEspecialpapernotice{(Working Version)}
 
-We are, then, left with the following question: how can the model quantitatively evaluate the quality of its current set of parameters?
-That could be achieved in a number of ways, but the most usual is using a \textbf{Cost Function} that, as the name suggests, measures the cost of each wrong classification of the model.
-The model then evolves in a way that minimizes the cost function.
-A usual choice for the cost function is the \textbf{sum of squares error}, given the Gaussian Noise assumption.
-Mathematically, if $y_{\theta}^i = y_{\theta}(x_i)$ is the prediction for the input data point $x_i$ with label $t_i$, given the current set of parameters $\theta$, the cost function using this metric is given by
+\begin{document}
 
-\begin{equation}\label{eq:costfunctionFund}
-	J(\theta) = \frac{1}{2} \sum_{i=1}^N \left( y_{\theta}^i - t_i \right)^2.
-\end{equation}
+\maketitle
 
-Sometimes, instead of applying the raw data to the model, we can apply some sort of \emph{preprocessing} to the data to extract the relevant features from it.
-For instance, instead of just feeding a raw image, we can perform several operations like edge detection or low-pass filtering, and apply them in parallel.
-In cases of highly dimensional data (i.e.
-each data vector has a very high number of features), we can apply techniques like \textbf{Principal Component Analysis} to reduce the feature space to a smaller dimension one, where the previous features were combined into two or three new features that pose themselves as the most relevant.
+\begin{abstract}
+This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
+to outperform software implementations, by exploiting the inherent parallelism in it.  
+The main design decisions are presented, along with the proposed network architecture. A description of the main
+building blocks of the network is also presented. The network is synthesized for various sizes,
+and performance results are presented and analysed. The synthesized network achieves a 195 times speed-up
+over a custom-built software network, proving the benefits of parallel computation for this kind of network.
+\end{abstract}
 
-The problems described above are, in fact, a subset of the problems that Machine Learning tries to address.
-These problems are called \textbf{classification problems}, because for each input data point, our model tries to fit it into the most appropriate class.
-But we can also address \textbf{regression problems} where the output is not limited to a discrete set of values but rather a continuous interval.
-On the other hand, the Neural Network that this work will implement addresses a special kind of classification problem, where the classification decision is influenced not only by the current input sample, but also by a given \emph{window of samples} that trail the current sample.
+\section{Introduction}\label{sec:intro}
+\IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
+particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain 
+time step are also fed as the input in the next time step, and that way their memory elements can make
+sense of patterns within data sequences, unlike classical recursive neural networks.
+These algorithms have been profusely implemented in software, and their practical applications are plentiful. 
+However, the benefits of the inherent parallelism offered by a dedicated hardware
+platform are not availed, and there are relatively few implementations of Machine Learning algorithms in 
+these kind of platforms.
 
-In summary, the most typical setting for a Machine Learning problem is having a large \textit{input dataset} which we use to \textit{train} our model (i.e.
-allowing him to dynamically adapt its set of parameters $\theta$), in order to produce an output label $y_i$ for each of them that minimizes a quality metric, the \textit{Cost Funcion}, which can be chosen to the sum of squared differences, the log-loss, or any other appropriate  mathematical relationship  between the estimate $y_i$ and the correct label $t_i$.
+Hardware platforms would achieve a considerable speed-up over software implementations, which could prove
+useful for high data throughput systems, where the calculation overhead is critical and limits the performance of the systems.
+Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
+of a given experiment faster than in a regular CPU, yielding an increase in scientific productivity.
 
-Now that the basic Machine Learning concepts have been presented, I will discuss, henceforth, one of the most important algorithms that address the supervised classification problems, the \emph{Artificial Neural Networks} that will be discussed in Section~\ref{sec:theorBack_ann} as a contextual introduction to the main theme of the thesis, which will be Recurrent Neural Networks (Section~\ref{sec:theorBack_rnn}), namely the \textbf{Long Short-Term Memory Networks} (Section~\ref{sec:theorBack_lstm}), both of which are improvements over the initial formulation of the ANNs.
-These two last networks branch even further from these set of problems, and are usually employed in \textit{Deep Learning} tasks, where we try to extract even higher level information from data at the expense of increased model complexity.
+Hitherto, there is only one implementation~\cite{Chang15}, but its performance is undermined
+by the external memory access. This implementation aims to make use of internal FPGA memory resources, and therefore
+achieve a higher speed performance.
 
-\subsection{Artificial Neural Networks}\label{sec:theorBack_ann}
+In overview, this article begins by explaining what are LSTM networks -- and their mathematical details -- in Section~\ref{sec:lstmnn}, and by providing a quick overview of their state of the art in Section~\ref{sec:relwork}.
+Section~\ref{sec:proparch} outlines the proposed hardware architecture and its main constituent modules, and the performance and synthesis results are reported in Section~\ref{sec:results}.
+A final concluding remark is given in Section~\ref{sec:concl}.
 
-Artificial Neural Networks (ANN) are mathematical structures that, as the name suggests, try to mimic the basic way of how a human brain works.
-ANN's building blocks, like their biological counterpart, model the high-level behaviour of biological neurons, in the sense that they neglect unnecessary biological aspects (such as modeling all the voltages across the neuron and all its electromagnetic interactions), and only retain its fundamental underlying mathematical function, which is a weighted linear combination of its inputs subject to a \emph{activation function}, i.e.
-a function that outputs a decision value depending on its inputs.
-Mathematically, we have
-
-\begin{equation}
-	y = f(\mb{w}^T \mb{x})
-\end{equation}
-where $\mb{w} = \left[ w_0 \; w_1 \; w_2 \;  w_3 \;  \dots \; w_n \right]$ is the input weight vector, $\mb{x} = \left[ x_0 \; x_1 \; x_2 \; x_3 \; \dots \; x_n \right]$ the input vector, $b_0$ is the bias factor and $f(\cdot)$ is the chosen activation function.
-Furthermore, we call the scalar quantity $a = \mb{w}^T \mb{x}$ the \textbf{activation}, since its value determines how the activation function will behave.
-Figure~\ref{fig:modelNeuron} exemplifies the roles of these variables within our neuron model, and compares each part of it with the biological counterpart.
-
-\begin{figure}[H]
-	\begin{subfigure}{0.5\linewidth}
-		\centering
-		\includegraphics[width=0.9\linewidth]{figures/neuron.eps}
-		\caption{Neural Network Node}
-		\label{fig:modelNeuron_a}
-	\end{subfigure}
-	\begin{subfigure}{0.5\linewidth}
-		\centering
-		\includegraphics[width=0.9\linewidth]{figures/MultipolarNeuron.png}
-		\caption{Biological Neuron Diagram}
-		\label{fig:modelNeuron_b}
-	\end{subfigure}
-
-    \caption[Representations of a Neuron in an ANN~\ref{fig:modelNeuron_a} and a biological multi polar neuron~\ref{fig:modelNeuron_b}]{In Figure~\ref{fig:modelNeuron_a}.
-each input feature $x_i$ is weighted by its corresponding weight $w_i$.
-During the training procedure, these weights are adjusted so that the output $y$ approaches the target value.
-In Figure~\ref{fig:modelNeuron_b}, we see the diagram of an actual multi polar neuron.
-The dendrites, where the stimuli are received, plays a role similar to that of the input nodes.
-The axon transmits the signal to the synaptic terminals, that are similar to the $y$ output.
-Source: \href{https://en.wikipedia.org/wiki/Neuron\#/media/File:Neuron\_Hand-tuned.sv://en.wikipedia.org/wiki/Neuron\#/media/File:Blausen\_0657\_MultipolarNeuron.png}{Wikipedia}}
-	\label{fig:modelNeuron}
-\end{figure}
-
-As far as the activation function is concerned, we can have several types.
-An immediate choice would be the \textbf{Binary Step Function} that  outputs -1 if the activation is \textbf{below} a given threshold and 1 otherwise.
-There can also be \textbf{real valued activation functions}, whose output is not binary, but rather that of a continuously differentiable function, such as the logistic sigmoid $\sigma(a) = \frac{1}{1 + \mathrm{e}^{-a}}$ or the hyperbolic tangent $\tanh(a)$.
-This aspect will prove useful for the usual training methods, that involve the computation of derivatives.
-In Figure~\ref{fig:activFunc} these activation functions are plotted.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.9\linewidth]{figures/activFunc.eps}
-    \caption[Possible activation functions of choice for a neuron]{Three different activation functions.
-As you can see, the hyperbolic tangent has the same extreme values as the sign step function, but has a smooth transition between them, which can be interpreted as a \emph{soft decision} in the more ambiguous middle region, reflecting the degree of uncertainty on the decision.
-On the other hand, the sigmoid function goes from zero to one, and is also smooth like the hyperbolic tangent.}
-	\label{fig:activFunc}
-\end{figure}
-
-A neuron by itself can be thought of as a simple linear regression, where we optimize the weight of each feature according to a target value, or function.
-While important in some applications, the main interest in ANN is to evaluate increasingly more complex models, and not a simple linear regression.
-This is achieved by \emph{chaining} nodes to one another, connecting the output of a given node to one of the inputs of another.
-We call \emph{layers} to a group of these nodes that occupy the same hierarchical position.
-There can be any number of layers, with any number of nodes, but most implementations generally have 3 layers: the \emph{input} layer, the \emph{hidden} layer (in the middle) and the \emph{output} layer.
-Figure~\ref{fig:neuralnet} suggests a possible structure for a 3 layer ANN.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.9\linewidth]{figures/ann.eps}
-    \caption[Three layer ANN]{A three layer ANN.
-We have omitted some of the connections in the hidden layer, for simplification purposes.
-$\mb{w}_1$ represents the weight matrix of the input layer, $\mb{w}_2$ the weight matrix of the connections between the input layer and the hidden layer, and $\mb{w}_3$ the weight connections between the hidden and the output layer.
-$f_{ij}(\dots)$ is the activation function of the $j$-th neuron of the $i$-th layer.
-Since they can be different, I chose different indexes to each.}
-	\label{fig:neuralnet}
-\end{figure}
-
-Regarding the training of ANNs, it is performed through a two-step process: first, a \textbf{feed-forward} step where the input is applied, and the activations are evaluated in succession up to the output neurons; then, we perform the \textbf{backpropagation} step, where we calculate the errors in each of the nodes (the so-called \textbf{deltas} of equation~\ref{eq:deltaj}), but now from the output to the input: the weights are updated and optimized using an iterative  method called \textit{Gradient Descent}, where if $\tau$ is the current time step, the next update on the weight matrix $\mb{w}^{(\tau+1)}$ is given by
-
-\begin{equation}\label{eq:gradDesc}
-	\mb{w}^{(\tau+1)} = \mb{w}^{(\tau)} - \eta \nabla E(\mb{w}^{(\tau)})
-\end{equation}
-where $E(\cdot)$ is the error function.
-As we can see, the weight matrix is moved in the direction that minimizes the error function the most, and $\eta$ controls how fast this is achieved, being the reason why it is called the \textbf{learning rate}.
-
-The computation of gradient of the error function comprises the evaluation of its derivatives with respect to each weight of all network connections, $w_{ij}$. They are
-
-\begin{equation}\label{eq:partialE}
-	\frac{\partial E}{\partial w_{ji}} = \delta_j f(a_i)
-\end{equation}
-where $f(\cdot)$ is the activation function of the neuron and
-
-\begin{equation}\label{eq:deltaj}
-	\delta_j = f'(a_j) \sum_k w_{kj} \delta_k .
-\end{equation}
-The interpretation of these equations is simple.
-If $w_{ji}$ is the weight of the connection between the neuron $j$ we are considering and a neuron $i$ in a previous layer, then the sum over $k$ relates to all the neurons in the \emph{next} layer to which $j$ connects: this way, since the update of $w_{ji}$, according to~\ref{eq:gradDesc}, is given by
-
-\begin{equation}\label{eq:update}
-	w_{ji}^{(\tau+1)} = w_{ji}^{(\tau)} - \eta \frac{\partial E}{\partial w_{ji}}
-\end{equation}
-we see that, from~\ref{eq:partialE}  it simply is the product of the error of the current neuron, $\delta_j$, with the output of the previous neuron $f(a_i)$.
-In turn, from~\ref{eq:deltaj}, we see the recurrence relationship between it and the weighted sum of all posterior neurons that connect to it.
-Hence, the name backpropagation is now clear: we are, in fact, propagating the errors backwards into the neuron of interest, weighted by the corresponding weight, but now \emph{backwards} instead of forward, as before.
-For the output units, the $\delta_j$ is simply the difference between the produced output and the corresponding label for that sample.
-This two-step process is performed for every data point in out dataset.
-%%Place where I can include an algorithmic overview?
-For a complete proof of the above formulas, see~\cite[chap. 5.3.1]{Bishop2006}.
-
-\subsection{Recurrent Neural Networks}\label{sec:theorBack_rnn}
-A Recurrent Neural Network (RNN) is, essentially, a regular ANN where some neurons (especially in the hidden layer) have \emph{feedback connections} to themselves, i.e.
-their outputs are fed as inputs.
-The relevance of this different structure is the possibility to retain \emph{sequence} information about the data.
-Before, each incoming data point only contributed to the training of the network, but the information about the correlation between themselves and the data points that preceded them did not influence the training step.
-Temporal relationship is disregarded and each data point considered conditionally independent of any other.
-This is obviously not necessarily true, and in fact there are many cases where the correlation between data points is high for those closely spaced in time, such as in video signals, audio signals, or other kinds of \emph{temporal sequences} of data.
-Therefore, the feedback connection of the neuron to himself acts as a kind of \emph{memory element} that takes into account in the present decision, the history of decisions previously taken, and hence the previous data.
-
-Figure~\ref{fig:recurrneuron} suggests a possible structure for a neuron of a hidden layer in an RNN, and also an alternate representation, where the structure is unfold through time.
-
-\begin{figure}[H]
-	\begin{subfigure}{0.5\linewidth}
-		\centering
-		\includegraphics[width=0.34\linewidth]{figures/recneuron.eps}
-		\caption{Recurrent Neuron}
-		\label{fig:recneuron}
-	\end{subfigure}
-	\begin{subfigure}{0.5\linewidth}
-		\centering
-		\includegraphics[width=0.9\linewidth]{figures/recneuron_unf.eps}
-		\caption{Recurrent Neuron, unfolded through time}
-		\label{fig:recneuron_unf}
-	\end{subfigure}
-
-    \caption[Representation of a Recurrent Neuron (Figure~\ref{fig:recneuron}) and a Recurrent Neuron unfolded through time (Figure~\ref{fig:recneuron_unf})]{In Figure~\ref{fig:recneuron}, a recurrent neuron is depicted, where the output is fed back to the input, weighted by $\theta_h$.
-Figure~\ref{fig:recneuron_unf}, an unfolding through time of that same neuron is performed.
-The basic idea is that feeding the output back to the input is similar to feeding the output to the input of the neuron \emph{at a previous time step}: this way, we linearize the structure.
-Considering a training epoch of $T$ samples, corresponds to having $T$ unfolded neurons/layers}
-	\label{fig:recurrneuron}
-\end{figure}
-
-The training of an RNN is usually performed using a variation of Backpropagation, called \textbf{Backpropagation Through Time} (BPTT), that as the name implies, performs the same backpropagation procedure discussed for the ANNs, but now taking into account the unfolding of the network through a fixed training epoch $T$ like Figure~\ref{fig:recneuron_unf}.
-Due to this very fact, this training procedure is memory and performance consuming, and so it will not be used in my final work, but instead a novel approach, the \textbf{Simultaneous Perturbation Stochastic Approximation} will be evaluated.
-
-Even though RNNs outperform static ANNs in sequence recognition problems~\cite{Bengio1991}, they fail to retain long-term dependencies.
-Of course that the weight training process is itself a form of memory, but the problem is that the weight update is much slower than the activations~\cite{Yoshua01}, and therefore this memory only retains short-term dependencies.
-This is because of the so-called \textbf{Vanishing Gradients Problem}~\cite{Yoshua94,Yoshua01}, where the error decays exponentially through time, and the impact of previous incoming data points on the training of the weights, and thus the current decision, quickly decreases.
-
-\subsection{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
-To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}.
-This section explains the main idea of this approach (Section ~\ref{sec:struct_lstm}), and also how it is trained (Section~\ref{sec:training_lstm}), serving as a support for the work of this thesis.
-Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}.
+\section{LSTM Neural Networks}\label{sec:lstmnn}
+Although originally formulated in~\cite{Hoch97}, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}.
 One of the inital proposers of LSTM, Prof.
@@
 
-\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
 A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}.
@@
 
-\begin{figure}[H]
+\begin{figure}[!t]
 	\centering
-	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
+	\includegraphics[width=2.5in]{figures/lstmneuron.png}
     \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
@@
 
-\begin{align}
-    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \notag\\
-    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \notag\\
-    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \notag\\
-    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \notag\\
-    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \notag \\
+\begin{eqnarray}
+    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
+    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
+    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
+    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
+    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
     \mb{y}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
-\end{align}
+\end{eqnarray}
 where $\odot$ is the Hadamard multiplication.
@@
 the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$.
-Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port.
-In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{*}$, $\mb{R}_{*}$ and the remaining.
 
-\subsubsection{Training -- SPSA}\label{sec:training_lstm}
-Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme.
-Since I am aiming at an hardware implementation, and although the memory resources of current FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system.
-That being said, we see that the usual training algorithms for LSTM cells -- i.e.
-Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance.
-A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.
-The main idea of this technique is, instead of explicitly evaluating the gradients for the cost function at each time step, to perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
-The weight update for the $i$-th weight at the time step $t$ is given by
+\section{Related Work}\label{sec:relwork}
 
-\begin{equation}\label{eq:spsa_weightUpdate}
-    \Delta w_i^{(t)} = \frac{J(\mb{w}^{(t)} + \beta \mb{s}^{(t)}) - J(\mb{w}^{(t)})}{\beta s_i^{(t)}}
-\end{equation}
-where $\beta$ is the magnitude of the perturbation to be introduced, $\mb{s}^{(t)}$ is a \textit{sign vector} whose $i$-th element, $s_i^{(t)}$, is either $-1$ or $1$.
-This way, we see that every weight is \textbf{randomly} incremented either by $-\beta$ or $\beta$, and we only need to keep a duplicate of the weight matrix with the perturbations, and we only need to evaluate the cost function twice per incoming sample.
-As for the \textit{update rule}, we have
+\section{Proposed Architecture}\label{sec:proparch}
 
-\begin{equation}\label{eq:spsa_updateRule}
-    w_i^{(t+1)} = w_i^{(t)} - \eta \Delta w_i^{(t)}
-\end{equation}
-where $\eta$ is the learning rate, and $\Delta w_i^{(t)}$ is the update for the $i$-th weight evaluated in~\ref{eq:spsa_weightUpdate}.
-According to the analysis performed in~\cite{Maeda05}, performing a two-term Taylor expansion at $\mb{w} = \mb{w}^{(t)}$, and taking the expected value of the equation, we get that there is a $\mb{w}_{S1}$ so that
+\subsection{Activation Function Calculation}
 
-\begin{equation}\label{eq:taylor}
-    \Delta w_i^{(t)} = s_i^{(t)} (\mb{s}^{(t)})^T \frac{\partial J(\mb{w}^{(t)})}{\partial \mb{w}} +
-                        \frac{c s_i^{(t)}}{2} (\mb{s}^{(t)})^T \frac{\partial^2 J(\mb{w}^{(t)})}{\partial \mb{w}^2}  (\mb{s}^{(t)}) 
-\end{equation}
-If the random sign vector $\mb{s}^{(t)}$ is chosen carefully so that the expected value of the vector at the $i$-th position is zero, $E(s_i^{(t)})$, and the signs of two different components $i$ and $j$, at different time-steps, are independent (i.e.
-$E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, where $\delta$ is the Kronecker fuction), taking the \emph{Expected Value} of equation~\ref{eq:taylor} yields
+\subsection{Matrix-vector Dot Product Calculation}
 
-\begin{equation}\label{eq:spsa_gradProof}
-    E(\Delta w_i^{(t)}) = \frac{\partial J(\mb{w}^{(t)})}{\partial w_i^{(t)}}
-\end{equation}
-that is, the \emph{expected value} of the weight update approximates the gradient of the cost function relative to that weight, and so the learning rule is a special form of \textit{Stochastic Gradient Descent}.
+\subsection{Weight storage}
 
-One last comment to be made concerning the update rule, is the hypothetical need for \textbf{limits} to the weight values, when the update rule exceeds $|w_{\text{max}}|$ (in that case, we set $w_i^{(t+1)} = \pm w_{\text{max}}$), which sometimes might be needed if the behaviour of the weight update is not appropriate~\cite{Maeda05}.
+\subsection{Network Architecture}
+
+\section{Results}\label{sec:results}
+
+\subsection{Validation}
+
+\subsection{Synthesis}
+
+\subsection{Performance}
+
+\section{Conclusion}\label{sec:concl}
+
+\bibliographystyle{IEEEtran}  
+\bibliography{bibliography}
+
+\end{document}

### 2016-07-05 paper/figures/activFunc.py

@@ -1,0 +1,26 @@
+import matplotlib.pyplot as plt
+from matplotlib import rc
+import numpy as np
+import math
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
+
+a       = np.arange(-6.0, 6.0, 0.1)
+sigmoid = 1.0 / (1.0 + np.exp(-1.0*a))
+tanh    = np.tanh(a)
+step    = np.sign(a)
+
+
+
+plt.plot(a, sigmoid, 'r', label=r"$f(a) = \sigma(a)$");
+plt.plot(a, tanh, 'b', label=r"$f(a) = \tanh(a)$");
+plt.plot(a, step, 'g', label=r"\rmfamily $f(a) = $ sign$(a)$");
+plt.grid(True);
+plt.xlabel(r"Activation $a$");
+plt.ylabel(r"$f(a)$");
+plt.legend(loc=2);
+plt.savefig(filename='activFunc.eps', ransparent=True);
+plt.show();
+
+

### 2016-07-05 paper/figures/dspplot.py

@@ -1,0 +1,30 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
+
+n4 = np.array([28, 20])
+x_n4 = np.array([1,2])
+n8 = np.array([56, 40, 32])
+x_n8 = np.array([1,2,3])
+n16 = np.array([112, 80, 64])
+x_n16 = np.array([1,2,3])
+n32 = np.array([160])
+x_n32 = np.array([2])
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+plt.plot(x_n4, n4, 'yo-', label="$N=4$")
+plt.plot(x_n8, n8, 'ro-', label="$N=8$")
+plt.plot(x_n16, n16, 'bo-', label="$N=16$")
+plt.plot(x_n32, n32, 'go-', label="$N=32$")
+ax.set_xticks([0,1,2,3,4])
+plt.axis([0.8,3.2,0,180])
+plt.grid(True);
+plt.xlabel(r"$\log_2(K_G$)");
+plt.ylabel(r"Number of DSP Slices used");
+plt.legend(loc=1);
+plt.savefig(filename='dspuse.eps', transparent=True);
+plt.show()

### 2016-07-05 paper/figures/maxfreqplot.py

@@ -1,0 +1,30 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
+
+n4 = np.array([140.154, 136.054])
+x_n4 = np.array([1,2])
+n8 = np.array([132.89, 131.579, 101.01])
+x_n8 = np.array([1,2,3])
+n16 = np.array([125.471, 105.263, 93.897])
+x_n16 = np.array([1,2,3])
+n32 = np.array([90.498])
+x_n32 = np.array([2])
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+plt.plot(x_n4, n4, 'yo-', label="$N=4$")
+plt.plot(x_n8, n8, 'ro-', label="$N=8$")
+plt.plot(x_n16, n16, 'bo-', label="$N=16$")
+plt.plot(x_n32, n32, 'go-', label="$N=32$")
+ax.set_xticks([0,1,2,3,4])
+plt.axis([0.8,3.2,85,145])
+plt.grid(True);
+plt.xlabel(r"$\log_2(K_G$)");
+plt.ylabel(r"Clock Speed in MHz");
+plt.legend(loc=1);
+plt.savefig(filename='maxfreq.eps', transparent=True);
+plt.show()

### 2016-07-05 paper/figures/mprop.py

@@ -1,0 +1,30 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
+
+py = 1/np.array([65e-6, 72e-6, 96e-6, 185e-6])/(1e6)
+x_py = np.array([2, 3, 4, 5])
+k2 = 1/np.array([292.535e-9, 368.725e-9, 518.05e-9])/(1e6)
+x_k2 = np.array([2,3,4])
+k4 = 1/np.array([360.15e-9, 494e-9, 921.5e-9, 1.78e-6])/(1e6)
+x_k4 = np.array([2,3,4,5])
+k8 = 1/np.array([960.3e-9, 1.715e-6])/(1e6)
+x_k8 = np.array([3,4])
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+plt.plot(x_py, py, 'yo-', label="Python")
+plt.plot(x_k2, k2, 'ro-', label="Hardware $K_G = 2$")
+plt.plot(x_k4, k4, 'go-', label="Hardware $K_G = 4$")
+plt.plot(x_k8, k8, 'bo-', label="Hardware $K_G = 8$")
+ax.set_xticks([2,3,4,5])
+plt.axis([1.8,5.2,-0.2,3.5])
+plt.grid(True);
+plt.xlabel(r"$\log_2(N)$");
+plt.ylabel(r"Million Classifications per Second");
+plt.legend(loc=1);
+plt.savefig(filename='Mclass-psec.eps', transparent=True);
+plt.show()

### 2016-07-06 paper/paper.tex
Paper ready, only missing introduction to architecture and results

@@
 \usepackage{graphicx}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{siunitx}
+
 %% Macros
@@
 
-In overview, this article begins by explaining what are LSTM networks -- and their mathematical details -- in Section~\ref{sec:lstmnn}, and by providing a quick overview of their state of the art in Section~\ref{sec:relwork}.
-Section~\ref{sec:proparch} outlines the proposed hardware architecture and its main constituent modules, and the performance and synthesis results are reported in Section~\ref{sec:results}.
-A final concluding remark is given in Section~\ref{sec:concl}.
+In overview, this article begins by explaining what are LSTM networks -- and their mathematical 
+details -- in Section~\ref{sec:lstmnn}, followed by providing a quick overview of their state 
+of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed 
+hardware architecture and its main constituent modules, and the performance and synthesis 
+results are reported in Section~\ref{sec:results}. A final concluding remark is given in Section~\ref{sec:concl}.
 
 \section{LSTM Neural Networks}\label{sec:lstmnn}
-Although originally formulated in~\cite{Hoch97}, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}.
-One of the inital proposers of LSTM, Prof.
-Jürgen Schmidhüber, did a survey on the most common variations of the model last year~\cite{Greff15}, and this will the basis of this short theoretical presentation, as well as the work that will be developed in this thesis.
+LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally 
+updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. 
+One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations 
+of the model last year~\cite{Greff15}, which is the reference for this short discussion.
 
-A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}.
-As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network.
-Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}.
-The main components, their role and relevance, are explained as follows
+A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As it can be seen in the picture, 
+we still have the recurrent connections from the regular RNNs, but now there are multiple entry points 
+that control the flow of information through the network. Although omitted from the picture, all the 
+gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}. The main components, their 
+role and relevance, are explained as follows.
 
@@
 \begin{itemize}
-    \item \textbf{Input Gate} -- this is the input gate, where the importance of each feature of the input vector at time $t$, $\mb{x}^t$, and the output vector at the previous time step $\mb{y}^{t-1}$ is weighed in, producing an output $\mb{i}^{t}$.
+    \item \textbf{Input Gate} -- this is the input gate, where the relative importance of each feature of 
+the input vector at time $t$, $\mb{x}^{(t)}$, and the output vector at the previous time step, $\mb{y}^{(t-1)}$, 
+are weighed in, producing an output $\mb{i}^{(t)}$.
 
-    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information from the input gate to the memory cell.
-It also receives the input vector and the previous output vector as inputs, but it does not have peephole connections and its dynamics are controlled by a different set of weights.
-The \textbf{activation function of this gate can be either}, but the most common choice is the \textbf{Hyperbolic Tangent}.
+    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information 
+from the input gate to the memory cell. It also receives the input vector and the previous output 
+set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either}, 
+but the most common choice is the \textbf{Hyperbolic Tangent}.
 
-    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to set them or reset them, using the \textit{Hadamard Elementwise} matrix multiplication of its output at time $t$, $\mb{c}^{(t)}$, with the contents of the memory unit at the previous time step, $\mb{c}^{(t-1)}$.
-The activation function of this gate is \textbf{always sigmoid}.
+    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to 
+set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time 
+$t$, $\mb{c}^{(t)}$, with the contents of the memory unit at the previous time step, $\mb{c}^{(t-1)}$. 
+The activation function of this gate is \textbf{always sigmoid}, and the resulting signal is $\mb{f}^{(t)}$.
 
-    \item \textbf{Output Block Gate} -- this gate has a role very similar to that of the Block Input Gate, but now it controls the information flow \textit{out} of the LSTM neuron, namely the activated Memory Cell output.
+    \item \textbf{Output Block Gate} -- this gate has a role very similar to that of the Block Input 
+Gate, but now it controls the information flow \textit{out} of the LSTM neuron, namely the activated 
+Memory Cell output. The control signal it produces is $\mb{o}^{(t)}$.
 
-    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron.
-This is the memory element of the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates that connect to it.
-Also, this is where the peephole connections come from.
+    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of 
+the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates 
+that connect to it. Also, this is where the peephole connections come from.
 
-    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation function that, as the gate activation function, can be any, but the \textit{hyperbolic tangent} is the most common choice.
+    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation 
+function that, as the gate activation function, can be any, but the \textit{hyperbolic tangent} 
+is the most common choice.
 
-    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates to `peep' at the states of the memory cell.
-They were added after the initial 1997 formulation, and their absence was proven to have a minimal performance impact~\cite{Greff15}.
+    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates 
+to `peep' at the states of the memory cell. They were added after the initial 1997 formulation, and their 
+absence was proven to have a minimal performance impact~\cite{Greff15}. For this reason, they were ommitted 
+in this architecture.
+
 \end{itemize}
 
-After these small conceptual definitions, that allow us to grasp some intuition on the operation of a \textit{single} LSTM cell, I can present the overview of a \textit{layer} of LSTM neurons and their formal mathematical formulation, that will be needed both for the high-level model and the HDL description.
 The operation of each set of gates of the layer is given by the following set of equations
-
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\odot$ is the Hadamard multiplication.
-The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer.
-Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e.
-the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$.
+where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds 
+to the value of the $i$-th neuron in the layer, which is a very convenient and compact representation of the 
+whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that 
+precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent 
+weight matrices $\mb{R}_*$ are squre matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$ 
+and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.
 
 \section{Related Work}\label{sec:relwork}
+LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is
+superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address 
+the same set of problems where LSTM are employed, namely predicting and producing classification decisions 
+from time-series data. A very comprehensive description of applications can be found in one of the initial 
+authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/~juergen/rnn.html}.
+
+The use of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass 
+HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for 
+instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark, 
+and also in \textbf{Large Scale Acoustic Modelling of Speech}~\cite{Sak14}. Other uses include 
+\textbf{Handwriting Synthesis}~\cite{Graves13_2}, \textbf{Translation}~\cite{Sustkever14}, \textbf{Biomedical Applications} such
+as protein homology detection~\cite{Hochreiter07}, \textbf{Music Analysis and Composition}, such as the transcription of piano music 
+to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
+in~\cite{Vinyals14},~\cite{Donahue14} and~\cite{Donahue14_2}.
+
+Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently 
+(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented,
+with 128 neurons each, which, for processing 1000 samples, yielded an execution time of $0.932 \, \si{\second}$. Assuming 
+that both layers are equal, and thus have the same execution time, this yields an approximate execution time of 
+$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$). 
+Therefore, this network is able to perform around 34.3 thousands of forward propagations per second. 
+Although the network is large, its performance is well below this work, because it does not have
+a full level of parallelism when compared to the proposed design of Section~\ref{sec:proparch}, and it
+makes use of external memory to store the weights. This last point is understandable, since the authors use
+the FPGA core as a co-processor for the main CPU.
 
 \section{Proposed Architecture}\label{sec:proparch}
+Before presenting the 
+\subsection{Activation Function Calculation}\label{sec:proprarch_af}
+In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).
 
-\subsection{Activation Function Calculation}
+The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as~\ref{eq:remezline}
 
-\subsection{Matrix-vector Dot Product Calculation}
+\begin{equation}\label{eq:remezline}
+    p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
+\end{equation}
+where $i \in \left[0, n+1\right]$.
+Of course, both functions have horizontal assymptotes, which can be used as the value in the edges, but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function as the value of the assymptote, I have further split the interval in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
+The algorithm was run using Python, and targeted second degree polynomials for each interval.
+After that, the Verilog model was described, where the correct coefficients are loaded according to the interval where the incoming operand is located.
+The evaluation of the polynomial was accomplished using the \textbf{Horner's Rule}, noting that
 
-\subsection{Weight storage}
+\begin{equation}\label{eq:factorPol}
+p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
+\end{equation}
+and, therefore, the calculation is simply the procedure of multiplying the operand by a value and adding a constant to the result,
+repeated two times. Again the internal machine state controls which values are multiplied and added according to the pipeline state.
+The output of the Sigmoid HDL block  and the $\tanh$ HDL block are presented in Figures~\ref{fig:nonlin-out-sig} 
+and~\ref{fig:nonlin-out-tanh}, respectively. Due to the internal datapath pipeline, the module takes 5 clock cycles to output a result.
 
-\subsection{Network Architecture}
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
+    \caption[Plot of the output of the Sigmoid Calculator HDL module]{Plot of the output of the Sigmoid Calculator HDL module}
+    \label{fig:nonlin-out-sig}
+\end{figure}
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
+    \caption[Plot of the output of the Hyperbolic Tangent HDL module]{Plot of the output of the Hyperbolic Tangent HDL module}
+    \label{fig:nonlin-out-tanh}
+\end{figure}
+
+\subsection{Matrix-vector Dot Product Calculation}\label{sec:proprarch_dot}
+From Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are 
+multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, 
+this block implements matrix-vector multiplication to perform those calculations, and is parameterized in 
+order to accommodate networks of various sizes, since $\mb{W}_*$ has size $N\times M$, and $\mb{R}_*$ 
+has size $N\times N$ -- $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$ 
+(the number of neurons in the layer). This way, if a layer with different parameters is needed, either in terms of 
+number of inputs or number of neurons, we only need to change the respective parameter before the synthesis stage, 
+instead of having to redesign the whole block for that particular size.
+
+The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}
+
+\begin{algorithm}
+\begin{algorithmic}
+\For {$i$ = $1:N$}
+    \For {$j$ = $1:M$}
+    \State $y_i := y_i + A_{ij} \cdot x_j$
+    \EndFor
+\EndFor
+\end{algorithmic}
+\caption{Matrix-vector multiplication of a matrix}
+\label{matvec-alg}
+\end{algorithm}
+This operation has a computational complexity of $O(n^2)$.
+Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix.
+Follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.
+
+Although this solution only requires one multiplication per row of the input matrix (i.e.
+$N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions.
+The solution found for this design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e.
+the number of columns); now, defining a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
+For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
+
+Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before.
+Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel.
+They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
+The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output register, respectively.
+In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$.
+After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.
+Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=2.5in]{figures/mem-array-prod.eps}
+    \caption[A column of the matrix that serves as input to the module]{A column of the matrix that serves as input to the module.
+The dark shaded part is for the first multiplier, and the light shaded is for the other, in parallel.
+The rowMux signal addresses the position within each shaded area}.
+    \label{fig:mem-arrayprod}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=2.5in]{figures/array-prod.eps}
+    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module.
+The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
+    \label{fig:array-prod}
+\end{figure}
+
+\subsection{Weight storage}\label{sec:proprarch_ram}
+The weights are stored in LUTRAM, and for that purpose, they are declared as a matrix of registers, where the access, both in terms
+of write and read operations, is made to each column. The write/read access can be performed at the same time, since the memory has
+separate communication ports for input/output, provided that the \textbf{addresses do not coincide}.
+
+\subsection{Gate Module}\label{sec:proparch_gate}
+The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$.
+If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices.
+This way, a Gate module needs to perform three tasks
+
+\begin{enumerate}
+    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
+    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
+    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
+\end{enumerate}
+Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
+This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.
+The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result.
+After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this module takes to produce an output is $6+K_G\cdot N$.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=2.5in]{figures/gate.eps}
+    \caption[Diagram of the hardware block that implements the Gate]{Diagram of the hardware block that implements the Gate}
+    \label{fig:gate}
+\end{figure}
+
+\subsection{Network Architecture}\label{sec:proprarch_net}
+Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel.
+This way we need four Gate Modules working in parallel, each one with its respective weigth RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation function calculator (detailed in Section~\ref{sec:nonlincalc}).
+There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
+
+Furthermore, we can avoid a naïve translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together.
+This way, instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
+The issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$.
+These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
+    \caption[Block diagram of the hardware LSTM Neural Network]{Block diagram of the hardware LSTM Neural Network}
+    \label{fig:network-opt}
+\end{figure}
+The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place.
+Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, another multiplexer was added before the elementwise multiplication, to select the $\mb{c}^{(t-1)}$ signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
+
+The flip-flops on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network's state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
+The first activated flip-flop is the middle one, which keeps the result from the elementwise mutiplication of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the $\tanh$-$\sigma$-$(\cdot)$wise structure, the bottom flip-flop saves the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal.
+Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.
+
+Now, since there is only a single elementwise multiplier and only two activation function calculators, the total requirement for DSP slices is simply
+\begin{equation}\label{eq:numdsp_network-opt}
+    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right).
+\end{equation}
+where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact.
+In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion.
+After adjustments to the state machine, and accounting for pipelining and synchronization within the datapath, the clock cycles needed after the gate module calculations are 27, so the total clock cycles needed are
+
+\begin{equation}\label{eq:numcc_network-opt}
+    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
+\end{equation}
+which is only 13 clock cycles more than a fully-parallel, naïve architecture.
+For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
 

### 2016-07-07 python/LSTMlayer.py
Found the actual bug in training

@@ -240,7 +240,7 @@ class LSTMlayer :
 
         # The Cost Function evaluation for this perturbation
         cost = float(np.sum(Jpert-J))
-
+        print("Cost: ", J, "PertCost:", Jpert, "Diff: ", cost) 
         # ****TRAINING**** The LSTM Layer 
         #print("Gradient Check: ", np.divide(cost, self.Wz_update))
         for i in range(self.Wz.shape[0]):

### 2016-07-08 python/LSTMlayer.py
Corrected bug in cost function

@@ -41,24 +41,24 @@ class LSTMlayer :
 
         # Initializing the matrix weights
         #LSTM Block
-        self.Wz = np.random.random((hiddenUnits, inputUnits)) - 0.5 
-        self.Wi = np.random.random((hiddenUnits, inputUnits)) - 0.5
-        self.Wf = np.random.random((hiddenUnits, inputUnits)) - 0.5  
-        self.Wo = np.random.random((hiddenUnits, inputUnits)) - 0.5
+        self.Wz = np.zeros((hiddenUnits, inputUnits)) + 0.5 
+        self.Wi = np.zeros((hiddenUnits, inputUnits)) + 0.5
+        self.Wf = np.zeros((hiddenUnits, inputUnits)) + 0.5  
+        self.Wo = np.zeros((hiddenUnits, inputUnits)) + 0.5
         
-        self.Rz = np.random.random((hiddenUnits, hiddenUnits))- 0.5
-        self.Ri = np.random.random((hiddenUnits, hiddenUnits))- 0.5
-        self.Rf = np.random.random((hiddenUnits, hiddenUnits))- 0.5
-        self.Ro = np.random.random((hiddenUnits, hiddenUnits))- 0.5
+        self.Rz = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
+        self.Ri = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
+        self.Rf = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
+        self.Ro = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
         
         self.pi = np.random.random((hiddenUnits, 1))- 0.5
         self.pf = np.random.random((hiddenUnits, 1))- 0.5
         self.po = np.random.random((hiddenUnits, 1))- 0.5
         
-        self.bz = np.random.random((hiddenUnits, 1))-0.5
-        self.bi = np.random.random((hiddenUnits, 1))-0.5
-        self.bo = np.random.random((hiddenUnits, 1))-0.5
-        self.bf = np.random.random((hiddenUnits, 1))-0.5
+        self.bz = np.zeros((hiddenUnits, 1))+0.5
+        self.bi = np.zeros((hiddenUnits, 1))+0.5
+        self.bo = np.zeros((hiddenUnits, 1))+0.5
+        self.bf = np.zeros((hiddenUnits, 1))+0.5
         
         # Updates
[…]

### 2016-07-08 python/testLSTM.py

@@ -115,6 +115,7 @@ for i in range(maxEpoch):
     
     
     print("Average Error:", epochError/testSamp)
+    print("Weight: ", lstmLayer1.Wz.item(0,0))
     #plt.scatter(i, epochError/testSamp, linestyle='-.')
     #plt.draw()
     epochError = 0

### 2016-07-09 python/LSTMlayer.py
Final perceptron is not loaded with correct weights. Training improves

@@ -5,6 +5,7 @@ LSTM Layer Class
 """
 import numpy as np
 import time
+import pickle
 
 # Defines the seed for random() as the current time in seconds
 np.random.seed(round(time.time()))
@@ -41,24 +42,24 @@ class LSTMlayer :
 
         # Initializing the matrix weights
         #LSTM Block
-        self.Wz = np.zeros((hiddenUnits, inputUnits)) + 0.5 
-        self.Wi = np.zeros((hiddenUnits, inputUnits)) + 0.5
-        self.Wf = np.zeros((hiddenUnits, inputUnits)) + 0.5  
-        self.Wo = np.zeros((hiddenUnits, inputUnits)) + 0.5
+        self.Wz = np.random.random((hiddenUnits, inputUnits)) + 0.5 
+        self.Wi = np.random.random((hiddenUnits, inputUnits)) + 0.5
+        self.Wf = np.random.random((hiddenUnits, inputUnits)) + 0.5  
+        self.Wo = np.random.random((hiddenUnits, inputUnits)) + 0.5
         
-        self.Rz = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
-        self.Ri = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
-        self.Rf = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
-        self.Ro = np.zeros((hiddenUnits, hiddenUnits))+ 0.5
+        self.Rz = np.random.random((hiddenUnits, hiddenUnits))+ 0.5
+        self.Ri = np.random.random((hiddenUnits, hiddenUnits))+ 0.5
+        self.Rf = np.random.random((hiddenUnits, hiddenUnits))+ 0.5
+        self.Ro = np.random.random((hiddenUnits, hiddenUnits))+ 0.5
         
         self.pi = np.random.random((hiddenUnits, 1))- 0.5
         self.pf = np.random.random((hiddenUnits, 1))- 0.5
         self.po = np.random.random((hiddenUnits, 1))- 0.5
         
-        self.bz = np.zeros((hiddenUnits, 1))+0.5
-        self.bi = np.zeros((hiddenUnits, 1))+0.5
-        self.bo = np.zeros((hiddenUnits, 1))+0.5
-        self.bf = np.zeros((hiddenUnits, 1))+0.5
+        self.bz = np.random.random((hiddenUnits, 1))+0.5
[…]

### 2016-07-14 paper/paper.tex
Conference Paper Concluded

@@
 %% COMMENT THIS FOR SUBMISSION
-\author{\IEEEauthorblockN{Jose Pedro Castro Fonseca} \\ 
-\IEEEauthorblockA{Faculty of Engineering University of Porto \\
-Porto, Portugal}}
-\IEEEspecialpapernotice{(Working Version)}
+%%\IEEEauthorblockA{Faculty of Engineering University of Porto \\
+%%Porto, Portugal}}
+%%\IEEEspecialpapernotice{(Working Version)}
 
@@
 This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
-to outperform software implementations, by exploiting the inherent parallelism in it.  
+to outperform software implementations, by exploiting its inherent parallelism.
 The main design decisions are presented, along with the proposed network architecture. A description of the main
-building blocks of the network is also presented. The network is synthesized for various sizes,
-and performance results are presented and analysed. The synthesized network achieves a 195 times speed-up
+building blocks of the network is also presented. The network is synthesized for various sizes and platforms,
+and the performance results are presented and analysed. The synthesized network achieves a 251 times speed-up
 over a custom-built software network, proving the benefits of parallel computation for this kind of network.
@@
 particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain
-time step are also fed as the input in the next time step, and that way their memory elements can make
+time step are also fed as the input in the next time step, and that their memory elements can make
 sense of patterns within data sequences, unlike classical recursive neural networks.
@@
 Hitherto, there is only one implementation~\cite{Chang15}, but its performance is undermined
-by the external memory access. This implementation aims to make use of internal FPGA memory resources, and therefore
-achieve a higher speed performance.
+by the external memory access. This implementation aims to make use of internal FPGA memory resources,
+and therefore achieve a higher speed performance.
 
@@
 hardware architecture and its main constituent modules, and the performance and synthesis
-results are reported in Section~\ref{sec:results}. A final concluding remark is given in Section~\ref{sec:concl}.
+results are reported in Section~\ref{sec:results}. A final concluding remark is given in
+Section~\ref{sec:concl}.
 
@@
 LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally
-updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. 
-One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations 
+updated in~\cite{Gers00} and~\cite{Gers2000}, and the most recent version is~\cite{Graves05}.
+One of the inital proposers of LSTM, Prof. Jurgen Schmidhuber, did a survey on the most common variations
 of the model last year~\cite{Greff15}, which is the reference for this short discussion.
@@
 set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either},
-but the most common choice is the \textbf{Hyperbolic Tangent}.
+but the most common choice is the \textbf{hyperbolic hangent}.
 
@@
 set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time
-$t$, $\mb{c}^{(t)}$, with the contents of the memory unit at the previous time step, $\mb{c}^{(t-1)}$. 
-The activation function of this gate is \textbf{always sigmoid}, and the resulting signal is $\mb{f}^{(t)}$.
+$t$, $\mb{c}^{(t-1)}$. The activation function of this gate is \textbf{always sigmoid}, and the resulting
+signal is $\mb{f}^{(t)}$.
 
@@
     \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation
-function that, as the gate activation function, can be any, but the \textit{hyperbolic tangent} 
+function that, as the gate activation function, can be any, but the \textbf{hyperbolic tangent}
 is the most common choice.
@@
 \begin{eqnarray}
-    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
-    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
-    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
-    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
-    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
-    \mb{y}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
+    \mb{z}^{(t)} & = & g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
+    \mb{i}^{(t)} & = & \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
+    \mb{f}^{(t)} & = & \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
+    \mb{o}^{(t)} & = & \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
+    \mb{c}^{(t)} & = & \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
+    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
 \end{eqnarray}
@@
 precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent
-weight matrices $\mb{R}_*$ are squre matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$ 
+weight matrices $\mb{R}_*$ are square matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$
 and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.
@@
 
-The use of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass 
+The uses of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass
 HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for
@@
 Therefore, this network is able to perform around 34.3 thousands of forward propagations per second.
-Although the network is large, its performance is well below this work, because it does not have
-a full level of parallelism when compared to the proposed design of Section~\ref{sec:proparch}, and it
-makes use of external memory to store the weights. This last point is understandable, since the authors use
-the FPGA core as a co-processor for the main CPU.
+Therefore ts performance is well below this work, which sits at \textbf{0.487 millions} of forward propagations per second.
+This is because the work in~\cite{Chang15} does not have a full level of parallelism when compared to the proposed design of
+Section~\ref{sec:proparch}, and it makes use of external memory to store the weights. This last point is understandable,
+since the authors use the FPGA core as a co-processor for the main CPU, and not as a standalone implementation.
 
 \section{Proposed Architecture}\label{sec:proparch}
-Before presenting the 
+
+The building blocks of the network in Section~\ref{sec:proprarch_net} are presented in Sections~\ref{sec:proprarch_af} through~\ref{sec:proparch_gate}.
+The number representation system used for this network was a \textbf{signed fixed-point} Q6.11 system in two's complement. The are 7 integer bits (one of which is
+the sign bit), and 11 fractional bits, adding to a total of \textit{18 bits}. The reason we use 18 bits, is to make full use of the DSP48E1 slices available
+in the FPGAs.
+
 \subsection{Activation Function Calculation}\label{sec:proprarch_af}
-In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).
+In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05},
+since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
+number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles
+needed to output a result).
 
-The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as~\ref{eq:remezline}
+The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where
+the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as Equation~\ref{eq:remezline}
 
-\begin{equation}\label{eq:remezline}
-    p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
-\end{equation}
-where $i \in \left[0, n+1\right]$.
-Of course, both functions have horizontal assymptotes, which can be used as the value in the edges, but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function as the value of the assymptote, I have further split the interval in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
-The algorithm was run using Python, and targeted second degree polynomials for each interval.
-After that, the Verilog model was described, where the correct coefficients are loaded according to the interval where the incoming operand is located.
-The evaluation of the polynomial was accomplished using the \textbf{Horner's Rule}, noting that
+\begin{eqnarray}\label{eq:remezline}
+    & p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow  \nonumber \\
+		& p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
+\end{eqnarray}
+where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal assymptotes, which can be used as the value in the edges,
+but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
+as the value of the assymptote, the interval was further split in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
+The algorithm was run using Python, and targeted second degree polynomials for each interval. After that, the Verilog model was described, where the correct
+coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial
+was accomplished using the \textbf{Horner's Rule}, noting that
 
@@
     \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
-    \caption[Plot of the output of the Sigmoid Calculator HDL module]{Plot of the output of the Sigmoid Calculator HDL module}
+    \caption{Plot of the output of the Sigmoid Calculator HDL module}
     \label{fig:nonlin-out-sig}
@@
     \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
-    \caption[Plot of the output of the Hyperbolic Tangent HDL module]{Plot of the output of the Hyperbolic Tangent HDL module}
+    \caption{Plot of the output of the Hyperbolic Tangent HDL module}
     \label{fig:nonlin-out-tanh}
@@
 
-The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}
+The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{alg:matvec-alg}
 
@@
 \caption{Matrix-vector multiplication of a matrix}
-\label{matvec-alg}
+\label{alg:matvec-alg}
 \end{algorithm}
-This operation has a computational complexity of $O(n^2)$.
-Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix.
-Follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.
+This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel},
+each only requiring the corresponding $i$-th line from the matrix. Following this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
+which is one of the great advantages of custom-tailored hardware solutions.
 
-Although this solution only requires one multiplication per row of the input matrix (i.e.
-$N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions.
-The solution found for this design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e.
-the number of columns); now, defining a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
-For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
+Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row
+size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must
+be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution found for this
+design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg},
+each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final
+result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, defining a
+parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the
+same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
+For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector
+elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are
+ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
 
-Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before.
-Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel.
-They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
-The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output register, respectively.
-In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$.
-After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.
-Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.
+Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication
+units within the module, respectively, where $K_G = 4$ and for the same matrix and vector sizes as before. Note that in this situation,
+we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod},
+that work in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding
+column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
+The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a
+fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
+of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output vector position, respectively.
+In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+
+is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time
+proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
+in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.
 
@@
     \includegraphics[width=2.5in]{figures/mem-array-prod.eps}
-    \caption[A column of the matrix that serves as input to the module]{A column of the matrix that serves as input to the module.
+    \caption{A column of the matrix that serves as input to the module.
 The dark shaded part is for the first multiplier, and the light shaded is for the other, in parallel.
@@
 
-\begin{figure}
+\begin{figure*}[!t]
     \centering
-    \includegraphics[width=2.5in]{figures/array-prod.eps}
-    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module.
+    \includegraphics[width=\linewidth]{figures/array-prod.eps}
+    \caption{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module.
 The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
     \label{fig:array-prod}
-\end{figure}
+\end{figure*}
 
@@
 \subsection{Gate Module}\label{sec:proparch_gate}
-The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$.
-If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices.
-This way, a Gate module needs to perform three tasks
+The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we
+note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network,
+we can omit them in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks
 
@@
 \end{enumerate}
-Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
+Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication
+in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
 This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.
 The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result.
-After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this module takes to produce an output is $6+K_G\cdot N$.
+After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
+of clock cycles this module takes to produce an output is $6+K_G\cdot N$.
 
@@
 \subsection{Network Architecture}\label{sec:proprarch_net}
-Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel.
-This way we need four Gate Modules working in parallel, each one with its respective weigth RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation function calculator (detailed in Section~\ref{sec:nonlincalc}).
-There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
+Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they
+operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel. This way we
+need four Gate Modules working in parallel, each one with its respective weight RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation
+function calculator (detailed in Section~\ref{sec:proprarch_af}). There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which
+can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
 
-Furthermore, we can avoid a naïve translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together.
-This way, instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
-The issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$.
-These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.
+Furthermore, we can avoid a naive translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers
+and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output
+of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. This way, instead of replicating
+these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in. The
+issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses
+between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which
+is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.
 
-\begin{figure}
+\begin{figure*}
     \centering
@@
     \label{fig:network-opt}
-\end{figure}
-The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place.
-Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, another multiplexer was added before the elementwise multiplication, to select the $\mb{c}^{(t-1)}$ signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
+\end{figure*}
+The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine,
+and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place.
+Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$,
+another multiplexer was added before the elementwise multiplication, to select the $\mb{c}^{(t-1)}$ signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
 
@@
 \end{equation}
-which is only 13 clock cycles more than a fully-parallel, naïve architecture.
-For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
+which is only 13 clock cycles more than a fully-parallel, naive architecture. For instance, an $N=32$ neuron network would
+require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
 
@@
 
-\subsection{Validation}
+\subsection{Validation}\label{sec:res-val}
 
-\subsection{Synthesis}
+The functionality of the network was verified against a Python model of an LSTM network that was developed as a reference,
+both for the forward propagation of the network, as well as for the training algorithm. Python and Numpy were used rather than MATLAB,
+since the former has higher performance at the same level of code complexity. The model was trained using the SPSA method~\cite{Spall98}.
+
+The learning problem presented to both the software and hardware network is the \textbf{addition} of two binary numbers of 8 bits. The $i$-th
+bit of each number is fed to the network as a vector, and the network outputs its prediction of the correct value of the $i$-th bit
+of the \emph{result}. After the whole number is processed, the memory cells of the LSTM network are reset and a new addition
+task can be presented to the network. Even though this seems a rather simple problem, it accounts for all the essential issues at which this
+network excels: first, this is a \emph{classification} problem in which the Machine Learning algorithm needs to output a prediction based on
+the input feature vector, and that prediction has to take into account the \emph{history} of predictions and inputs so far, because the current
+bit is the sum of not only the bits of the two operands, but also the \emph{carry} generated at the last few positions -- this is where the
+memory cells of this special Recursive Neural Network come into effect.
+
+After the software network was trained, the trained weights were loaded into the network and several addition problems
+were fed to the Network in the testbench, which yielded no errors.
+
+\subsection{Synthesis}\label{sec:results_synth}
+The proposed network was first synthesized aiming a Xilinx XC7Z020 SoC, for sizes $N$ of 4, 8, 16 and 32, varying the resource sharing parameter $K_G$, and the number of inputs set
+to $M=2$. For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were
+successfully synthesized. This is because of the complexity of the input multiplexer of the Matrix-vector Dot Product Module in~\ref{sec:proprarch_dot}. To synthesize the
+design for sizes $N$ of 64, 128 and 256, a Virtex-7 VC707 board was used.
+
+\subsubsection{Maximum Frequency}\label{sec:results_synth_maxfreq}
+The maximum clock frequency that does not cause timing violations of any sort is plotted in Figure~\ref{fig:maxfreq}. For $N=4$, since there are
+only 4 rows to be multiplied, the maximum value of $K_G$ is 4, and hence no synthesis was performed for $K_G = 8$ (N.A.); also,
+since $K_G = 2$ and $N=32$, a network such as theses would use $32(8/2+3) = 224$ DSP slices, that exceeds the 220 slices
+available in the XC7Z020, so there is no synthesis data for that value (N.D.), as well.
+
+It is clear that with increasing $K_G$, the maximum clock speed decreases, and that decrease is steeper for larger values
+of $K_G$. This means that there is a critical path in the Matrix-Vector multiplication unit, whose multiplexer becomes
+increasingly complex for higher values of $K_G$. On the other hand, when $K_G$ is the same, smaller networks are faster
+than larger networks. The fastest design is an $N=4$ and $K_G = 2$ network, with a clock frequency of $158.228$ MHz, and
+the slowest one is an $N=32$ and $K_G=4$ network, clocked at $101.523$ MHz. The reference design used for validation in
+Section~\ref{sec:res-val} is an $N=8$ and $K_G=2$ network is clocked at $154.321$ MHz, which yields a clock period of
+$6.48$ ns.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/maxfreq.eps}
+    \caption{The maximum achievable clock frequency for feasible network sizes in the XC7Z020}
+    \label{fig:maxfreq}
+\end{figure}
+
+As previously stated, a synthesis targeting a Virtex-7 VC707 platform was performed, in order to synthesize networks of
+larger size that, due to the LUT usage restriction, could not be synthesized in the XC7Z020. The maximum clock achievable
+for each network size is reported in Table~\ref{tab:maxfreq-virtx7}
+
+\begin{table}
+	\caption{Maximum clock frequencies for the VC707}
+	\label{tab:maxfreq-virtx7}
+    \centering
+  \begin{tabular}{ | l | c | }
+    \hline
+     & $Frequency$  \\
+    \hline
+    $N=64$ & 140.854 MHz \\
+    \hline
+    $N=128$ & 140.854 MHz\\
+    \hline
+    $N=256$ &  ??? MHz \\
+\hline
+  \end{tabular}
+
+\end{table}
+
+\subsubsection{DSP Slices Usage}\label{sec:results_synth_dsp}
+The estimates made in Equation~\ref{eq:numdsp_network-opt} were proven to be accurate, as Figure~\ref{fig:dspused} confirms.
+The reference network design with $N=8$ uses 56 DSP slices, which corresponds to $25.45\%$ of the total number of DSP slices available.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/dspuse.eps}
+    \caption{The number of DSP slices used for several Network Sizes $N$ and resource sharing level $K_G$}
+    \label{fig:dspused}
+\end{figure}
+The DSP usage for the VC707 was also accurate.
+
+\subsubsection{Power Consumption}\label{sec:res-synth-power}
+Another important metric of the performance of a design is its \textbf{power consumption}. All  XC7Z020 designs yielded a
+constant baseline value for the \emph{static} power consumption of around 120 mW, and the power consumption reported
+in Figure~\ref{fig:power} refers to the \emph{total} power consumptions, i.e. both the baseline static power and the dynamic power consumption.
+It is clear that the smaller the network is, the less power is consumed, as one would expect. Furthermore, an increasing
+level of resource sharing yields a substantially lower power consumption figure: this is because less DSP slices are used as $K_G$ increases.
+Of course, even though the power consumption is lower in that case, that comes at the expense of a lower clock frequency and more clock cycles
+elapsed per forward propagation.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/power.eps}
+    \caption{Power Consumption estimates for several Network Sizes $N$ and resource sharing level $K_G$}
+    \label{fig:power}
+\end{figure}
+
+For the synthesized networks on the VC707, the power consumption figures are reported in Table~\ref{tab:power-virtx7}. Being a high-end
+system, and since the network size is larger, the power consumption is higher than before, but still within reasonable levels given
+the complexity of the system.
+
+\begin{table}
+	\caption{Power Consumption estimates for several Network Sizes $N$ in the VC707}
+	\label{tab:power-virtx7}
+    \centering
+  \begin{tabular}{ | l | c | }
+    \hline
+     & $Power$  \\
+    \hline
+    $N=64$ & 1.34 W \\
+    \hline
+    $N=128$ & 1.509 W\\
+    \hline
+    $N=256$ & ??? W \\
+\hline
+  \end{tabular}
+
+\end{table}
+
+\subsubsection{Other Resources Usage}\label{sec:res-synth-otheres}
+The LUT, LUTRAM and Flip-Flop usages are discussed here. In Table~\ref{tab:lut}, the usage of LUTs is reported, where we can see
+that although there is not a clear trend on how the LUT usage varies with increasing $K_G$, it is clear, and expectable, that
+the LUT usage increases with the size of the network by an approximate factor of 2, from $N=2$ to $N=16$. As for $N=32$, the
+usage does not follow this apparent trend, and rises sharply to 91\%. For $K_G = 8$ the usage surpasses the maximum amount
+of LUTs available in the XC7Z020.
+
+\begin{table}
+	\caption{LUT usage for different $N$ and $K_G$ in the XC7Z020}
+	\label{tab:lut}
+    \centering
+  \begin{tabular}{ | l | c | c | c |}
+    \hline
+    & $K_G=2$  & $K_G=4$ & $K_G=8$ \\
+    \hline
+    $N=4$ & 6.87\% & 6.04\% & N.A. \\
+    \hline
+    $N=8$ & 14.64\% & 13.03\% & 14.11\% \\
+    \hline
+    $N=16$ & 28.97\% & 27.72\% & 29.85\% \\
+    \hline
+    $N=32$ & N.A. & 91.09\% & N.D. \\
+\hline
+  \end{tabular}
+
+\end{table}
+
+As for LUTs, the FF usage also scales accordingly to a $2\times$ factor. In terms of LUTRAM, used to store the network weights,
+the amount used increases by 2 with increasing values of $N$, as before, and \textbf{does not} depend on $K_G$. This is because
+the amount of weights only depends on the network sizes $M$ and $N$, and not on $K_G$. Furthermore, unlike LUTs, it scales well
+with increasing network sizes, and does not pose a limitation on the network size. The usage results for the XC7Z020 are reported in Table~\ref{tab:ff},
+and in Table~\ref{tab:ff-virtx7} the usages for the VC707.
+
+\begin{table}
+	\caption{Flip-flop and LUTRAM usage for the XC7Z020}
+	\label{tab:ff}
+    \centering
+  \begin{tabular}{ | l | c | c |}
+    \hline
+         & LUTRAM  & FF  \\
+    \hline
+    $N=4$ & 0.18\%  & 3.39\% \\
+    \hline
+    $N=8$ & 4.41\%  & 6.7\% \\
+    \hline
+    $N=16$ & 8.83\%  & 13.36\% \\
+    \hline
+    $N=32$ & 17.66\% &  26.45\% \\
+\hline
+  \end{tabular}
+
+\end{table}
+
+\begin{table}
+	\caption{Flip-flop, LUTRAM and LUT usage for the VC707}
+	\label{tab:ff-virtx7}
+    \centering
+  \begin{tabular}{ | l | c | c | c |}
+    \hline
+         & LUTRAM  & FF & LUT \\
+    \hline
+    $N=64$ & 24.22\%  & 9.14\% & 24.22\% \\
+    \hline
+    $N=128$ & 14.09\%  & 18.3\% & 41.82\%\\
+    \hline
+    $N=256$ & 8.83\%  & 13.36\% & ??.?? \%\\
+\hline
+  \end{tabular}
+
+\end{table}
 
 \subsection{Performance}
+To evaluate the performance, a metric was defined based on how many predictions it can produce per second (i.e. produce a new result bit in the output sequence),
+in \textbf{millions}. The metric on~\cite{Chang15}, as discussed in Section~\ref{sec:relwork}, is pretty much pointless, since
+although a system can perform many calculations per second, if those operations are redundant, that metric has no relevant information
+regarding \textbf{how fast} the system can \textbf{perform the actual task it was meant to do}, which in this case is a complete Forward
+Propagation through the network. The prediction time is the time elapsed since a new input vector is applied \emph{to} the moment the LSTM
+Network produces an output vector. Hence, we multiply the number of clock cycles yielded by Equation~\ref{eq:numcc_network-opt} by the equivalent clock period
+from the synthesis clock report of Figure~\ref{fig:maxfreq}.
+This result is epitomized in Table~\ref{tab:process-time}, where the calculation time of the Python module's Forward Propagation function is
+also included. This time was measured using the \textbf{timeit}\footnote{https://docs.python.org/3.5/library/timeit.html}  module, that allows
+the evaluation of the execution time of small pieces of code as well as complete functions with arguments. The Python code was run on a Linux System,
+powered by an i7-3770k Intel Processor, running at 4.2GHz.
+
+\begin{table}
+	\caption{Total processing time for a single forward propagation on the XC7Z020}
+	\label{tab:process-time}
+    \centering
+    \begin{tabular}{ | l | c | c | c | c | c | }
+    \hline
+    & $K_G=8$  & $K_G=4$ & $K_G=2$ & Python & Speed-up \\
+    \hline
+    $N=4$ & N.A.  & 309.68 ns  & 259.12 ns & 65 $\mu$s & $\times251$ \\
+    \hline
+    $N=8$ & 793.46 ns  & 421.12 ns  &  317.52 ns & 72 $\mu$s & $\times228$ \\
+    \hline
+    $N=16$ & 1.497 $\mu$s  & 738.19 ns  & 461.336 ns & 96 $\mu$s & $\times208$ \\
+    \hline
+    $N=32$ & N.D.          & 1.586 $\mu$s & N.A.       & 185 $\mu$s  &  $\times117$ \\
+\hline
+  \end{tabular}
+\end{table}
+The performance increase is impressive, even for the slowest of the designs (the $N=32$ network). The hardware network is, at best, $\times251$ faster than the software counterpart,
+and at worst $\times117$ faster. Also, it is noticeable that increasing the level of resource sharing increases the computation time, since the level of parallelism is lower.
+
+Since the previous values are the time needed for a \emph{single} forward propagation, to know how many forward propagations we can perform \emph{per second},
+we only need to invert the previous values. These values are presented in Figure~\ref{fig:Mclass-psec}. While the $N=8$ and $K_G=2$ network is able to
+perform around 3.15 million predictions per second, the Python model can only output around 14 thousand predictions, which is a very significant result that
+proves the relevance of this implementation.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/Mclass-psec.eps}
+    \caption[Millions of classifications per second of each design according to the network size $N$]{Millions of classifications per second of each design according to the network size $N$.
+The comparison is between the software Python model and 3 networks of different levels of resource sharing $K_G$.}
+    \label{fig:Mclass-psec}
+\end{figure}
+
+As for the larger-sized networks synthesized in the VC707, the results are also very promising. For network sizes of $N=64$ and $N=128$, a complete
+forward propagation takes 1.14 $\mu$s and 2.052 $\mu$s respectively, and for both the maximum clock frequency achievable was 140.845 MHz. Since the
+design in~\cite{Chang15}, for $N=128$, takes an estimated 29.13 $\mu$s (see Section~\ref{sec:relwork}), our design yields an improvement of $14\times$
+over it.
 
 \section{Conclusion}\label{sec:concl}
+The LSTM Hardware architecture presented surpassed the performance of current software implementations
+and also the only current hardware implementation by $14\times$, and making use solely of internal FPGA
+resources, unlike the other hardware implementation available, making it a general architecture that can
+be used in any FPGA, and not only limited to a particular one. Furthermore, the design proved to scale relatively
+well with the network size, being able to accommodate networks of sizes up to 256.
+
+Given these results, this architecure places itself as the current state of the art system in LSTM Neural Networks
+implementations in software.
 

### 2016-07-14 finalText/figures/maxfreqplot.py

@@ -5,13 +5,13 @@ from matplotlib import rc
 rc('text', usetex=True);
 rc('font', family='serif', serif='Times');
 
-n4 = np.array([140.154, 136.054])
+n4 = np.array([158.228, 158.228])
 x_n4 = np.array([1,2])
-n8 = np.array([132.89, 131.579, 101.01])
+n8 = np.array([154.321, 154.321, 122.249])
 x_n8 = np.array([1,2,3])
-n16 = np.array([125.471, 105.263, 93.897])
+n16 = np.array([140.895, 131.406, 107.527])
 x_n16 = np.array([1,2,3])
-n32 = np.array([90.498])
+n32 = np.array([101.523])
 x_n32 = np.array([2])
 
 fig = plt.figure()

### 2016-07-14 finalText/figures/powerplot.py

@@ -5,13 +5,13 @@ from matplotlib import rc
 rc('text', usetex=True);
 rc('font', family='serif', serif='Times');
 
-n4 = np.array([176, 149])
+n4 = np.array([178, 153])
 x_n4 = np.array([1,2])
-n8 = np.array([189, 174, 161])
+n8 = np.array([203, 185, 170])
 x_n8 = np.array([1,2,3])
-n16 = np.array([386, 296, 200])
+n16 = np.array([421, 342, 212])
 x_n16 = np.array([1,2,3])
-n32 = np.array([482])
+n32 = np.array([526])
 x_n32 = np.array([2])
 
 fig = plt.figure()

### 2016-07-14 paper/figures/maxfreqplot.py

@@ -5,13 +5,16 @@ from matplotlib import rc
 rc('text', usetex=True);
 rc('font', family='serif', serif='Times');
 
-n4 = np.array([140.154, 136.054])
+n4 = np.array([158.228, 158.228])
 x_n4 = np.array([1,2])
-n8 = np.array([132.89, 131.579, 101.01])
+
+n8 = np.array([154.321, 154.321, 122.249])
 x_n8 = np.array([1,2,3])
-n16 = np.array([125.471, 105.263, 93.897])
+
+n16 = np.array([140.895, 131.406, 107.527])
 x_n16 = np.array([1,2,3])
-n32 = np.array([90.498])
+
+n32 = np.array([101.523])
 x_n32 = np.array([2])
 
 fig = plt.figure()
@@ -21,7 +24,7 @@ plt.plot(x_n8, n8, 'ro-', label="$N=8$")
 plt.plot(x_n16, n16, 'bo-', label="$N=16$")
 plt.plot(x_n32, n32, 'go-', label="$N=32$")
 ax.set_xticks([0,1,2,3,4])
-plt.axis([0.8,3.2,85,145])
+plt.axis([0.8,3.2,90,160])
 plt.grid(True);
 plt.xlabel(r"$\log_2(K_G$)");
 plt.ylabel(r"Clock Speed in MHz");

### 2016-07-14 paper/figures/mprop.py

@@ -7,11 +7,14 @@ rc('font', family='serif', serif='Times');
 
 py = 1/np.array([65e-6, 72e-6, 96e-6, 185e-6])/(1e6)
 x_py = np.array([2, 3, 4, 5])
-k2 = 1/np.array([292.535e-9, 368.725e-9, 518.05e-9])/(1e6)
+
+k2 = 1/np.array([259.12e-9, 317.52e-9, 461.336e-9])/(1e6)
 x_k2 = np.array([2,3,4])
-k4 = 1/np.array([360.15e-9, 494e-9, 921.5e-9, 1.78e-6])/(1e6)
+
+k4 = 1/np.array([309.68e-9, 421.12e-9, 738.17e-9, 1.5858e-6])/(1e6)
 x_k4 = np.array([2,3,4,5])
-k8 = 1/np.array([960.3e-9, 1.715e-6])/(1e6)
+
+k8 = 1/np.array([793.46e-9, 1.4973e-6])/(1e6)
 x_k8 = np.array([3,4])
 
 fig = plt.figure()
@@ -21,7 +24,7 @@ plt.plot(x_k2, k2, 'ro-', label="Hardware $K_G = 2$")
 plt.plot(x_k4, k4, 'go-', label="Hardware $K_G = 4$")
 plt.plot(x_k8, k8, 'bo-', label="Hardware $K_G = 8$")
 ax.set_xticks([2,3,4,5])
-plt.axis([1.8,5.2,-0.2,3.5])
+plt.axis([1.8,5.2,-0.2,4.5])
 plt.grid(True);
 plt.xlabel(r"$\log_2(N)$");
 plt.ylabel(r"Million Classifications per Second");

### 2016-07-14 verilog/network/Synth/Synth.runs/impl_10/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/impl_10'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log top_network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source top_network.tcl -notrace
+
+

### 2016-07-14 verilog/network/Synth/Synth.runs/synth_8/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth/Synth.runs/synth_8'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log top_network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source top_network.tcl

### 2016-07-14 verilog/network/Synth_virtex/Synth_virtex.runs/impl_1/rundef.js

@@ -1,0 +1,44 @@
+//
+// Vivado(TM)
+// rundef.js: a Vivado-generated Runs Script for WSH 5.1/5.6
+// Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+//
+
+echo "This script was generated under a different operating system."
+echo "Please update the PATH variable below, before executing this script"
+exit
+
+var WshShell = new ActiveXObject( "WScript.Shell" );
+var ProcEnv = WshShell.Environment( "Process" );
+var PathVal = ProcEnv("PATH");
+if ( PathVal.length == 0 ) {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;";
+} else {
+  PathVal = "/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64;/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64;/opt/Xilinx/Vivado/2015.4/bin;" + PathVal;
+}
+
+ProcEnv("PATH") = PathVal;
+
+var RDScrFP = WScript.ScriptFullName;
+var RDScrN = WScript.ScriptName;
+var RDScrDir = RDScrFP.substr( 0, RDScrFP.length - RDScrN.length - 1 );
+var ISEJScriptLib = RDScrDir + "/ISEWrap.js";
+eval( EAInclude(ISEJScriptLib) );
+
+
+// pre-commands:
+ISETouchFile( "init_design", "begin" );
+ISEStep( "vivado",
+         "-log network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source network.tcl -notrace" );
+
+
+
+
+
+function EAInclude( EAInclFilename ) {
+  var EAFso = new ActiveXObject( "Scripting.FileSystemObject" );
+  var EAInclFile = EAFso.OpenTextFile( EAInclFilename );
+  var EAIFContents = EAInclFile.ReadAll();
+  EAInclFile.Close();
+  return EAIFContents;
+}

### 2016-07-14 verilog/network/Synth_virtex/Synth_virtex.runs/impl_1/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth_virtex/Synth_virtex.runs/impl_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source network.tcl -notrace
+
+

### 2016-07-14 verilog/network/Synth_virtex/Synth_virtex.runs/impl_2/runme.sh

@@ -1,0 +1,43 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth_virtex/Synth_virtex.runs/impl_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+# pre-commands:
+/bin/touch .init_design.begin.rst
+EAStep vivado -log network.vdi -applog -m64 -messageDb vivado.pb -mode batch -source network.tcl -notrace
+
+

### 2016-07-14 verilog/network/Synth_virtex/Synth_virtex.runs/synth_1/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth_virtex/Synth_virtex.runs/synth_1'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source network.tcl

### 2016-07-14 verilog/network/Synth_virtex/Synth_virtex.runs/synth_2/runme.sh

@@ -1,0 +1,39 @@
+#!/bin/sh
+
+# 
+# Vivado(TM)
+# runme.sh: a Vivado-generated Runs Script for UNIX
+# Copyright 1986-2015 Xilinx, Inc. All Rights Reserved.
+# 
+
+if [ -z "$PATH" ]; then
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin
+else
+  PATH=/opt/Xilinx/SDK/2015.4/bin:/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/bin/lin64:/opt/Xilinx/Vivado/2015.4/bin:$PATH
+fi
+export PATH
+
+if [ -z "$LD_LIBRARY_PATH" ]; then
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64
+else
+  LD_LIBRARY_PATH=/opt/Xilinx/Vivado/2015.4/ids_lite/ISE/lib/lin64:$LD_LIBRARY_PATH
+fi
+export LD_LIBRARY_PATH
+
+HD_PWD='/home/josefonseca/Documents/thesis/verilog/network/Synth_virtex/Synth_virtex.runs/synth_2'
+cd "$HD_PWD"
+
+HD_LOG=runme.log
+/bin/touch $HD_LOG
+
+ISEStep="./ISEWrap.sh"
+EAStep()
+{
+     $ISEStep $HD_LOG "$@" >> $HD_LOG 2>&1
+     if [ $? -ne 0 ]
+     then
+         exit
+     fi
+}
+
+EAStep vivado -log network.vds -m64 -mode batch -messageDb vivado.pb -notrace -source network.tcl

### 2016-07-25 finalText/abstract.tex
Final Version of the Text and Short Paper

@@
 
-Esta dissertação tem como objetivo a implementação em hardware, numa plataforma
-FPGA, de uma Rede Neuronal \textit{Long Short-Term Memory} (LSTM).
+Esta dissertação tem como objetivo a implementação em FPGA de uma Rede Neuronal \textit{Long Short-Term Memory} (LSTM).
 
-As redes neuronais são uma das técnicas mais usadas na área da Aprendizagem Computacional Profunda para tarefas em
-que a programação explícita de \textit{software}, para além de ser impraticavelmente complexa, é necessário que haja uma qualidade
-de tomada de decisão semelhante ou superior à inteligência Humana. Esta rede em particular, a LSTM, é uma rede recursiva, na medida em que a saída
-da cada neurónio, num instante de tempo, \emph{também} serve de entrada no instante seguinte, e assim
-os elementos de memória nela presentes conseguem encontrar padrões também em sequências de dados,
+As redes neuronais são uma das técnicas mais usadas na área da Aprendizagem Computacional Profunda.
+Esta rede em particular, a LSTM, é uma rede recursiva, na medida em que a saída
+da cada neurónio, num instante de tempo, \emph{também} serve de entrada no instante de tempo seguinte, e
+os seus elementos de memória conseguem recordar padrões em sequências de dados,
 apresentando vantagens claras, neste campo, face a redes neuronais convencionais.
 
-As aplicações desta rede são inúmeras, como se poderá atestar no capítulo
+As aplicações das Redes Neuronais são inúmeras, como se poderá atestar no capítulo
 que faz o levantamento do estado-da-arte. Embora a implementação em \textit{software}
 seja a solução comum para todas estas arquitecturas, as implementações em \textit{hardware}
-são ainda poucas e os benefícios do paralelismo inerente a uma plataforma de \textit{hardware} dedicado (no caso deste trabalho, uma FPGA) não
-são aproveitados. É nesse enquadramento que este trabalho se posiciona, apresentando uma implementação inédita de LSTM em FPGA, fazendo apenas uso de 
-recursos internos à mesma, que tem uma melhoria de rapidez de processamento de cerca de 195 vezes face 
-a uma implementação \textit{software}, assim como a proposta de um circuito que permita fazer o 
-treino \textit{on-chip} da rede, usando um método de perturbações estocásticas simultâneas (SPSA).
+são ainda poucas e os benefícios do paralelismo inerente a uma plataforma de \textit{hardware} dedicado não
+são aproveitados. É nesse enquadramento que este trabalho se posiciona, apresentando uma implementação inédita de LSTM em FPGA,
+fazendo apenas uso de recursos internos à mesma, apresentando um \textit{speed-up} de 251 vezes face
+a uma implementação \textit{software} --- executada num Computador de alta \textit{performance} --- assim como a
+proposta de um circuito auxiliar que permita fazer o treino \textit{on-chip} da rede, usando um método de perturbações
+estocásticas simultâneas (SPSA).
 
@@
 
-Neural Networks are one of the most commonly used techniques in Deep Learning for tasks
-where explicit programming of software, besides impractically complex, it is necessary that
-the quality of the decision making is similar, or even better, than that of Human intelligence.
-This particular type of network, LSTM, is a recursive network, given that the neuron outputs in a 
-given time is \emph{also} fed as the input in the next time instant, and that way their memory elements can
+Neural Networks are one of the most commonly used techniques in Deep Learning.
+This particular type of network, as LSTM Network, is a recursive network, given that the neuron outputs in a
+given time are \emph{also} fed as the input in the next time step and, that way, their memory elements can
 make sense of patterns within data sequences, unlike conventional neural networks.
@@
 in software, and their practical applications are plentiful, as one can attest
-in the State of the Art chapter. However, the benefits of the inherent parallelism offered by a dedicated hardware platform
-(in this work, an FPGA) are not availed, and there are relatively few the implementation of Machine Learning algorithms in
-these kind of platforms.
-It is within this grounding that this work positions itself, implementing an LSTM Network in FPGA, using only its internal resources, and achieving an $\times195$ increase in processing time when compared to a software implementation, and also proposes a circuit that will allow on-chip training for this network, using simultaneous stochastic perturbations (SPSA).
+in the State of the Art chapter. However, there are only few FPGA implementation of these algorithms, and the benefits of the inherent
+parallelism offered by a dedicated hardware platform are not availed. This work, then, implements an LSTM Network in FPGA, using only
+its internal resources, and achieving a $251 \times$ \textit{speed-up} in processing time when compared to a software implementation,
+running on a high-end Desktop. A circuit that will allow on-chip training for this network, using simultaneous stochastic perturbations (SPSA),
+is also proposed.

### 2016-07-25 finalText/acknows.tex
Final Version of the Text and Short Paper

@@
 
-Agradeço a todos os professores que se cruzaram comigo na minha vida e que, através da sua ação
+Agradeço a todos os professores que se cruzaram comigo na minha vida e que, através da sua ação,
 me transmitiram os conhecimentos necessários para me formar como Engenheiro e produzir este trabalho,
@@
 ao meu orientador deste trabalho, o Prof. João Canas Ferreira, que desde o primeiro momento acolheu
-esta proposta de trabalho e soube ver nela o seu potencial devido, e ao co-orientador, o Ivo Timóteo
+esta proposta de trabalho e soube ver nela o seu potencial devido, e ao co-orientador, o Ivo Timóteo,
 que mais do que minha referência para a parte teórica deste trabalho, é um bom amigo.
@@
 
-Uma última palavra de apreço para as pessoas especiais da minha vida, os meus pais, que me formaram como Pessoa e me
-incutiram força em todos os momentos da vida, do qual esta tese não é excepção, e à minha namorada, a Mafalda, que sempre 
-se entusiasmou e interessou por todos os pequenos progressos que fui atingindo ao longo destes meses.
+Uma última palavra de apreço para as pessoas especiais da minha vida. Os meus pais, que me formaram como Pessoa e me
+incutiram força em todos os momentos da vida, do qual esta tese não é excepção, e à minha namorada, a Mafalda, que apesar
+de não ser da minha área, sempre se entusiasmou e interessou por todos os progressos que fui atingindo ao longo destes meses.
 

### 2016-07-25 finalText/conclusion.tex
Final Version of the Text and Short Paper

\chapter{Conclusion}\label{chap:conclusion}
The LSTM Hardware architecture presented surpassed the performance of the custom-built software implementation by $251\times$, at best,
and also the only current hardware implementation by $14\times$, and solely making use of internal FPGA
resources, achieving a higher level of parallelism. The higher levels of parallelism of this work are achieved at the cost of increasing
design complexity, which undermines its scalability for higher sized networks, unlike the implementation of Chang et al.~\cite{Chang15}.
On the other hand, the HDL description of this work is parameterized, and is thus very flexible for networks of any size, not requiring
a redesign of the system every time a different sized network is required. Furthermore, making use of internal memory makes it suitable
for including an on-chip learning system that can perform training on the network weights.

Given these results, this architecture places itself as the current state of the art system in LSTM Neural Networks
implementations in hardware, providing the most efficient implementation to date.

A possible follow-up of the work of this thesis, is the integration of the training circuit in Section~\ref{sec:train-circ}, providing
an LSTM Network with on-chip training capabilities.

### 2016-07-25 finalText/intro.tex
Final Version of the Text and Short Paper

@@
 In this work, a hardware implementation of an LSTM Neural Network is presented, along with a proposal for a digital circuit that performs on-chip training.
-In the reference example, the proposed hardware network was 195 times
-faster than the reference software implementation. The training convergence, although simulated, is not conclusive.
+In the reference example, the proposed hardware network was 251 times faster than the reference software implementation running in a high-end Desktop.
+The training convergence, although simulated, was not conclusive.
 
 \section{Background}\label{sec:intro_back}
-Artificial Neural Networks are one of the most popular models in the field of Machine Learning.
-As the name suggests, their operation is inspired by the operation of the building blocks of our brain, the \textbf{neurons}.
-In spite of its high performance, one of their shortcomings is the fact that they cannot retain temporal dependences among incoming data samples, thus not being suitable to process time-series data, such as audio, video or other kinds of time-varying data streams, where current inputs have a high temporal dependence with previous and future inputs.
+Artificial Neural Networks are one of the most popular algorithms in the field of Machine Learning. As the name suggests, their operation is inspired by
+the operation of the building blocks of our brain, the \textbf{neurons}. In spite of its high performance, one of their shortcomings is the fact that
+they cannot retain temporal dependences in a stream of data samples, thus not being suitable to process time-series data, such as audio, video or other
+kinds of time-varying data streams, where inputs at a given time step have a high temporal dependence with previous and future inputs.
 
-To address this issue, several algorithms have been used, such as Hidden Markov Models (HMM) or Recurrent Neural Networks (RNN), but both of these methods fail to recall temporal dependences that extend over an large period of time, for the reasons that we will understand in Sections~\ref{sec:theorBack_rnn} and~\ref{sec:theorBack_lstm}.Thus, in 1997 Hochreiter et al.
-proposed a novel RNN structure~\cite{Hoch97}, the Long Short-Term Network (LSTM), where a memory cell was introduced, and the input/output/read/write access is controlled by individual gates that are activated both by the incoming data samples, but also by the outputs from the previous time-step (it is an RNN after all).
-They are one of the state of the art methods in Deep Learning nowadays, as we can attest in Section~\ref{sec:sa_apps}.
+To address this issue, several algorithms have been used, such as Hidden Markov Models (HMM) or Recurrent Neural Networks (RNN), but both of these methods fail
+to recall temporal dependences that extend over a large period of time, for the reasons that we will understand in Sections~\ref{sec:theorBack_rnn} and~\ref{sec:theorBack_lstm}.
+In 1997 Hochreiter et al. proposed a novel RNN structure~\cite{Hoch97}, the Long Short-Term Network (LSTM), where a memory cell was introduced, and
+the input/output/read/write access is controlled by individual gates that are activated both by the incoming data samples, but also by the outputs
+from the previous time-step (it is an RNN after all). They are one of the state of the art methods in Deep Learning nowadays, as we can attest in Section~\ref{sec:sa_apps}.
 
 \section{Motivation}\label{sec:intro_mot}
-Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be sped up by a dedicated hardware structure.
+Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based. But the parallel nature of the
+structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that has only recently been done
+once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve
+software solution, it makes use of external memory to store the network parameters, which undermines the network performance. Furthermore, the
+design is not flexible to accommodate networks of different sizes, and lacks the ability to perform on-chip learning.
 
-All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to now generally have between 2-8 cores (mobile devices and general use personal computers),
+All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real
+parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to now generally have between 2-8 cores
+(mobile devices and general use personal computers),
 
-In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the porting process is not entirely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details.
-Therefore, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance.
-Furthermore, when processing an incoming stream of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.
+In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a
+Field-Programmable Gate Array (FPGA), but the porting process is not entirely automatic and to have the least performance drop possible, it has
+to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and
+low-level details. Therefore, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of
+networks on demand, for a particular task that requires them, achieving a lower computation time, and unburdening the CPU from running it,
+thus saving performance for running other tasks related with the Operating System, for instance. Furthermore, when processing an incoming stream
+of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.
 
 \section{Objectives}\label{sec:intro_obj}
-Taking into account the considerations done in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
-Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate some of the applications portrayed in~\ref{sec:sa_apps}.
-Lastly, I will work to make this structure reconfigurable on the go, and trying to minimize the reconfiguration time.
+Taking into account the considerations done in~\ref{sec:intro_mot}, a hardware implementation of an
+LSTM Network is proposed, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}.
+Moreover, the HDL description of the network is parameterizable, meaning that to synthesize a network for a given size, we only need to
+change the appropriate parameter at synthesis time, therefore requiring no redesign effort whatsoever, unlike~\cite{Chang15}.
 
 \section{People Involved}\label{sec:intro_people}
-Besides me, the candidate to the Master's Degree, there are two more people involved, namely
+Besides myself, the candidate to the Master's Degree, there are two more people involved, namely
 
@@
 \section{Overview of the Document}\label{sec:intro_overview}
-Chapter~\ref{chap:theorBack}, the theoretical foundations will be layed out.
-Section~\ref{sec:tb_basicCon} presents the basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs.
-Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final solution, which will be outlined in Section~\ref{chap:propSol}.
+In Chapter~\ref{chap:theorBack}, the theoretical foundations will be laid out. Section~\ref{sec:tb_basicCon} presents the
+basic concepts of Machine Learning, and in the following sections, the theoretical details of ANNs, RNNs and LSTMs.
+Section~\ref{sec:training_lstm} provides a quick explanation of the training algorithm that will be used in the final
+solution, which will be outlined in Section~\ref{sec:train-circ}.
 
-In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications~\ref{sec:sa_apps}, their hardware implementations~\ref{sec:sa_hardware} and the current work regarding the chosen training algorithm~\ref{sec:sa_training} are presented.
+In Chapter~\ref{chap:stateArt}, the state of the art of LSTMs, their applications (Section~\ref{sec:sa_apps}), their
+hardware implementations (Section~\ref{sec:sa_hardware}) and the current work regarding the chosen training algorithm (Section~\ref{sec:sa_training})
+are presented.
 
-In Chapter~\ref{chap:propSol} the proposed architecture of the hardware LSTM network is explained in detail, as well as details related to each of its constituent blocks.
+In Chapter~\ref{chap:propSol} the proposed architecture of the hardware LSTM network is explained in detail, as well as details
+related to each of its constituent blocks.
 
 Finally, in Chapter~\ref{chap:results}, I present the results achieved with the hardware network of Chapter~\ref{chap:propSol}.
-
-\textbf{Disclaimer}: since this is a very innovative work, I will not present here any Verilog source code to protect my Intellectual Property.
-However, you can ask permission to have it by \href{mailto:ee11126@fe.up.pt}{emailing me}.

### 2016-07-25 finalText/proposedSol.tex
Paper ready, only missing introduction to architecture and results
Final Version of the Text and Short Paper

@@
 \subsection{Finite-precision representation system}
-Before discussing any other details concerned with the actual hardware implementation, I will layout some of the design choices that were made regarding how the numbers involved in the calculations are represented.
-For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point representation systems, as well as the bitwidth and precision chosen for then number representation system of the network, and Section~\ref{sec:convrulesfp} states how the conversion between real and fixed-point numbers is performed.
-Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic imposes.
-The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.
+Before discussing any other details concerned with the actual hardware implementation, some of the design choices that were
+made regarding how the numbers involved in the calculations are represented are laid out. For that purpose, Section~\ref{sec:precbit} discusses the fundamentals of fixed-point
+representation systems, as well as the bitwidth and precision chosen for the number representation system of the network, and Section~\ref{sec:convrulesfp} states
+how the conversion between real and fixed-point numbers is performed. Finally, Section~\ref{sec:arithrulesfp} explains the special rules that fixed-point arithmetic
+imposes. The considerations made in this chapter can be found on~\cite{Yates13}, which is a good reference for fixed-point arithmetic theory.
 
 \subsubsection{Precision bitwidth}\label{sec:precbit}
-Since we are dealing with real numbers, and I plan to make use of the DSP48 slices within the FPGA, I chose to use an 18-bit signed fixed-point system, with the sign information coded as 2's complement.
-Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit.
-In this way, the value of the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is $-2^{n-1}$.
-In terms of range of representation, the maximum positive number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real number (the decimal point is at the $m$-th bit),
+Since we are dealing with real numbers, and the plan is to make use of the DSP48 slices within the FPGA, an 18-bit signed fixed-point system was chosen, with the sign
+information coded as 2's complement. Fixed-point systems are usually addressed in the $Qn.m$ form, where $n$ is the bitwidth of the of the integer part (excluding
+the sign bit) and $m$ is the bitwidth of the fractional part, and so the total bitwidth is $N=m+n+1$ to account for the sign bit. In this way, the value of
+the $i$-th position bit is $2^{i-m}$, and since this is a 2's complement system, the last bit is worth $-2^{n-1}$. In terms of range of representation, the maximum positive
+number that can be represented corresponds to all bits set to one, except for the last ($2^{n+m+1}-1$), but shifted by $m$ bits to the right to yield the correct real
+number (the decimal point is at the $m$-th bit), i.e.
 
@@
 \begin{equation}\label{eq:minneg}
-    \text{Small. Negative Number} = -\frac{2^{N-1}}{2^m} = -2^{N-1-m}
+    \text{Small. Negative Number} = -\frac{2^{N-1}}{2^m} = -2^{N-1-m}.
 \end{equation}
 
-Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, our fractional part precision should be, at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$.
-On the other hand, given that in the Python experiments I have attested that the weight values generally do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to be large enough to accommodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m=11$.
-Since $18=n+m+1$, then $n = 17 - m = 6$ and the representation system to be used is $Q6.11$.
-According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented with this choice of $n$ and $m$ are in the range
+Assuming that the smallest perturbation used in the training system will be $2^{-9}$, according to the previous Python experiments, the fractional part precision should be,
+at least, greater than this, so a sensible choice would be either $m=10$ or $m=11$. On the other hand, given that, in the Python experiments, it was attested that the
+weight values generally do not (and should not) exceed values around the first power of ten, there is no need for large values of $n$ (although it is advisable to be large
+enough to accommodate the intermediate calculations, avoiding overflow), and so we should choose to have as much precision as possible, so $m$ was set to 11. Since $18=n+m+1$, then
+$n = 17 - m = 6$ and the representation system to be used is $Q6.11$. According to Equations~\ref{eq:maxpos} and~\ref{eq:minneg}, the real values $x$ than can be represented
+with this choice of $n$ and $m$ are in the range
 
@@
 \subsubsection{Conversion between real and fixed-point}\label{sec:convrulesfp}
-In order to find the real number equivalent of a $Qn.m$ fixed-point system, and \textit{vice-versa}, we need to take into account that, according to Section~\ref{sec:precbit}, the $i$-th position bit is worth $2^{i-m}$, and therefore the decimal point in this fixed-point system is at position $i=m$, since $2^{m-m} = 1$.
+In order to find the real number equivalent of a $Qn.m$ fixed-point system, and \textit{vice-versa}, we need to take into account that, according to
+Section~\ref{sec:precbit}, the $i$-th position bit is worth $2^{i-m}$, and therefore the decimal point in this fixed-point system is at position $i=m$, since $2^{m-m} = 1$.
 This way, the rules are as follows
@@
 \subsubsection{Fixed-point arithmetic rules}\label{sec:arithrulesfp}
-The three main operations needed in my network design are \textbf{signed sums}, \textbf{signed multiplications} and \textbf{arithmetic shifts} (i.e.
-the ones that preserve the sign of the MSB) to implement divisions/multiplications by powers of two.
-In terms of signed sums, the rule is simple: both numbers need to be scaled to the same base, with their $m$'s being same, so that the decimal point is in the same place in both numbers ($n$, however, can be different, since that only means that one number is longer than the other, and the missing bits in the smaller one can be interpreted as zeros).
+The three main operations needed in my network design are \textbf{signed sums}, \textbf{signed multiplications} and \textbf{arithmetic shifts} (i.e. the ones that preserve the sign of the MSB)
+to implement divisions/multiplications by powers of two. In terms of signed sums, the rule is simple: both numbers need to be scaled to the same base, with their $m$'s being same,
+so that the decimal point is in the same place in both numbers ($n$, however, can be different, since that only means that one number is longer than the other, and the missing bits
+in the smaller one are interpreted as zeros).
 
-For signed multiplication, since both operands are in fixed-point $Qn.m$, and are thus scaled by $2^m$, we need to scale the result by multiplying it by $2^{-m}$ (or perform an arithmetic right shift of $m$ bits).
-This is because, if $a$ and $b$ the real numbers to be multiplied, we have that in fixed-point arithmetic, the result $c$ is
+For signed multiplication, since both operands are in fixed-point $Qn.m$, and are thus scaled by $2^m$, we need to scale the result by multiplying it
+by $2^{-m}$ (or perform an arithmetic right shift of $m$ bits). This is because, if $a$ and $b$ are the real numbers to be multiplied, and $c$ is the
+result, we have that, in fixed-point arithmetic, the result is
 
@@
 \end{equation}
-Since in $Qn.m$, all numbers are represented as $r 2^{m}$ with respect to their real counterpart $r$, we need to scale back the decimal point in the result of Equation~\ref{eq:multfp}.
-We can see that it can be easily achieved by dividing by $2^m$, as stated in the previous paragraph.
+Since in $Qn.m$, all numbers are represented as $r 2^{m}$ with respect to their real counterpart $r$, we need to scale back the decimal point in the result of Equation~\ref{eq:multfp}. We can see
+that it can be easily achieved by dividing by $2^m$, as stated in the previous paragraph.
 
-\subsection{Non-linearity Calculator}\label{sec:nonlincalc}
-In order to evaluate the non-linear activation functions $\sigma(x)$ and $\tanh(x)$, since there is no algorithm that can directly compute them, I had to find a suitable way to compute them accurately using a finite number of elementary operations, multiplications and additions, that can be performed efficiently by specially tailored blocks within the FPGA (DSP48 slices for multiplication, for instance).
-For that purpose, after using~\cite{Muller05} as a reference on elementary function approximation algorithms, I decided to use \textbf{Polynomial Approximations}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).
+\subsection{Activation Function Calculator}\label{sec:nonlincalc}
+In order to evaluate the non-linear activation functions $\sigma(x)$ and $\tanh(x)$, since there is no algorithm that can directly compute them,
+a suitable way to compute them accurately had to be found, using a finite number of elementary operations, multiplications and additions, that can be performed
+efficiently by specially tailored blocks within the FPGA (DSP48 slices for multiplication, for instance). For that purpose, after using~\cite{Muller05} as a
+reference on elementary function approximation algorithms, \textbf{Polynomial Approximations} were used, since evaluating a polynomial does not have high memory usage
+needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction
+both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).
 
 \subsubsection{Theoretical considerations on the approximation method}
-The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of degree $n$, in order to meet the optimization criteria chosen \textit{a priori}.
-This optimization criteria can be either the well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation.
-Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error does not exceed the precision limit of this representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded.
-Weierstrass's Theorem~\cite{Weierstrass1885}, from 1885, guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$.
-Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynomial approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in at least $n+2$ points, and the sign of approximation error alternates from one interval to the other, and thus the error is not \emph{biased}.
-This leads to a linear system of $n+2$ equations whose $i$-th line is given by
+The polynomial approximation methods aim to approximate some function $f(x)$ in an interval $\left[a, b\right]$ using a polynomial $p^*_n \in \mathcal{P}$ of
+degree $n$, in order to meet the optimization criteria chosen \textit{a priori}. This optimization criteria can be either the
+well-known \textbf{Least Squares Approximation} procedure, where we minimize the \emph{average quadratic error} $\left[f(x)-p^*(x)\right]^2$, or
+the \textbf{Least Maximum Approximation}, where we \emph{minimize} the \emph{maximum possible error}, also commonly called a \textit{minimax} approximation.
+Since we are operating in $Q6.11$ fixed-point arithmetic, we need to guarantee that the maximum approximation error is close to the precision limit of this
+representation, $2^{-11}$, and so a \textit{minimax} approach is desirable, since it guarantees that a given maximum error is not exceeded. Weierstrass's
+Theorem guarantees that there is always a polynomial that can approximate any continuous function f with error $\epsilon > 0$.
+Chebyshev also proved~\cite{Muller05} that, in a \textit{minimax} polynomial approximation of degree $n$, the minimum approximation error $\epsilon$ is achieved in
+at least $n+2$ points, and the sign of the approximation error alternates from one interval to the other, meaning that the error is not \emph{biased} and has zero average.
+This leads to a linear system of $n+2$ equations, whose $i$-th line is given by
 
@@
 
-On one hand, the system of~\ref{eq:remezline} of Step 2 can be written in matricial notation as
+On one hand, the system of~\ref{eq:remezline} of Step 2 can be written in matrix notation as
 
@@
 \end{equation}
-and therefore the solution vector $p = \left[p_0 \, p_1 \, p_2 \, \cdots \, p_n \, \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the matrix on the left-hand side of the equation and $b$ is the vector on the right-hand side.
+and therefore the solution vector $p = \left[p_0 \; p_1 \; p_2 \; \cdots \; p_n \; \epsilon \right]$ for Step 2 of Remez's Algorithm is simply given by $p = A^{-1}b$, where $A$ is the matrix on the left-hand side of the equation and $b$ is the vector on the right-hand side.
 The Python implementation of this algorithm is presented in Listing~\ref{lst:remezpy} of Appendix~\ref{ap1:remez-pyth}.
 
-Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, I chose to partition the domain of the activation functions in \textbf{6 intervals}, and approximate each of those intervals using polynomials of degree $n=2$.
-This proved to yield a lower overall approximation error, as expected (the interval on which to perform the approximation is smaller).
-Also, since both the $\sigma(x)$ and $\tanh(x)$ have horizontal asymptotes in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value equal to the corresponding asymptote.
-The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are
+Instead of using a \emph{single} polynomial of higher degree ($n \geq 3$) for the whole domain, the domain of the activation functions was partitioned
+in \textbf{6 intervals}, and polynomials of degree $n=2$ were used to approximate the function in each of those intervals. This proved to yield a lower overall approximation error,
+as expected (the interval on which to perform the approximation is smaller). Also, since both $\sigma(x)$ and $\tanh(x)$ have horizontal asymptotes
+in $\{0,1\}$ and $\{-1,1\}$ respectively, the far-left and far-right intervals do not need a polynomial approximation, and can be assigned a constant value
+equal to the corresponding asymptote. The resulting \textit{minimax} approximation polynomials yielded by running the code in Listing~\ref{lst:remezpy} are
 
 \begin{equation}\label{eq:coefs_sigm}
-\hat{\sigma(x)} = \left\{ 
+\sigma(x) \approx \left\{
 \begin{array}{lc}
@@
 \begin{equation}\label{eq:coefs_tanh}
-\hat{\tanh(x)} = \left\{ 
+\tanh(x) \approx \left\{
 \begin{array}{lc}
@@
 \subsubsection{Hardware Implementation}
-The hardware module that implements these non-linearities is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$.
+The hardware module that implements these activation functions is, essentially, a 2nd degree polynomial calculator, where the coefficients of the polynomial are chosen accordingly with the value of the input $x$.
 This last functionality
@@
 \end{equation}
-where we can note that this operation can be divided in a two-step procedure of a simultaneous multiplication of the operand by a constant, and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand by this last result and summing $p_0$.
-The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}.
-Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (I used Python3's Numpy as reference) for both activation functions.
+where we can note that this operation can be divided in a two-step procedure of a simultaneous multiplication of the operand by a constant,
+and a subsequent addition of another constant: first, by multiplying the operand by $p_2$ and summing $p_1$, and then by multiplying the operand
+by this last result and summing $p_0$. The block diagram of the hardware implementation of the non-linearity calculator is presented in Figure~\ref{fig:nonlin}.
+Also, in Figure~\ref{fig:nonlin-out}, the output of the Verilog module that implements this design is compared with the actual output (Python3's Numpy was used as reference)
+for both activation functions. The maximum error for the approximation of the $\sigma(x)$ was of 0.001408 and for the $\tanh(x)$ was 0.0121.
 
@@
     \includegraphics[width=0.9\linewidth]{figures/non-lin.eps}
-    \caption[Block Diagram of the Non-linearity Calculator]{Block Diagram of the Non-linearity Calculator using a single multiplication.
+    \caption[Block Diagram of the Activation Function Calculator]{Block Diagram of the Activation Function Calculator using a single multiplication.
 The multiplexer state is controlled by the flip-flip and sum block, that change state every clock cycle.
@@
 \subsection{Matrix-vector Dot Product Unit}\label{sec:dotprod_sec}
-From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively.
-This way, we need a HDL block that implements matrix-vector multiplication, and that block must be parameterized in order to accommodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has length $N$ (the number of neurons in the layer), and so $\mb{R}_*$ now has size $N\times N$.
-Plus, if we need a layer with different parameters, either in terms of number of inputs or number of neurons, it advisable to only change the respective parameter at the synthesis stage instead of having to redesign the whole block for that particular size.
+From the set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are multiplied by the input
+vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, we need an HDL block that implements matrix-vector multiplication,
+and the description of that block must be parameterized in order to accommodate different matrices/vectors of various sizes: note that $x$ has length $M$ (the number
+of inputs to the layer), and so $\mb{W}_*$ has size $N\times M$, while $y$ has length $N$ (the number of neurons in the layer), and so $\mb{R}_*$
+now has size $N\times N$. Plus, a layer with different parameters is needed, either in terms of number of inputs or number of neurons, it is advisable
+to only change the respective parameter at the synthesis stage, instead of having to redesign the whole block for that particular size.
 
-The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}
+The matrix-vector dot product of a matrix $A$ of size $N \times M$ with a vector $\mb{x}$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}
 
@@
 \end{algorithm}
-This operation has a computational complexity of $O(n^2)$.
-It can be seen that each of the $i$-th component of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix.
-If we follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.
+This operation has a computational complexity of $O(n^2)$. It can be seen that each of the $i$-th components of the output vector $\mb{y}$ can be calculated \textbf{in parallel},
+each only requiring the corresponding $i$-th line from the matrix. Following this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which
+is one of the great advantages of custom-tailored hardware solutions.
 
-Although this solution only requires one multiplication per row of the input matrix (i.e.
-$N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions.
-The solution I have found for this issue was to \emph{share} the multiplication slice between rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e.
-the number of columns); now, if we define a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
-Suppose that we have a $8\times2$ matrix, and that $K_G = 2$; in this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
+Although this solution only requires one multiplication per row of the input matrix (i.e. $N$ multiplications), if the row size is large, we may quickly run out of resources in the FPGA
+as $N$ increases; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accommodate networks of larger dimensions.
+The solution found for this issue was to \emph{share} the multiplication slice between rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each
+multiplication slice was responsible for producing the $i$-th element of the output vector $\mb{y}$ (of size $N$), therefore the final result for the vector would be ready in $M$
+clock cycles (i.e. the number of columns); now, if defining a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$, the number of rows that share the
+same multiplier, each multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles. Suppose that we
+have a $8\times2$ matrix, and that $K_G = 2$; in this way, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$
+clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
 
-Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication and units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before.
-Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
-The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to write to the correct output register, respectively.
-In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$.
-After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.
-Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency, we should add a 3 clock cycle overhead to the previous estimate.
+Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication units within the module, respectively,
+where $K_G = 4$ was set, for the same matrix and vector sizes as before. Note that in this situation there would be only 2 multipliers, and therefore the module would be composed of two
+multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel: they address a particular column using the signal \verb+colSel+, which is used by the RAM
+module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
+The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the
+control unit of the module, and essentially operates the left Mux and right Demux of Figure~\ref{fig:array-prod} that allows to choose the proper position of the weight column and to
+write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$.
+After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$:
+since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock
+frequency, we should add a 3 clock cycle overhead to the previous estimate.
 
@@
     \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module.
-The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
+The flip-flop accumulates the sum, and the output Demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
     \label{fig:array-prod}
@@
 \subsection{Dual-port weight RAM}
-Even though I have chosen to keep the bias weight vectors in normal registers and let the synthesis tool decide how to place them in the FPGA, for the weight matrices, the case is different.
-Since the network size can be large, it is good to make sure that they are placed in the RAMs available in the FPGA.
-In Section~\ref{sec:dotprod_sec}, we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column by column access to the weight matrix, and therefore this block outputs a given column in the \verb+rowOut+ output port terminal, at the \emph{negative edge} of the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock.
-As far as \textbf{writing} to the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{negative edge} of the clock, and is placed at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.
+Even though bias weight vectors, with size $N\times 1$, were kept as normal registers and the synthesis tool was left to decide how to place them in the FPGA,
+for the weight matrices the case is different. Since the network size can be large, it is good to make sure that they are placed in the RAMs available
+in the FPGA. In Section~\ref{sec:dotprod_sec} we see that, because of the architecture of the matrix-vector multiplier, it is convenient to have a column
+by column access to the weight matrix, and therefore this block was designed to output a given column in the \verb+rowOut+ output port terminal, at the \emph{positive edge} of
+the clock, selected by the value present in the input \verb+addressOut+ at the immediately previous \emph{positive edge} of the clock. As far as \textbf{writing} to
+the memory is concerned, the process is identical but now the column of weights at the input \verb+rowIn+ is sampled at \emph{positive edge} of the clock, and is placed
+at the column specified by the input \verb+addressIn+ at the previous \emph{positive edge} of the clock.
 
-Note that, for an $N \times M$ matrix, since the memory outputs and writes one column at a time, both the input and output port terminals will carry $N$ weights, and a total bitwidth of $\verb+BITWIDTH+ \cdot N$.
+Note that, for an $N \times M$ matrix, since the memory outputs and writes one column at a time, both the input and output port terminals will carry $N$ weights,
+and a total bitwidth of $\verb+BITWIDTH+ \cdot N$.
 
-The Verilog coding followed the Verilog Coding Guidelines in Xilinx's \href{http://www.xilinx.com/support/documentation/sw_manuals/xilinx2015_2/ug901-vivado-synthesis.pdf}{UG901} that, in page 50, recommends that the \verb+RAM_STYLE+ parameter be set to \verb+block+.
-This can be done by adding the follwing compiler directive before the register definition, as follows
+The Verilog coding followed the Verilog Coding Guidelines in Xilinx's \href{http://www.xilinx.com/support/documentation/sw_manuals/xilinx2015_2/ug901-vivado-synthesis.pdf}{UG901 Document} that,
+in page 50, recommends that the \verb+RAM_STYLE+ parameter be set to \verb+block+. This can be done by adding the following compiler directive before the register definition, as follows
 
@@
 
-After performing synthesis, I have noticed that using block RAM led to a slightly poorer performance, so I have settled with the use of \emph{Distributed RAM}, storing the weight matrices in LUTRAM.
+After performing synthesis, it was noticed that using block RAM led to a slightly poorer speed performance, so I have settled with the use of \emph{Distributed RAM}, storing the weight matrices in LUTRAM.
 
 \subsection{Gate Module}\label{sec:gatemod}
-The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$.
-If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices.
-This way, a Gate module needs to perform three tasks
+The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we note that, according
+to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, they were omitted
+in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks
 
@@
 The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result.
-After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this module takes to output a result is $6+K_G\cdot N$.
+After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this
+module takes to output a result is $6+K_G\cdot N$.
 
@@
 \section{Hardware LSTM Network -- Forward Propagation}\label{sec:hLSTM-fp}
-After detailing the design of each of the constituent modules of the LSTM Network, I will now show how they are interconnected
-to implement a hardware version of Equations~\ref{eq:equationsLSTM}.
-This is performed in Section~\ref{sec:hLSTM-struct}, where the design decisions and compromises that were taken are detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, the naïve solution (~\ref{sec:struct-fullpar}) and an optimized version of the former (~\ref{sec:struct-shared}) that exploits some of the hardware redundancies by multiplexing in time the usage of those redundant structures.
+After detailing the design of each of the constituent modules of the LSTM Network, it will now be shown how they are interconnected
+to implement a hardware version of Equations~\ref{eq:equationsLSTM}. This is performed in Section~\ref{sec:hLSTM-struct}, where the design decisions
+and compromises that were taken are detailed, along with some estimates on the resource usage and calculation time for each of the design iteration versions, namely
+the \textit{naïve solution} (~\ref{sec:struct-fullpar}) and an \textit{optimized version} of the former (~\ref{sec:struct-shared}) that exploits some
+of the hardware redundancies by multiplexing in time the usage of those redundant structures.
 
 \subsection{Network Structure}\label{sec:hLSTM-struct}
-Looking at equations~\ref{eq:equationsLSTM}, we note that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- we see that they can be calculated in parallel, meaning that we need four Gate Modules (see Section~\ref{sec:gatemod} working in parallel, each one with its respective two Block RAMs for $\mb{W}_*$ and $\mb{R}_*$, each followed by the respective activation function calculator (detailed in Section~\ref{sec:nonlincalc}).
-There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
+Looking at Equations~\ref{eq:equationsLSTM}, we note that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other, because
+they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$. This way, they can be calculated in parallel, meaning that we need
+four Gate Modules (see Section~\ref{sec:gatemod} working in parallel, each one with its respective two RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation
+function calculator (detailed in Section~\ref{sec:nonlincalc}), \textbf{one for each} of the $N$ vector elements. There are three elementwise multiplications, two for producing
+signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
 
 \subsubsection{Fully Parallel}\label{sec:struct-fullpar}
-By implementing directly the ideas outlined previously, we get the network of Figure~\ref{fig:network}.
-The memory element of the network is the array of flip-flops that keep the value of the vector $\mb{c}^{(t)}$, which is activated everytime we have a new incoming signal in order to store the last value, which is now $\mb{c}^{(t)}$
+By implementing directly the ideas outlined previously, we get the network of Figure~\ref{fig:network}. The memory element of the network is the array of flip-flops that
+keep the value of the vector $\mb{c}^{(t)}$, which is activated every time we have a new incoming signal in order to store the last value, which is now $\mb{c}^{(t)}$
 
-(Talk about RAM usage)
-
-In terms of DSP slice usage, this proposed design comprises 3 elementwise multiplications and 5 activation function Modules.
-Since the number of elementwise multiplications is equal to the network size $N$, we need $3N$ DSP slices for each elementwise block.
-As far as the activation function module is concerned, since we need to apply it to each one of the elements of the layer, $N$, and each module uses one multiplication, the ensemble of all 5 modules needs $5N$ DSP slices.
-Therefore, and noting that each Gate has $\frac{2N}{K_G}$ DSP slices, the total number of DSP slices used is
+In terms of DSP slice usage, this proposed design comprises 3 elementwise multiplications and 5 activation function Modules. Since the number of elementwise multiplications is
+equal to the network size $N$, we need $3N$ DSP slices for each elementwise block. As far as the activation function module is concerned, since we need to apply it to each one
+of the $N$ elements of the Gate's output vectors, and each module uses one multiplication, the ensemble of all 5 modules needs $5N$ DSP slices. Therefore, and noting that each
+Gate has $\frac{2N}{K_G}$ DSP slices, the total number of DSP slices used is
 
@@
 \end{equation}
-In terms of time performance, we can measure it by estimating the number of clock cycles needed to perform a complete forward propagation.
-Since the gate module outputs a result in $K_G \cdot N$ cycles, the activation function evaluators need always 5, and the elementwise multiplicators need only 2 cycles (one cycle for the pipeline, and another for the actual calculation), and noting that the two elementwise multiplicators that sum to produce $\mb{c}^{(t)}$ can work in parallel, the estimated number of clock cycles needed would be proportional to
+In terms of time performance, we can measure it by estimating the number of clock cycles needed to perform a complete forward propagation. Since the gate module
+outputs a result in $K_G \cdot N + 7$ cycles, the activation function evaluators need always 5, and the elementwise multipliers need only 2 cycles (one cycle
+for the pipeline, and another for the actual calculation), and noting that the two elementwise multipliers that sum to produce $\mb{c}^{(t)}$ can work in parallel,
+the estimated number of clock cycles needed would be proportional to
 
@@
 \subsubsection{Shared Elementwise Multiplication Block}\label{sec:struct-shared}
-In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplier preceeding the flip-flops, all of them follow a similar structure: one of the operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$.
-An improvement over the last network architecture is to instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
-Besides, the right elementwise multiplier of Figure~\ref{fig:network} is not used as the same time as any other, so it is a perfect waste of resources.
-The issue about the elementwise multiplier that precedes the flip-flops can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$.
-These ideas resulted in the improved network design of Figure~\ref{fig:network-opt}.
+In Figure~\ref{fig:network}, we see that, apart from the elementwise multiplier preceding the register, all of them follow a similar structure: one of the
+operands is the output of a $\tanh(\mb{x})$ block and the other from a $\sigma(\mb{x})$. An improvement over the last network architecture is to, instead of
+replicating these `$\tanh$-$\sigma$-$(\cdot)$wise' structures, use a \emph{single one} and choose the input operands accordingly to the state that the network is
+currently in. Besides, the right elementwise multiplier of Figure~\ref{fig:network} is not used as the same time as any other, so it is a waste of resources.
+The issue about the elementwise multiplier that precedes the register can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$
+module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the improved network design of Figure~\ref{fig:network-opt}.
 
@@
 \end{figure}
-The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise  structure: this is where the time multiplexing of the structure takes place.
-Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplicator (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, I added another multiplexer before the elementwise multiplication that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
+The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine and its value
+is incremented after each complete usage of the `$\tanh$-$\sigma$-$(\cdot)$wise' structure: this is where the time multiplexing of the structure takes place.
+Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the register in the previous design) is the signal $\mb{c}^{(t-1)}$,
+another multiplexer was added before the elementwise multiplication, that selects that signal in that particular case, and the output from the $\tanh(\mb{x})$ block otherwise.
 
-The flip-flops on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
-The first activated flip-flop is the middle one, which keeps the result from the operation of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the elementwise multiplier, the bottom flip-flop to save the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal.
-Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.
+The registers on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network state machine that enable the appropriate
+register, placing the output from the elementwise multiplicator in the correct place. The first activated register is the middle one, which keeps the result from the
+operation of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the elementwise multiplier, the bottom register is enabled to save the other portion of
+the sum that evaluates to the $\mb{c}^{(t)}$ signal. Lastly, the top register saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$
+and is used by the Gate modules in this next batch of calculations.
 
@@
 \end{equation}
-where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact.
-In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion.
-After adjustments to the state machine, and accounting for pipelining and synchronization within the datapath, the clock cycles needed after the gate calculations are 27, so the total clock cycles needed are
+where we see that we saved $5N$ multipliers when comparing to the previous design, which for a large value of $N$ can have a decisive impact.
+In terms of speed performance, although the Gate calculation time remains the same, now the `$\tanh$-$\sigma$-$(\cdot)$wise' structure runs
+for 3 consecutive times, in a non parallel fashion. After adjustments to the state machine, and accounting for pipelining and synchronization
+within the datapath, the clock cycles needed after the gate calculations are 27, so the total clock cycles needed are
 
@@
 \end{equation}
-which is only 13 clock cycles more than the fully-parallel architecture.
-For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
+which is only 13 clock cycles more than the fully-parallel architecture. For instance, an $N=32$ neuron network using the fully-parallel design would require 320 DSP slices,
+while this new architecture would only require 160. Only at the expense of 13 clock cycle overhead.
 
 \section{Hardware LSTM Network -- SPSA Training}
-In this section, I propose a digital circuit that aims to train the Network whose details were explained in the previous Section by implementing the training algorithm of Section~\ref{sec:training_lstm}.
-The training consists in changing the weight values in the RAMs accordingly with the error of the current prediction against the correct outcome.
-This envolves \textbf{two} forward propagations: one, where we run the network with the weights unchanged, and another where we \emph{perturbate} the weights by a parameter $\pm \beta$ (the actual sign for each weight is generated at the beginning of the first forward propagation).
-After these two forward propagations, there is a small time frame where the weights are updated, row by row, before the next incoming sample.
-This is explained in Section~\ref{sec:train-circ}.
-The way how the numbers are randomly generated in hardware is presented in Section~\ref{sec:randnum-circ}.
+In this section, a digital circuit is proposed that aims to train the Network whose details were explained in the previous Section by implementing the training
+algorithm of Section~\ref{sec:training_lstm}. The training consists in changing the weight values in the RAMs accordingly with the error of the current prediction
+against the correct outcome. This involves \textbf{two} forward propagations: one, where we run the network with the weights unchanged, and another where we \emph{perturbate}
+the weights by a parameter $\pm \beta$ (the actual sign for each weight is generated at the beginning of the first forward propagation). After these two forward propagations,
+there is a small time frame where the weights are updated, row by row, before the next incoming sample. This is explained in Section~\ref{sec:train-circ}. The way how the numbers
+are randomly generated in hardware is presented in Section~\ref{sec:randnum-circ}.
 
@@
 While a True Random Number Generator (TRNG) generates data from a physical entropy source, a Pseudo-random Number Generator (PRNG) generates a sequence of numbers by applying a given
-function to an initial seed. This means that, if one knows that seed and the function that generates the sequence, one can predict the previous and future numbers given the current state. While this
-is not acceptable for high security cryptographic systems, it serves the purpose of providing a sequence of random sign vectors (from whom both the signal of the perturbation and the update will be choosen from).
-The remaining issue is to find a PRNG with good statistical qualities, so instead of implementing a simple linear-feedback shift register (LFSR), I did some research on the topic, and eventually chose the solution
-of~\cite{Tkacik2003}, based on a 43-bit LFSR and a 37-bit Cellular Automata, by selecting 32 bits from each, permuting the bits and performing a bitwise XOR -- the output has, also, 32 bits.
-The quality of the
+function, consecutively, to an initial seed. This means that, if one knows that initial seed and the function that generates the sequence, one can predict the previous and
+future numbers given the current state. While this is not acceptable for high security cryptographic systems, it serves the purpose of providing a sequence of random sign
+vectors (from whom both the signal of the perturbation and the update will be chosen from). The remaining issue is to find a PRNG with good statistical qualities, so instead
+of implementing a simple linear-feedback shift register (LFSR), I did some research on the topic, and eventually chose the solution of~\cite{Tkacik2003}, based on a 43-bit LFSR
+and a 37-bit Cellular Automata, by selecting 32 bits from each, permuting the bits and performing a bitwise XOR -- the output has, also, 32 bits. The quality of the
 statistical properties are guaranteed, since it passed the DIEHARD test battery, as well as others, according to~\cite{Tkacik2003}.
@@
 
-The output of my hardware implementation of this PRNG is presented in Figure~\ref{fig:prng_out}.
-It is a state-time diagram, where the contents of the registers at each time are represented by black (ones) and white (zeros), and the left column represents the output of the LFSR, the center column is the output of the Cellular Automata and the right column is their combined output.
-As one can see, the time dependence of the LFSR is very high, which is not desirable.
-The Cellular Automata, although uncorrelated, still retains some of the patterns that emerge from CAs.
-On the other hand, the combined output is the one that most resembles white noise, which is a desirable characteristic.
+The output of the hardware implementation of this PRNG is presented in Figure~\ref{fig:prng_out}. It is a state-time diagram, where the contents of the
+registers at each time are represented by black (ones) and white (zeros). Tthe left column represents the output of the LFSR, the center column is the
+output of the Cellular Automata and the right column is their combined output. As one can see, the time dependence of the LFSR is very high, which is not desirable.
+The Cellular Automata, although uncorrelated, still retains some of the patterns that emerge from CAs. On the other hand, the combined output is the one that most resembles
+white noise, which is a desirable characteristic.
 
@@
 \subsection{Training Algorithm Circuit}\label{sec:train-circ}
-Although the convergence results are not conclusive, here is presented the proposed digital circuit of the SPSA training for the network.
-The circuit of Figure~\ref{fig:train} is replicated for each of the eight weight RAMs, and the Sign RAM holds a single bit for each of the weight positions of the main weight RAM.
-This bit keeps the sign of the perturbation to be applied to that particular weight, both on the second forward propagation and the training step.
-During the first forward propagation, and since the weights are not perturbed in this run, we can use this time to propagate the sign RAM with new values for the next training stage.
-For this purpose, the PRNG is activated by the signal \verb+genRandNum+, generated by the state machine, and we write columns of bits to the sign RAM, re-using the addressing signal produced by the Gate, namely \verb+colAddressRead+ (\verb+weightUpdate+ is set to zero at this stage).
-At the second forward propagation, \verb+pertWeights+ is set to one, and the Perturbation Mux selects the input where the weights are summed with the perturbation $\beta$, whose sign is selected by the sign RAM.
+Here is presented the proposed digital circuit of the SPSA training for the network. The circuit of Figure~\ref{fig:train} is replicated for each of the
+eight weight RAMs, and the Sign RAM holds a single bit for each of the weight positions of the weight RAMs. This bit keeps the sign of the perturbation
+to be applied to that particular weight, both on the second forward propagation and on the training stage. During the first forward propagation, and since
+the weights are not perturbed in this run, we can use this time to populate the sign RAMs with new values for the next training stage, coming from the
+PRNG of Section~\ref{sec:randnum-circ}. For this purpose, the PRNG is activated by the signal \verb+genRandNum+, generated by the state machine, and we
+write columns of bits to the sign RAM, re-using the addressing signal produced by the Gate, namely \verb+colAddressRead+ (\verb+weightUpdate+ is set to
+zero at this stage). At the second forward propagation, \verb+pertWeights+ is set to one, and the Perturbation Mux selects the input where the weights
+are summed with the perturbation $\beta$, whose sign is selected by the sign RAM.
 
-After this stage is completed, and before we start this new training cycle for the new incoming sample, we perform the actual training.
-For that, \verb+weightUpdate+ is set to one, and we enable writing to the Weight RAM.
-Also, the machine state is now responsible for addressing both RAMs, and we begin a simultaneous \emph{pipelined} write to the Weight RAM: this is done by delaying the the Write address in respect to the Read address.
-This way, at the first clock cyle while we read Column 0, in the next clock cycle we will read Column 1 \emph{and} write the sign-changed column 0 plus the update coming from the difference between the outputs in the two forward propagations.
-In order to simplify the update rule of Equation~\ref{eq:spsa_updateRule}, I have fixed $\alpha$ and $\beta$ as negative powers of two, and this replaces the multiplication and division by a simple arithmetic right shift -- this idea is also proposed by~\cite{Maeda05}.
-
-The learning, although working in the Python model developed, is still not converging in the simulations of the Verilog code for this circuit.
-This might have to deal with an overflow problem at the weight update.
-This is a good question to be addressed in future work.
+After this stage is completed, and before the network starts a new training cycle for the new incoming sample, we perform the actual training. For that purpose,
+\verb+weightUpdate+ is set to one, and we enable writing to the Weight RAM. Also, the machine state is now responsible for addressing both RAMs, and we
+begin a simultaneous \emph{pipelined} write to the Weight RAM: this is done by delaying the Write address in respect to the Read address by one clock cycle. This way, at
+the first clock cycle, while we read Column 0, the next clock cycle will read Column 1 \emph{and} write the sign-changed column 0 plus the update coming
+from the difference between the outputs in the two forward propagations. In order to simplify the update rule of Equation~\ref{eq:spsa_updateRule},
+$\alpha$ and $\beta$ were fixed to negative powers of two, and this replaces the multiplication and division by a simple arithmetic right shift -- this idea
+is also proposed by~\cite{Maeda05} -- considerably simplifying the design.
 

### 2016-07-25 finalText/results.tex
Final Version of the Text and Short Paper

@@
 \chapter{Results}\label{chap:results}
-The optimized design of Section~\ref{sec:struct-shared} was synthesized for different network parameters, using Xilinx's Vivado 2015.4 Design Suite, and the synthesis results are reported in Section~\ref{sec:res-synth}, along with their discussion.
-The performance of the synthesized networks is very promising, showing an approximate $195\times$ speed-up over the LSTM Python model, which was running on a high-end desktop computer with an over clocked CPU.
-Although the FPGA aims for an Embedded System environment usage, the fairness of this comparison still holds, since a Python/Numpy implementation is slower than a lower-level language implementation (such as in C) and this level of performance would be similar to a C implementation in a low-power, embedded CPU platform.
-These results are reported in Section~\ref{sec:res-valid}.
+The optimized design of Section~\ref{sec:struct-shared} was synthesized for different network parameters, using Xilinx's Vivado 2015.4 Design Suite,
+and the synthesis results are reported in Section~\ref{sec:res-synth}, along with their discussion. The performance of the synthesized networks is very promising,
+showing an approximate $251\times$ speed-up over the LSTM Python model, which was running on a high-end desktop computer with an overclocked CPU, and a $14\times$ speed-up
+over the most recent hardware implementation~\cite{Chang15} Although the FPGA aims for an Embedded System environment usage, the fairness of the comparison with a Desktop computer
+still holds, since a Python/Numpy implementation is slower than a lower-level language implementation (such as in C) and this level of performance would be similar to a C implementation
+in a low-power, embedded CPU platform. These results are reported in Section~\ref{sec:res-valid}.
 
 \section{Synthesis}\label{sec:res-synth}
-In order to assess the impact of the network size on the performance and resource usage of the proposed design, I have synthesized the Verilog description of the proposed network for sizes $N$ of 4, 8, 16 and 32, and for each of these the resource sharing parameter $K_G$ was also varied (as will be explained further ahead, the number of inputs is set to $M=2$).
-Since performance is the key factor for using an FPGA implementation, the synthesis strategy used was the \verb+Flow_PerfOptimized_high+, and the implementation strategy was \verb+Performance_ExplorePostRoutePhysOpt+ to perform a post-route optimization stage that proved useful in many cases to meet the clock specifications when the first Place\&Route pass failed to meet them.
-For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were successfully synthesized.
+The proposed network was first synthesized aiming a Xilinx XC7Z020 SoC, for sizes $N$ of 4, 8, 16 and 32, varying the resource sharing parameter $K_G$, and the number of inputs set
+to $M=2$. For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were
+successfully synthesized. This is because of the complexity of the input multiplexer of the Matrix-vector Dot Product Module in~\ref{sec:dotprod_sec}. To synthesize the
+design for sizes $N$ of 64, 128 and 256, a Virtex-7 VC707 board was used.
+
+In order to assess the impact of the network size on the performance and resource usage of the proposed design, the Verilog description of the proposed network was first synthesized
+aiming a Xilinx XC7Z020 SoC platform, for sizes $N \in \left\{4, 8, 16, 32 \right\}$, and for each of these the resource sharing parameter $K_G$ was also varied (as will be explained further ahead,
+the number of inputs is set to $M=2$). Since performance is the key factor for using an FPGA implementation, the synthesis strategy used was the \verb+Flow_PerfOptimized_high+, and the implementation
+strategy was \verb+Performance_ExplorePostRoutePhysOpt+ to perform a post-route optimization stage that proved useful in many cases to meet the clock specifications when the first Place\&Route pass
+failed to meet them.
+For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were successfully synthesized (N.D.
+in the tables).
+Networks of sizes $N \in \left\{64, 128\right\}$ were also synthesized for a Virtex-7 VC707 board.
 
 \subsection{Maximum Frequency}\label{sec:res-synth-maxfreq}
-Using the synthesis strategy outlined in the previous paragraph, the maximum clock frequency that does not cause timing violations of any sort is plotted in Figure~\ref{fig:maxfreq}.
-For $N=4$, since there are only 4 rows to be multiplied, the maximum value of $K_G$ is 4, and hence no synthesis was performed for $K_G = 8$; also, since $K_G = 2$ and $N=32$ would use $32(8/2+3) = 224$ DSP slices, that exceeds the 220 slices available in the Zynq 7020, so there is no synthesis data for that value.
+Using the synthesis strategy outlined in the previous paragraph, the maximum clock frequency that does not cause timing violations of any sort is plotted in
+Figure~\ref{fig:maxfreq}. For $N=4$, since there are only 4 rows to be multiplied, the maximum value of $K_G$ is 4, and hence no synthesis was performed for $K_G = 8$ (N.A. in the tables);
+also, since $K_G = 2$ and $N=32$ would use $32(8/2+3) = 224$ DSP slices, that exceeds the 220 slices available in the Zynq 7020, so there is no synthesis data for that value.
 
-First, for all network sizes, it is clear that with increasing $K_G$, the maximum clock speed decreases.
-This means that there is a critical path in the Matrix-Vector multiplication unit, whose multiplexer becomes increasingly complex for higher values of $K_G$.
-On the other hand, when $K_G$ is the same, smaller networks are faster than larger networks.
-The fastest design is an $N=4$ and $K_G = 2$ network, with a clock frequency of $140.154$ MHz, and the slowest one is an $N=32$ and $K_G=4$ network, clocked at $90.498$ MHz.
-The reference design used for validation in Section~\ref{sec:res-valid} is an $N=8$ and $K_G=2$ network is clocked at $132.89$ MHz, which yields a clock period of $7.525$ ns.
+First, for all network sizes, it is clear that with increasing $K_G$, the maximum clock speed decreases. This means that there is a critical path in the Matrix-Vector multiplication unit,
+whose multiplexer becomes increasingly complex for higher values of $K_G$. On the other hand, when $K_G$ is the same, smaller networks are faster than larger networks. The fastest design
+is an $N=4$ and $K_G = 2$ network, with a clock frequency of $158.228$ MHz, and the slowest one is an $N=32$ and $K_G=4$ network, clocked at $101.523$ MHz. The reference design used for
+validation in Section~\ref{sec:res-valid} is an $N=8$ and $K_G=2$ network is clocked at $154.321$ MHz, which yields a clock period of $6.48$ ns.
 
@@
 \subsection{DSP Slices Usage}\label{sec:res-synth-dsp}
-The estimates made in Equation~\ref{eq:numdsp_network-opt} were proven to be accurate, as Figure~\ref{fig:dspused} confirms.
-The reference network design uses 56 DSP slices, which corresponds to $25.45\%$ of the total number of DSP slices available.
+The estimates made in Equation~\ref{eq:numdsp_network-opt} were proven to be accurate, as Figure~\ref{fig:dspused} confirms. The reference network design uses 56 DSP slices, which
+corresponds to $25.45\%$ of the total number of DSP slices available.
+
 \begin{figure}
@@
 
-\subsection{Power Usage}\label{sec:res-synth-power}
-Another important metric of the performance of a design is its \textbf{power consumption}.
-All designs yielded a constant baseline value for the \emph{static} power consumption of around 120 mW, and the power consumption reported in Figure~\ref{fig:power} refers to the \emph{total} power consumptions, i.e.
-both the baseline static power and the dynamic power consumption.
-It is clear that the smaller the network is, the less power is consumed, as one would expect.
-Furthermore, an increasing level of resource sharing yields a substantially lower power consumption figure: this is because we use less DSP slices when we increase $K_G$.
-Of course, even though the power consumption is lower in that case, that comes at the expense of a lower clock frequency and more clock cycles elapsed per forward propagation.
+\subsection{Power Consumption}\label{sec:res-synth-power}
+Another important metric of the performance of a design is its \textbf{power consumption}. All designs yielded a constant baseline value for the \emph{static} power consumption of around 120 mW,
+and the power consumption reported in Figure~\ref{fig:power} refers to the \emph{total} power consumptions, i.e. both the baseline static power and the dynamic power consumption.
+It is clear that the smaller the network is, the less power is consumed, as one would expect. Furthermore, an increasing level of resource sharing, $K_G$, yields a substantially lower
+power consumption figure: this is because less DSP slices are used when $K_G$ is increased. Of course, even though the power consumption is lower in that case, that comes at the
+expense of a lower clock frequency and a higher computation overhead per forward propagation.
 
@@
 \subsection{Other Resources Usage}\label{sec:res-synth-otheres}
-Besides DSP slices, it is also worth mentioning the usage of LUT, LUTRAMs and Flip-Flops.
-In Table~\ref{tab:lut}, the usage of LUTs is reported, where we can see that although there is not a clear trend on how the LUT usage varies with increasing $K_G$, it is clear, and expectable, that the LUT usage increases with the size of the network by an approximate factor of 2, from $N=2$ to $N=16$.
-As for $N=32$, the usage does not follow this apparent trend, and rises sharply to 91\%.
-For $K_G = 8$ the usage even surpasses the maximum amount of LUTs available.
+The LUT, LUTRAM and Flip-Flop usages are discussed here. In Table~\ref{tab:lut}, the usage of LUTs is reported, where we can see
+that although there is not a clear trend on how the LUT usage varies with increasing $K_G$, it is clear, and expectable, that
+the LUT usage increases with the size of the network by an approximate factor of 2, from $N=2$ to $N=16$. As for $N=32$, the
+usage does not follow this apparent trend, and rises sharply to 91\%. For $K_G = 8$ the usage surpasses the maximum amount
+of LUTs available in the XC7Z020.
+
 \begin{table}
+
     \centering
@@
   \end{tabular}
-    \caption{LUT usage for different $N$ and $K_G$}
+  \caption{LUT usage for different $N$ and $K_G$ in the XC7Z020}
   \label{tab:lut}
@@
 
-As for LUT, the FF usage also scales accordingly to a $2\times$ factor.
-Furthermore the LUTRAM usage scales well and does not pose a limitation on the network size.
-The usage results are reported in Table~\ref{tab:ff}.
+As for LUTs, the FF usage also scales accordingly to a $2\times$ factor. In terms of LUTRAM, used to store the network weights,
+the amount used increases by 2 with increasing values of $N$, as before, and \textbf{does not} depend on $K_G$. This is because
+the amount of weights only depends on the network sizes $M$ and $N$, and not on $K_G$. Furthermore, unlike LUTs, it scales well
+with increasing network sizes, and does not pose a limitation on the network size. The usage results for the XC7Z020 are reported in Table~\ref{tab:ff},
+and in Table~\ref{tab:ff-virtx7} the usages for the VC707.
 
@@
   \end{tabular}
-    \caption{Flip-flop usage for different $N$ and $K_G$}
+  \caption{Flip-flop and LUTRAM usage for the XC7Z020}
 	\label{tab:ff}
@@
 
+\begin{table}
+
+    \centering
+  \begin{tabular}{ | l | c | c | c |}
+    \hline
+         & LUTRAM  & FF & LUT \\
+    \hline
+    $N=64$ & 24.22\%  & 9.14\% & 24.22\% \\
+    \hline
+    $N=128$ & 14.09\%  & 18.3\% & 41.82\%\\
+    \hline
+  \end{tabular}
+  \caption{Flip-flop, LUTRAM and LUT usage for the VC707}
+  \label{tab:ff-virtx7}
+\end{table}
+
 \section{Validation and Comparison}\label{sec:res-valid}
-Over the course of this Section, the functionality of the network is verified against the Python reference module of~\ref{sec:res-pyth} that I have developed as a reference, both for the forward propagation of the network, as well as for the training algorithm.
-The methodology of this procedure is outlined in Section~\ref{sec:res-meth}.
-On Section~\ref{sec:res-valid-fpps}, I will assess how the performance results of Section~\ref{sec:res-synth-maxfreq} translate into how many classifications per second this design achieves, and how that figure compares with the capabilities of the Python module.
+Over the course of this Section, the functionality of the network is verified against the Python reference module of~\ref{sec:res-pyth} that was developed as a
+reference, both for the forward propagation of the network, as well as for the training algorithm. The methodology of this procedure is outlined in Section~\ref{sec:res-meth}.
+On Section~\ref{sec:res-valid-fpps}, the performance results of Section~\ref{sec:res-synth-maxfreq} is translated into how many classifications per second this design
+achieves, and how that figure compares with the capabilities of the Python module and the current existing hardware implementation~\cite{Chang15}.
 
 \subsection{Reference Python Module}\label{sec:res-pyth}
-Before performing the Verilog description of the network, I have tested the ideas outlined in Section~\ref{chap:theorBack} first by building a software version of an LSTM Network.
-Since I wanted to quickly test the ideas without much effort, I decided to use Python and Numpy rather than MATLAB, since the former have higher performance at the same level of code complexity.
-The code for this Python class is listed in Appendix~\ref{ap1:lstm-pyth}.
-It supports both the forward propagation of an already trained network, but also provides functions to train a model using the SPSA method outlined in Section~\ref{sec:training_lstm} and also Backpropagation Through-Time as presented in~\cite{Greff15}.
-As the reader can attest by running the code himself, the SPSA proves to be a valid approach for training these networks, achieving convergence for the Machine Learning task that will be presented in the next paragraph, although it converges slower than Backpropagation Through Time.
+Before performing the Verilog description of the network, the ideas outlined in Section~\ref{chap:theorBack} were tested first by building a software version of an LSTM Network.
+Since we want to quickly test the ideas without much effort, Python and Numpy were used rather than MATLAB, since Python/Numpy has higher performance at the same level of
+code complexity. The code for this Python class is listed in Appendix~\ref{ap1:lstm-pyth}. It supports both the forward propagation of an already trained network, but also provides
+functions to train a model using the SPSA method outlined in Section~\ref{sec:training_lstm} and also Backpropagation Through-Time as presented in~\cite{Greff15}. As the reader can attest
+by running the code himself, the SPSA proves to be a valid approach for training these networks, achieving convergence for the Machine Learning task that will be presented in the next paragraph,
+although it converges slower than Backpropagation Through Time.
 
-The learning problem presented to both the software and hardware network is the sum of two binary numbers of 8 bits.
-The bits of each number at position $i$ are input to the network as a vector, and the network outputs its prediction of the correct value of the $i$-th bit of the \emph{result} number.
-After the whole number number is processed, the memory cells of the LSTM network are reset and a new addition task can be presented to the network.
-Even though this seems a rather simple problem, it accounts for all the essential issues at which this network excels: first, this is a \emph{classification} problem in which the Machine Learning algorithm needs to output a prediction based on the input feature vector, and that prediction has to take into account the \emph{history} of predictions so far, because the current bit is the sum of not only the bits of the two operands, but also the \emph{carry} generated at the last few positions -- this is where the memory cells of this special Neural Network come into effect.
+The learning problem presented to both the software and hardware network is the sum of two binary numbers of 8 bits. The bits of each number at position $i$ are input to the network as a vector,
+and the network outputs its prediction of the correct value of the $i$-th bit of the \emph{result} number. After the whole number number is processed, the memory cells of the LSTM network are reset
+and a new addition task can be presented to the network. Even though this seems a rather simple problem, it accounts for all the essential issues at which this network excels: first, this is
+a \emph{classification} problem in which the Machine Learning algorithm needs to output a prediction based on the input feature vector, and that prediction has to take into account the \emph{history}
+of predictions so far, because the current bit is the sum of not only the bits of the two operands, but also the \emph{carry} generated at the last few positions -- this is where the memory cells of
+this special Neural Network come into effect.
 
 \subsection{Methodology}\label{sec:res-meth}
-The Python testbench that both trains and verifies the predictions of the software network of Appendix~\ref{ap1:lstm-pyth} is listed in Appendix~\ref{ap1:testbench-pyth}.
-This script, after achieving convergence, stores the trained network weights in \href{https://docs.python.org/3/library/pickle.html}{Pickle} format.
-These weights are, then, converted to $Q6.11$ fixed-point format and directly loaded into the Verilog model by the Verilog testbench.
-The Verilog testbench then applies in order the bits of randomly generated numbers, captures the output bit and checks whether it is correct or not, and counts the number of incorrect bits.
+The Python testbench that both trains and verifies the predictions of the software network of Appendix~\ref{ap1:lstm-pyth} is listed in Appendix~\ref{ap1:testbench-pyth}. The scripts train the
+network for a maximum of 100 epochs, and in each epoch, the network is presented with 5000 8 bit additions for training, and then 100 8 bit additions for testing, where the prediction error is
+calculated, and the training convergence is verified (when there is zero error for two consecutive times). After achieving convergence, the script stores the trained network weights in
+\href{https://docs.python.org/3/library/pickle.html}{Pickle} format. These weights are, then, converted to $Q6.11$ fixed-point format and directly loaded into the Verilog model by the Verilog
+testbench. The Verilog testbench then applies in order the bits of randomly generated numbers, captures the output bit and checks
+whether it is correct or not, and counts the number of incorrect bits.
 
-This network has, then, two inputs ($M=2$) and eight neurons ($N=8$).
-Furthermore, the output of the eight LSTM neurons is reduced to a single output by means of a simple neuron, as the one described in Figure~\ref{fig:modelNeuron_a}.
-This introduces a further 5 clock cycle overhead that is not accounted in Section~\ref{sec:res-valid-fpps}, since it is not actually part of the LSTM Network and, although it was placed in series in the datapath, after the LSTM network, it can run in parallel with the LSTM network, introducing no overhead whatsoever.
+This network has, then, two inputs ($M=2$) and eight neurons ($N=8$). Furthermore, the output of the eight LSTM neurons is reduced to a single output by means of a simple neuron, as the one
+described in Figure~\ref{fig:modelNeuron_a}. This introduces a further 5 clock cycle overhead that is not accounted in Section~\ref{sec:res-valid-fpps}, since it is not actually part of the
+LSTM Network. Although it was placed in series in the datapath, after the LSTM network, it can be run in parallel with the LSTM network, introducing no overhead whatsoever.
 
-A Python script was written (see Appendix~\ref{ap1:runtb-lstm}) to automatically compile the Verilog code, generate the input and output data, and call Questasim to run the simulation of the HDL code.
-Figure~\ref{fig:res-meth-wave} shows the waveform viewer window of a run of the Verilog testbench, and in Appendix~\ref{ap1:transcript-cmd} shows the transcript of a successful run of this testbench, proving that the learning ability of the Hardware network still retains the predicting capability of the software network.
+A Python script was written (see Appendix~\ref{ap1:runtb-lstm}) to automatically compile the Verilog code, generate the input and output data, and call Questasim to run the simulation of the
+HDL code. Figure~\ref{fig:res-meth-wave} shows the waveform viewer window of a run of the Verilog testbench, and in Appendix~\ref{ap1:transcript-cmd} shows the transcript of a successful
+run of this testbench, proving that the learning ability of the Hardware network still retains the predicting capability of the software network.
 
-\begin{figure}
+\begin{sidewaysfigure}
     \centering
@@
     \label{fig:res-meth-wave}
-\end{figure}
+\end{sidewaysfigure}
+
 \subsection{Performance}\label{sec:res-valid-fpps}
-To evaluate the performance of the network in terms of how fast it can produce data, I have defined a metric based on how many predictions it can produce per second (i.e.
-produce a new result bit in the output sequence), in \textbf{millions}.
-The prediction time is the time elapsed since a new input vector is applied \emph{to} the moment the LSTM Network produces an output vector (as discussed previously, the five clock cycles of the simple neuron at the output are \textbf{not} considered).
-To do this, we multiply the number of clock cycles yielded by Equation~\ref{eq:numcc_network-opt} by the equivalent clock period from the synthesis clock report of Figure~\ref{fig:maxfreq}.
-This result is epitomized in Table~\ref{tab:process-time}, where the calculation time of the Python module forward propagation function is also included.
-This time was measured using the \href{https://docs.python.org/3.5/library/timeit.html}{\textbf{timeit}} module, that allows the evaluation of the execution time of small pieces of code as well as complete functions with arguments.
-The Python code was run on a Linux System, powered by an i7-3770k Intel Processor, running at 4.2GHz.
+To evaluate the performance, a metric was defined based on how many predictions it can produce per second (i.e. produce a new result bit in the output sequence),
+in \textbf{millions}. The metric on~\cite{Chang15} is not very informative, since
+although a system can perform many calculations per second, if those operations are redundant, that metric has no relevant information
+regarding \textbf{how fast} the system can \textbf{perform the actual task it was meant to do}, which in this case is a complete Forward
+Propagation through the network. The prediction time is the time elapsed since a new input vector is applied \emph{to} the moment the LSTM
+Network produces an output vector. Hence, we multiply the number of clock cycles yielded by Equation~\ref{eq:numcc_network-opt} by the equivalent clock period
+from the synthesis clock report of Figure~\ref{fig:maxfreq}.
+This result is epitomized in Table~\ref{tab:process-time}, where the calculation time of the Python module's Forward Propagation function is
+also included. This time was measured using the \textbf{timeit}\footnote{https://docs.python.org/3.5/library/timeit.html}  module, that allows
+the evaluation of the execution time of small pieces of code as well as complete functions with arguments. The Python code was run on a Linux System,
+powered by an i7-3770k Intel Processor, running at 4.2GHz.
 
@@
     \hline
-    $N=4$ & N.A.  & 360.15 ns  & 292.535 ns & 65 $\mu$s & $\times180$ \\   
+    $N=4$ & N.A.  & 309.68 ns  & 259.12 ns & 65 $\mu$s & $\times251$ \\
     \hline
-    $N=8$ & 960.3 ns  & 494 ns  &  368.725 ns & 72 $\mu$s & $\times194$ \\    
+    $N=8$ & 793.46 ns  & 421.12 ns  &  317.52 ns & 72 $\mu$s & $\times228$ \\
     \hline
-    $N=16$ & 1.715 $\mu$s  & 921.5 ns  & 518.05 ns & 96 $\mu$s & $\times185$ \\
+    $N=16$ & 1.497 $\mu$s  & 738.19 ns  & 461.336 ns & 96 $\mu$s & $\times208$ \\
     \hline
-    $N=32$ & N.D. & 1.78 $\mu$s  & N.A. & 185 $\mu$s  &  $\times104$ \\    
+    $N=32$ & N.D.          & 1.586 $\mu$s & N.A.       & 185 $\mu$s  &  $\times117$ \\
 		\hline
   \end{tabular}
-    \caption{Total processing time for a single forward propagation, for hardware networks of various $K_G$ and the Python software network of equivalent size.
-The speed-up figure is calculated between the Python processing time versus the best hardware network time of the same size}
+  \caption{Total processing time for a single forward propagation on the XC7Z020}
   \label{tab:process-time}
 \end{table}
-The performance increase is impressive, even for the slowest of the designs (the $N=32$ network).
-The hardware network is, at best, $\times200$ faster than the software counterpart, and at worst $\times100$ faster.
-Also, it is noticeable that increasing the level of resource sharing increases the computation time, since the level of parallelism is lower.
+The performance increase is significant, even for the slowest of the designs (the $N=32$ network). The hardware network is, at best, $\times251$ faster than the software counterpart,
+and at worst $\times117$ faster. Also, it is noticeable that increasing the level of resource sharing increases the computation time, since the level of parallelism is lower.
 
- Since the previous values are the time needed for a \emph{single} forward propagation, to know how many forward propagations we can perform \emph{per second}, we only need to invert the previous values.
-These values are presented in Figure~\ref{fig:Mclass-psec}.
-While the $N=8$ and $K_G=2$ network is able to perform around 3.2 million predictions per second, the Python model can only output around 14 thousand predictions, which is a very significant result that proves the relevance of this implementation.
+Since the previous values are the time needed for a \emph{single} forward propagation, to know how many forward propagations we can perform \emph{per second},
+we only need to invert the previous values. These values are presented in Figure~\ref{fig:Mclass-psec}. While the $N=8$ and $K_G=2$ network is able to
+perform around 3.15 million predictions per second, the Python model can only output around 14 thousand predictions, which is a very significant result that
+proves the relevance of this implementation.
+
 \begin{figure}
@@
 \end{figure}
+
+As for the larger-sized networks synthesized in the VC707, the results are also very promising. For network sizes of $N=64$ and $N=128$, a complete
+forward propagation takes 1.14 $\mu$s and 2.052 $\mu$s respectively, and for both the maximum clock frequency achievable was 140.845 MHz. Since the
+design~\cite{Chang15}, for $N=128$, takes an estimated 29.13 $\mu$s, this design yields an improvement of $14\times$ over it.

### 2016-07-25 finalText/stateArt.tex
Final Version of the Text and Short Paper

@@
 \begin{itemize}
-    \item \textbf{Handwriting Recognition} -- an LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}.
+    \item \textbf{Handwriting Recognition} -- an LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieving a recognition rate of up to 91\%~\cite{ICDAR09}.
 LSTMs have also proven to surpass HMM-based models in terms of optical character recognition of printed text, as~\cite{Breuel13} suggests.
@@
 \section{Hardware Implementations of LSTM}\label{sec:sa_hardware}
-Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently (November 2015) by Chang et al.~\cite{Chang15} in the Computing Research Repository (CoRR).
-It consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered~ Zedboard implementation.
-It uses a character-level language model from Andrej Karpathy~\footnote{\href{https://github.com/karpathy/char-rnn}{https://github.com/karpathy/char-rnn}}, written in Lua using the Torch7 framework~\footnote{\href{http://torch.ch/}{http://torch.ch/}} (the Lua calls are implemented in C, so no performance is lost).
+Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
+(Januray 2016) by Chang et al.~\cite{Chang15} in the international Journal of Advanced Research in
+Electrical, Electronics and Instrumentation Engineering. It consists on the proposal of an LSTM network with $N=128$ for dedicated hardware,
+targeting a Xilinx\textregistered~ Zedboard implementation. It uses a character-level language model from Andrej
+Karpathy~\footnote{\href{https://github.com/karpathy/char-rnn}{https://github.com/karpathy/char-rnn}}, written in
+Lua using the Torch7 framework~\footnote{\href{http://torch.ch/}{http://torch.ch/}}.
 
-Although the article is not clear on whether there is active learning by the ARM CPU -- the authors refer that the CPU loads the weights before operation, and that it changes them during operation, although how and why that change is done is not even clearly explained, neither mathematically nor conceptually --, \textbf{there is no on-chip learning module} in the FPGA according to the description provided.
+Although the article is not clear on whether there is active learning by the ARM CPU -- the authors refer that the CPU
+loads the weights before operation, and that it changes them during operation, although how and why that change is done
+is not even clearly explained, neither mathematically nor conceptually --, \textbf{there is no on-chip learning module}
+in the FPGA according to the description provided. For that reason, the system stores the weights in external memory,
+and accesses them using Direct Memory Access (DMA). Furthermore, the level of parallelism is low, since for an
+$N=128$ network, the design uses only $22.73\%$ of the DSP48's available in the board. This is because the matrix multiplications
+are serialized, and only one DSP is used for each matrix-vector product (aside from other uses).
 
-The implementation itself makes an extensive use of the Multiply-and-Accumulate units (MAC) of the FPGA, which since they are limited in number in the target platform, limits the number of neurons that we can deploy in parallel.
-The authors report a nearly 75\% usage for only 8 LSTM neurons.
-Although an apparently excessive number, I am left to scrutinize whether the usage of MACs is compulsory to obtain a better performance than a CPU or if it can be discarded, allowing for smaller cells.
-This way, we can have a layer of more neurons occupying the same area and, hopefully, the same, or even fewer, resources within the FPGA.
+The implementation is not flexible, since there is nothing that suggests a parameterization of the HDL description.
+Hence, it is only possible to synthesize networks of size $N=128$, meaning that there needs to be a redesign of the network
+for other network sizes. On the other hand, this is a valid approach for special applications that require networks of
+this size, and where we only need a co-processor for LSTM networks, and not a dedicated standalone system.
 
 \section{Training Algorithms and Implementation}\label{sec:sa_training}
-As stated in~\ref{sec:sa_hardware}, the work of~\cite{Chang15} does not feature on-chip learning at FPGA level, although there are a handful proposed solutions for it in recent literature.
-I will particularly look into the ones that use SPSA (see Section~\ref{sec:training_lstm}), since that is the training algorithm of choice for my proposed solution, and also to the ones that particularly apply SPSA to the training of Neural Networks.
+As stated in~\ref{sec:sa_hardware}, the work of~\cite{Chang15} does not feature on-chip learning at FPGA level, although there
+are a handful proposed solutions for it in recent literature. I will particularly look into the ones that use SPSA (see
+Section~\ref{sec:training_lstm}), since that is the training algorithm of choice for my proposed solution, and also to the
+ones that particularly apply SPSA to the training of Neural Networks~\cite{Maeda05}.
 
-The SPSA algorithm was initially published by Spall in~\cite{Spall98}, and its theoretical details are outlined in Section~\ref{sec:training_lstm}.
-As to its applications to the training of general Neural Networks, the earliest examples come from 1995 and 1996 in~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time where the memory resources of digital circuitry were limited, and so most of these structures were analog-based.
-Its adequacy was also established for \textbf{control problems}, such as those proposed in~\cite{Figueiredo97}, where a Time Delay Neural Network is used to control an unknown plant in a linear feedback system.
+The SPSA algorithm was initially published by Spall in~\cite{Spall98}, and its theoretical details, are outlined in
+Section~\ref{sec:training_lstm}. As to its applications to the training of general Neural Networks, the earliest examples
+come from 1995 and 1996 in~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time
+where the memory resources of digital circuitry were limited, and so most of these structures were analog-based. Its adequacy
+was also established for \textbf{control problems}, such as those proposed in~\cite{Figueiredo97}, where a Time Delay Neural
+Network is used to control an unknown plant in a linear feedback system.
 
-In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network in an FPGA (and thus a digital system), achieving promising results in an Altera FPGA.
-The article carefully delineates the approach taken, and also the hardware architecture designed, so it is a very good reference for the design that I will have to implement.
+In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network
+in an FPGA (and thus a digital system), achieving promising results in an Altera FPGA. The article carefully delineates the approach
+taken, and also the hardware architecture designed.
 
-Furthermore, a 2013 article by Tavear et al.~\cite{Tavear13} proposes, for the first time, using SPSA to train LSTM Neurons, although the article focuses on proving the suitability of SPSA to LSTM, and no actual hardware implementation is done or proposed.
-The authors simply demonstrate the suitability using conceptual arguments and by building a software model of an SPSA-trained LSTM network, and by comparing both the performance and computing speed of their model with the results achieved by Hochreiter et al.
-in~\cite{Hochreiter07}.
-Since the forward phase in both regular LSTM and SPSA-trained LSTM is the same, the computation time suffers no performance penalty whatsoever and the learning ability is preserved to a high degree, showing that SPSA is a valid alternative do BPTT and other similar and more common training schemes.
+Furthermore, a 2013 article by Tavear et al.~\cite{Tavear13} proposes, for the first time, using SPSA to train LSTM Neurons, although
+the article focuses on proving the suitability of SPSA to LSTM, and no actual hardware implementation is done or proposed. The authors
+simply demonstrate the suitability using conceptual arguments and by building a software model of an SPSA-trained LSTM network, and by
+comparing both the performance and computing speed of their model with the results achieved by Hochreiter et al. in~\cite{Hochreiter07}.
+Since the forward phase in both regular LSTM and SPSA-trained LSTM is the same, the computation time suffers no performance penalty whatsoever
+and the learning ability is preserved to a high degree, showing that SPSA is a valid alternative do BPTT and other similar and more common training schemes.
 
 \section{Final Overview}\label{sec:overview}
-As we could attest from this small literature survey, although there is already an hardware implementation of LSTM, there is still a good deal of room for improvement by adding on chip-learning to the system, and also to restrict it to a smaller use of the FPGA resources, allowing it to accomodate a more complex network.
-Furthermore,~\cite{Tavear13} shows that, at least for that particular case, the LSTM network doesn't suffer a great performance impact from using SPSA training, as opposed to the more common BPTT, and~\cite{Maeda05} showed that an SPSA hardware implementation is feasible.
-These three conclusions combined indicate that the idea of a hardware LSTM network with on-chip learning using SPSA is viable.
-
-Besides the proposed hardware implementation, it would be advantageous to perform \emph{benchmarking} to see how well it compares with software solutions, and Section~\ref{sec:sa_apps} suggests a handful of examples to test my final solution.
+As we could attest from this small literature survey, although there is already an hardware implementation of LSTM, there is still a good deal
+of room for improvement by adding on chip-learning to the system, and also to improve its performance and flexibility.
+Furthermore,~\cite{Tavear13} shows that, at least for that particular case, the LSTM network doesn't suffer a great performance impact from
+using SPSA training, as opposed to the more common BPTT, and~\cite{Maeda05} showed that an
+SPSA hardware implementation is feasible.

### 2016-07-25 finalText/theorBack.tex
Final Version of the Text and Short Paper

@@
 \subsubsection{Training -- SPSA}\label{sec:training_lstm}
-Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme.
-Since I am aiming at an hardware implementation, and although the memory resources of current FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system.
-That being said, we see that the usual training algorithms for LSTM cells -- i.e.
-Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance.
-A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.
-The main idea of this technique is, instead of explicitly evaluating the gradients for the cost function at each time step, to perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk.
-The weight update for the $i$-th weight at the time step $t$ is given by
+Since this work aims to propose an on-chip learning circuit, it is important to find a suitable learning scheme. Since I am aiming at an hardware 
+implementation, and although the memory resources of current FPGAs are abundant, one must find an algorithm that uses as less memory as possible 
+(at the smallest performance cost possible). That being said, we see that the usual training algorithms for LSTM cells --- i.e. Real-Time Recurrent 
+Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} --- usually involve 
+the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both 
+in terms of memory and performance. A most efficient approach to training times series dependant structures like LSTM is the use 
+of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}.  The main idea of this technique is, instead of explicitly evaluating 
+the gradients for the cost function at each time step, to perform a random walk in the neighbourhood of the current weight matrix, in the weight space, and 
+approximate the new weight update by the approximation of the gradient of the cost function that resulted from that random walk. The weight update for the $i$-th 
+weight at the time step $t$ is given by
 
@@
 If the random sign vector $\mb{s}^{(t)}$ is chosen carefully so that the expected value of the vector at the $i$-th position is zero, $E(s_i^{(t)})$, and the signs of two different components $i$ and $j$, at different time-steps, are independent (i.e.
-$E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, where $\delta$ is the Kronecker fuction), taking the \emph{Expected Value} of equation~\ref{eq:taylor} yields
+$E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, where $\delta$ is the Kronecker function), taking the \emph{Expected Value} of equation~\ref{eq:taylor} yields
 

### 2016-07-25 paper/paper.tex
Final Version of the Text and Short Paper

@@
 \usepackage{siunitx}
+\usepackage[utf8]{inputenc}
 
@@
 \newcommand{\mb}[1]{\mathbf{#1}}
+\renewcommand\IEEEkeywordsname{Keywords}
 
@@
 \begin{abstract}
-This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
+Our work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
 to outperform software implementations, by exploiting its inherent parallelism.
@@
 building blocks of the network is also presented. The network is synthesized for various sizes and platforms,
-and the performance results are presented and analysed. The synthesized network achieves a 251 times speed-up
-over a custom-built software network, proving the benefits of parallel computation for this kind of network.
+and the performance results are presented and analyzed. Our synthesized network achieves a 251 times speed-up
+over a custom-built software network, running on an i7-3770k Desktop computer, proving the benefits of parallel computation
+for this kind of network.
 \end{abstract}
+
+\begin{IEEEkeywords}
+Neural Networks, Long Short-Term, FPGA, Reconfigurable Hardware, Machine Learning
+\end{IEEEkeywords}
 
@@
 \IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
-particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain
-time step are also fed as the input in the next time step, and that their memory elements can make
+particular type of network, a Long Short-Term Memory (LSTM) Network, is a recursive network, in which the neuron outputs in a certain
+time step are also fed as inputs in the next time step, and since it possesses memory, the system can make
 sense of patterns within data sequences, unlike classical recursive neural networks.
@@
 However, the benefits of the inherent parallelism offered by a dedicated hardware
-platform are not availed, and there are relatively few implementations of Machine Learning algorithms in
+platform are not exploited, and there are relatively few implementations of Machine Learning algorithms in
 these kind of platforms.
 
-Hardware platforms would achieve a considerable speed-up over software implementations, which could prove
-useful for high data throughput systems, where the calculation overhead is critical and limits the performance of the systems.
+Hardware platforms could achieve a considerable speed-up over software implementations, which would prove
+useful for high data throughput systems, where the calculation overhead is critical and limits the performance.
 Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
@@
 
-Hitherto, there is only one implementation~\cite{Chang15}, but its performance is undermined
-by the external memory access. This implementation aims to make use of internal FPGA memory resources,
-and therefore achieve a higher speed performance.
+Hitherto, there is only one implementation, the one of~\cite{Chang15}, but its performance is undermined
+by the external memory access. Our implementation aims to make use of internal FPGA memory resources,
+and therefore achieve a higher throughput.
 
-In overview, this article begins by explaining what are LSTM networks -- and their mathematical
-details -- in Section~\ref{sec:lstmnn}, followed by providing a quick overview of their state
+Summarizing, this article begins by explaining what are LSTM networks -- and their mathematical
+details -- in Section~\ref{sec:lstmnn}, followed by a quick overview of their state
 of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed
 hardware architecture and its main constituent modules, and the performance and synthesis
-results are reported in Section~\ref{sec:results}. A final concluding remark is given in
+results are reported in Section~\ref{sec:results}. Concluding remarks are given in
 Section~\ref{sec:concl}.
@@
 \section{LSTM Neural Networks}\label{sec:lstmnn}
-LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally
+LSTM Networks were originally formulated in~\cite{Hoch97}, but their formulation has been incrementally
 updated in~\cite{Gers00} and~\cite{Gers2000}, and the most recent version is~\cite{Graves05}.
-One of the inital proposers of LSTM, Prof. Jurgen Schmidhuber, did a survey on the most common variations
-of the model last year~\cite{Greff15}, which is the reference for this short discussion.
+One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations
+of the model~\cite{Greff15}, which is the reference for this short discussion.
 
-A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As it can be seen in the picture,
+A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As can be seen in the picture,
 we still have the recurrent connections from the regular RNNs, but now there are multiple entry points
 that control the flow of information through the network. Although omitted from the picture, all the
-gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}. The main components, their
-role and relevance, are explained as follows.
+gates are biased, as defined in Equations~\ref{eq:equationsLSTM}. The role and relevance of the main components can
+be summarized as follows.
 
@@
 	\centering
-	\includegraphics[width=2.5in]{figures/lstmneuron.png}
+	\includegraphics[width=1.1\linewidth]{figures/lstmneuron.png}
     \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
@@
 the input vector at time $t$, $\mb{x}^{(t)}$, and the output vector at the previous time step, $\mb{y}^{(t-1)}$,
-are weighed in, producing an output $\mb{i}^{(t)}$.
+are weighed, producing an output $\mb{i}^{(t)}$.
 
@@
 from the input gate to the memory cell. It also receives the input vector and the previous output
-set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either},
-but the most common choice is the \textbf{hyperbolic hangent}.
+set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can both} the logistic
+sigmoid, $\sigma(x) = 1 \over (1+e^{-x})$, or the hyperbolic tangent, $\tanh(x)$ but the most common choice is the
+\textbf{hyperbolic hangent}.
 
     \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to
-set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time
+set or reset them, using the \textit{Hadamard} vector multiplication of its output at time
 $t$, $\mb{c}^{(t-1)}$. The activation function of this gate is \textbf{always sigmoid}, and the resulting
@@
     \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of
-the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates
+the neuron, where the previous state is kept, and updated according to the dynamics of the gates
 that connect to it. Also, this is where the peephole connections come from.
@@
     \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation
-function that, as the gate activation function, can be any, but the \textbf{hyperbolic tangent}
-is the most common choice.
+function that, as the gate activation function, is the \textbf{hyperbolic tangent}.
 
-    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates
+    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell to the gates, that allow them
 to `peep' at the states of the memory cell. They were added after the initial 1997 formulation, and their
-absence was proven to have a minimal performance impact~\cite{Greff15}. For this reason, they were ommitted
+absence was proven to have a minimal performance impact~\cite{Greff15}. For this reason, they were omitted
 in this architecture.
@@
 
-The operation of each set of gates of the layer is given by the following set of equations
+The operation of each set of gates of the layer is given by the following set of equations, where vectors are
+represented by bold, lower-case letters, and matrices are bold, upper-case letters
 \begin{eqnarray}
@@
     \mb{c}^{(t)} & = & \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
-    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
+    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM},
 \end{eqnarray}
@@
 precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent
-weight matrices $\mb{R}_*$ are square matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$
-and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.
+weight matrices $\mb{R}_*$ are square matrices of size $N \times N$, and that the bias weight vectors $\mb{b}_*$
+and the vectors $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N$.
 
 \section{Related Work}\label{sec:relwork}
-LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is
+LSTM Networks are nowadays one of the state-of-the-art algorithms in Deep Learning, and their performance is
 superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address
@@
 from time-series data. A very comprehensive description of applications can be found in one of the initial
-authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/~juergen/rnn.html}.
+authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/\~{}juergen/rnn.html}.
 
 The uses of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass
-HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for
+HMM-based models for optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for
 instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark,
@@
 to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
-in~\cite{Vinyals14},~\cite{Donahue14} and~\cite{Donahue14_2}.
+in~\cite{Vinyals14, Donahue14, Donahue14_2}.
 
 Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
-(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented,
-with 128 neurons each, which, for processing 1000 samples, yielded an execution time of $0.932 \, \si{\second}$. Assuming
+(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented in FPGA,
+with 128 neurons each, which, for processing 1000 samples, yielded an execution time of \SI{0.932}{\second}. Assuming
 that both layers are equal, and thus have the same execution time, this yields an approximate execution time of
 $29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$).
-Therefore, this network is able to perform around 34.3 thousands of forward propagations per second.
-Therefore ts performance is well below this work, which sits at \textbf{0.487 millions} of forward propagations per second.
+Therefore, the network~\cite{Chang15} is able to perform around \num{34.3d3} forward propagations per second.
+Although the network is large, its performance is well below this work, which achieves \num{487d3} forward propagations per second, for that network size.
 This is because the work in~\cite{Chang15} does not have a full level of parallelism when compared to the proposed design of
-Section~\ref{sec:proparch}, and it makes use of external memory to store the weights. This last point is understandable,
-since the authors use the FPGA core as a co-processor for the main CPU, and not as a standalone implementation.
+Section~\ref{sec:proparch}, and it makes use of external memory to store the weights. On the other hand, the authors use it as a co-processor for the main CPU, and not as a
+standalone implementation.
 
@@
 In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05},
-since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
-number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles
+since evaluating a polynomial does not have high memory usage requirements (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
+number of multiplications needed is small enough to not pose a restriction both on resources (now DSP slices, and not memory) and on speed (number of clock cycles
 needed to output a result).
@@
 The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where
-the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as Equation~\ref{eq:remezline}
+the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which produces a system of $n+2$ linear equations such as Equation~\ref{eq:remezline}
 
@@
 \end{eqnarray}
-where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal assymptotes, which can be used as the value in the edges,
-but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
-as the value of the assymptote, the interval was further split in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
-The algorithm was run using Python, and targeted second degree polynomials for each interval. After that, the Verilog model was described, where the correct
+where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal asymptotes, which can be used as the value in the edges.
+Instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
+as the value of the asymptote, the interval was further split in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
+The algorithm was run using Python, and targeted second degree polynomials for each interval. The coefficients achieved for the Sigmoid and Hyperbolic Tangent functions
+are reported in Tables~\ref{tab:coefs-sigm} and ~\ref{tab:coefs-tanh}. The maximum approximation errors were, respectively, \num{1.408d-3} and \num{1.21d-2}.
+
+\begin{table}
+	\caption{Polynomial Coefficients for the Sigmoid approximation}
+	\label{tab:coefs-sigm}
+    \centering
+  \begin{tabular}{ | c | c | c | c | }
+    \hline
+    $p_0$      &  $p_1$     &  $p_2$        &  Interval  \\
+		\hline
+    0          & 0          &  0             & $x < -6$ \\
+    \hline
+    0.20323428 & 0.0717631  & 0.00642858     & $-6 \leq x \leq -3$ \\
+    \hline
+		0.50195831 & 0.27269294 & 0.04059181     & $-3 \leq  x \leq 0$ \\
+		\hline
+		0.49805785 & 0.27266221 & 0.04058115     &  $0 \leq  x \leq 3$ \\
+		\hline
+		0.7967568 & 0.07175359 & 0.00642671      & $3 \leq  x \leq 6$ \\
+		\hline
+		1         &  0         &   0             & $x > 6$ \\
+		\hline
+  \end{tabular}
+\end{table}
+
+\begin{table}
+	\caption{Polynomial Coefficients for the $\tanh(x)$ approximation}
+	\label{tab:coefs-tanh}
+    \centering
+  \begin{tabular}{ | c | c | c | c | }
+    \hline
+    $p_0$      &  $p_1$     &  $p_2$        &  Interval  \\
+		\hline
+    -1          & 0          &  0             & $x < -3$ \\
+    \hline
+    -0.39814608 & 0.46527859 &  0.09007576   & $-3 \leq x \leq -1$ \\
+    \hline
+		0.0031444 & 1.08381219 & 0.31592922      & $-1 \leq  x \leq 0$ \\
+		\hline
+		-0.00349517 & 1.08538355 & -0.31676793   &  $0 \leq  x \leq 1$ \\
+		\hline
+		0.39878032 & 0.46509003 & 0.09013554     & $1 \leq  x \leq 3$ \\
+		\hline
+		1         &  0         &   0             & $x > 3$ \\
+		\hline
+  \end{tabular}
+\end{table}
+The Verilog model was described, where the correct
 coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial
-was accomplished using the \textbf{Horner's Rule}, noting that
+was accomplished using the \textbf{Horner's Rule}, that is
 
@@
 repeated two times. Again the internal machine state controls which values are multiplied and added according to the pipeline state.
-The output of the Sigmoid HDL block  and the $\tanh$ HDL block are presented in Figures~\ref{fig:nonlin-out-sig}
-and~\ref{fig:nonlin-out-tanh}, respectively. Due to the internal datapath pipeline, the module takes 5 clock cycles to output a result.
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
-    \caption{Plot of the output of the Sigmoid Calculator HDL module}
-    \label{fig:nonlin-out-sig}
-\end{figure}
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
-    \caption{Plot of the output of the Hyperbolic Tangent HDL module}
-    \label{fig:nonlin-out-tanh}
-\end{figure}
+Due to the internal datapath pipeline, each module takes 5 clock cycles to output one result.
 
 \subsection{Matrix-vector Dot Product Calculation}\label{sec:proprarch_dot}
-From Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are
-multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way,
-this block implements matrix-vector multiplication to perform those calculations, and is parameterized in
+From the Set of Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are
+multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This block implements
+matrix-vector multiplication to perform those calculations, and its description is parameterized in
 order to accommodate networks of various sizes, since $\mb{W}_*$ has size $N\times M$, and $\mb{R}_*$
-has size $N\times N$ -- $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$
-(the number of neurons in the layer). This way, if a layer with different parameters is needed, either in terms of
+has size $N\times N$, because $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$
+(the number of neurons in the layer). If a layer with different parameters is needed, either in terms of
 number of inputs or number of neurons, we only need to change the respective parameter before the synthesis stage,
@@
 This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel},
-each only requiring the corresponding $i$-th line from the matrix. Following this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
+each one only requiring the corresponding $i$-th line from the matrix. Using this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
 which is one of the great advantages of custom-tailored hardware solutions.
@@
 size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must
-be used to ensure the flexibility of the solution to accomodate networks of larger dimensions. The solution found for this
-design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg},
-each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final
+be used to ensure the flexibility of the solution to accommodate networks of larger dimensions. The solution found for this
+design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{alg:matvec-alg},
+each multiplication slice is responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final
 result for the vector would be ready in $M$ clock cycles (i.e. the number of columns); now, defining a
-parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the
+parameter $K_G$, such that
+
+\begin{equation}\label{eq:kg}
+K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}.
+\end{equation}
+This parameters represents the number of rows that share the same multiplier. The
 same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
@@
 
-Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication
+Figures~\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication
 units within the module, respectively, where $K_G = 4$ and for the same matrix and vector sizes as before. Note that in this situation,
@@
 column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
-The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a
-fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
-of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output vector position, respectively.
-In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+
+The dark shaded part of the memory is used by the first multiplier, and the light shaded one is used by the other, in parallel for a
+fixed \verb+rowMux+. This signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
+of Figure~\ref{fig:array-prod}, that allows choosing the proper position of the weight column and writing to the correct output vector position, respectively.
+In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluates $y_0$ and $y_4$. After this, \verb+rowMux+
 is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time
-proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
-in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.
+proportional to $K_G \times M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
+in the \emph{next clock cycle}, and this module is pipelined both at the input and at the output, in order to increase the maximum clock frequency.
 
@@
 \subsection{Gate Module}\label{sec:proparch_gate}
-The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we
-note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network,
-we can omit them in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks
+The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. This way,
+each Gate module needs to perform three tasks
 
@@
 Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication
-in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
-This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.
-The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result.
+in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles. This way, tasks 1 and 2 can be performed in parallel, and we can use the extra
+time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time. The module is triggered by a
+\verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded,
+and the value at the output is the final result.
 After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
@@
 Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they
-operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel. This way we
+operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel. For that, we
 need four Gate Modules working in parallel, each one with its respective weight RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation
@@
 Furthermore, we can avoid a naive translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers
-and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output
-of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. This way, instead of replicating
-these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in. The
+and activation function calculators) and require more area to save a negligible number of clock cycles, by noting that one of the operands is the output
+of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. Instead of replicating
+these `$\tanh$-$\sigma$-$(\cdot)$wise' structures, we use a \emph{single one} and choose the operand according to the state that the network is currently in. The
 issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses
@@
 The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine,
-and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place.
+and its value is incremented after each complete usage of the `$\tanh$-$\sigma$-$(\cdot)$wise' structure: this is where the time multiplexing of the structure takes place.
 Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$,
@@
 
-The flip-flops on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network's state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
-The first activated flip-flop is the middle one, which keeps the result from the elementwise mutiplication of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the $\tanh$-$\sigma$-$(\cdot)$wise structure, the bottom flip-flop saves the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal.
-Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.
+The registers on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network's state machine that enable the appropriate register, placing the
+output from the elementwise multiplier in the correct place. The first activated register is the middle one, which keeps the result from the elementwise mutiplication of the $\mb{z}^{(t)}$
+and $\mb{i}^{(t)}$ vectors, then, after a full operation of the `$\tanh$-$\sigma$-$(\cdot)$wise' structure, the bottom register saves the other portion of the sum that evaluates to the $\mb{c}^{(t)}$
+signal. Lastly, the top register saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.
 
@@
 \begin{equation}\label{eq:numdsp_network-opt}
-    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right).
+    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right),
 \end{equation}
-where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact.
-In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion.
-After adjustments to the state machine, and accounting for pipelining and synchronization within the datapath, the clock cycles needed after the gate module calculations are 27, so the total clock cycles needed are
+where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact. In terms of speed performance, although the Gate
+calculation time remains the same, now the `$\tanh$-$\sigma$-$(\cdot)$wise' structure runs for 3 consecutive times. After adjustments to the state machine, and accounting for
+pipelining and synchronization within the datapath, the number of clock cycles needed after the gate module calculations is 27, so the total clock cycles needed to perform a
+complete forward propagation are
 
@@
 \end{equation}
-which is only 13 clock cycles more than a fully-parallel, naive architecture. For instance, an $N=32$ neuron network would
-require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
+which is only 13 clock cycles more than a fully-parallel, naïve architecture. For instance, an $N=32$ neuron network would
+require 320 DSP slices, while this new architecture use 160, at the expense of 13 more clock cycles.
 
@@
 both for the forward propagation of the network, as well as for the training algorithm. Python and Numpy were used rather than MATLAB,
-since the former has higher performance at the same level of code complexity. The model was trained using the SPSA method~\cite{Spall98}.
+since the former has higher performance for the same level of code complexity. The model was trained using the SPSA method~\cite{Spall98}.
 
@@
 
-After the software network was trained, the trained weights were loaded into the network and several addition problems
+The software network was trained for 50 epochs. In each epoch, 5000 training sequences were presented, and then tested with 100 sequences, where the
+prediction error was evaluated for that epoch. The trained weights were loaded into the network and several addition problems
 were fed to the Network in the testbench, which yielded no errors.
@@
 \subsection{Synthesis}\label{sec:results_synth}
-The proposed network was first synthesized aiming a Xilinx XC7Z020 SoC, for sizes $N$ of 4, 8, 16 and 32, varying the resource sharing parameter $K_G$, and the number of inputs set
+The proposed network was first synthesized for a Xilinx XC7Z020 SoC, for sizes $N \in \left\{4, 8, 16, 32\right\}$, varying the resource sharing parameter $K_G$, while keeping the number of inputs set
 to $M=2$. For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were
-successfully synthesized. This is because of the complexity of the input multiplexer of the Matrix-vector Dot Product Module in~\ref{sec:proprarch_dot}. To synthesize the
-design for sizes $N$ of 64, 128 and 256, a Virtex-7 VC707 board was used.
+successfully synthesized. This is because of the complexity of the sharing logic for the multipliers of the Matrix-vector Dot Product Module in~\ref{sec:proprarch_dot}. To synthesize the
+design for sizes $N \in \left\{64, 128 \right\}$ a Virtex-7 VC707 board was used, which as a XC7VX485T (speed grade -2) FPGA core.
 
@@
 the slowest one is an $N=32$ and $K_G=4$ network, clocked at $101.523$ MHz. The reference design used for validation in
-Section~\ref{sec:res-val} is an $N=8$ and $K_G=2$ network is clocked at $154.321$ MHz, which yields a clock period of
-$6.48$ ns.
+Section~\ref{sec:res-val} is an $N=8$ and $K_G=2$ network clocked at $154.321$ MHz, which yields a clock period of
+\SI{6.48}{\nano\second}. The maximum clock achievable for the VC707 networks is reported in Table~\ref{tab:maxfreq-virtx7}
 
@@
 \end{figure}
-
-As previously stated, a synthesis targeting a Virtex-7 VC707 platform was performed, in order to synthesize networks of
-larger size that, due to the LUT usage restriction, could not be synthesized in the XC7Z020. The maximum clock achievable
-for each network size is reported in Table~\ref{tab:maxfreq-virtx7}
 
@@
     \hline
-    $N=256$ &  ??? MHz \\
 \hline
@@
 \subsubsection{DSP Slices Usage}\label{sec:results_synth_dsp}
-The estimates made in Equation~\ref{eq:numdsp_network-opt} were proven to be accurate, as Figure~\ref{fig:dspused} confirms.
+The estimates made in Equation~\ref{eq:numdsp_network-opt} were shown to be accurate, as Figure~\ref{fig:dspused} confirms.
 The reference network design with $N=8$ uses 56 DSP slices, which corresponds to $25.45\%$ of the total number of DSP slices available.
@@
 
-\subsubsection{Power Consumption}\label{sec:res-synth-power}
-Another important metric of the performance of a design is its \textbf{power consumption}. All  XC7Z020 designs yielded a
-constant baseline value for the \emph{static} power consumption of around 120 mW, and the power consumption reported
-in Figure~\ref{fig:power} refers to the \emph{total} power consumptions, i.e. both the baseline static power and the dynamic power consumption.
-It is clear that the smaller the network is, the less power is consumed, as one would expect. Furthermore, an increasing
-level of resource sharing yields a substantially lower power consumption figure: this is because less DSP slices are used as $K_G$ increases.
-Of course, even though the power consumption is lower in that case, that comes at the expense of a lower clock frequency and more clock cycles
-elapsed per forward propagation.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{figures/power.eps}
-    \caption{Power Consumption estimates for several Network Sizes $N$ and resource sharing level $K_G$}
-    \label{fig:power}
-\end{figure}
-
-For the synthesized networks on the VC707, the power consumption figures are reported in Table~\ref{tab:power-virtx7}. Being a high-end
-system, and since the network size is larger, the power consumption is higher than before, but still within reasonable levels given
-the complexity of the system.
-
-\begin{table}
-	\caption{Power Consumption estimates for several Network Sizes $N$ in the VC707}
-	\label{tab:power-virtx7}
-    \centering
-  \begin{tabular}{ | l | c | }
-    \hline
-     & $Power$  \\
-    \hline
-    $N=64$ & 1.34 W \\
-    \hline
-    $N=128$ & 1.509 W\\
-    \hline
-    $N=256$ & ??? W \\
-\hline
-  \end{tabular}
-
-\end{table}
-
 \subsubsection{Other Resources Usage}\label{sec:res-synth-otheres}
-The LUT, LUTRAM and Flip-Flop usages are discussed here. In Table~\ref{tab:lut}, the usage of LUTs is reported, where we can see
+The LUT, LUTRAM and Flip-Flop usages are discussed here. In Table~\ref{tab:lut}, the usage of LUTs is reported. We can see
 that although there is not a clear trend on how the LUT usage varies with increasing $K_G$, it is clear, and expectable, that
@@
 
-As for LUTs, the FF usage also scales accordingly to a $2\times$ factor. In terms of LUTRAM, used to store the network weights,
+As for LUTs, the FF usage also scales according to a $2\times$ factor. In terms of LUTRAM, used to store the network weights,
 the amount used increases by 2 with increasing values of $N$, as before, and \textbf{does not} depend on $K_G$. This is because
 the amount of weights only depends on the network sizes $M$ and $N$, and not on $K_G$. Furthermore, unlike LUTs, it scales well
-with increasing network sizes, and does not pose a limitation on the network size. The usage results for the XC7Z020 are reported in Table~\ref{tab:ff},
-and in Table~\ref{tab:ff-virtx7} the usages for the VC707.
+with increasing network sizes, and does not pose a limitation on the network size. The usage results for the XC7Z020 are reported in Table~\ref{tab:ff};
+the results for the VC707 are in Table~\ref{tab:ff-virtx7}.
 
@@
 \hline
-    $N=256$ & 8.83\%  & 13.36\% & ??.?? \%\\
-\hline
   \end{tabular}
-
 \end{table}
@@
 \subsection{Performance}
-To evaluate the performance, a metric was defined based on how many predictions it can produce per second (i.e. produce a new result bit in the output sequence),
-in \textbf{millions}. The metric on~\cite{Chang15}, as discussed in Section~\ref{sec:relwork}, is pretty much pointless, since
+To evaluate the throughput of the system, a metric was defined based on how many predictions it can produce per second (i.e. produce a new result bit in the output sequence),
+in \textbf{millions}. The metric in~\cite{Chang15}, as discussed in Section~\ref{sec:relwork}, is not very informative, since
 although a system can perform many calculations per second, if those operations are redundant, that metric has no relevant information
 regarding \textbf{how fast} the system can \textbf{perform the actual task it was meant to do}, which in this case is a complete Forward
-Propagation through the network. The prediction time is the time elapsed since a new input vector is applied \emph{to} the moment the LSTM
+Propagation through the network. The prediction time is the time elapsed from the moment a new input vector is applied \emph{to} the moment the LSTM
 Network produces an output vector. Hence, we multiply the number of clock cycles yielded by Equation~\ref{eq:numcc_network-opt} by the equivalent clock period
-from the synthesis clock report of Figure~\ref{fig:maxfreq}.
-This result is epitomized in Table~\ref{tab:process-time}, where the calculation time of the Python module's Forward Propagation function is
-also included. This time was measured using the \textbf{timeit}\footnote{https://docs.python.org/3.5/library/timeit.html}  module, that allows
+from the synthesis clock report of Figure~\ref{fig:maxfreq}. Nevertheless, a comparison on the number of Millions of operations both systems do will be presented further ahead.
+This result is reported in Table~\ref{tab:process-time}, where the calculation time of the Python module's Forward Propagation function is
+also included. This time was measured using the \textbf{timeit}\footnote{https://docs.python.org/3.5/library/timeit.html}  module, which allows
 the evaluation of the execution time of small pieces of code as well as complete functions with arguments. The Python code was run on a Linux System,
-powered by an i7-3770k Intel Processor, running at 4.2GHz.
+powered by an i7-3770k Intel Processor, running at 4.2GHz, with 8GB of RAM.
 
@@
 forward propagation takes 1.14 $\mu$s and 2.052 $\mu$s respectively, and for both the maximum clock frequency achievable was 140.845 MHz. Since the
-design in~\cite{Chang15}, for $N=128$, takes an estimated 29.13 $\mu$s (see Section~\ref{sec:relwork}), our design yields an improvement of $14\times$
+design~\cite{Chang15}, for $N=128$, takes an estimated \SI{29.13}{\micro\second} (see Section~\ref{sec:relwork}), our design yields an improvement of $14\times$
 over it.
+In terms of millions of operations per second (Mops), for an $N=128$ network as the one of~\cite{Chang15}, our work achieves \num{4534.8} Mops per second
+while theirs only achieves \num{264.4} Mops per second. This is because, as stated in~\cite{Chang15}, a network of this size performs \num{132.1d3} Mops, and since
+our work outputs a sample every $1 \over \SI{29.13}{\micro\second}$, we have that 4534.8 Mops per sec. $=\frac{\num{132.1d3} \; \text{Mops}}{\SI{29.13}{\micro\second}}$.
+
+\subsubsection{Power Consumption}\label{sec:res-synth-power}
+Another important metric of the performance of a design is its \textbf{power consumption}. These power consumption estimates are post Place\&Route, and
+have a \textbf{medium} confidence level. All  XC7Z020 designs yielded a constant baseline value for the \emph{static} power consumption of around 120 mW,
+and the power consumption reported
+in Figure~\ref{fig:power} refers to the \emph{total} power consumptions, i.e. both the baseline static power and the dynamic power consumption.
+It is clear that the smaller the network is, the less power is consumed, as one would expect. Furthermore, an increasing
+level of resource sharing yields a substantially lower power consumption figure: this is because less DSP slices are used as $K_G$ increases.
+Of course, even though the power consumption is lower in that case, that comes at the expense of a lower clock frequency and more clock cycles
+elapsed per forward propagation.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/power.eps}
+    \caption{Power Consumption estimates for several Network Sizes $N$ and resource sharing level $K_G$}
+    \label{fig:power}
+\end{figure}
+
+For the networks synthesized on the VC707, the power consumption figures are reported in Table~\ref{tab:power-virtx7}. Since the device is bigger,
+and since the network size is larger, the power consumption is higher than before, but still within reasonable levels given
+the complexity of the system.
+
+\begin{table}
+	\caption{Power Consumption estimates for several Network Sizes $N$ in the VC707}
+	\label{tab:power-virtx7}
+    \centering
+  \begin{tabular}{ | l | c | }
+    \hline
+     & $Power$  \\
+    \hline
+    $N=64$ & 1.34 W \\
+    \hline
+    $N=128$ & 1.509 W\\
+\hline
+  \end{tabular}
+
+\end{table}
+
+All these estimates were provided by Vivado, and have a \textbf{medium} confidence level.
 
 \section{Conclusion}\label{sec:concl}
-The LSTM Hardware architecture presented surpassed the performance of current software implementations
-and also the only current hardware implementation by $14\times$, and making use solely of internal FPGA
-resources, unlike the other hardware implementation available, making it a general architecture that can
-be used in any FPGA, and not only limited to a particular one. Furthermore, the design proved to scale relatively
-well with the network size, being able to accommodate networks of sizes up to 256.
+The LSTM Hardware architecture presented surpassed the performance of the custom-built software implementation by $251\times$, at best,
+and also the only current hardware implementation by $14\times$, and solely making use of internal FPGA
+resources, achieving a higher level of parallelism. The higher levels of parallelism of this work are achieved at the cost of increasing
+design complexity, which limits its scalability to higher sized networks, unlike the implementation of Chang et al.~\cite{Chang15}.
+On the other hand, the HDL description of this work is parameterized, and is thus very flexible for networks of any size, not requiring
+a redesign of the system every time a differently sized network is required. Furthermore, making use of internal memory makes it suitable
+for including an on-chip learning system that can perform training on the network weights.
 
-Given these results, this architecure places itself as the current state of the art system in LSTM Neural Networks
-implementations in software.
+Given these results, this architecture advances the current state of the art in LSTM Neural Networks
+hardware implementations, providing the most efficient implementation to date.
 

### 2016-07-25 shortPaper/paper.tex
Final Version of the Text and Short Paper

--- 2016-07-06 paper/paper.tex
@@
 \newcommand{\mb}[1]{\mathbf{#1}}
+\renewcommand\IEEEkeywordsname{Keywords}
 
-\title{An FPGA Implementation of a Long Short-Term Memory Neural Network}
+\title{FPGA Implementation of a LSTM Neural Network}
 
 %% COMMENT THIS FOR SUBMISSION
-\author{\IEEEauthorblockN{Jose Pedro Castro Fonseca} \\ 
-\IEEEauthorblockA{Faculty of Engineering University of Porto \\
-Porto, Portugal}}
-\IEEEspecialpapernotice{(Working Version)}
+\author{
+\IEEEauthorblockN{José Fonseca\IEEEauthorrefmark{1}, João Canas Ferreira\IEEEauthorrefmark{1}, Ivo Timóteo\IEEEauthorrefmark{2}}\\
+\IEEEauthorblockA{FEUP\IEEEauthorrefmark{1}, University of Cambridge\IEEEauthorrefmark{2}}
+}
+%%\IEEEspecialpapernotice{(Working Version)}
 
@@
 \begin{abstract}
-This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
-to outperform software implementations, by exploiting the inherent parallelism in it.  
+Our work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
+to outperform software implementations, by exploiting its inherent parallelism.
 The main design decisions are presented, along with the proposed network architecture. A description of the main
-building blocks of the network is also presented. The network is synthesized for various sizes,
-and performance results are presented and analysed. The synthesized network achieves a 195 times speed-up
-over a custom-built software network, proving the benefits of parallel computation for this kind of network.
+building blocks of the network is also presented. The network is synthesized for various sizes and platforms,
+and the performance results are presented and analyzed. Our synthesized network achieves a 251 times speed-up
+over a custom-built software network, running on a Desktop computer, and a 14 times speed up over the current state of the art.
 \end{abstract}
+
+\begin{IEEEkeywords}
+Neural Networks, Long Short-Term, FPGA, Reconfigurable Hardware, Machine Learning
+\end{IEEEkeywords}
 
@@
 \IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
-particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain 
-time step are also fed as the input in the next time step, and that way their memory elements can make
-sense of patterns within data sequences, unlike classical recursive neural networks.
+particular type of network, a Long Short-Term Memory (LSTM) Network, is a recursive neural network, with memory.
 These algorithms have been profusely implemented in software, and their practical applications are plentiful.
 However, the benefits of the inherent parallelism offered by a dedicated hardware
-platform are not availed, and there are relatively few implementations of Machine Learning algorithms in 
+platform are not exploited, and there are relatively few implementations of Machine Learning algorithms in
 these kind of platforms.
 
-Hardware platforms would achieve a considerable speed-up over software implementations, which could prove
-useful for high data throughput systems, where the calculation overhead is critical and limits the performance of the systems.
-Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
-of a given experiment faster than in a regular CPU, yielding an increase in scientific productivity.
+\section{LSTM Neural Networks}\label{sec:lstmnn}
+LSTM Networks were originally formulated in~\cite{Hoch97}, and their operation is detailed in Equations~\ref{eq:equationsLSTM}.
 
-Hitherto, there is only one implementation~\cite{Chang15}, but its performance is undermined
-by the external memory access. This implementation aims to make use of internal FPGA memory resources, and therefore
-achieve a higher speed performance.
-
-In overview, this article begins by explaining what are LSTM networks -- and their mathematical 
-details -- in Section~\ref{sec:lstmnn}, followed by providing a quick overview of their state 
-of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed 
-hardware architecture and its main constituent modules, and the performance and synthesis 
-results are reported in Section~\ref{sec:results}. A final concluding remark is given in Section~\ref{sec:concl}.
-
-\section{LSTM Neural Networks}\label{sec:lstmnn}
-LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally 
-updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. 
-One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations 
-of the model last year~\cite{Greff15}, which is the reference for this short discussion.
-
-A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As it can be seen in the picture, 
-we still have the recurrent connections from the regular RNNs, but now there are multiple entry points 
-that control the flow of information through the network. Although omitted from the picture, all the 
-gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}. The main components, their 
-role and relevance, are explained as follows.
-
-\begin{figure}[!t]
-	\centering
-	\includegraphics[width=2.5in]{figures/lstmneuron.png}
-    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
-	\label{fig:lstmneuron}
-\end{figure}
-
-\begin{itemize}
-    \item \textbf{Input Gate} -- this is the input gate, where the relative importance of each feature of 
-the input vector at time $t$, $\mb{x}^{(t)}$, and the output vector at the previous time step, $\mb{y}^{(t-1)}$, 
-are weighed in, producing an output $\mb{i}^{(t)}$.
-
-    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information 
-from the input gate to the memory cell. It also receives the input vector and the previous output 
-set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either}, 
-but the most common choice is the \textbf{Hyperbolic Tangent}.
-
-    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to 
-set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time 
-$t$, $\mb{c}^{(t)}$, with the contents of the memory unit at the previous time step, $\mb{c}^{(t-1)}$. 
-The activation function of this gate is \textbf{always sigmoid}, and the resulting signal is $\mb{f}^{(t)}$.
-
-    \item \textbf{Output Block Gate} -- this gate has a role very similar to that of the Block Input 
-Gate, but now it controls the information flow \textit{out} of the LSTM neuron, namely the activated 
-Memory Cell output. The control signal it produces is $\mb{o}^{(t)}$.
-
-    \item \textbf{Memory Cell} -- the cornerstone of the LSTM neuron. This is the memory element of 
-the neuron, where the previous state is kept, and updated accordingly to the dynamics of the gates 
-that connect to it. Also, this is where the peephole connections come from.
-
-    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation 
-function that, as the gate activation function, can be any, but the \textit{hyperbolic tangent} 
-is the most common choice.
-
-    \item \textbf{Peepholes} -- direct connections \textit{from} the memory cell that allow for gates 
-to `peep' at the states of the memory cell. They were added after the initial 1997 formulation, and their 
-absence was proven to have a minimal performance impact~\cite{Greff15}. For this reason, they were ommitted 
-in this architecture.
-
-\end{itemize}
-
-The operation of each set of gates of the layer is given by the following set of equations
 \begin{eqnarray}
-    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
-    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
-    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
-    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
-    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
-    \mb{y}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
+    \mb{z}^{(t)} & = & g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
+    \mb{i}^{(t)} & = & \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
+    \mb{f}^{(t)} & = & \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
+    \mb{o}^{(t)} & = & \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
+    \mb{c}^{(t)} & = & \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
+    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM},
 \end{eqnarray}
-where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds 
-to the value of the $i$-th neuron in the layer, which is a very convenient and compact representation of the 
-whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that 
-precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent 
-weight matrices $\mb{R}_*$ are squre matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$ 
-and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.
-
-\section{Related Work}\label{sec:relwork}
-LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is
-superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address 
-the same set of problems where LSTM are employed, namely predicting and producing classification decisions 
-from time-series data. A very comprehensive description of applications can be found in one of the initial 
-authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/~juergen/rnn.html}.
-
-The use of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass 
-HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for 
-instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark, 
-and also in \textbf{Large Scale Acoustic Modelling of Speech}~\cite{Sak14}. Other uses include 
-\textbf{Handwriting Synthesis}~\cite{Graves13_2}, \textbf{Translation}~\cite{Sustkever14}, \textbf{Biomedical Applications} such
-as protein homology detection~\cite{Hochreiter07}, \textbf{Music Analysis and Composition}, such as the transcription of piano music 
-to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
-in~\cite{Vinyals14},~\cite{Donahue14} and~\cite{Donahue14_2}.
+where $\odot$ is the Hadamard multiplication. A layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that
+precedes this).
 
 Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
-(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented,
-with 128 neurons each, which, for processing 1000 samples, yielded an execution time of $0.932 \, \si{\second}$. Assuming 
-that both layers are equal, and thus have the same execution time, this yields an approximate execution time of 
-$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$). 
-Therefore, this network is able to perform around 34.3 thousands of forward propagations per second. 
-Although the network is large, its performance is well below this work, because it does not have
-a full level of parallelism when compared to the proposed design of Section~\ref{sec:proparch}, and it
-makes use of external memory to store the weights. This last point is understandable, since the authors use
-the FPGA core as a co-processor for the main CPU.
+(March 2016) by Chang et al.~\cite{Chang15}.
 
 \section{Proposed Architecture}\label{sec:proparch}
-Before presenting the 
-\subsection{Activation Function Calculation}\label{sec:proprarch_af}
-In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).
 
-The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as~\ref{eq:remezline}
+\subsection{Network Architecture}\label{sec:proprarch_net}
+Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they
+operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel. Furthermore, we
+can avoid a naive translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers
+and activation function calculators) and require more area to save a negligible number of clock cycles, by noting that one of the operands is the output
+of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. Instead of replicating
+these `$\tanh$-$\sigma$-$(\cdot)$wise' structures, we use a \emph{single one} and choose the operand according to the state that the network is currently in. The
+issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses
+between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$.
 
-\begin{equation}\label{eq:remezline}
-    p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
+The total requirement for DSP slices is
+\begin{equation}\label{eq:numdsp_network-opt}
+    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right).
 \end{equation}
-where $i \in \left[0, n+1\right]$.
-Of course, both functions have horizontal assymptotes, which can be used as the value in the edges, but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function as the value of the assymptote, I have further split the interval in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
-The algorithm was run using Python, and targeted second degree polynomials for each interval.
-After that, the Verilog model was described, where the correct coefficients are loaded according to the interval where the incoming operand is located.
-The evaluation of the polynomial was accomplished using the \textbf{Horner's Rule}, noting that
 
-\begin{equation}\label{eq:factorPol}
-p(x) = p_0 + p_1x + p_2x^2 = p_0 + x(p_1 + xp_2)
+On the other hand, the number of clock cycles needed to output a result is simply
+\begin{equation}\label{eq:numcc_network-opt}
+    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G.
 \end{equation}
-and, therefore, the calculation is simply the procedure of multiplying the operand by a value and adding a constant to the result,
-repeated two times. Again the internal machine state controls which values are multiplied and added according to the pipeline state.
-The output of the Sigmoid HDL block  and the $\tanh$ HDL block are presented in Figures~\ref{fig:nonlin-out-sig} 
-and~\ref{fig:nonlin-out-tanh}, respectively. Due to the internal datapath pipeline, the module takes 5 clock cycles to output a result.
 
-\begin{figure}[!t]
+\section{Results}\label{sec:results}
+\subsection{Validation}\label{sec:res-val}
+
+The functionality of the network was verified against a Python model of an LSTM network that was developed as a reference,
+both for the forward propagation of the network, as well as for the training algorithm.
+
+The learning problem presented to both the software and hardware network is the \textbf{addition} of two binary numbers of 8 bits. The $i$-th
+bit of each number is fed to the network as a vector, and the network outputs its prediction of the correct value of the $i$-th bit
+of the \emph{result}. After the whole number is processed, the memory cells of the LSTM network are reset and a new addition
+task can be presented to the network.
+
+\subsection{Synthesis}\label{sec:results_synth}
+The proposed network was first synthesized for a Xilinx XC7Z020 SoC, for sizes $N \in \left\{4, 8, 16, 32\right\}$, varying the resource sharing parameter $K_G$, while keeping $M=2$.
+For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were
+successfully synthesized. To synthesize the design for sizes $N \in \left\{64, 128 \right\}$ a Virtex-7 VC707 board was used, which has a XC7VX485T (speed grade -2) FPGA core.
+
+\subsubsection{Maximum Frequency}\label{sec:results_synth_maxfreq}
+For $N=4$, since there are only 4 rows to be multiplied, the maximum value of $K_G$ is 4, and hence no synthesis was performed for $K_G = 8$ (N.A.); also,
+since $K_G = 2$ and $N=32$, it would use $32(8/2+3) = 224$ DSP slices, and that exceeds the 220 slices
+available in the XC7Z020, so there is no synthesis data (N.D.), as well.
+
+Increasing $K_G$, the maximum clock speed decreases, and that decrease is steeper for larger values
+of $K_G$. This means that there is a critical path in the Matrix-Vector multiplication unit, whose multiplexer becomes
+increasingly complex. On the other hand, when $K_G$ is the same, smaller networks are faster
+than larger networks. The fastest design is an $N=4$ and $K_G = 2$ network, with a clock frequency of $158.228$ MHz, and
+the slowest one is an $N=32$ and $K_G=4$ network, clocked at $101.523$ MHz. The reference design used for validation in
+Section~\ref{sec:res-val} is an $N=8$ and $K_G=2$ network clocked at $154.321$ MHz, which yields a clock period of
+\SI{6.48}{\nano\second}. The maximum clock frequency for the VC707 was 140.854 MHz for both $N=64$ and $N=128$.
+
+\subsection{Performance}
+To evaluate the throughput of the system, a metric was defined based on how many predictions it can produce per second (i.e. produce a new result bit in the output sequence),
+in \textbf{millions}. Hence, we multiply the number of clock cycles yielded by Equation~\ref{eq:numcc_network-opt} by the equivalent clock period
+from the synthesis clock report for each network. The Python code was run on an i7-3770k Intel Processor, running at 4.2GHz, with 8GB of RAM.
+
+\begin{table}
+	\caption{Total processing time for a single forward propagation on the XC7Z020}
+	\label{tab:process-time}
     \centering
-    \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
-    \caption[Plot of the output of the Sigmoid Calculator HDL module]{Plot of the output of the Sigmoid Calculator HDL module}
-    \label{fig:nonlin-out-sig}
-\end{figure}
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
-    \caption[Plot of the output of the Hyperbolic Tangent HDL module]{Plot of the output of the Hyperbolic Tangent HDL module}
-    \label{fig:nonlin-out-tanh}
-\end{figure}
-
-\subsection{Matrix-vector Dot Product Calculation}\label{sec:proprarch_dot}
-From Equations~\ref{eq:equationsLSTM}, we see that the weight matrices $\mb{W}_*$ and $\mb{R}_*$ are 
-multiplied by the input vector $\mb{x}$ and the layer output vector $\mb{y}$, respectively. This way, 
-this block implements matrix-vector multiplication to perform those calculations, and is parameterized in 
-order to accommodate networks of various sizes, since $\mb{W}_*$ has size $N\times M$, and $\mb{R}_*$ 
-has size $N\times N$ -- $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$ 
-(the number of neurons in the layer). This way, if a layer with different parameters is needed, either in terms of 
-number of inputs or number of neurons, we only need to change the respective parameter before the synthesis stage, 
-instead of having to redesign the whole block for that particular size.
-
-The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}
-
-\begin{algorithm}
-\begin{algorithmic}
-\For {$i$ = $1:N$}
-    \For {$j$ = $1:M$}
-    \State $y_i := y_i + A_{ij} \cdot x_j$
-    \EndFor
-\EndFor
-\end{algorithmic}
-\caption{Matrix-vector multiplication of a matrix}
-\label{matvec-alg}
-\end{algorithm}
-This operation has a computational complexity of $O(n^2)$.
-Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix.
-Follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.
-
-Although this solution only requires one multiplication per row of the input matrix (i.e.
-$N$ multiplications), if the row size is large, we may run out of resources in the FPGA; therefore, some sort of \textit{resource multiplexing} strategy must be used to ensure the flexibility of the solution to accomodate networks of larger dimensions.
-The solution found for this design was to \emph{share} the multiplication slice among the rows of the matrix: in a direct implementation of Algorithm~\ref{matvec-alg}, each multiplication slice was responsible for producing the $i$-th element of the output vector $y$ (of size $N$), therefore the final result for the vector would be ready in $M$ clock cycles (i.e.
-the number of columns); now, defining a parameter $K_G = \frac{\text{Number of rows}}{\text{Number of multipliers}}$ -- the number of rows that share the same multiplier -- the same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
-For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector elements $y_0$, $y_2$, $y_4$ and $y_6$ would be ready after $M=2$ clock cycles, and the remaining -- $y_1$, $y_3$, $y_5$ and $y_7$ --  are ready after another two clock cycles, that is $2M = 4$ clock cycles after the calculations began.
-
-Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before.
-Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel.
-They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
-The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output register, respectively.
-In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$.
-After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$.
-Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.
+    \begin{tabular}{ | l | c | c | c | c | c | }
+    \hline
+    & $K_G=8$  & $K_G=4$ & $K_G=2$ & Python & Speed-up \\
+    \hline
+    $N=4$ & N.A.  & 309.68 ns  & 259.12 ns & 65 $\mu$s & $\times251$ \\
+    \hline
+    $N=8$ & 793.46 ns  & 421.12 ns  &  317.52 ns & 72 $\mu$s & $\times228$ \\
+    \hline
+    $N=16$ & 1.497 $\mu$s  & 738.19 ns  & 461.336 ns & 96 $\mu$s & $\times208$ \\
+    \hline
+    $N=32$ & N.D.          & 1.586 $\mu$s & N.A.       & 185 $\mu$s  &  $\times117$ \\
+		\hline
+  \end{tabular}
+\end{table}
+The hardware network is, at best, $\times251$ faster than the software counterpart,
+and at worst $\times117$ faster.
+Inverting the values in Table~\ref{tab:process-time}, we know how many forward propagations the network can produce, per second. These values are presented in Figure~\ref{fig:Mclass-psec}.
+While the $N=8$ and $K_G=2$ network is able to perform around 3.15 million predictions per second, the Python model can only output around 14 thousand predictions.
 
@@
     \centering
-    \includegraphics[width=2.5in]{figures/mem-array-prod.eps}
-    \caption[A column of the matrix that serves as input to the module]{A column of the matrix that serves as input to the module.
-The dark shaded part is for the first multiplier, and the light shaded is for the other, in parallel.
-The rowMux signal addresses the position within each shaded area}.
-    \label{fig:mem-arrayprod}
+    \includegraphics[width=\linewidth]{figures/Mclass-psec.eps}
+    \caption[Millions of classifications per second of each design according to the network size $N$]{Millions of classifications per second of each design according to the network size $N$.
+The comparison is between the software Python model and 3 networks of different levels of resource sharing $K_G$.}
+    \label{fig:Mclass-psec}
 \end{figure}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=2.5in]{figures/array-prod.eps}
-    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module.
-The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
-    \label{fig:array-prod}
-\end{figure}
-
-\subsection{Weight storage}\label{sec:proprarch_ram}
-The weights are stored in LUTRAM, and for that purpose, they are declared as a matrix of registers, where the access, both in terms
-of write and read operations, is made to each column. The write/read access can be performed at the same time, since the memory has
-separate communication ports for input/output, provided that the \textbf{addresses do not coincide}.
-
-\subsection{Gate Module}\label{sec:proparch_gate}
-The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$.
-If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices.
-This way, a Gate module needs to perform three tasks
-
-\begin{enumerate}
-    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
-    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
-    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
-\end{enumerate}
-Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles.
-This way, task 2 and task 1 can be performed in parallel, and we can use the extra time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time.
-The module is triggered by a \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded, and the value at the output is the actual final result.
-After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number of clock cycles this module takes to produce an output is $6+K_G\cdot N$.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=2.5in]{figures/gate.eps}
-    \caption[Diagram of the hardware block that implements the Gate]{Diagram of the hardware block that implements the Gate}
-    \label{fig:gate}
-\end{figure}
-
-\subsection{Network Architecture}\label{sec:proprarch_net}
-Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel.
-This way we need four Gate Modules working in parallel, each one with its respective weigth RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation function calculator (detailed in Section~\ref{sec:nonlincalc}).
-There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).
-
-Furthermore, we can avoid a naïve translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together.
-This way, instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in.
-The issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$.
-These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
-    \caption[Block diagram of the hardware LSTM Neural Network]{Block diagram of the hardware LSTM Neural Network}
-    \label{fig:network-opt}
-\end{figure}
-The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place.
-Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, another multiplexer was added before the elementwise multiplication, to select the $\mb{c}^{(t-1)}$ signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.
-
-The flip-flops on the right hand side of Figure~\ref{fig:network-opt} are activated by signals generated within the network's state machine that enable the appropriate flip-flop, placing the output from the elementwise multiplicator in the correct place.
-The first activated flip-flop is the middle one, which keeps the result from the elementwise mutiplication of the $\mb{z}^{(t)}$ and $\mb{i}^{(t)}$ vectors, then, after a full operation of the $\tanh$-$\sigma$-$(\cdot)$wise structure, the bottom flip-flop saves the other portion of the sum that evaluates to the $\mb{c}^{(t)}$ signal.
-Lastly, the top flip-flop saves the network output $\mb{y}^{(t)}$, which in the next incoming sample becomes $\mb{y}^{(t-1)}$ and is used by the Gate modules in this next batch of calculations.
-
-Now, since there is only a single elementwise multiplier and only two activation function calculators, the total requirement for DSP slices is simply
-\begin{equation}\label{eq:numdsp_network-opt}
-    4\frac{2N}{K_G} + 2N + N = N \left( \frac{8}{K_G} + 3 \right).
-\end{equation}
-where we see that we saved $5N$ multipliers, which for a large value of $N$ can have a decisive impact.
-In terms of speed performance, although the Gate calculation time remains the same, now the $\tanh$-$\sigma$-$(\cdot)$wise structure runs for 3 consecutive times, in a non parallel fashion.
-After adjustments to the state machine, and accounting for pipelining and synchronization within the datapath, the clock cycles needed after the gate module calculations are 27, so the total clock cycles needed are
-
-\begin{equation}\label{eq:numcc_network-opt}
-    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
-\end{equation}
-which is only 13 clock cycles more than a fully-parallel, naïve architecture.
-For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.
-
-\section{Results}\label{sec:results}
-
-\subsection{Validation}
-
-\subsection{Synthesis}
-
-\subsection{Performance}
+As for the larger-sized networks synthesized in the VC707, the results are also very promising. For network sizes of $N=64$ and $N=128$, a complete
+forward propagation takes 1.14 $\mu$s and 2.052 $\mu$s respectively, and for both the maximum clock frequency achievable was 140.845 MHz. Since the
+design~\cite{Chang15}, for $N=128$, takes an estimated \SI{29.13}{\micro\second}, our design yields an improvement of $14\times$
+over it.
+%In terms of millions of operations per second (Mops), for an $N=128$ network as the one of~\cite{Chang15}, our work achieves \num{4534.8} Mops per second
+%while theirs only achieves \num{264.4} Mops per second. This is because, as stated in~\cite{Chang15}, a network of this size performs \num{132.1d3} Mops, and since
+%our work outputs a sample every $1 \over \SI{29.13}{\micro\second}$, we have that 4534.8 Mops per sec. $=\frac{\num{132.1d3} \; \text{Mops}}{\SI{29.13}{\micro\second}}$.
 
 \section{Conclusion}\label{sec:concl}
+The LSTM Hardware architecture presented surpassed the performance of the custom-built software implementation by $251\times$, at best,
+and also the only current hardware implementation by $14\times$, and solely making use of internal FPGA
+resources, achieving a higher level of parallelism. The higher levels of parallelism of this work are achieved at the cost of increasing
+design complexity, which limits its scalability to higher sized networks, unlike the implementation of Chang et al.~\cite{Chang15}.
+On the other hand, the HDL description of this work is parameterized, and is thus very flexible for networks of any size, not requiring
+a redesign of the system every time a differently sized network is required. Furthermore, making use of internal memory makes it suitable
+for including an on-chip learning system that can perform training on the network weights.
+
+Given these results, this architecture advances the current state of the art in LSTM Neural Networks
+hardware implementations, providing the most efficient implementation to date.
 

### 2016-07-25 finalText/figures/maxfreqplot-virtx7.py

@@ -1,0 +1,22 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib import rc
+
+rc('text', usetex=True);
+rc('font', family='serif', serif='Times');
+
+f = np.array([65e-6, 72e-6, 96e-6, 185e-6])/(1e6)
+N = np.array([2, 3, 4, 5])
+
+fig = plt.figure()
+ax = fig.add_subplot(1,1,1)
+plt.plot(f, N, 'yo-', label="$N=4$")
+ax.set_xticks([2,3,4,5])
+plt.axis([0.8,3.2,90,160])
+plt.grid(True);
+plt.xlabel(r"$\log_2(N$)");
+plt.ylabel(r"Clock Speed in MHz");
+plt.legend(loc=1);
+plt.savefig(filename='maxfreq-virtx7.eps', transparent=True);
+plt.show()
+

### 2016-07-25 python/minimaxApprox/remez.py

@@ -13,7 +13,7 @@ from scipy.signal import argrelmax
 
 # The target approximation function
 def f(x):
-    return np.tanh(x)
+    return 1/(1+np.exp(-x))
 
 # Nber of iterations
 numIter = 8

### 2016-07-25 verilog/dot_prod/genData.py

@@ -15,7 +15,7 @@ def Qnm_to_real(real, n, m):
 def sign_ext(value, newSize, oldSize):
     if(value >= 2**(oldSize-1)):
         return value + 2**(newSize) - 2**(oldSize)
-    else: 
+    else:
         return value
 
 NUM_MATRICES = 100
@@ -46,7 +46,7 @@ for n in range(NUM_MATRICES) :
             Wq[i,j] = real_to_Qnm(W[i,j],QN,QM)
             fin_W.write("{0:018b}\n".format(int(Wq[i,j])))
 
-    y  = np.dot(W,x)    
+    y  = np.dot(W,x)
     yq = np.zeros_like(y)
 
     for i in range(NROW):
@@ -55,13 +55,13 @@ for n in range(NUM_MATRICES) :
 
     #yq = np.dot(Wq,xq)
     yrec = np.zeros_like(yq)
- 
+
     for i in range(NROW):
         #yq[i,0]   = int(yq[i,0]/(2**11)) & int(0x3ffff)
         yrec[i,0] = Qnm_to_real(yq[i,0],QN,QM)
         #fout.write("{0:018b}\n".format(int(yq[i,0]) & int(2**(QN+QM+1)-1)))
         fout.write("{0:018b}\n".format(real_to_Qnm(y[i,0],QN,QM)))
-    
+
     quantError += sum(y-yrec)
 
 print("Quantizaion Error: ", quantError/(NUM_MATRICES*NROW))

### 2016-07-25 verilog/gate/genData.py

@@ -15,11 +15,11 @@ def Qnm_to_real(real, n, m):
 def sign_ext(value, newSize, oldSize):
     if(value >= 2**(oldSize-1)):
         return value + 2**(newSize) - 2**(oldSize)
-    else: 
+    else:
         return value
 
 NUM_MATRICES = 50
-NROW         = 32
+NROW         = 16
 NCOL         = 4
 QN           = 6
 QM           = 11
@@ -52,7 +52,7 @@ for n in range(NUM_MATRICES) :
         xq[i,0] = real_to_Qnm(x[i,0],QN,QM)
         fin_x.write("{0:018b}\n".format(int(xq[i,0])))
 
-    for i in range(NROW) : 
+    for i in range(NROW) :
         bq[i,0] = real_to_Qnm(b[i,0],QN,QM)
         yq[i,0] = real_to_Qnm(y[i,0],QN,QM)
         fin_b.write("{0:018b}\n".format(int(bq[i,0])))
@@ -64,7 +64,7 @@ for n in range(NUM_MATRICES) :
             Wxq[i,j] = real_to_Qnm(Wx[i,j],QN,QM)
             fin_Wx.write("{0:018b}\n".format(int(Wxq[i,j])))
 
-    out  = np.dot(Wx,x) + np.dot(Wy,y) + b    
+    out  = np.dot(Wx,x) + np.dot(Wy,y) + b
     outq = np.zeros_like(out)
 
     for i in range(NROW):
@@ -75,13 +75,13 @@ for n in range(NUM_MATRICES) :
 
     #yq = np.dot(Wq,xq)
     outrec = np.zeros_like(outq)
- 
+
     for i in range(NROW):
         #yq[i,0]   = int(yq[i,0]/(2**11)) & int(0x3ffff)
         outrec[i,0] = Qnm_to_real(outq[i,0],QN,QM)
         #fout.write("{0:018b}\n".format(int(yq[i,0]) & int(2**(QN+QM+1)-1)))
         fout.write("{0:018b}\n".format(real_to_Qnm(out[i,0],QN,QM)))
-    
+
     quantError += sum(out-outrec)
 
 print("Quantizaion Error: ", quantError/(NUM_MATRICES*NROW))

### 2016-10-09 paper/paper.tex
Conference version of paper

@@
 sigmoid, $\sigma(x) = 1 \over (1+e^{-x})$, or the hyperbolic tangent, $\tanh(x)$ but the most common choice is the
-\textbf{hyperbolic hangent}.
+\textbf{hyperbolic tangent}.
 
@@
 with 128 neurons each, which, for processing 1000 samples, yielded an execution time of \SI{0.932}{\second}. Assuming
-that both layers are equal, and thus have the same execution time, this yields an approximate execution time of
-$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$).
-Therefore, the network~\cite{Chang15} is able to perform around \num{34.3d3} forward propagations per second.
-Although the network is large, its performance is well below this work, which achieves \num{487d3} forward propagations per second, for that network size.
+that both layers are equal, this yields an approximate execution time of $466 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$), and
+if the computation time increases linearly with time, an 8 neuron layer would have an execution time of $29.13 \, \si{\micro\second}$.
+Therefore, an 8-neuron network of~\cite{Chang15} would be able to perform around \num{34.3d3} forward propagations per second, while this work achieves \num{487d3} forward propagations per second, for that network size.
 This is because the work in~\cite{Chang15} does not have a full level of parallelism when compared to the proposed design of
@@
 \end{equation}
-This parameters represents the number of rows that share the same multiplier. The
+This parameters represent the number of rows that share the same multiplier. The
 same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
@@
 bit is the sum of not only the bits of the two operands, but also the \emph{carry} generated at the last few positions -- this is where the
-memory cells of this special Recursive Neural Network come into effect.
+memory cells of this special Recurrent Neural Network come into effect.
 
@@
     $N=128$ & 140.854 MHz\\
-    \hline
     \hline

### 2016-10-11 paper/paper.tex
Final paper version

@@
 \usepackage[utf8]{inputenc}
-
+\usepackage[T1]{fontenc}
 %% Macros
@@
 %%\IEEEspecialpapernotice{(Working Version)}
+\author{
+    \IEEEauthorblockN{Joao Canas Ferreira\IEEEauthorrefmark{1}, Jose Fonseca\IEEEauthorrefmark{2}}\\
+    \IEEEauthorblockA{\IEEEauthorrefmark{1}INESC TEC and Faculty of Engineering of the University of Porto, Portugal, jcf@fe.up.pt}\\
+    \IEEEauthorblockA{\IEEEauthorrefmark{2}CERN, Switzerland, jose.pedro.castro.fonseca@cern.ch}
+}
 
@@
 we still have the recurrent connections from the regular RNNs, but now there are multiple entry points
-that control the flow of information through the network. Although omitted from the picture, all the
+that control the flow of information through the network. Furthermore, all the
 gates are biased, as defined in Equations~\ref{eq:equationsLSTM}. The role and relevance of the main components can
@@
 LSTM Networks are nowadays one of the state-of-the-art algorithms in Deep Learning, and their performance is
-superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address
-the same set of problems where LSTM are employed, namely predicting and producing classification decisions
-from time-series data. A very comprehensive description of applications can be found in one of the initial
+superior to that of other kinds of RNNs and Hidden Markov Models. A very comprehensive description of applications can be found in one of the initial
 authors webpage dedicated to the subject~\footnote{http://people.idsia.ch/\~{}juergen/rnn.html}.
@@
 Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
-(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented in FPGA,
-with 128 neurons each, which, for processing 1000 samples, yielded an execution time of \SI{0.932}{\second}. Assuming
-that both layers are equal, this yields an approximate execution time of $466 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$), and
+(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network was implemented,
+with 128 neurons each, which, for processing 1000 samples, had an execution time of \SI{0.932}{\second}. Assuming
+two equal layers, this yields an approximate execution time of $466 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$), and
 if the computation time increases linearly with time, an 8 neuron layer would have an execution time of $29.13 \, \si{\micro\second}$.
@@
 In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05},
-since evaluating a polynomial does not have high memory usage requirements (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
+since evaluating a polynomial does not have high memory usage requirements and, if the polynomial degree is sufficiently low, the
 number of multiplications needed is small enough to not pose a restriction both on resources (now DSP slices, and not memory) and on speed (number of clock cycles
@@
 \end{eqnarray}
-where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal asymptotes, which can be used as the value in the edges.
+where $i \in \left[0, n+1\right]$. Both functions have horizontal asymptotes, which were used as the value in the edges.
 Instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
@@
 coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial
-was accomplished using the \textbf{Horner's Rule}, that is
+was accomplished using \textbf{Horner's Rule}, that is
 
@@
 and, therefore, the calculation is simply the procedure of multiplying the operand by a value and adding a constant to the result,
-repeated two times. Again the internal machine state controls which values are multiplied and added according to the pipeline state.
-Due to the internal datapath pipeline, each module takes 5 clock cycles to output one result.
+repeated two times. An internal machine state controls which values are multiplied and added according to the pipeline state, and due to it,
+each module takes 5 clock cycles to output one result.
 
@@
 has size $N\times N$, because $x$ has length $M$ (the number of inputs to the layer), while $y$ has length $N$
-(the number of neurons in the layer). If a layer with different parameters is needed, either in terms of
-number of inputs or number of neurons, we only need to change the respective parameter before the synthesis stage,
-instead of having to redesign the whole block for that particular size.
+(the number of neurons in the layer). If a layer with different parameters is needed, we only need to change the respective parameter before the synthesis stage,
+instead of having to redesign the whole block.
 
@@
 This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel},
-each one only requiring the corresponding $i$-th line from the matrix. Using this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
-which is one of the great advantages of custom-tailored hardware solutions.
+each one only requiring the corresponding $i$-th line from the matrix. Using this approach, matrix-vector multiplication can now be performed in \textbf{linear time}.
 
@@
 \end{equation}
-This parameters represent the number of rows that share the same multiplier. The
-same multiplier is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
+This parameter represents the number of rows that share the same multiplier: it is responsible for producing several $i$-th elements of the output vector, in consecutive time slots of $M$ clock cycles.
 For instance, in an $8\times2$ matrix scenario, with $K_G = 2$, we would have 4 multipliers, and the output vector
@@
 we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod},
-that work in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding
+working in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding
 column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
 The dark shaded part of the memory is used by the first multiplier, and the light shaded one is used by the other, in parallel for a
-fixed \verb+rowMux+. This signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
+fixed \verb+rowMux+. This signal is produced by the control unit of the module, and operates the left multiplexer and right demultiplexer
 of Figure~\ref{fig:array-prod}, that allows choosing the proper position of the weight column and writing to the correct output vector position, respectively.
@@
 proportional to $K_G \times M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
-in the \emph{next clock cycle}, and this module is pipelined both at the input and at the output, in order to increase the maximum clock frequency.
+in the \emph{next clock cycle}, and this module is pipelined both at the input and at the output.
 
@@
 The weights are stored in LUTRAM, and for that purpose, they are declared as a matrix of registers, where the access, both in terms
-of write and read operations, is made to each column. The write/read access can be performed at the same time, since the memory has
+of write and read operations, is made to each column. The write/read access can be performed simultaneously, since the memory has
 separate communication ports for input/output, provided that the \textbf{addresses do not coincide}.
@@
 time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time. The module is triggered by a
-\verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded,
-and the value at the output is the final result.
+\verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded.
 After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
@@
 
+\smallskip
+\emph{\textbf{This work is financed by the ERDF – European Regional Development Fund through the Operational Programme for Competitiveness and Internationalisation - COMPETE 2020 Programme within project \guillemotleft POCI-01-0145-FEDER-006961\guillemotright, and by National Funds through the FCT – Fundação para a Ciência e a Tecnologia (Portuguese Foundation for Science and Technology) as part of project \guillemotleft UID/EEA/50014/2013\guillemotright.}}
+
 \bibliographystyle{IEEEtran}

### 2016-10-12 paper/paper.tex
8 pages only

@@
 Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
-of a given experiment faster than in a regular CPU, yielding an increase in scientific productivity.
+of a given experiment faster than in a regular CPU, increasing scientific productivity.
 
@@
 
-Summarizing, this article begins by explaining what are LSTM networks -- and their mathematical
-details -- in Section~\ref{sec:lstmnn}, followed by a quick overview of their state
+This article begins by explaining what are LSTM networks  in Section~\ref{sec:lstmnn}, followed by a quick overview of their state
 of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed
@@
 \verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded. 
-After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
+Taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
 of clock cycles this module takes to produce an output is $6+K_G\cdot N$.
@@
 and activation function calculators) and require more area to save a negligible number of clock cycles, by noting that one of the operands is the output
-of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. Instead of replicating
+of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied, elementwise, together. Instead of replicating
 these `$\tanh$-$\sigma$-$(\cdot)$wise' structures, we use a \emph{single one} and choose the operand according to the state that the network is currently in. The
@@
 which is only 13 clock cycles more than a fully-parallel, naïve architecture. For instance, an $N=32$ neuron network would
-require 320 DSP slices, while this new architecture use 160, at the expense of 13 more clock cycles.
+require 320 DSP slices, while this new architecture uses 160, at the expense of 13 more clock cycles.
 
@@
 
-Since the previous values are the time needed for a \emph{single} forward propagation, to know how many forward propagations we can perform \emph{per second},
+To know how many forward propagations we can perform \emph{per second},
 we only need to invert the previous values. These values are presented in Figure~\ref{fig:Mclass-psec}. While the $N=8$ and $K_G=2$ network is able to
@@
 
-\smallskip
-\emph{\textbf{This work is financed by the ERDF – European Regional Development Fund through the Operational Programme for Competitiveness and Internationalisation - COMPETE 2020 Programme within project \guillemotleft POCI-01-0145-FEDER-006961\guillemotright, and by National Funds through the FCT – Fundação para a Ciência e a Tecnologia (Portuguese Foundation for Science and Technology) as part of project \guillemotleft UID/EEA/50014/2013\guillemotright.}}
+\section*{Acknowledgements}
+This work is financed by the ERDF – European Regional Development Fund through the Operational Programme for Competitiveness and Internationalisation - COMPETE 2020 Programme within project \guillemotleft POCI-01-0145-FEDER-006961\guillemotright, and by National Funds through the FCT – Fundação para a Ciência e a Tecnologia (Portuguese Foundation for Science and Technology) as part of project \guillemotleft UID/EEA/50014/2013\guillemotright.
 

### 2016-10-12 paper/paper.tex
Conference Ready

@@
-\documentclass{IEEEtran}
+\documentclass[conference]{IEEEtran}
 
@@
 \usepackage[T1]{fontenc}
+\usepackage{eso-pic}
+
 %% Macros
@@
     \IEEEauthorblockA{\IEEEauthorrefmark{1}INESC TEC and Faculty of Engineering of the University of Porto, Portugal, jcf@fe.up.pt}\\
-    \IEEEauthorblockA{\IEEEauthorrefmark{2}CERN, Switzerland, jose.pedro.castro.fonseca@cern.ch}
+    \IEEEauthorblockA{\IEEEauthorrefmark{2}Faculty of Engineering of the University of Porto, Portugal, jose.pedro.castro.fonseca@cern.ch}
 }
+
+\AddToShipoutPicture*{\small\sffamily\raisebox{1.8cm}{\hspace{1.8cm}978-1-5090-3707-0/16/\$31.00 \copyright2016 European Union}}
 
@@
 
-A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As can be seen in the picture,
-we still have the recurrent connections from the regular RNNs, but now there are multiple entry points
+The structure of an LSTM neuron is presented in~\cite{Greff15}. LSTM Networks retain recurrent connections from the regular RNNs, but now there are multiple entry points
 that control the flow of information through the network. Furthermore, all the
@@
 be summarized as follows.
-
-\begin{figure}[!t]
-	\centering
-	\includegraphics[width=1.1\linewidth]{figures/lstmneuron.png}
-    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}. Source:~\cite{Greff15}}
-	\label{fig:lstmneuron}
-\end{figure}
 
@@
 Section~\ref{sec:res-val} is an $N=8$ and $K_G=2$ network clocked at $154.321$ MHz, which yields a clock period of
-\SI{6.48}{\nano\second}. The maximum clock achievable for the VC707 networks is reported in Table~\ref{tab:maxfreq-virtx7}
+\SI{6.48}{\nano\second}. The maximum clock achievable for the VC707 networks is 140.854 MHz, for $N=64$ and $N=128$.
 
@@
 \end{figure}
-
-\begin{table}
-	\caption{Maximum clock frequencies for the VC707}
-	\label{tab:maxfreq-virtx7}
-    \centering
-  \begin{tabular}{ | l | c | }
-    \hline
-     & $Frequency$  \\
-    \hline
-    $N=64$ & 140.854 MHz \\
-    \hline
-    $N=128$ & 140.854 MHz\\
-    \hline
-  \end{tabular}
-
-\end{table}
 
@@
 
-For the networks synthesized on the VC707, the power consumption figures are reported in Table~\ref{tab:power-virtx7}. Since the device is bigger,
-and since the network size is larger, the power consumption is higher than before, but still within reasonable levels given
-the complexity of the system.
-
-\begin{table}
-	\caption{Power Consumption estimates for several Network Sizes $N$ in the VC707}
-	\label{tab:power-virtx7}
-    \centering
-  \begin{tabular}{ | l | c | }
-    \hline
-     & $Power$  \\
-    \hline
-    $N=64$ & 1.34 W \\
-    \hline
-    $N=128$ & 1.509 W\\
-\hline
-  \end{tabular}
-
-\end{table}
-
-All these estimates were provided by Vivado, and have a \textbf{medium} confidence level.
+For the networks synthesized on the VC707, the power consumption was 1.34 Watt for $N=64$, and 1.51 Watt for $N=128$.
+Since the network size is larger, the power consumption is higher than before, but still within reasonable levels given the complexity of the system.
+These estimates were provided by Vivado, and have a \textbf{medium} confidence level.
 
